### 2016-02-10 paper/paper.tex
Basic template for paper about cyclone.

\documentclass[twocolumn]{article}
\usepackage{times}

\begin{document}
\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
introduction of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. Datacenter applications typically use storage devices to survive
faults that take one more machines offline by combining the persistence offered by
durable storage with availability through replication. Building such fault
tolerant applications is hard and users are forced to either build
them from scratch or modify their applications to fit the interfaces of solutions
such as zookeeper. In addition to a loss of flexibility such applications today
are unable to take advantage of the speed of directly attached non-volatile
memory as opposed to older slower storage in the form of disks and SSDs.

This paper presents Cyclone - fault tolerance middleware that exposes a
replicated and durable heap. This heap can be directly manipulated using loads and
stores. Cyclone transparently provides both failure atomicity as well as
strongly consistent replication thereby giving strong guarantees to programmers
with minimal effort on their part. Under the hood Cyclone decouples modification
of local persistent memory from replication to maximize performance. In
addition, unlike traditional systems that provide strong replication, Cyclone
does not take checkpoints since NVM is directly manipulated by programmers and
therefore up to date till the last executed transaction. Instead, Cyclone
provides a novel NVM copy mechanism to bring up new nodes added to the cluster.
\end{abstract}

\section{Introduction}

\section{Programming Model}
Cyclone provides users with a replicated

\section{Failure Atomicity}

\section{Replication}

\section{Evauation}

\subsection{Baseline}

\subsection{Configuration Management}

\subsection{Transaction Processing}

\subsection{Network Function Virtualization}

\end{document}

### 2016-02-10 paper/paper.tex
More text in paper.

@@
 \section{Programming Model}
-Cyclone provides users with a replicated
+Cyclone provides users with a replicated and durable heap that can be
+manipulated via a set of library routines. The library provides special
+functions to allocate and free persistent memory as well as fat pointers that
+wrap pointers to persistent memory to ensure that they do not change across
+restarts - since the virtual memory ranges into which persistent memory is
+mapped may change.
+
+\begin{figure}
+\begin{verbatim}
+
+struct ll_node {
+  int value;
+  TOID(ll_node) next;
+};
+
+void insert_after(TOID(ll_node) prev, TOID(ll_node) new_node)
+{
+  D_RW(new_node)->next = D_RO(prev)->next;
+  D_RW(prev)->next = new_node;
+}
+
+\end{verbatim}
+\caption{A Persistent Linked List of Integers}
+\label{fig:example}
+\end{figure}
+
+Figure~\ref{fig:example} shows the example of a linked list node and a function
+for inserting a new node after a given one. The {\tt TOID} declaration wraps the
+linked list node type to produce a fat pointer to it. A {\tt TOID(ll\_node)}
+type therefore is a pointer to a linked list node stored in persistent memory.
+The {\tt insert\_after} function used the {\tt D\_RO} and {\tt D\_RW} calls to
+convert fat pointers into directly accessible read-only and read-write pointers
+respectively. Other than the mechanics of using fat pointers the code entirely
+resembles that which a programmer might write for a standard linked list in
+DRAM.
 
 \section{Failure Atomicity}
+Programmers typically find it hard to reason about failures when writing
+code. This is a problem when directly manipulating data structures in persistent
+memory. Depending on the specifics of the system architecture, updates to
+persistent memory can be persisted out of order thereby leaving  the data
+structure in an inconsistent state. To illustrate this consider a failure during
+execution of the code to add a node to a linked list in
+Figure~\ref{fig:example}. If either of the two updates to the next fields of the
+previous or new node fail to be persisted, the linked list is left in a
+disconnected state with one or more connected nodes inaccessible.
+
+Cyclone provides tools for programmers to write code that is failure
+atomic. Portions of the code are delimited as transactions that are either
+completely executed in terms of its effects on persistent memory or not executed
+at all. Under the hood, Cyclone tracks all updates to persistent memory in a
+persistent undo log - the log entries are persisted before the update to the
+actual memory is allowed. On a failure, the undo log is used to roll back
+incomplete transactions. A transaction is considered complete when it completes
+execution and the updated objects are flushed (automatically by Cyclone) to
+persistent memory. The log entries for the transaction are then immediately
+deleted. The only burden placed on a programmer for failure atomicity is that
+they must inform the Cyclone runtime about objects that are
+updated. Figure~\ref{fig:example_fatomic} shows how to convert the function for
+adding a node to a linked list into a failure atomic one. The only additions are
+the transaction delimiters and the {\tt TX\_ADD} calls to mark the objects being
+updated \emph{before} the actual updates to them.
+
+\begin{figure}
+\begin{verbatim}
+
+struct ll_node {
+  int value;
+  TOID(ll_node) next;
+};
+
+void insert_after(TOID(ll_node) prev, TOID(ll_node) new_node)
+{
+  TX_BEGIN {
+    TX_ADD(new_node);
+    D_RW(new_node)->next = D_RO(prev)->next;
+    TX_ADD(prev);
+    D_RW(prev)->next = new_node;
+  } TX_END
+}
+
+\end{verbatim}
+\caption{A Persistent Linked List of Integers}
+\label{fig:example}
+\end{figure}
 

### 2016-02-11 paper/paper.tex
More updates to paper.

@@
 \begin{figure}
+{\scriptsize
 \begin{verbatim}
@@
 \end{verbatim}
+}
 \caption{A Persistent Linked List of Integers}
@@
 the transaction delimiters and the {\tt TX\_ADD} calls to mark the objects being
-updated \emph{before} the actual updates to them.
+updated \emph{before} the actual updates to them. Figure~\ref{fig:undo_log}
+shows how Cyclone maintains the undo-log of objects.
 
 \begin{figure}
+{\scriptsize
 \begin{verbatim}
@@
 \end{verbatim}
+}
 \caption{A Persistent Linked List of Integers}
-\label{fig:example}
+\label{fig:example_fatomic}
 \end{figure}
+
+\begin{figure}
+{\scriptsize
+\begin{verbatim}
+TX_BEGIN()
+{
+  Initialize undo log
+}
+
+TX_ADD(o)
+{
+  if(o not already in undo log) {
+    Append <o, value of o> to undo log
+  }
+}
+
+TX_END()
+{
+  for each entry <o, v> in undo log {
+    flush o to persistent storage
+  }
+  truncate undo log
+}
+
+Recover()
+{
+  for each entry <o, v> in undo log {
+    set o to v
+    flush o to persistent storage
+  }   
+}
+
+\end{verbatim}
+}
+\caption{Undo Log in Cyclone}
+\label{fig:undo_log}
+\end{figure}
+
+A crucial feature of Cyclone is that it abstracts the details of the persistent
+mechanism away from programmers. This allows the same code to be deployed on
+clusters with a persistent storage medium such as 3D XPoint or on clusters with
+battery backed NVDIMMs for example. In the former case, explicit {\tt PCOMMIT}
+instructions are necessary for persistence [ref PMFS paper], while in the latter
+case no such instructions are necessary as on failure the CPU caches and DRAM
+are flushed to stable storage under battery (or capacitor) power. Even in the
+NVDIMM case however, the undo log is necessary to protect against OS or
+application crashes to ensure that the persistent memory is always left in a
+consistent state.
 
@@
 
+\section{Related Work}
+
 \end{document}

### 2016-02-11 test/rbtree_map_coordinator_driver.cpp
Add a coordinator driver test.

@@ -1,0 +1,102 @@
+/*
+ * Copyright (c) 2015, Intel Corporation
+ *
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted provided that the following conditions
+ * are met:
+ *
+ *     * Redistributions of source code must retain the above copyright
+ *       notice, this list of conditions and the following disclaimer.
+ *
+ *     * Redistributions in binary form must reproduce the above copyright
+ *       notice, this list of conditions and the following disclaimer in
+ *       the documentation and/or other materials provided with the
+ *       distribution.
+ *
+ *     * Neither the name of Intel Corporation nor the names of its
+ *       contributors may be used to endorse or promote products derived
+ *       from this software without specific prior written permission.
+ *
+ * THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
+ * "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
+ * LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
+ * A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
+ * OWNER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
+ * SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
+ * LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
+ * DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
+ * THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
+ * (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+ * OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+ */
+
+#include <stdio.h>
+#include <sys/stat.h>
+#include <stdlib.h>
+#include <string.h>
+#include <unistd.h>
+#include <time.h>
+#include <assert.h>
+#include "rbtree_map_coordinator.hpp"
[…]

### 2016-02-13 paper/paper.tex
More updates to paper.

@@
 
-\section{Programming Model}
+\section{Single Node Programming Model}
 Cyclone provides users with a replicated and durable heap that can be
@@
 \section{Replication}
+Cyclone replicates the state of the NVM heap across multiple machines in a
+cluster. Cyclone provides strong consistency - a majority of machines in the
+quorum of replicas always agrees on the current state of the NVM heap. Unlike
+traditional approaches to strong replication however, Cyclone is designed to
+decouple replication from execution of user code modifying NVM. It does so by
+encapsulating changes in to the NVM heap in RPC calls driven by client requests.
+We chose this model since it closely reflects the use cases most suitable for
+NVM-based clusters.
+
+Cyclone replicates state \emph{by executing the same sequence of RPC calls on
+  all members of the quorum}. Unlike traditional replication therefore Cyclone
+does not replicate logs of changes to apply to replicas. Instead it replicates
+the RPC call and its arguments, executing it on every replica. By executing the
+same sequence of RPC calls on all machines we end up with the same
+state. Cyclone uses RAFT under the hood for consensus on the contents of the log
+of RPC calls - although any other similar state machine replication protocol
+such as Paxos is also suitable. Cyclone provides the means to achieve exactly
+once semantics to clients by numbering RPC calls and remembering the results of
+the last executed RPC call.
+
+\subsection{RPC Call Numbering}
+Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
+from a client are numbered and Cyclone remembers the results of the last RPC
+call made by a client. A particular client can exist on any machine - cyclone
+treats all instances of the client as a single client. Cyclone rejects RPC calls
+that are not in sequence. Any RPC call is executed exactly once across the
+quorum - regardless of failure and reconfigurations. This provides the means for
+clients to easily achieve exactly once semantics, provided the client does not
+crash. Cyclone can supply the last executed RPC call number and its result to
+assist recovering and stateless clients.
+
+\subsection{Asynchronous Replication}
+By default, Cyclone decouples execution of the RPC call on any replica from its
+replication. This is particularly beneficial to performance since it allows
+overlapping the update of non-volatile state with the network I/O associated
+with replication. Persisting RAFT log entries, making the upcall to the
+application and executing the RPC call in a failure atomic transaction takes a
+significant amount of time and therefore it makes sense to overlap this with the
+round-trip delay on the network for replication. This is possible in Cyclone and
+not other traditional replication systems since we use failure atomic
+transactions that cna be rolled back. Replication can fail due to various
+reasons such as failure of nodes, a network partition or a new leader getting
+elected in the underlying replication protocol (RAFT). Since failure atomic
+transactions can be rolled back, we are able to speculatively apply the changes
+to NVM without waiting for replication to finish.
+
+The runtime steps associated with asynchronous replication are shown in
+Figure~\ref{fig:async_rep}.
+
+\begin{figure}
+{\scriptsize
+\begin{verbatim}
+Initiate replication of RPC call and arguments
+TX_BEGIN {
+  Execute RPC call and modify local NVM
+  while(!replication_complete);
+  if(replication_failed) {
+    TX_ABORT;
+  }
+} TX_END
+\end{verbatim}
+}
+\caption{Asynchronous Replication}
+\label{fig:async_rep}
+\end{figure}
+
+\subsection{Synchronous Replication}
+
+\section{Multi-node Programming Model}
+
+\subsection{Example: Distributed Transactions}
 

### 2016-02-16 paper/paper.tex
paper updates

@@
 \subsection{Synchronous Replication}
+Asynchronous replication provides good performance but lacks some guarantees
+necessary for certain distributed systems paradigms such as output commit and
+the ability to integrate code that has non-deterministic execution paths or side
+effects - such as reading the time. Cyclone therefore provides programmers the
+ability to closely synchronize execution across the replicas using synchronous
+replication. Synchronous replication extends the programming model provided by
+cyclone to users - neccessitating some awareness on the part of the user about
+the difference between the leader and follower replicas. We show later
+(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
+fault tolerant application with little effort on the part of the programmer.
 
-\section{Multi-node Programming Model}
+Figure~\ref{fig:sync_rep} shows how synchronous replication operates. Execution
+of the RPC call is only initiated when the call and its arguments have been
+replicated. This ensures that any side effects such as calls to other machines
+only happen after it is guaranteed that the effects of the RPC call will be
+committed to the replicas. This provides the output commit property necessary
+for some of the applications we discuss below. Next the RPC call is executed on
+the leader and the result of the RPC call is replicated to the followers. The
+followers then proceed to execute the RPC call locally taking into account the
+result from the leader. This result can include data to synchronize
+non-deterministic steps such as reading the time on the replica. This process is
+always conditional on the view (current configuration of replicas and choice
+leader) staying unchanged. On a view change the process is reexcuted to ensure
+than only one set of RPC call results is propagated across the replicas.
 
-\subsection{Example: Distributed Transactions}
+\begin{figure}
+{\scriptsize
+\begin{verbatim}
+Initiate replication of RPC call and arguments
+TX_BEGIN {
+  while(!replication_complete);
+  if(replication_failed) {
+    TX_ABORT;
+  }  
+  while(true) {
+    if(is_leader && !have_execution_result) {
+      Execute RPC call and modify local NVM
+      replicate_rpc_call_result in current view;
+      if(result replication failed) {
+        continue;
+      }
+      while(!result replication complete && !view change);
+      if(view change) {
+        continue;
+      }
+    }
+    else {
+      while(!result replication complete && !view change);
+      if(view change) {
+        continue;
+      }
+      Execute RPC call to modify local NVM with result from leader
+    }
+  }
+} TX_END
+\end{verbatim}
+}
+\caption{Synchronous Replication}
+\label{fig:sync_rep}
+\end{figure}
+
+\section{Example Applications}
 

### 2016-03-01 test/rbtree_map_server.cpp
Add support for cookie at server end.

@@ -76,32 +76,30 @@ int callback(const unsigned char *data,
 {
   struct proposal *req = (struct proposal *)data;
   int code = req->fn;
+  *return_value  = malloc(sizeof(struct proposal));
+  struct proposal *rep  = (struct proposal *)*return_value;
+  memcpy(rep->cookie, req->cookie, sizeof(cookie_t));
+
   if(code == FN_LOCK) {
     uint64_t *vptr = D_RW(version_table) +
       key_to_vtable_index(req->kv_data.key);
-    *return_value  = malloc(sizeof(struct proposal));
-    struct proposal *rep  = (struct proposal *)*return_value;
     uint64_t version = *vptr;
     rep->code = version;
     if(version == req->kv_data.value) { // UNLOCKED ?
       pmemobj_tx_add_range_direct(vptr, sizeof(uint64_t));
       *vptr = *vptr + 1; // LOCK
     }
-    return sizeof(struct proposal);
   }
   else if(code == FN_GET_VERSION) {
     uint64_t *vptr = D_RW(version_table) +
       key_to_vtable_index(req->k_data.key);
-    *return_value  = malloc(sizeof(struct proposal));
     struct proposal *rep  = (struct proposal *)*return_value;
     uint64_t version = *vptr;
     rep->code = version;
-    return sizeof(struct proposal);
   }
   else if(code == FN_UNLOCK) {
     uint64_t *vptr = D_RW(version_table) +
       key_to_vtable_index(req->kv_data.key);
-    *return_value  = malloc(sizeof(struct proposal));
     struct proposal *rep  = (struct proposal *)*return_value;
     uint64_t version = *vptr;
     rep->code = version;
@@ -109,13 +107,15 @@ int callback(const unsigned char *data,
       pmemobj_tx_add_range_direct(vptr, sizeof(uint64_t));
       *vptr = *vptr + 1; // UNLOCK
     }
-    return sizeof(struct proposal);
   }
   else if(code == FN_INSERT) {
     PMEMoid item = rbtree_map_get(pop, the_tree, req->kv_data.key);
+    rep->code = CODE_OK;
+    rep->kv_data.key = req->kv_data.key;
     if(!OID_IS_NULL(item)) {
       pmemobj_tx_add_range(item, 0, sizeof(uint64_t));
       uint64_t *ptr = (uint64_t *)pmemobj_direct(item);
+      rep->kv_data.value = *ptr;
[…]

### 2016-03-01 test/tree_map.hpp

@@ -53,6 +53,20 @@ const int CODE_NOK  = 1;
 
 struct k {uint64_t key;};
 struct kv {uint64_t key; uint64_t value;};
+
+
+typedef struct cookie_st {
+  int phase;
+  int locks_taken;
+  int index;
+} cookie_t;
+
+const int COOKIE_PHASE_LOCK     = 0;
+const int COOKIE_PHASE_VERSION  = 1;
+const int COOKIE_PHASE_INSERT   = 2;
+const int COOKIE_PHASE_DELETE   = 3;
+const int COOKIE_PHASE_UNLOCK   = 4;
+
 struct proposal {
   union {
     int fn;
@@ -65,6 +79,7 @@ struct proposal {
     struct kv kv_data;
     struct k k_data;
   };
+  cookie_t cookie;
 };
 
 

### 2016-03-01 test/rbtree_map_server.cpp
Compile fix.

@@ -78,7 +78,7 @@ int callback(const unsigned char *data,
   int code = req->fn;
   *return_value  = malloc(sizeof(struct proposal));
   struct proposal *rep  = (struct proposal *)*return_value;
-  memcpy(rep->cookie, req->cookie, sizeof(cookie_t));
+  memcpy(&rep->cookie, &req->cookie, sizeof(cookie_t));
 
   if(code == FN_LOCK) {
     uint64_t *vptr = D_RW(version_table) +

### 2016-03-01 core/dispatch_client.cpp
Add API call to get last response.

@@ -83,6 +83,63 @@ typedef struct rpc_client_st {
     return packet_in->last_client_txid;
   }
 
+  int retrieve_response(void **response, int txid)
+  {
+    packet_out->client_id   = me;
+    packet_out->client_txid = txid;
+    packet_out->timestamp   = clock.current_time();
+    while(true) {
+      packet_out->code        = RPC_REQ_STATUS_BLOCK;
+      retcode = cyclone_tx_timeout(router->output_socket(server), 
+				   (unsigned char *)packet_out, 
+				   sizeof(rpc_t), 
+				   timeout_msec*1000,
+				   "PROPOSE");
+      if(retcode == -1) {
+	update_server("tx timeout");
+	continue;
+      }
+
+      while(true) {
+	resp_sz = cyclone_rx_timeout(router->input_socket(server), 
+				     (unsigned char *)packet_in, 
+				     DISP_MAX_MSGSIZE, 
+				     timeout_msec*1000,
+				     "RESULT");
+	if(resp_sz != -1 && 
+	   packet_in->client_txid != txid &&
+	   packet_in->code != RPC_REP_INVSRV) {
+	  continue; // Ignore response
+	}
+	break;
+      }
+      if(resp_sz == -1) {
+	update_server("rx timeout");
+	continue;
+      }
+      if(packet_in->code == RPC_REP_INVSRV) {
+	if(packet_in->master == -1) {
+	  BOOST_LOG_TRIVIAL(info) << "Unknown master !";
+	  update_server("unknown master");
+	}
+	else {
+	  server = packet_in->master;
+	  set_server();
+	}
+	continue;
+      }
+      if(packet_in->code == RPC_REP_PENDING) {
+	continue;
+      }
+      break;
+    }
+    if(packet_in->code == RPC_REP_OLD) {
+      return RPC_EOLD;
+    }
+    *response = (void *)(packet_in + 1);
+    return (int)(resp_sz - sizeof(rpc_t));
+  }
+  
   int make_rpc(void *payload, int sz, void **response, int txid, int flags)
   {
     int retcode;
@@ -242,4 +299,10 @@ int get_last_txid(void *handle)
 {
   rpc_client_t *client = (rpc_client_t *)handle;
   return client->get_last_txid();
+}
+
+int get_response(void *handle, void **response, int txid)
+{
+  rpc_client_t *client = (rpc_client_t *)handle;
+  return client->retrieve_response(response, txid);
 }

### 2016-03-01 core/libcyclone.hpp

@@ -66,6 +66,10 @@ int make_rpc(void *handle,
 	     int rpc_flags);
 // Get last accepred txid
 int get_last_txid(void *handle);
+// Get the last response
+int get_response(void *handle,
+		 void **response,
+		 int txid);
 
 // Possible flags 
 static const int RPC_FLAG_SYNCHRONOUS   = 1; // Synchronous execution across replicas

### 2016-03-01 core/dispatch_client.cpp
Compile fixes.

@@ -85,6 +85,9 @@ typedef struct rpc_client_st {
 
   int retrieve_response(void **response, int txid)
   {
+    rtc_clock clock;
+    int retcode;
+    int resp_sz;
     packet_out->client_id   = me;
     packet_out->client_txid = txid;
     packet_out->timestamp   = clock.current_time();

### 2016-03-31 paper/paper.tex
Additions to the paper.
Updates to paper.

--- 2016-02-13 paper/paper.tex
@@
 replicated and durable heap. This heap can be directly manipulated using loads and
-stores. Cyclone transparently provides both failure atomicity as well as
-strongly consistent replication thereby giving strong guarantees to programmers
+stores. Cyclone provides both failure atomicity as well as
+strongly consistent replication when updates to the heap are enclosed in
+software transactions thereby giving strong guarantees to programmers
 with minimal effort on their part. Under the hood Cyclone decouples modification
@@
 addition, unlike traditional systems that provide strong replication, Cyclone
-does not take checkpoints since NVM is directly manipulated by programmers and
-therefore up to date till the last executed transaction. Instead, Cyclone
-provides a novel NVM copy mechanism to bring up new nodes added to the cluster.
+does require programs to be provisioned for checkpoints since NVM is directly
+manipulated by programmers and therefore up to date till the last executed
+transaction. Instead, Cyclone provides a novel NVM copy mechanism to bring up
+new nodes added to the cluster.
+
 \end{abstract}
@@
 \begin{figure}
-{\scriptsize
+{ \scriptsize
 \begin{verbatim}
@@
 Figure~\ref{fig:example}. If either of the two updates to the next fields of the
-previous or new node fail to be persisted, the linked list is left in a
+previous and new node fail to be persisted, the linked list is left in a
 disconnected state with one or more connected nodes inaccessible.
@@
 
-A crucial feature of Cyclone is that it abstracts the details of the persistent
+A crucial feature of Cyclone is that it abstracts the details of the persistence
 mechanism away from programmers. This allows the same code to be deployed on
@@
 are flushed to stable storage under battery (or capacitor) power. Even in the
-NVDIMM case however, the undo log is necessary to protect against OS or
+NVDIMM case however, the undo log can be desirable to protect against OS or
 application crashes to ensure that the persistent memory is always left in a
-consistent state.
+consistent state. Cyclone's programmer visible features are designed to enable
+the construction of applications that work regardless of the underlying
+persistence mechancisms.
 
@@
 decouple replication from execution of user code modifying NVM. It does so by
-encapsulating changes in to the NVM heap in RPC calls driven by client requests.
+encapsulating changes to the NVM heap in RPC calls driven by client requests.
 We chose this model since it closely reflects the use cases most suitable for
@@
 Cyclone replicates state \emph{by executing the same sequence of RPC calls on
-  all members of the quorum}. Unlike traditional replication therefore Cyclone
-does not replicate logs of changes to apply to replicas. Instead it replicates
-the RPC call and its arguments, executing it on every replica. By executing the
-same sequence of RPC calls on all machines we end up with the same
-state. Cyclone uses RAFT under the hood for consensus on the contents of the log
-of RPC calls - although any other similar state machine replication protocol
-such as Paxos is also suitable. Cyclone provides the means to achieve exactly
-once semantics to clients by numbering RPC calls and remembering the results of
-the last executed RPC call.
+  all members of the quorum}. By executing the same sequence of RPC calls on all
+machines we end up with the same state\footnote{We discuss how to handle
+  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
+  raft paper] under the hood for consensus on the contents of the log of RPC
+calls - although any other similar state machine replication protocol such as
+Paxos is also suitable. Cyclone provides the means to achieve exactly once
+semantics to clients by numbering RPC calls and remembering the result of the
+last executed RPC call.
 
@@
 significant amount of time and therefore it makes sense to overlap this with the
-round-trip delay on the network for replication. This is possible in Cyclone and
-not other traditional replication systems since we use failure atomic
-transactions that cna be rolled back. Replication can fail due to various
-reasons such as failure of nodes, a network partition or a new leader getting
-elected in the underlying replication protocol (RAFT). Since failure atomic
-transactions can be rolled back, we are able to speculatively apply the changes
-to NVM without waiting for replication to finish.
+round-trip delay on the network for replication. This is possible in Cyclone
+since we use failure atomic transactions that can be rolled back. Replication
+can fail due to various reasons such as failure of nodes, a network partition or
+a new leader getting elected in the underlying replication protocol
+(RAFT). Since failure atomic transactions can be rolled back, we are able to
+speculatively apply the changes to NVM without waiting for replication to
+finish.
 
-The runtime steps associated with asynchronous replication are shown in
+The runtime steps for asynchronous replication are shown in
 Figure~\ref{fig:async_rep}.
@@
   Execute RPC call and modify local NVM
-  while(!replication_complete);
-  if(replication_failed) {
+  // Block till replication result known
+  if(replication failed) {
     TX_ABORT;
@@
 \subsection{Synchronous Replication}
+\label{sec:synch}
+Asynchronous replication provides good performance but lacks guarantees
+occasionally necessary for distributed systems paradigms such as output commit and
+the ability to integrate code that has non-deterministic execution paths or side
+effects - such as reading the time. Cyclone therefore provides programmers the
+ability to closely synchronize execution across the replicas using synchronous
+replication. Synchronous replication extends the programming model provided by
+cyclone - but neccessitates some awareness on the part of the user about
+the difference between the leader and follower replicas. We show later
+(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
+fault tolerant applications with little effort on the part of the programmer.
 
-\section{Multi-node Programming Model}
+Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
+of the RPC call \emph{waits until the call and its arguments have been
+  replicated.} This crucial difference from asynchronous RPC execution ensures
+that any externally visible side effects of the RPC such as calls to other
+machines only happen after it is guaranteed that the effects of the RPC call
+will be committed to the replicas. This provides the output commit property
+necessary for some fault tolerance features in applications. Next the RPC call
+is executed on the leader and the result of the RPC call is replicated to the
+followers. The followers then proceed to execute the RPC call locally taking
+into account the result from the leader. This result can include data to
+synchronize non-deterministic steps such as reading the time on the replica.
 
-\subsection{Example: Distributed Transactions}
+This process is made conditional on the view (current configuration of
+replicas and choice of leader) staying unchanged. To successfully complete
+execution, the leader must be able to execute the RPC call and replicate the
+result within the current view. This ensures that in the event of a network
+partition and later repair an old execution of a previous leader cannot commit a
+log record with results from a synchronously executed RPC call.
+
+\begin{figure*}
+{ \scriptsize
+\begin{verbatim}
+Initiate replication of RPC call and arguments
+while(true) {
+  TX_BEGIN {
+    // Block till replication result known
+    if(replication failed) {
+      TX_ABORT;
+    }  
+    if(is_leader && !have_execution_result) {
+      Execute RPC call and modify local NVM
+      replicate rpc call result in current view;
+      // Block till replication result known
+      if(result replication failed) {
+        TX_ABORT;
+      }
+    }
+    else {
+      // Block till replication done or view changes
+      if(!result replication complete) {
+        TX_ABORT;
+      }
+      Execute RPC call to modify local NVM with result from leader
+    }
+  }
+} TX_END
+\end{verbatim}
+}
+\caption{Synchronous Replication}
+\label{fig:sync_rep}
+\end{figure*}
+
+\section{Example Applications}
+\label{sec:examples}
 

### 2016-05-04 core/dispatcher.cpp
Bugfixes.

@@ -314,7 +314,7 @@ void exec_rpc_internal_synchronous(rpc_info_t *rpc)
 	  free(follower_data);
 	  while(!rpc->rep_follower_success &&
 		cyclone_get_term(cyclone_handle) == execution_term);
-	  if(cyclone_get_term(cyclone_handle) != execution_term) {
+	  if(!rpc->rep_follower_success) {
 	    repeat = true;
 	    pmemobj_tx_abort(-1);
 	  }
@@ -460,12 +460,15 @@ static void gc_pending_rpc_list(bool is_master)
   tmp = pending_rpc_head;
   while(tmp) {
     if(tmp->req_follower_data_active) {
-      cyclone_add_entry_term(cyclone_handle,
-			     tmp->req_follower_data,
-			     tmp->req_follower_data_size,
-			     tmp->req_follower_term);
+      cookie = cyclone_add_entry_term(cyclone_handle,
+				      tmp->req_follower_data,
+				      tmp->req_follower_data_size,
+				      tmp->req_follower_term);
       __sync_synchronize();
       tmp->req_follower_data_active = false;
+      if(cookie != NULL) {
+	free(cookie);
+      }
     }
     tmp = tmp->next;
   }

### 2016-05-04 core/dispatcher_exec.hpp

@@ -12,7 +12,7 @@ typedef struct rpc_info_st {
   void *ret_value;
   void *follower_data;
   int follower_data_size;
-  bool have_follower_data;
+  volatile bool have_follower_data;
   char * volatile req_follower_data;
   volatile int req_follower_data_size;
   volatile int req_follower_term;

### 2016-05-04 paper/paper.tex

@@
 with minimal effort on their part. Under the hood Cyclone decouples modification
-of local persistent memory from replication to maximize performance. In
-addition, unlike traditional systems that provide strong replication, Cyclone
-does require programs to be provisioned for checkpoints since NVM is directly
-manipulated by programmers and therefore up to date till the last executed
-transaction. Instead, Cyclone provides a novel NVM copy mechanism to bring up
-new nodes added to the cluster.
-
+of local persistent memory from replication to maximize performance. Unlike
+traditional systems that provide strong replication, the ``in-memory'' image is
+persistent and there is no need for checkpoints - except when new replicas are
+added. In addition linearizable reads can be directly executed without the need to
+combine pending updates from a log.
 \end{abstract}

### 2016-05-04 core/dispatcher.cpp
Bugfix.

@@ -274,17 +274,19 @@ void exec_rpc_internal_synchronous(rpc_info_t *rpc)
   volatile int execution_term;
   volatile bool is_leader;
   volatile bool have_data;
- while(!rpc->rep_success && !rpc->rep_failed);
+  while(!rpc->rep_success && !rpc->rep_failed);
   if(rpc->rep_success) {
     while(repeat) {
-      execution_term = cyclone_get_term(cyclone_handle);
+      execution_term = cyclone_get_term(cyclone_handle); // get current view
+      if(cyclone_get_leader(cyclone_handle) == -1) { // Make sure its stable
+	continue;
+      }
       is_leader = cyclone_is_leader(cyclone_handle);
       have_data = rpc->have_follower_data;
       user_tx_aborted = false;
       __sync_synchronize();
-      if(cyclone_get_term(cyclone_handle) != execution_term ||
-	 cyclone_get_leader(cyclone_handle) == -1) {
-	continue;
+      if(cyclone_get_term(cyclone_handle) != execution_term) {
+	continue; // Make sure view hasn't changed
       }
       repeat = false;
       TX_BEGIN(state) {

### 2016-05-04 test/rbtree_map_coordinator.cpp
Bugfixes.

@@ -10,7 +10,7 @@ void **quorum_handles;
 int me;
 int quorums;
 int clients;
-int *ctr;
+volatile int *ctr;
 
 TOID(char) nvheap_setup(TOID(char) recovered,
 			PMEMobjpool *state)
@@ -91,9 +91,9 @@ int leader_callback_recovery(const unsigned char *data,
 	  client_resp->tx_status = 0;
 	}
       }
-    }
-    if(response->cookie.txnum != *D_RO(txnum)) {
-      pmemobj_tx_abort(-1);
+      if(response->cookie.txnum != *D_RO(txnum)) {
+	pmemobj_tx_abort(-1);
+      }
     }
   }
 
@@ -129,6 +129,11 @@ int leader_callback_recovery(const unsigned char *data,
     req.kv_data = *info;
     req.src       = clients - 1;
     req.order     = 0;
+    req.cookie.phase       = COOKIE_PHASE_LOCK;
+    req.cookie.locks_taken = i;
+    req.cookie.index       = i;
+    req.cookie.success     = client_resp->tx_status;
+    req.cookie.txnum       = *D_RO(txnum);
 
     int sz = make_rpc(quorum_handles[partition],
 		      &req,
@@ -167,6 +172,13 @@ int leader_callback_recovery(const unsigned char *data,
     req.k_data.key = info->key;
     req.src       = clients - 1;
     req.order     = 0;
+    req.cookie.phase       = COOKIE_PHASE_VERSION;
+    req.cookie.locks_taken = tx->num_locks;
+    req.cookie.index       = i;
+    req.cookie.success     = client_resp->tx_status;
+    req.cookie.txnum       = *D_RO(txnum);
+
+
     int sz = make_rpc(quorum_handles[partition],
 		      &req,
 		      sizeof(struct proposal),
@@ -201,6 +213,12 @@ int leader_callback_recovery(const unsigned char *data,
     req.kv_data = *info;
     req.src       = clients - 1;
     req.order     = 0;
+    req.cookie.phase       = COOKIE_PHASE_INSERT;
+    req.cookie.locks_taken = tx->num_locks;
+    req.cookie.index       = i;
+    req.cookie.success     = client_resp->tx_status;
+    req.cookie.txnum       = *D_RO(txnum);
+
 
     int sz = make_rpc(quorum_handles[partition],
 		      &req,
@@ -227,6 +245,12 @@ int leader_callback_recovery(const unsigned char *data,
     req.kv_data = *info;
[…]

### 2016-05-04 core/clock.hpp
Cleanup tracing infra.

@@ -7,7 +7,7 @@
 class rtc_clock {
   unsigned long start_time;
   unsigned long elapsed_useconds;
-  unsigned long get_current_rtc()
+  static unsigned long get_current_rtc()
   {
     struct timeval tm;
     gettimeofday(&tm, NULL);
@@ -35,7 +35,7 @@ class rtc_clock {
   {
     return elapsed_useconds;
   }
-  unsigned long current_time()
+  static unsigned long current_time()
   {
     return get_current_rtc();
   }

### 2016-05-04 core/cyclone.cpp

@@ -4,14 +4,6 @@
 #include "timeouts.hpp"
 #include "checkpoint.hpp"
 
-#ifdef TRACING
-extern void trace_send_cmd(void *data, const int size);
-extern void trace_pre_append(void *data, const int size);
-extern void trace_post_append(void *data, const int size);
-extern void trace_send_entry(void *data, const int size);
-#endif
-
-
 void *cyclone_control_socket_out(void *cyclone_handle, 
 				 int replica)
 {
@@ -80,7 +72,6 @@ static int __send_appendentries(raft_server_t* raft,
   void *socket      = raft_node_get_udata(node);
   msg_t *msg = (msg_t *)cyclone_handle->cyclone_buffer_out;
   unsigned char *ptr = (unsigned char *)(msg + 1);
-  rtc_clock clock;
   int nodeidx = raft_node_get_id(node);
   bool same_log_point  =
     (m->prev_log_term == cyclone_handle->throttles[nodeidx].prev_log_term &&
@@ -88,12 +79,12 @@ static int __send_appendentries(raft_server_t* raft,
   bool prev_was_heartbeat = (cyclone_handle->throttles[nodeidx].prev_entries == 0);
   bool current_is_heartbeat = (m->n_entries == 0);
   if(same_log_point && (prev_was_heartbeat == current_is_heartbeat)) {
-    if((clock.current_time() - cyclone_handle->throttles[nodeidx].last_tx_time) <=
+    if((rtc_clock::current_time() - cyclone_handle->throttles[nodeidx].last_tx_time) <=
        cyclone_handle->throttles[nodeidx].timeout) {
       return 0; // throttle retx
     }
     else {
-      cyclone_handle->throttles[nodeidx].last_tx_time = clock.current_time();
+      cyclone_handle->throttles[nodeidx].last_tx_time = rtc_clock::current_time();
       cyclone_handle->throttles[nodeidx].timeout = 2*cyclone_handle->throttles[nodeidx].timeout;
     }
   }
@@ -101,7 +92,7 @@ static int __send_appendentries(raft_server_t* raft,
     cyclone_handle->throttles[nodeidx].prev_log_term = m->prev_log_term;
[…]

### 2016-05-04 core/cyclone_context.hpp

@@ -61,11 +61,6 @@ typedef struct
   };
 } msg_t;
 
-#ifdef TRACING
-extern void trace_recv_entry(void *data, const int len);
-extern void trace_recv_cmd(void *data, const int size);
-#endif
-
 struct throttle_st {
   int prev_log_term;
   int prev_log_idx;
@@ -444,9 +439,6 @@ typedef struct cyclone_st {
       memcpy(msg->ae.entries[i].data.buf, 
 	     ptr, 
 	     msg->ae.entries[i].data.len);
-#ifdef TRACING
-      trace_recv_entry(ptr, msg->ae.entries[i].data.len);
-#endif
       ptr += msg->ae.entries[i].data.len;
     }
     e = raft_recv_appendentries(raft_handle, 
@@ -464,9 +456,6 @@ typedef struct cyclone_st {
 					   &msg->aer);
       break;
     case MSG_CLIENT_REQ:
-#ifdef TRACING
-      trace_recv_cmd(msg->client.ptr, msg->client.size);
-#endif
       if(!cyclone_is_leader(this)) {
 	client_rep = NULL;
 	cyclone_tx(router->request_in(),
@@ -494,9 +483,6 @@ typedef struct cyclone_st {
       }
       break;
      case MSG_CLIENT_REQ_BATCH:
-#ifdef TRACING
-      trace_recv_cmd(msg->client.ptr, msg->client.size);
-#endif
       if(!cyclone_is_leader(this)) {
 	void * cookies = NULL;
 	cyclone_tx(router->request_in(),
@@ -534,9 +520,6 @@ typedef struct cyclone_st {
       }
       break;
     case MSG_CLIENT_REQ_CFG:
-#ifdef TRACING
-      trace_recv_cmd(msg->client.ptr, msg->client.size);
-#endif
       if(!cyclone_is_leader(this)) {
 	client_rep = NULL;
 	cyclone_tx(router->request_in(),
@@ -564,9 +547,6 @@ typedef struct cyclone_st {
       }
       break;
     case MSG_CLIENT_REQ_TERM:
-#ifdef TRACING
-      trace_recv_cmd(msg->client.ptr, msg->client.size);
-#endif
       if(!cyclone_is_leader(this)) {
 	client_rep = NULL;
 	cyclone_tx(router->request_in(),

### 2016-05-04 core/dispatch_client.cpp

@@ -101,9 +101,6 @@ typedef struct rpc_client_st {
     while(true) {
       packet_out->code        = RPC_REQ_LAST_TXID;
       packet_out->client_id   = me;
-#ifdef TRACING
-      packet_out->timestamp   = clock.current_time();
-#endif
       packet_out->client_txid = (int)packet_out->timestamp;
       packet_out->channel_seq = channel_seq++;
       packet_out->requestor   = me_mc;
@@ -160,9 +157,6 @@ typedef struct rpc_client_st {
     while(true) {
       packet_out->code        = RPC_REQ_NODEDEL;
       packet_out->client_id   = me;
-#ifdef TRACING
-      packet_out->timestamp   = clock.current_time();
-#endif
       packet_out->client_txid = txid;
       packet_out->channel_seq = channel_seq++;
       packet_out->requestor   = me_mc;
@@ -220,9 +214,6 @@ typedef struct rpc_client_st {
     while(true) {
       packet_out->code        = RPC_REQ_NODEADD;
       packet_out->client_id   = me;
-#ifdef TRACING      
-      packet_out->timestamp   = clock.current_time();
-#endif
       packet_out->client_txid = txid;
       packet_out->channel_seq = channel_seq++;
       packet_out->requestor   = me_mc;
@@ -280,9 +271,6 @@ typedef struct rpc_client_st {
     int resp_sz;
     packet_out->client_id   = me;
     packet_out->client_txid = txid;
-#ifdef TRACING    
-    packet_out->timestamp   = clock.current_time();
-#endif
     packet_out->channel_seq  = channel_seq++;
     packet_out->requestor   = me_mc;
     while(true) {
@@ -352,9 +340,6 @@ typedef struct rpc_client_st {
       packet_out->client_txid = txid;
       packet_out->channel_seq = channel_seq++;
       packet_out->requestor   = me_mc;
-#ifdef TRACING	
-      packet_out->timestamp   = clock.current_time();
-#endif
       memcpy(packet_out + 1, payload, sz);
       retcode = cyclone_tx_timeout(router->output_socket(server), 
 				   (unsigned char *)packet_out, 

### 2016-05-04 core/dispatcher.cpp

@@ -912,9 +912,8 @@ struct dispatcher_loop {
   
   void operator ()()
   {
-    rtc_clock clock;
     bool is_master = false;
-    unsigned long last_gc = clock.current_time();
+    unsigned long last_gc = rtc_clock::current_time();
     unsigned long last_batch = last_gc;
     int requests = 0;
     unsigned char * ptr;
@@ -930,10 +929,11 @@ struct dispatcher_loop {
 				  ptr,
 				  DISP_MAX_MSGSIZE,
 				  "DISP RCV");
-      unsigned long mark = clock.current_time();
+      unsigned long mark = rtc_clock::current_time();
       if(sz != -1) {
 	rx_sizes[requests++] = sz;
 	req = (rpc_t *)ptr;
+	req->timestamp = mark;
 	ptr += sz;
 	if(req->code == RPC_REQ_NODEADD || req->code == RPC_REQ_NODEDEL) {
 	  if(requests > 1) {

### 2016-05-04 core/clock.hpp
Overhauled timing.

@@ -7,6 +7,7 @@
 class rtc_clock {
   unsigned long start_time;
   unsigned long elapsed_useconds;
+  const char *msg;
   static unsigned long get_current_rtc()
   {
     struct timeval tm;
@@ -15,34 +16,13 @@ class rtc_clock {
   }
 
  public:
- rtc_clock()
-   :elapsed_useconds(0)
-    {
-    }
-  void start()
+ rtc_clock(const char *msg_in)
+   :msg(msg_in)
   {
-    start_time = get_current_rtc();
-  }
-  void stop()
-  {
-    elapsed_useconds += (get_current_rtc() - start_time);
-  }
-  void reset()
-  {
-    elapsed_useconds = 0;
-  }
-  unsigned long elapsed_time()
-  {
-    return elapsed_useconds;
   }
   static unsigned long current_time()
   {
     return get_current_rtc();
-  }
-  void print(const char header[])
-  {
-    double elapsed_seconds = (double)elapsed_useconds/1000000;
-    BOOST_LOG_TRIVIAL(info) << header << " " << elapsed_seconds << " seconds";
   }
 };
 

### 2016-05-04 core/cyclone_comm.hpp

@@ -36,9 +36,8 @@ static int cyclone_tx_timeout(void *socket,
 			      unsigned long timeout_usecs,
 			      const char *context) 
 {
-  rtc_clock clock;
   int rc;
-  clock.start();
+  unsigned long mark = rtc_clock::current_time();
   while (true) {
     rc = zmq_send(socket, data, size, ZMQ_NOBLOCK);
     if(rc >= 0) {
@@ -53,11 +52,9 @@ static int cyclone_tx_timeout(void *socket,
 	exit(-1);
       }
     }
-    clock.stop();
-    if(clock.elapsed_time() >= timeout_usecs) {
+    if((rtc_clock::current_time() - mark) >= timeout_usecs) {
       break;
     }
-    clock.start();
   }
   return rc;
 }
@@ -94,8 +91,7 @@ static int cyclone_rx_timeout(void *socket,
 			      const char *context)
 {
   int rc;
-  rtc_clock clock;
-  clock.start();
+  unsigned long mark = rtc_clock::current_time();
   while (true) {
     rc = zmq_recv(socket, data, size, ZMQ_NOBLOCK);
     if(rc >= 0) {
@@ -108,11 +104,9 @@ static int cyclone_rx_timeout(void *socket,
 	<< zmq_strerror(zmq_errno());
       exit(-1);
     }
-    clock.stop();
-    if(clock.elapsed_time() >= timeout_usecs) {
+    if((rtc_clock::current_time() - mark) >= timeout_usecs) {
       break;
     }
-    clock.start();
   }
   return rc;
 }

### 2016-05-04 core/cyclone_context.hpp

@@ -625,8 +625,8 @@ struct cyclone_monitor {
   
   void operator ()()
   {
-    rtc_clock timer;
-    timer.start();
+    unsigned long mark = rtc_clock::current_time();
+    unsigned long elapsed_time;
     while(!terminate) {
       // Handle any outstanding requests
       for(int i=0;i<cyclone_handle->replicas;i++) {
@@ -647,14 +647,11 @@ struct cyclone_monitor {
 	  cyclone_handle->handle_incoming(sz);
 	}
       }
-      timer.stop();
-      int elapsed_time = timer.elapsed_time();
       // Handle periodic events -- - AFTER any incoming requests
-      if(elapsed_time >= PERIODICITY) {
-	raft_periodic(cyclone_handle->raft_handle, elapsed_time);
-	timer.reset();
+      if((elapsed_time = rtc_clock::current_time() - mark) >= PERIODICITY) {
+	raft_periodic(cyclone_handle->raft_handle, (int)elapsed_time);
+	mark = rtc_clock::current_time();
       }
-      timer.start();
     }
   }
 };

### 2016-05-04 core/dispatch_client.cpp

@@ -97,7 +97,6 @@ typedef struct rpc_client_st {
   {
     int retcode;
     int resp_sz;
-    rtc_clock clock;
     while(true) {
       packet_out->code        = RPC_REQ_LAST_TXID;
       packet_out->client_id   = me;
@@ -153,7 +152,6 @@ typedef struct rpc_client_st {
   {
     int retcode;
     int resp_sz;
-    rtc_clock clock;
     while(true) {
       packet_out->code        = RPC_REQ_NODEDEL;
       packet_out->client_id   = me;
@@ -210,7 +208,6 @@ typedef struct rpc_client_st {
   {
     int retcode;
     int resp_sz;
-    rtc_clock clock;
     while(true) {
       packet_out->code        = RPC_REQ_NODEADD;
       packet_out->client_id   = me;
@@ -266,7 +263,6 @@ typedef struct rpc_client_st {
 
   int retrieve_response(void **response, int txid)
   {
-    rtc_clock clock;
     int retcode;
     int resp_sz;
     packet_out->client_id   = me;
@@ -331,7 +327,6 @@ typedef struct rpc_client_st {
   {
     int retcode;
     int resp_sz;
-    rtc_clock clock;
     while(true) {
       // Make request
       packet_out->code        = RPC_REQ_FN;
@@ -401,7 +396,6 @@ void* cyclone_client_init(int client_id,
 			  const char *config_server,
 			  const char *config_client)
 {
-  rtc_clock clock;
   rpc_client_t * client = new rpc_client_t();
   client->me = client_id;
   client->me_mc = client_mc;
@@ -423,7 +417,7 @@ void* cyclone_client_init(int client_id,
   buf = new char[DISP_MAX_MSGSIZE];
   client->packet_out_doorbell = (rpc_t *)buf;
   client->replicas = replicas;
-  client->channel_seq = clock.current_time();
+  client->channel_seq = client_id*client_mc*rtc_clock::current_time();
   client->rung_doorbell = new bool[replicas];
   for(int i=0;i<replicas;i++) {
     client->rung_doorbell[i] = false;

### 2016-05-04 core/dispatcher_exec.cpp

@@ -36,52 +36,4 @@ void exec_send_checkpoint(void *socket, void *handle)
   ioService2.post(boost::bind(send_checkpoint, socket, handle));
 }
 
-static void print(const char *prefix,
-		  const void *data,
-		  const int size)
-{
-  rtc_clock timer;
-  const rpc_t *rpc = (const rpc_t *)data;
-  if(rpc->code == RPC_REQ_MARKER) {
-    BOOST_LOG_TRIVIAL(info)
-      << "KICKER";
-    return;
-  }
-  BOOST_LOG_TRIVIAL(info)
-    << prefix << " "
-    << rpc->client_id  << " "
-    << rpc->client_txid << " "
-    << (timer.current_time() - rpc->timestamp);
-}
-
-void trace_send_cmd(void *data, const int size)
-{
-  print("SERVER: SEND_CMD", data, size);
-}
-
-void trace_recv_cmd(void *data, const int size)
-{
-  print("SERVER: RECV_CMD", data, size);
-}
-
-void trace_pre_append(void *data, const int size)
-{
-  print("SERVER: PRE_APPEND", data, size);
-}
-
-void trace_post_append(void *data, const int size)
-{
-  print("SERVER: POST_APPEND", data, size); 
-}
-
-void trace_send_entry(void *data, const int size)
-{
-  print("SERVER: SEND_ENTRY", data, size);
-}
-
-void trace_recv_entry(void *data, const int size)
-{
-  print("SERVER: RECV_ENTRY", data, size);
-}
-
 

### 2016-05-04 test/cyclone_test.cpp

@@ -7,13 +7,11 @@
 #include "../core/logging.hpp"
 #include "../core/timeouts.hpp"
 
-rtc_clock timer;
-
 void print(const char *prefix,
 	   void *data,
 	   const int size)
 {
-  unsigned long elapsed_usecs = timer.current_time();
+  unsigned long elapsed_usecs = rtc_clock::current_time();
   char *buf = (char *)data;
   if(size != 16) {
     BOOST_LOG_TRIVIAL(fatal) << "SERVER: Incorrect record size";
@@ -88,7 +86,6 @@ int main(int argc, char *argv[])
     printf("Usage %s node_id replicas\n", argv[0]);
     exit(-1);
   }
-  timer.start();
   unsigned int node_id = atoi(argv[1]);
   unsigned int replicas = atoi(argv[2]);
   unsigned char entry[16];
@@ -108,7 +105,7 @@ int main(int argc, char *argv[])
     *(unsigned int *)(entry + 4) = ctr;
     void *cookie;
     do {
-      *(unsigned long *)(entry + 8) = timer.current_time();
+      *(unsigned long *)(entry + 8) = rtc_clock::current_time();
       cookie = cyclone_add_entry(cyclone_handle, entry, 16);
       if(cookie == NULL) {
 	BOOST_LOG_TRIVIAL(info) << "CLIENT: Not on master";
@@ -126,7 +123,7 @@ int main(int argc, char *argv[])
       }
     } while(true);
     free(cookie);
-    unsigned long elapsed_usecs = timer.current_time();
+    unsigned long elapsed_usecs = rtc_clock::current_time();
     if(result == 1) {
       print("CLIENT: ACCEPT ", (void *)entry, 16);
     }

### 2016-05-04 test/dispatch_test_client.cpp

@@ -4,13 +4,11 @@
 #include <libcyclone.hpp>
 #include <stdio.h>
 
-rtc_clock timer;
-
 static void print(const char *prefix,
 		  const void *data,
 		  const int size)
 {
-  unsigned long elapsed_usecs = timer.current_time();
+  unsigned long elapsed_usecs = rtc_clock::current_time();
   char *buf = (char *)data;
   if(size != 16) {
     BOOST_LOG_TRIVIAL(fatal) << "CLIENT: Incorrect record size";
@@ -44,7 +42,7 @@ int main(int argc, char *argv[])
   while(true) {
     *(unsigned int *)&proposal[0] = me;
     *(unsigned int *)&proposal[4] = ctr;
-    *(unsigned long *)&proposal[8] = timer.current_time();
+    *(unsigned long *)&proposal[8] = rtc_clock::current_time();
     void *resp;
     print("PROPOSE", proposal, 16);
     int sz = make_rpc(handle, proposal, 16, &resp, ctr, 0);

### 2016-05-04 test/rbtree_map_add_node.cpp

@@ -46,7 +46,6 @@
 
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 9) {
     printf("Usage: %s client_id mc_id replicas clients server_prefix client_prefix partition replica\n", argv[0]); 
     exit(-1);

### 2016-05-04 test/rbtree_map_client.cpp

@@ -48,7 +48,6 @@ static uint64_t nkeys;
 static uint64_t keys[MAX_INSERTS];
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 4) {
     printf("Usage: %s client_id replicas clients\n", argv[0]);
     exit(-1);
@@ -76,26 +75,26 @@ int main(int argc, const char *argv[]) {
     insert_data.value = rand();
     prop->fn = FN_INSERT;
     prop->kv_data = insert_data;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
     sz = make_rpc(handle, buffer, sizeof(struct proposal), &resp, ctr, 0);
     ctr++;
     BOOST_LOG_TRIVIAL(info) << "RPC TIME = " 
-			    << (clock.current_time() - prop->timestamp);
+			    << (rtc_clock::current_time() - prop->timestamp);
     keys[nkeys++] = insert_data.key;
   }
   for (int i = 0; i < MAX_INSERTS; ++i) {
     query_data.key = keys[i];
     prop->fn = FN_LOOKUP;
     prop->k_data = query_data;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
     sz = make_rpc(handle, buffer, sizeof(struct proposal), &resp, ctr, 0);
     ctr++;
     BOOST_LOG_TRIVIAL(info) << "RPC TIME = " 
-			    << (clock.current_time() - prop->timestamp);
+			    << (rtc_clock::current_time() - prop->timestamp);
     struct proposal *rep = (struct proposal *)resp;
     if(rep->code != CODE_OK) {
       BOOST_LOG_TRIVIAL(error) << "Key not found !";

### 2016-05-04 test/rbtree_map_coordinator_driver.cpp

@@ -48,7 +48,6 @@
 #define KEYS 191
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 10) {
     printf("Usage: %s client_id replicas clients sleep_usecs server_config  client_config partitions quorum_server_prefix quorum_client_prefix\n", argv[0]);
     exit(-1);
@@ -87,7 +86,7 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   BOOST_LOG_TRIVIAL(info) << "Connecting to coordinator";
   int ctr = get_last_txid(handle) + 1;
@@ -100,7 +99,7 @@ int main(int argc, const char *argv[]) {
     struct proposal vquery;
     vquery.fn = FN_GET_VERSION;
     vquery.k_data.key   = me*KEYS + key;
-    vquery.timestamp = clock.current_time();
+    vquery.timestamp = rtc_clock::current_time();
     vquery.src       = me;
     vquery.order     = 0;
     int q = vquery.k_data.key % partitions;
@@ -118,7 +117,7 @@ int main(int argc, const char *argv[]) {
     infolock->value = ((struct proposal *)resp)->code;
     infok->key   = me*KEYS + key;
     infok->value = 0xdeadbeef;
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     sz = make_rpc(handle, buffer, size_tx(tx), &resp, ctr, RPC_FLAG_SYNCHRONOUS);
     if(sz != sizeof(tx_client_response)) {
       fprintf(stderr, "unexepected response to tx of size %d", sz);
@@ -129,17 +128,17 @@ int main(int argc, const char *argv[]) {
       exit(-1);
     }
     ctr++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
[…]

### 2016-05-04 test/rbtree_map_coordinator_load.cpp

@@ -48,7 +48,6 @@
 #define KEYS 100
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 7) {
     printf("Usage: %s client_id replicas clients sleep_usecs server_config client_config\n", argv[0]);
     exit(-1);
@@ -70,7 +69,7 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   int ctr = get_last_txid(handle) + 1;
   
@@ -79,20 +78,20 @@ int main(int argc, const char *argv[]) {
     struct kv *infok = inserts_list(tx, 0);
     infok->key   = me*KEYS + i;
     infok->value = me*KEYS + i;
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     sz = make_rpc(handle, buffer, sizeof(rbtree_tx_t), &resp, ctr, RPC_FLAG_SYNCHRONOUS);
     ctr++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			      << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			      << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			      << " tx/sec "
 			      << "LATENCY = "
 			      << ((double)total_latency)/tx_block_cnt
 			      << " us ";
-      tx_block_begin = clock.current_time();
+      tx_block_begin = rtc_clock::current_time();
       tx_block_cnt   = 0;
       total_latency  = 0;
     }
@@ -100,7 +99,7 @@ int main(int argc, const char *argv[]) {
 
   total_latency = 0;
   tx_block_cnt  = 0;
-  tx_block_begin = clock.current_time();
+  tx_block_begin = rtc_clock::current_time();
   BOOST_LOG_TRIVIAL(info) << "Complete";
   return 0;
 }

### 2016-05-04 test/rbtree_map_driver.cpp

@@ -46,7 +46,6 @@
 #define KEYS 100
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 5) {
     printf("Usage: %s client_id replicas clients sleep_usecs\n", argv[0]);
     exit(-1);
@@ -67,7 +66,7 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   int ctr = get_last_txid(handle) + 1;
   
@@ -75,23 +74,23 @@ int main(int argc, const char *argv[]) {
     prop->fn = FN_INSERT;
     prop->kv_data.key   = me*KEYS + i;
     prop->kv_data.value = me*KEYS + i;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     sz = make_rpc(handle, buffer, sizeof(struct proposal), &resp, ctr, 0);
     ctr++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			      << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			      << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			      << " tx/sec "
 			      << "LATENCY = "
 			      << ((double)total_latency)/tx_block_cnt
 			      << " us ";
-      tx_block_begin = clock.current_time();
+      tx_block_begin = rtc_clock::current_time();
       tx_block_cnt   = 0;
       total_latency  = 0;
     }
@@ -100,29 +99,29 @@ int main(int argc, const char *argv[]) {
 
   total_latency = 0;
   tx_block_cnt  = 0;
[…]

### 2016-05-04 test/rbtree_map_partitioned_driver.cpp

@@ -49,7 +49,6 @@
 #define KEYS 191
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 8) {
     printf("Usage: %s client_id replicas clients sleep_usecs partitions server_config_prefix client_config_prefix\n", argv[0]);
     exit(-1);
@@ -80,7 +79,7 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   
   int ctr[partitions];
@@ -94,10 +93,10 @@ int main(int argc, const char *argv[]) {
     prop->fn = FN_INSERT;
     prop->kv_data.key   = me*KEYS + i;
     prop->kv_data.value = me*KEYS + i;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     int partition = prop->kv_data.key % partitions;
     sz = make_rpc(handles[partition],
 		  buffer,
@@ -106,17 +105,17 @@ int main(int argc, const char *argv[]) {
 		  ctr[partition],
 		  0);
     ctr[partition]++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			      << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			      << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			      << " tx/sec "
 			      << "LATENCY = "
 			      << ((double)total_latency)/tx_block_cnt
 			      << " us ";
-      tx_block_begin = clock.current_time();
+      tx_block_begin = rtc_clock::current_time();
       tx_block_cnt   = 0;
[…]

### 2016-05-04 test/rbtree_map_ro_driver.cpp

@@ -46,7 +46,6 @@
 #define KEYS 100
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 5) {
     printf("Usage: %s client_id replicas clients sleep_usecs\n", argv[0]);
     exit(-1);
@@ -67,30 +66,30 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   int ctr = get_last_txid(handle) + 1;
   for(int i=0;i<KEYS;i++) {
     prop->fn = FN_INSERT;
     prop->kv_data.key   = me*KEYS + i;
     prop->kv_data.value = me*KEYS + i;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     sz = make_rpc(handle, buffer, sizeof(struct proposal), &resp, ctr, 0);
     ctr++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			      << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			      << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			      << " tx/sec "
 			      << "LATENCY = "
 			      << ((double)total_latency)/tx_block_cnt
 			      << " us ";
-      tx_block_begin = clock.current_time();
+      tx_block_begin = rtc_clock::current_time();
       tx_block_cnt   = 0;
       total_latency  = 0;
     }
@@ -98,16 +97,16 @@ int main(int argc, const char *argv[]) {
 
   total_latency = 0;
   tx_block_cnt  = 0;
[…]

### 2016-05-04 test/rbtree_map_server.cpp

@@ -75,25 +75,24 @@ int callback(const unsigned char *data,
 	     void **return_value)
 {
   // Heartbeat
-  rtc_clock clock;
   static unsigned long tx_block_cnt   = 0;
-  static unsigned long tx_block_begin = clock.current_time();
+  static unsigned long tx_block_begin = rtc_clock::current_time();
   static unsigned long total_latency = 0;
 
-  if(clock.current_time() - tx_block_begin >= 5000000 &&
+  if(rtc_clock::current_time() - tx_block_begin >= 5000000 &&
      tx_block_cnt > 0) {
     BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			    << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			    << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			    << " tx/sec "
 			    << "LATENCY ="
 			    << ((double)total_latency)/tx_block_cnt
 			    << " us ";
-    tx_block_begin = clock.current_time();
+    tx_block_begin = rtc_clock::current_time();
     tx_block_cnt   = 0;
     total_latency  = 0;
   }
   tx_block_cnt++;
-  unsigned long tx_begin_time = clock.current_time();
+  unsigned long tx_begin_time = rtc_clock::current_time();
   //////////////
 
   struct proposal *req = (struct proposal *)data;
@@ -191,7 +190,7 @@ int callback(const unsigned char *data,
     BOOST_LOG_TRIVIAL(fatal) << "Tree: unknown fn !";
     exit(-1);
   }
-  total_latency += (clock.current_time() - tx_begin_time);
+  total_latency += (rtc_clock::current_time() - tx_begin_time);
   return sizeof(struct proposal);
 }
 

### 2016-05-04 test/rbtree_map_synchronous_driver.cpp

@@ -46,7 +46,6 @@
 #define KEYS 100
 
 int main(int argc, const char *argv[]) {
-  rtc_clock clock;
   if(argc != 5) {
     printf("Usage: %s client_id replicas clients sleep_usecs\n", argv[0]);
     exit(-1);
@@ -67,31 +66,31 @@ int main(int argc, const char *argv[]) {
   void *resp;
   unsigned long order = 0;
   unsigned long tx_block_cnt   = 0;
-  unsigned long tx_block_begin = clock.current_time();
+  unsigned long tx_block_begin = rtc_clock::current_time();
   unsigned long total_latency  = 0;
   int ctr = get_last_txid(handle) + 1;
   for(int i=0;i<KEYS;i++) {
     prop->fn = FN_INSERT;
     prop->kv_data.key   = me*KEYS + i;
     prop->kv_data.value = me*KEYS + i;
-    prop->timestamp = clock.current_time();
+    prop->timestamp = rtc_clock::current_time();
     prop->src       = me;
     prop->order     = (order++);
-    unsigned long tx_begin_time = clock.current_time();
+    unsigned long tx_begin_time = rtc_clock::current_time();
     sz = make_rpc(handle, buffer, sizeof(struct proposal), &resp, ctr,
 		  RPC_FLAG_SYNCHRONOUS);
     ctr++;
-    total_latency += (clock.current_time() - tx_begin_time);
+    total_latency += (rtc_clock::current_time() - tx_begin_time);
     usleep(sleep_time);
     tx_block_cnt++;
-    if(clock.current_time() - tx_block_begin >= 10000) {
+    if(rtc_clock::current_time() - tx_block_begin >= 10000) {
       BOOST_LOG_TRIVIAL(info) << "LOAD = "
-			      << ((double)1000000*tx_block_cnt)/(clock.current_time() - tx_block_begin)
+			      << ((double)1000000*tx_block_cnt)/(rtc_clock::current_time() - tx_block_begin)
 			      << " tx/sec "
 			      << "LATENCY = "
 			      << ((double)total_latency)/tx_block_cnt
 			      << " us ";
-      tx_block_begin = clock.current_time();
+      tx_block_begin = rtc_clock::current_time();
       tx_block_cnt   = 0;
       total_latency  = 0;
     }
@@ -99,30 +98,30 @@ int main(int argc, const char *argv[]) {
 
   total_latency = 0;
   tx_block_cnt  = 0;
[…]

### 2016-05-04 core/clock.hpp
Add sampling to clock.

@@ -5,9 +5,10 @@
 
 //! Real time clock
 class rtc_clock {
-  unsigned long start_time;
-  unsigned long elapsed_useconds;
   const char *msg;
+  unsigned long dump_interval;
+  unsigned long accumulator;
+  unsigned long samples;
   static unsigned long get_current_rtc()
   {
     struct timeval tm;
@@ -16,13 +17,26 @@ class rtc_clock {
   }
 
  public:
- rtc_clock(const char *msg_in)
-   :msg(msg_in)
+  rtc_clock(const char *msg_in, 
+	    unsigned long dump_interval_in)
+    :msg(msg_in),
+     dump_interval(dump_interval_in)
   {
+    accumulator = 0;
+    samples = 0;
   }
   static unsigned long current_time()
   {
     return get_current_rtc();
+  }
+  void sample(unsigned long sample)
+  {
+    accumulator += sample;
+    samples++;
+    if(samples >= dump_interval) {
+      BOOST_LOG_TRIVIAL(info) << msg
+			      << ((double)accumulator/samples);
+    }
   }
 };
 

### 2016-05-05 core/dispatcher.cpp
Aggressive batching.

@@ -935,6 +935,9 @@ struct dispatcher_loop {
 	req = (rpc_t *)ptr;
 	req->timestamp = mark;
 	ptr += sz;
+	if(requests == 1) {
+	  last_batch = mark;
+	}
 	if(req->code == RPC_REQ_NODEADD || req->code == RPC_REQ_NODEDEL) {
 	  if(requests > 1) {
 	    handle_batch_rpc(rx_sizes, requests - 1);
@@ -944,7 +947,6 @@ struct dispatcher_loop {
 	  handle_batch_rpc(rx_sizes, 1);
 	  requests = 0;
 	  ptr = rx_buffers;
-	  last_batch = mark;
 	}
 	else if(req->code == RPC_DOORBELL) {
 	  router->ring_doorbell(zmq_context,
@@ -968,7 +970,6 @@ struct dispatcher_loop {
 	  handle_batch_rpc(rx_sizes, requests);
 	  requests = 0;
 	  ptr = rx_buffers;
-	  last_batch = mark;
 	}
       }
       else if(requests > 0 && 
@@ -976,7 +977,6 @@ struct dispatcher_loop {
 	handle_batch_rpc(rx_sizes, requests);
 	requests = 0;
 	ptr = rx_buffers;
-	last_batch = mark;
       }
 
       if((mark - last_gc) >= PERIODICITY) {

### 2016-05-09 paper/paper.tex
Updates to writeup.

@@
 \usepackage{times}
-
+\usepackage{parskip}
 \begin{document}
@@
 
-\section{Introduction}
+%\section{Introduction}
 
@@
 Cyclone provides users with a replicated and durable heap that can be
-manipulated via a set of library routines. The library provides special
-functions to allocate and free persistent memory as well as fat pointers that
-wrap pointers to persistent memory to ensure that they do not change across
-restarts - since the virtual memory ranges into which persistent memory is
-mapped may change.
+manipulated via a set of library routines. Cyclone builds on top of Intel's Non
+Volatile Memory Library (NVML) and provides transparent replication to programs
+written to use NVML for persistence on a single node. This section therefore
+describes NVML's programming model.
+
+NVML provides special functions to allocate and free persistent memory as well
+as fat pointers to persistent memory to ensure that referential integrity in the
+heap is maintained across restarts - since the virtual memory ranges into which
+persistent memory is mapped may change.
 
 \begin{figure}
-{ \scriptsize
+{\scriptsize
 \begin{verbatim}
@@
 
-void insert_after(TOID(ll_node) prev, TOID(ll_node) new_node)
+void insert_after(TOID(ll_node) prev, 
+                  TOID(ll_node) new_node)
 {
@@
 type therefore is a pointer to a linked list node stored in persistent memory.
-The {\tt insert\_after} function used the {\tt D\_RO} and {\tt D\_RW} calls to
+The {\tt insert\_after} function uses the {\tt D\_RO} and {\tt D\_RW} calls to
 convert fat pointers into directly accessible read-only and read-write pointers
@@
 
-Cyclone provides tools for programmers to write code that is failure
-atomic. Portions of the code are delimited as transactions that are either
-completely executed in terms of its effects on persistent memory or not executed
-at all. Under the hood, Cyclone tracks all updates to persistent memory in a
-persistent undo log - the log entries are persisted before the update to the
-actual memory is allowed. On a failure, the undo log is used to roll back
-incomplete transactions. A transaction is considered complete when it completes
-execution and the updated objects are flushed (automatically by Cyclone) to
-persistent memory. The log entries for the transaction are then immediately
-deleted. The only burden placed on a programmer for failure atomicity is that
-they must inform the Cyclone runtime about objects that are
-updated. Figure~\ref{fig:example_fatomic} shows how to convert the function for
-adding a node to a linked list into a failure atomic one. The only additions are
-the transaction delimiters and the {\tt TX\_ADD} calls to mark the objects being
-updated \emph{before} the actual updates to them. Figure~\ref{fig:undo_log}
-shows how Cyclone maintains the undo-log of objects.
+NVML provides tools for programmers to write code that is failure
+atomic by allowing a block of code to be delimited as a transaction. A
+transaction is either completely executed in terms of its effects on
+persistent memory or not executed at all. Under the hood, NVML tracks all
+updates to persistent memory in a persistent undo log - the log entries are
+persisted before the update to the actual memory is allowed. On a failure, the
+undo log is used to roll back incomplete transactions. A transaction is
+considered complete when it completes execution and the updated objects are
+flushed (automatically by NVML) to persistent memory. The log entries for the
+transaction are then immediately deleted. The only burden placed on a programmer
+for failure atomicity is to delimit transactions and that they must inform NVML
+about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
+convert the function for adding a node to a linked list into a failure atomic
+one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
+calls to mark the objects being updated \emph{before} the actual updates to
+them. Figure~\ref{fig:undo_log} shows how Cyclone maintains the undo-log of
+objects.
 
@@
 }
-\caption{A Persistent Linked List of Integers}
+\caption{Failure Atomic Peristent Linked List}
 \label{fig:example_fatomic}
@@
 
-A crucial feature of Cyclone is that it abstracts the details of the persistence
+A crucial feature of NVML is that it abstracts the details of the persistence
 mechanism away from programmers. This allows the same code to be deployed on
-clusters with a persistent storage medium such as 3D XPoint or on clusters with
-battery backed NVDIMMs for example. In the former case, explicit {\tt PCOMMIT}
-instructions are necessary for persistence [ref PMFS paper], while in the latter
-case no such instructions are necessary as on failure the CPU caches and DRAM
-are flushed to stable storage under battery (or capacitor) power. Even in the
-NVDIMM case however, the undo log can be desirable to protect against OS or
-application crashes to ensure that the persistent memory is always left in a
-consistent state. Cyclone's programmer visible features are designed to enable
-the construction of applications that work regardless of the underlying
-persistence mechancisms.
+clusters with a range of persistence mechanisms - mapping pages from secondary
+storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
+save DRAM state to SSD on failure [FARM] or using directly accessed persistent
+memory [PMFS]. Each of these options has different requirements. For example the
+NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
+instructions. NVDIMMs would conceivably require only a CLFLUSH to flush the
+cacheline and ensure that the line enters the asynchronous data refresh
+domain. Using secondary storage is also an option with faster 3DXPoint storage
+behind an NVMe interface - this would require an {\tt msync} and OS pagecache
+intervention to persist data to secondary storage. NVML seamlessly uses the
+right persistence mechanism without requiring code using the library to be aware
+of these details.
 
 \section{Replication}
-Cyclone replicates the state of the NVM heap across multiple machines in a
+Cyclone replicates the state of the NVML heap across multiple machines in a
 cluster. Cyclone provides strong consistency - a majority of machines in the
@@
 decouple replication from execution of user code modifying NVM. It does so by
-encapsulating changes to the NVM heap in RPC calls driven by client requests.
-We chose this model since it closely reflects the use cases most suitable for
-NVM-based clusters.
+encapsulating changes to the NVM heap in server side code driven by client RPC
+requests. We chose this model since it closely reflects the use cases most
+suitable for NVM-based clusters.
 
@@
   non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
-  raft paper] under the hood for consensus on the contents of the log of RPC
-calls - although any other similar state machine replication protocol such as
-Paxos is also suitable. Cyclone provides the means to achieve exactly once
-semantics to clients by numbering RPC calls and remembering the result of the
-last executed RPC call.
+  raft paper] under the hood for consensus on the log of RPC calls - although
+any other similar state machine replication protocol such as
+Paxos is also suitable. Cyclone provides the means to clients to achieve exactly
+once semantics for the RPC calls they make by numbering RPC calls and
+remembering the number and result of the last executed RPC call at the server end.
 
@@
 Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
-from a client are numbered and Cyclone remembers the results of the last RPC
-call made by a client. A particular client can exist on any machine - cyclone
-treats all instances of the client as a single client. Cyclone rejects RPC calls
-that are not in sequence. Any RPC call is executed exactly once across the
-quorum - regardless of failure and reconfigurations. This provides the means for
-clients to easily achieve exactly once semantics, provided the client does not
-crash. Cyclone can supply the last executed RPC call number and its result to
-assist recovering and stateless clients.
+from a client are numbered and Cyclone remembers (in a persistent manner) the
+number and result of the last RPC call made by a client. A particular client can
+exist on any machine - cyclone treats all instances of the client as a single
+client. Cyclone rejects RPC calls that are not in sequence. Any RPC call is
+executed exactly once across the quorum - regardless of failure and
+reconfigurations. This provides the means for clients to easily achieve exactly
+once semantics, provided the client does not crash. Cyclone can supply the last
+executed RPC call number and its result to assist recovering and stateless
+clients.
 
@@
 overlapping the update of non-volatile state with the network I/O associated
-with replication. Persisting RAFT log entries, making the upcall to the
-application and executing the RPC call in a failure atomic transaction takes a
+with replication. Making the call to the application and executing the it in a
+failure atomic transaction takes a
 significant amount of time and therefore it makes sense to overlap this with the
@@
 a new leader getting elected in the underlying replication protocol
-(RAFT). Since failure atomic transactions can be rolled back, we are able to
-speculatively apply the changes to NVM without waiting for replication to
-finish.
+(we refer to this as a view change). Since failure atomic transactions can be
+rolled back, we are able to speculatively apply the changes to NVM without
+waiting for replication to finish.
 
@@
   Execute RPC call and modify local NVM
-  // Block till replication result known
+  Block till replication result known
   if(replication failed) {
@@
 Asynchronous replication provides good performance but lacks guarantees
-occasionally necessary for distributed systems paradigms such as output commit and
-the ability to integrate code that has non-deterministic execution paths or side
-effects - such as reading the time. Cyclone therefore provides programmers the
-ability to closely synchronize execution across the replicas using synchronous
+occasionally necessary for distributed systems paradigms such as output commit
+and the ability to integrate code that has non-deterministic execution paths
+(such as being dependent on the current time) or side effects - such as making
+further requests. Cyclone therefore provides programmers the ability to closely
+synchronize execution across the replicas using synchronous
 replication. Synchronous replication extends the programming model provided by
-cyclone - but neccessitates some awareness on the part of the user about
-the difference between the leader and follower replicas. We show later
+cyclone - but neccessitates some awareness on the part of the user about the
+difference between the leader and follower replicas. We show later
 (Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
@@
 will be committed to the replicas. This provides the output commit property
-necessary for some fault tolerance features in applications. Next the RPC call
-is executed on the leader and the result of the RPC call is replicated to the
-followers. The followers then proceed to execute the RPC call locally taking
-into account the result from the leader. This result can include data to
-synchronize non-deterministic steps such as reading the time on the replica.
+necessary in some applications.
+
+Next, the RPC call is executed on the leader and the result of the RPC call is
+replicated to the followers. The followers then proceed to execute the RPC call
+locally taking into account the result from the leader. This result can include
+data to synchronize non-deterministic steps such as reading the time on the
+replica.
 
@@
 result within the current view. This ensures that in the event of a network
-partition and later repair an old execution of a previous leader cannot commit a
+partition and later repair, an old execution of a previous leader cannot commit a
 log record with results from a synchronously executed RPC call.
@@
 \begin{figure*}
+\centering
+\begin{minipage}{0.7\textwidth}  
 { \scriptsize
@@
 Initiate replication of RPC call and arguments
-while(true) {
+Block till replication of RPC call returns
+if(replication success) {  
+  while (true) { // Must complete 
+    view = current RAFT view (term)
     TX_BEGIN {
-    // Block till replication result known
-    if(replication failed) {
-      TX_ABORT;
-    }  
       if(is_leader && !have_execution_result) {
         Execute RPC call and modify local NVM
-      replicate rpc call result in current view;
-      // Block till replication result known
+        replicate result in view;
+        Block till result replication returns
         if(result replication failed) {
@@
       else {
-      // Block till replication done or view changes
-      if(!result replication complete) {
+        Block till committed result replication or view changes
+        if(!committed result replication) {
           TX_ABORT;
@@
       }
+    } TX_ONABORT {
+      continue; // Try again
+    } TX_END
+    break; // All done
   }
-} TX_END
+}
 \end{verbatim}
 }
+\end{minipage}
 \caption{Synchronous Replication}

### 2016-05-10 core/cyclone.cpp
Remove unnecessary allocation and copying.

@@ -260,7 +260,6 @@ static int __raft_logentry_offer(raft_server_t* raft,
 				   ety->term);
     
   }
-  free(chunk); // release temporary memory
   return result;
 }
 
@@ -299,7 +298,6 @@ static int __raft_logentry_offer_batch(raft_server_t* raft,
 				     e->term);
 
     }
-    free(e->data.buf);
     e->data.buf = spot;
   }
   persist_to_circular_log(cyclone_handle->pop_raft_state, log,

### 2016-05-10 core/cyclone_context.hpp

@@ -435,10 +435,7 @@ typedef struct cyclone_st {
     msg->ae.entries = (msg_entry_t *)payload;
     ptr = (char *)(payload + msg->ae.n_entries*sizeof(msg_entry_t));
     for(int i=0;i<msg->ae.n_entries;i++) {
-      msg->ae.entries[i].data.buf = malloc(msg->ae.entries[i].data.len);
-      memcpy(msg->ae.entries[i].data.buf, 
-	     ptr, 
-	     msg->ae.entries[i].data.len);
+      msg->ae.entries[i].data.buf = ptr;
       ptr += msg->ae.entries[i].data.len;
     }
     e = raft_recv_appendentries(raft_handle, 
@@ -465,10 +462,7 @@ typedef struct cyclone_st {
       }
       else {
 	client_req.id = rand();
-	client_req.data.buf = malloc(msg->client.size);
-	memcpy(client_req.data.buf, 
-	       msg->client.ptr, 
-	       msg->client.size);
+	client_req.data.buf = msg->client.ptr;
 	client_req.data.len = msg->client.size;
 	client_req.type = RAFT_LOGTYPE_NORMAL;
 	// TBD: Handle error
@@ -497,10 +491,7 @@ typedef struct cyclone_st {
 	for(int i=0;i<msg->client.size;i++) {
 	  int msg_size = msg->client.batch_sizes[i];
 	  messages[i].id = rand();
-	  messages[i].data.buf = malloc(msg_size); 
-	  memcpy(messages[i].data.buf, 
-		 ptr, 
-		 msg_size);
+	  messages[i].data.buf = ptr;
 	  messages[i].data.len = msg_size;
 	  ptr = ptr + msg_size;
 	  messages[i].type = RAFT_LOGTYPE_NORMAL;
@@ -529,10 +520,7 @@ typedef struct cyclone_st {
       }
       else {
 	client_req.id = rand();
-	client_req.data.buf = malloc(msg->client.size);
-	memcpy(client_req.data.buf, 
-	       msg->client.ptr, 
-	       msg->client.size);
+	client_req.data.buf = msg->client.ptr;
 	client_req.data.len = msg->client.size;
 	client_req.type = msg->client.type;
 	// TBD: Handle error
@@ -563,10 +551,7 @@ typedef struct cyclone_st {
       }
       else {
 	client_req.id = rand();
-	client_req.data.buf = malloc(msg->client.size);
[…]

### 2016-05-10 core/dispatcher.cpp
Switch to a custom runqueue.

@@ -53,22 +53,6 @@ static volatile unsigned long list_lock   = 0;
 static volatile unsigned long result_lock = 0;
 static volatile bool building_image = false;
 
-static void lock(volatile unsigned long *lockp)
-{
-  // TEST + TEST&SET
-  do {
-    while((*lockp) != 0);
-  } while(!__sync_bool_compare_and_swap(lockp, 0, 1));
-  __sync_synchronize();
-}
-
-static void unlock(volatile unsigned long *lockp)
-{
-  __sync_synchronize();
-  __sync_bool_compare_and_swap(lockp, 1, 0);
-  __sync_synchronize();
-}
-
 static void lock_rpc_list()
 {
   lock(&list_lock);

### 2016-05-10 core/dispatcher_exec.cpp

@@ -4,32 +4,72 @@
 #include "dispatcher_exec.hpp"
 #include "checkpoint.hpp"
 
-boost::asio::io_service ioService;
-boost::asio::io_service::work work(ioService);
 boost::asio::io_service ioService2;
 boost::asio::io_service::work work2(ioService2);
+
+static rpc_info_t *volatile run_queue_head = NULL;
+static rpc_info_t *volatile run_queue_tail = NULL;
+static unsigned long runqueue_lock = 0;
+
+void add_to_runqueue(rpc_info_t *work)
+{
+  work->next_issue = NULL;
+  lock(&runqueue_lock);
+  if(run_queue_head == NULL) {
+    run_queue_head = run_queue_tail = work;
+  }
+  else {
+    run_queue_tail->next_issue = work;
+    run_queue_tail = work;
+  }
+  unlock(&runqueue_lock);
+}
+
+rpc_info_t * get_from_runqueue()
+{
+  rpc_info_t *work;
+  while(run_queue_head == NULL);
+  lock(&runqueue_lock);
+  work = run_queue_head;
+  run_queue_head = work->next_issue;
+  if(run_queue_head == NULL) {
+    run_queue_tail = NULL;
+  }
+  unlock(&runqueue_lock);
+  return work;
+}
+
+static struct executor_st {
+  void operator()
+  {
+    while(true) {
+      rpc_info_t *rpc = get_from_runqueue();
+      if(rpc->rpc->flags & RPC_FLAG_RO) {
+	exec_rpc_internal_ro(rpc);
+      }
+      else if(rpc->rpc->flags & RPC_FLAG_SYNCHRONOUS) {
+	exec_rpc_internal_synchronous(rpc);
+      }
+      else {
+	exec_rpc_internal(rpc);
+      }
+    }
+  }
+} executor;
+
 boost::thread_group threadpool;
+boost::thread executor_thread;
 void dispatcher_exec_startup()
 {
-  threadpool.create_thread
-    (boost::bind(&boost::asio::io_service::run, &ioService));
+  executor_thread = new boost::thread(boost::ref(executor));
   threadpool.create_thread
     (boost::bind(&boost::asio::io_service::run, &ioService2));
 }
 
 void exec_rpc(rpc_info_t *rpc)
 {
-  if(rpc->rpc->flags & RPC_FLAG_RO) {
[…]

### 2016-05-10 core/dispatcher_exec.hpp

@@ -24,6 +24,7 @@ typedef struct rpc_info_st {
   volatile unsigned long pending_lock;
   volatile int client_blocked;
   struct rpc_info_st *next;
+  struct rpc_info_st *volatile next_issue;
 } rpc_info_t;
 
 
@@ -33,4 +34,21 @@ extern void exec_rpc_internal(rpc_info_t *rpc);
 extern void exec_rpc_internal_synchronous(rpc_info_t *rpc);
 extern void exec_rpc_internal_ro(rpc_info_t *rpc);
 extern void exec_send_checkpoint(void *socket, void *handle);
+
+static void lock(volatile unsigned long *lockp)
+{
+  // TEST + TEST&SET
+  do {
+    while((*lockp) != 0);
+  } while(!__sync_bool_compare_and_swap(lockp, 0, 1));
+  __sync_synchronize();
+}
+
+static void unlock(volatile unsigned long *lockp)
+{
+  __sync_synchronize();
+  __sync_bool_compare_and_swap(lockp, 1, 0);
+  __sync_synchronize();
+}
+
 #endif

### 2016-05-10 paper/paper.tex
Updates to paper.

@@
 
-\section{Example Applications}
+\section{Example Application}
 \label{sec:examples}
+We now demonstrate how Cyclone can be used in practise by putting together an
+example of a linearizable set of counters. The problem is to construct a sparse
+set of counters indexed by a 64 bit key and support \emph{linearizable}
+increment, decrement and an atomic set increment. The atomic set increment takes
+as input a set of counter indexes and their current value and atomically
+increments the counters from the the given values.
 
-\section{Evauation}
+\begin{figure*}
+\begin{tabular}{cc}    
+\begin{minipage}{0.5\textwidth}  
+{ \scriptsize
+\begin{verbatim}
+bool is_stable(uint64_t value)
+{
+  if(value is even) {
+    return true
+  }
+  else {
+    return false
+  }
+}
 
-\subsection{Baseline}
+Get(uint64_t index)
+{
+  if(not is_stable(counters[index]) {
+    return EUNSTABLE
+  }
+  else {
+    return counters[index]
+  }
+}
 
-\subsection{Configuration Management}
+Increment(uint64_t index)
+{
+  if(not is_stable(counters[index]) {
+    return EUNSTABLE
+  }
+  else {
+    counters[index] = counters[index] + 2
+    return SUCCESS
+  }
+}
+\end{verbatim}
+}
+\end{minipage} &
+\begin{minipage}{0.5\textwidth}  
+{ \scriptsize
+\begin{verbatim}
+Prepare(uint64_t index, uint64_t value, bool success)
+{
+  if(counters[index] == value) {
+    success = success AND TRUE
+  }
+  else {
+    success = success AND FALSE
+  }
+  // Make unstable
+  counters[index] = counters[index] + 1
+}
 
-\subsection{Transaction Processing}
+Commit(uint64_t index, bool success)
+{
+  if(success == TRUE) {
+    // Update
+    counters[index] = counters[index] + 1
+  }
+  else {
+    // Rollback
+    counters[index] = counters[index] - 1
+  }
+}
 
-\subsection{Network Function Virtualization}
+AtomicIncrementSet(uint64_t indices[],
+                   uint64_t pre_values[],
+                   int count)
+{
+  bool success = TRUE
+  for i = 1 to count
+    success =
+      success AND
+        Prepare(indices[i],
+                pre_values[i],
+                success)
+  for i = 1 to count
+    Commit(indices[i], success)       
+}
+\end{verbatim}
+}
+\end{minipage}
+\end{tabular}
+\caption{Counter Set}
+\label{fig:counter_set}
+\end{figure*}
 

### 2016-05-10 utils-global/config_generator.py
Updates to scripts.

@@ -43,7 +43,6 @@ logsize=config.getint('meta','logsize')
 heapsize=config.getint('meta','heapsize')
 coord_heapsize=config.getint('meta','coord_heapsize')
 server_baseports = {}
-client_baseports = {}
 
 cond_abs_dir(output)
 
@@ -58,8 +57,6 @@ ports=5
 for q in range(0, quorums):
     qstring='quorum' + str(q)
     server_baseports[str(q)] = baseport
-    baseport = baseport + ports
-    client_baseports[str(q)] = baseport
     baseport = baseport + ports
     config_name=output + '/' + 'config' + str(q) + '.ini'
     f=open(config_name, 'w')
@@ -87,7 +84,6 @@ for q in range(0, quorums):
             index=index+1
     f.write('[dispatch]\n')
     f.write('server_baseport=' + str(server_baseports[str(q)]) + '\n')
-    f.write('client_baseport=' + str(client_baseports[str(q)]) + '\n')
     f.write('filepath=' + str(filepath) + '\n')
     f.write('heapsize=' + str(heapsize) + '\n')
     f.close()
@@ -187,7 +183,6 @@ for q in range(0, quorums):
         f.write('iface' + str(i) + '=' + iface + '\n')
     f.write('[dispatch]\n')
     f.write('server_baseport=' + str(server_baseports[str(q)]) + '\n')
-    f.write('client_baseport=' + str(client_baseports[str(q)]) + '\n')
     f.close()
 
 
@@ -241,7 +236,7 @@ for m in range(0, machines):
             cmd=cmd + str(co_replicas) + ' '
             cmd=cmd + str(clients - 1) + ' 0 '
             cmd=cmd + str(quorums) + ' '
-            cmd=cmd + 'config config_client &> client_log' + str(c) + '&\n'
+            cmd=cmd + 'config config_client &> client_log' + str(c) + '\n'
             f_preload.write(cmd)
             cmd='./counter_driver '
             cmd=cmd + str(c) + ' '

### 2016-05-10 utils-global/deploy_configs.sh

@@ -1,0 +1,38 @@
+#!/bin/bash
+if [ $# -ne 3 ]
+    then echo "Usage $0 output_dir deploy_dir group"
+    exit
+fi
+output_dir=$1
+deploy_dir=$2
+GROUP=$3
+echo "output dir = $output_dir"
+echo "deploy dir = $deploy_dir"
+echo "group = $GROUP"
+echo '#!/bin/bash' > exec_servers.sh
+echo "export LD_LIBRARY_PATH=/usr/local/lib:/usr/lib" >> exec_servers.sh
+echo "cd $deploy_dir" >> exec_servers.sh
+cp exec_servers.sh exec_coord.sh
+cp exec_servers.sh exec_inactive_servers.sh
+echo "source launch_servers" >> exec_servers.sh
+echo "source launch_inactive_servers" >> exec_inactive_servers.sh
+echo "source launch_coord" >> exec_coord.sh
+chmod u+x exec_servers.sh
+chmod u+x exec_inactive_servers.sh
+chmod u+x exec_coord.sh
+clush -w ${GROUP} rm -f ${deploy_dir}/launch_servers
+clush -w ${GROUP} rm -f ${deploy_dir}/launch_inactive_servers
+clush -w ${GROUP} rm -f ${deploy_dir}/launch_coord
+for i in ${output_dir}/* 
+do
+    if [ -d "$i" ] ; then
+	mc=$(basename $i)
+	echo "deploying to $mc"
+	scp ${i}/* ${mc}:${deploy_dir}
+	scp exec_servers.sh ${mc}:
+	scp exec_inactive_servers.sh ${mc}:
+	scp exec_coord.sh ${mc}:
+	scp ${output_dir}/*.ini ${mc}:${deploy_dir}
+	
+    fi
+done

### 2016-05-10 utils-global/deploy_driver.sh

@@ -10,7 +10,6 @@ echo "group = $GROUP"
 echo '#!/bin/bash' > exec_client.sh
 echo "export LD_LIBRARY_PATH=/usr/local/lib:/usr/lib" >> exec_client.sh
 echo "cd $deploy_dir" >> exec_client.sh
-echo "ulimit -n 10000" >> exec_client.sh
 cp exec_client.sh exec_preload.sh
 cp exec_client.sh exec_tx_client.sh
 chmod u+x exec_client.sh

### 2016-05-10 utils-global/deploy_services.sh

@@ -6,42 +6,10 @@ fi
 output_dir=$1
 deploy_dir=$2
 GROUP=$3
-echo "output dir = $output_dir"
-echo "deploy dir = $deploy_dir"
-echo "group = $GROUP"
-echo '#!/bin/bash' > exec_servers.sh
-echo "export LD_LIBRARY_PATH=/usr/local/lib:/usr/lib" >> exec_servers.sh
-echo "cd $deploy_dir" >> exec_servers.sh
-echo "ulimit -n 10000" >> exec_servers.sh
-cp exec_servers.sh exec_coord.sh
-cp exec_servers.sh exec_inactive_servers.sh
-echo "source launch_servers" >> exec_servers.sh
-echo "source launch_inactive_servers" >> exec_inactive_servers.sh
-echo "source launch_coord" >> exec_coord.sh
-chmod u+x exec_servers.sh
-chmod u+x exec_inactive_servers.sh
-chmod u+x exec_coord.sh
-clush -w ${GROUP} rm -f ${deploy_dir}/launch_servers
-clush -w ${GROUP} rm -f ${deploy_dir}/launch_inactive_servers
-clush -w ${GROUP} rm -f ${deploy_dir}/launch_coord
-for i in ${output_dir}/* 
-do
-    if [ -d "$i" ] ; then
-	mc=$(basename $i)
-	echo "deploying to $mc"
-	scp ${i}/* ${mc}:${deploy_dir}
-	scp exec_servers.sh ${mc}:
-	scp exec_inactive_servers.sh ${mc}:
-	scp exec_coord.sh ${mc}:
-	scp ${output_dir}/*.ini ${mc}:${deploy_dir}
-	
-    fi
-done
-
 clush -w ${GROUP} 'rm -rf /dev/shm/*.cyclone*'
 clush -w ${GROUP} cp ${deploy_dir}/cyclone.git/test/counter_server ${deploy_dir}
 clush -w ${GROUP} cp ${deploy_dir}/cyclone.git/test/counter_coordinator ${deploy_dir}
 clush -w ${GROUP} ./exec_servers.sh
-echo "Deployed servers sleeping 10 sec ..."
-sleep 10
+echo "Deployed servers sleeping 5 sec ..."
+sleep 5
 clush -w ${GROUP} ./exec_coord.sh

### 2016-05-10 paper/paper.tex

@@
 
+Figure~\ref{fig:counter_set} shows a linearizable implementation executing on
+servers. Odd counter values are considered unstable and even values stable.
+Clients make RPC calls to servers the trigger the execution of the
+appropriate function. Any client can make calls for the get and increment
+operations shown on the left hand side. A client retries an operation until it
+finds a stable counter value.
+
+A single distinguished client called the co-ordinator runs any atomic increment
+set operations on behalf of other clients - shown on the right hand side of
+Figure~\ref{fig:counter_set}. We chose to have a single
+distinguished co-ordinator to avoid making the example complex by introducting
+concurrency.
+
+The operations in Figure~\ref{fig:counter_set} do not comprehend failures.
+Cyclone includes a client library to automatically sequence RPC calls at the client
+side, retry them, connect to a new quorum leader on a view change and finally
+take care of unrealiable networks. At the server side the results of increment,
+prepare and commit RPC calls are always persisted and replicated before the
+response to the client. For individual get, increment, prepare and commit
+operations therefore persistence, replication and failover is transparently
+handled.
+
+The AtomicIncrementSet operation presents a special case however. Clients can
+fail and therefore a client executing the steps of the operation can fail midway
+leaving some counters in an unstable state forever. The solution is the
+replicate the distinguished co-ordinator client too. The co-ordinator therefore
+becomes both a server, receiving AtomicIncrementSet operations from clients as
+well as a client, making individual Prepare and Commit calls to the quorum
+leader. The implementation of the AtomicIncrementSet however requires guarantees
+beyond that of simple asynchronous replication
+
+\begin{itemize}
+  \item Output Commit: AtomicIncrementSet has externally visible side effects at
+    other servers when execting. It should therefore be guaranteed to finish
+    executing once it starts
+  \item Exactly once: Regardless of restarts AtomicIncrementSet should be
+    preceived to have executed exactly once by an external observer.
+\end{itemize}
+
+The solution is to simply run AtomicIncrementSet in synchronous replication
+mode. This guarantees the first property. It also guarantees that exactly once
+execution will succced in ``leader'' mode with followers simply remembering the
+success status of a transaction. To ensure that due to a failover we do not
+prepare or commit a counter more than once, we use exactly once RPCs that
+prevent RPC calls from executing more than once. Finally, since only
+one AtomicIncrementSet is executed at a time, we include the current step (the
+place in the for loop) in RPC calls to the quorum leader and have it reflect the
+information back in the response (a cookie of sorts). By querying the results of
+the last made call, the co-ordinator can simply fast forward to the correct
+index in the for loop during recovery and use the value of the ``success'' field
+in the response to update its own local success variable.
+
 \begin{figure*}
@@
 \begin{verbatim}
-Prepare(uint64_t index, uint64_t value, bool success)
+Prepare(uint64_t index, 
+        uint64_t value, 
+        bool success)
 {
@@
     Commit(indices[i], success)       
+  return success
 }

### 2016-06-18 paper/paper.tex
Updates to paper.

@@
 
+\section{Concurrency}
+\subsection{Network}
+The single leader in RAFT must send a message to every replica and receive a
+response from every replica to replicate a log entry. The number of
+messages the leader must handle grows therefore with the number of replicas
+while follower replicas only handle a fixed number of
+messages for each log entry. This imbalance causes the leader to become a
+bottleneck at high throughputs for even a small number of replicas.
+
+In order to ease this bottleneck, Cyclone allows clients to assist in
+replication. To do so with minimum changes to the RAFT protocol, the leader
+continues to replicate entries to clients on a timeout, with the clients
+providing best effort replication. The flow for assisted replication is
+simple. For any RPC call added to the log, the leader sends back a message to
+the client with a list of replicas. The client then sends the RPC call to
+follower replicas on behalf of the leader. It then then waits for a timeout,
+gathers and sends back a message to the leader with the list of replicas that
+have responded within the timeout. Client assisted replication is therefore best
+effort. The leader also continues to run the normal RAFT protocol - in the event
+that a log entry remains unreplicated at a follower it sends it the log entry
+for replication as usual.
+
+The only change necessary for client assisted replication from the perspective
+of RAFT is for follower replicas to be able to handle out of order arrival of
+log entries. This is accomplished by an in memory cache of replicated log
+entries. When an arriving entry completes a sequence of log entries starting at
+the next desired index, the cache attempts to append the log entries to the
+actual RAFT log. For each entry the appropriate RAFT response (accept/reject) is
+sent back to the client that had sent the entry on behalf of the leader.
+
 \section{Example Application}

### 2016-06-21 paper/paper.tex

@@
 providing best effort replication. The flow for assisted replication is
-simple. For any RPC call added to the log, the leader sends back a message to
-the client with a list of replicas. The client then sends the RPC call to
+as follows. For any RPC call added to the log, the leader sends back a message
+to the client with a list of replicas. The client then sends the RPC call to
 follower replicas on behalf of the leader. It then then waits for a timeout,
-gathers and sends back a message to the leader with the list of replicas that
-have responded within the timeout. Client assisted replication is therefore best
-effort. The leader also continues to run the normal RAFT protocol - in the event
+during which it gathers responses from clients and finally sends back a message
+to the leader with the list of replicas that have responded within the
+timeout. Client assisted replication is therefore best effort. The leader also
+continues to run the normal RAFT protocol - in the event
 that a log entry remains unreplicated at a follower it sends it the log entry
@@
 log entries. This is accomplished by an in memory cache of replicated log
-entries. When an arriving entry completes a sequence of log entries starting at
-the next desired index, the cache attempts to append the log entries to the
-actual RAFT log. For each entry the appropriate RAFT response (accept/reject) is
-sent back to the client that had sent the entry on behalf of the leader.
+entries - a structure we dub the log cache. When an arriving entry completes a
+sequence of log entries starting at the next desired index, the cache attempts
+to append the log entries to the actual RAFT log. For each entry the appropriate
+RAFT response (accept/reject and the next desired log entry in the case of
+accept) is sent back to the client that had sent the entry on behalf of the
+leader. Pseudocode for the log cache is shown in Figure~\ref{fig:logcache}.
+
+\begin{figure}
+{ \scriptsize
+\begin{verbatim}
+on ReceiveEntry(term, entry) {
+  if (term < logcache_term) {
+    drop entry
+  }
+  else if(term > logcache_term){
+    drop all entries from logcache
+    logcache_term = term
+  }
+  while(entry = canAdd(raftLog, logcache)) {
+    AddEntry(raftLog, entry);
+    RemoveEntry(logcache, entry)
+    Send response to client for entry;
+  }
+}
+\end{verbatim}
+}
+\caption{Logcache}
+\label{fig:logcache}
+\end{figure}
+
+Any client assists in the replication of exactly one entry at a time, dropping
+the previous entry is it receives a new entry from the leader to replicate. The
+client needs to therefore uniquely identify the entry for which a response is
+received from a replica. In RAFT, entries being replicated are uniqely
+identified by the term of the leader, the term of the entry and the index of the
+previous entry. These pieces of identifying information are carried in
+replication messages to allow RAFT to tolerate out of order and unreliable
+messages on the network. The client uses these parameters to filter out all
+reponses from replicas other than the one for the entry it is attempting to
+replicate.
+
+Finally, we need to show that the logcache does not lead to a departure from the
+RAFT protocol. This follows from the fact that in RAFT there is exactly one
+leader for any given term number (stemming from the safety properties of the
+election protocol). Further, the leader always extens its log, it never rolls
+back log entries - entries always flow from leader to followers in RAFT (unlike
+other protocols such as Paxos). Therefore any log entry added to the logcache is
+always present at the leader in that term. We ensure stale entries do not
+survive leadership changes by flushing the cache on a change in term. Hence any
+entry appended to the actual RAFT log at the follower is one that is present at
+the leader - the protocol therefore behaves no differently than if the leader
+had actually transmitted the log entry directly to the follower replica.
 

### 2016-06-22 core/cyclone.hpp
Make doorbell implicit.

@@ -62,6 +62,7 @@ typedef struct rpc_st {
   int code;
   int flags;
   int client_id;
+  int client_port;
   union {
     unsigned long client_txid;
     int parent_raft_idx;
@@ -86,10 +87,9 @@ static const int RPC_REQ_MARKER         = 3; // Dispatcher internal (do not use)
 static const int RPC_REQ_DATA           = 4; // Dispatcher internal (do not use)
 static const int RPC_REQ_NODEADD        = 5; // Add a replica (non blocking)
 static const int RPC_REQ_NODEDEL        = 6; // Delete a replica (non blocking)
-static const int RPC_DOORBELL           = 7; // Ring doorbell
-static const int RPC_REP_COMPLETE       = 8; // DONE 
-static const int RPC_REP_PENDING        = 9; // PENDING 
-static const int RPC_REP_UNKNOWN        = 10; // UNKNOWN RPC
-static const int RPC_REP_INVSRV         = 11; // WRONG master  -- master set in reply
-static const int RPC_REP_OLD            = 12; // RPC too old to cache results
+static const int RPC_REP_COMPLETE       = 7; // DONE 
+static const int RPC_REP_PENDING        = 8; // PENDING 
+static const int RPC_REP_UNKNOWN        = 9; // UNKNOWN RPC
+static const int RPC_REP_INVSRV         = 10; // WRONG master  -- master set in reply
+static const int RPC_REP_OLD            = 11; // RPC too old to cache results
 #endif

### 2016-06-22 core/dispatch_client.cpp

@@ -14,55 +14,10 @@ typedef struct rpc_client_st {
   int me_mc;
   client_switch *router;
   rpc_t *packet_out;
-  rpc_t *packet_out_doorbell;
   rpc_t *packet_in;
   int server;
   int replicas;
   unsigned long channel_seq;
-  bool *rung_doorbell;
-
-  void ring_doorbell()
-  {
-    while(true) {
-      if(rung_doorbell[server]) {
-	break;
-      }
-      BOOST_LOG_TRIVIAL(info) << "Ringing doorbell";
-      packet_out_doorbell->code = RPC_DOORBELL;
-      packet_out_doorbell->client_id   = me;
-      packet_out_doorbell->channel_seq = channel_seq++;
-      packet_out_doorbell->requestor   = me_mc;
-      packet_out_doorbell->port      = router->input_port(server);
-      cyclone_tx(router->output_socket(server), 
-		 (unsigned char *)packet_out_doorbell, 
-		 sizeof(rpc_t), 
-		 "PROPOSE");
-      int resp_sz;
-      while(true) {
-	resp_sz = cyclone_rx_timeout(router->input_socket(server), 
-				     (unsigned char *)packet_in, 
-				     DISP_MAX_MSGSIZE, 
-				     timeout_msec*1000,
-				     "RESULT");
-	if(resp_sz == -1) {
-	  break;
-	}
-	if(packet_in->channel_seq != (channel_seq - 1)) {
-	  continue;
-	}
-	break;
-      }
-
-      if(resp_sz == -1) {
-	BOOST_LOG_TRIVIAL(info) << "Timeout doorbell";
-	server = (server + 1)%replicas;
-	BOOST_LOG_TRIVIAL(info) << "CLIENT SET NEW LEADER " << server;
-	continue;
-      }
-
-      rung_doorbell[server] = true;
-    }
-  }
   
   void update_server(const char *context)
   {
@@ -73,17 +28,11 @@ typedef struct rpc_client_st {
       << context;
     server = (server + 1)%replicas;
     BOOST_LOG_TRIVIAL(info) << "CLIENT SET NEW LEADER " << server;
-    if(!rung_doorbell[server]) {
-      ring_doorbell();
-    }
   }
 
   void set_server()
   {
     BOOST_LOG_TRIVIAL(info) << "CLIENT SETTING LEADER " << server;
-    if(!rung_doorbell[server]) {
-      ring_doorbell();
-    }
   }
 
[…]

### 2016-06-22 core/dispatcher.cpp

@@ -29,6 +29,7 @@ struct client_ro_state_st {
   volatile unsigned long committed_txid;
   char * last_return_value;
   int last_return_size;
+  int client_port;
   bool inflight;
 } client_ro_state [MAX_CLIENTS];
 static unsigned long ro_result_lock = 0;
@@ -935,6 +936,12 @@ struct dispatcher_loop {
 	rx_sizes[requests] = sz;
 	req = (rpc_t *)ptr;
 	req->timestamp = mark;
+	if(client_ro_state[req->client_id].client_port != req->client_port) {
+	  router->ring_doorbell(req->requestor,
+				req->client_id,
+				req->client_port,
+				0);
+	}
 	if(!client_ro_state[req->client_id].inflight) { // Only one inflight per client
 	  ptr += sz;
 	  space_left -= sz;
@@ -952,24 +959,6 @@ struct dispatcher_loop {
 	    requests = 0;
 	    ptr = rx_buffers;
 	    space_left = BUFSPACE;
-	  }
-	  else if(req->code == RPC_DOORBELL) {
-	    router->ring_doorbell(req->requestor,
-				  req->client_id,
-				  req->port,
-				  0);
-	    rpc_t *rpc_rep = (rpc_t *)tx_buffer;
-	    rpc_rep->code = RPC_REP_COMPLETE;
-	    rpc_rep->client_id   = req->client_id;
-	    rpc_rep->channel_seq = req->channel_seq;
-	    router->send_data(req->requestor,
-			      req->client_id, 
-			      tx_buffer, 
-			      sizeof(rpc_t),
-			      0);
-	    ptr -= sz;
-	    space_left += sz;
-	    requests--;
 	  }
 	  else if(requests == adaptive_batch_size) {
 	    handle_batch_rpc(rx_sizes, requests);
@@ -1156,6 +1145,7 @@ void dispatcher_start(const char* config_server_path,
     client_ro_state[i].last_return_size  = 0;
     client_ro_state[i].last_return_value = NULL;
     client_ro_state[i].inflight = false;
+    client_ro_state[i].client_port = -1;
   }
 
   committed_raft_log_idx = -1;

### 2016-06-22 core/cyclone_comm.hpp
Refactor our client tracking.

@@ -206,6 +206,68 @@ static void cyclone_bind_endpoint(void *socket, const char *endpoint)
   }
 }
 
+struct client_paths {
+  void **sockets_out;
+  int *socket_out_ports;
+  int clients;
+  int client_machines;
+  boost::property_tree::ptree *saved_pt_client;
+
+  int index(int mc, int client)
+  {
+    return mc*clients + client;
+  }
+
+  void ring_doorbell(int mc,
+		     int client,
+		     int port)
+  {
+    if(sockets_out[index(mc, client)] != NULL && 
+       socket_out_ports[index(mc, client)] == port) {
+      return; // Already setup
+    }
+    if(sockets_out[index(mc, client)] != NULL) {
+      zmq_close(sockets_out[index(mc, client)]);
+    }
+    std::stringstream key; 
+    std::stringstream addr;
+    // output wire to client
+    void *socket = cyclone_socket_out(saved_context);
+    sockets_out[index(mc, client)] = socket;
+    key.str("");key.clear();
+    addr.str("");addr.clear();
+    key << "machines.addr" << mc;
+    addr << "tcp://";
+    addr << saved_pt_client->get<std::string>(key.str().c_str());
+    addr << ":" << port;
+    cyclone_connect_endpoint(socket, addr.str().c_str());
+    socket_out_ports[index(mc, client)] = port;
+  }
+  
+  void init()
+  {
+    sockets_out = new void *[client_machines*clients];
+    socket_out_ports = new int[client_machines*clients];
+    for(int i=0;i<client_machines;i++) {
+      for(int j=0;j<clients;j++) {
+	sockets_out[index(i,j)] = NULL;
+      }
+    }
+  }
+
+  void teardown()
+  {
+    for(int i=0;i<client_machines;i++) {
+      for(int j=0;j<clients;j++) {
+	if(sockets_out[index(i, j)] != NULL) {
+	  zmq_close(sockets_out[index(i, j)]);
+	}
+      }
+    }
+  }
+  
+};
+
 class raft_switch {
   void **sockets_out;
   void **control_sockets_out;
@@ -342,19 +404,9 @@ typedef struct mux_state_st{
 
 class server_switch {
   void *saved_context;
-  void **sockets_out;
-  int *socket_out_ports;
   void *socket_in;
-  int client_machines;
-  int server_machines;
   int clients;
-  boost::property_tree::ptree *saved_pt_client;
-
[…]

### 2016-06-22 core/cyclone_comm.hpp
Fixes.

@@ -207,6 +207,7 @@ static void cyclone_bind_endpoint(void *socket, const char *endpoint)
 }
 
 struct client_paths {
+  void *saved_context;
   void **sockets_out;
   int *socket_out_ports;
   int clients;
@@ -243,6 +244,11 @@ struct client_paths {
     cyclone_connect_endpoint(socket, addr.str().c_str());
     socket_out_ports[index(mc, client)] = port;
   }
+
+  void *socket(int mc, int client)
+  {
+    return sockets_out[index(mc, client)];
+  }
   
   void init()
   {
@@ -264,6 +270,7 @@ struct client_paths {
 	}
       }
     }
+    delete[] sockets_out;
   }
   
 };
@@ -435,7 +442,8 @@ public:
     
     cpaths.clients  = clients_in;
     cpaths.saved_pt_client = pt_client;
-
+    cpaths.saved_context = saved_context;
+    
     cpaths.init();
 
     // Input wire
@@ -520,18 +528,12 @@ public:
 		       sizeof(mux_state_t),
 		       "demux");
       if(cmd.op == RING_DOORBELL) {
-	cpaths.ring_doorbell_backend(cmd.mc,
-				     cmd.client,
-				     cmd.port);
+	cpaths.ring_doorbell(cmd.mc,
+			     cmd.client,
+			     cmd.port);
       }
       else {
-	if(sockets_out[index(cmd.mc, cmd.client)] == NULL) {
-	  BOOST_LOG_TRIVIAL(info) << "received rpc req without doorbell "
-				  << " mc = " << cmd.mc
-				  << " client = " << cmd.client;
-	  exit(-1);
-	}
-	cyclone_tx(sockets_out[index(cmd.mc, cmd.client)],
+	cyclone_tx(cpaths.socket(cmd.mc, cmd.client),
 		   (const unsigned char *)cmd.data,
 		   cmd.size,
 		   "demux");
@@ -547,7 +549,6 @@ public:
   {
     cpaths.teardown();
     zmq_close(socket_in);
-    delete[] sockets_out;
    }
 };
 

### 2016-06-22 core/cyclone.cpp
Add client tracker to raft switch.

@@ -584,12 +584,14 @@ static struct cyclone_img_load_st {
 } cyclone_image_loader;
 
 void* cyclone_boot(const char *config_path,
+		   const char *config_path_client,
 		   cyclone_callback_t cyclone_rep_callback,
 		   cyclone_callback_t cyclone_pop_callback,
 		   cyclone_callback_t cyclone_commit_callback,
 		   cyclone_build_image_t cyclone_build_image_callback,
 		   int me,
 		   int replicas,
+		   int clients,
 		   void *user_arg)
 {
   cyclone_t *cyclone_handle;
@@ -603,6 +605,7 @@ void* cyclone_boot(const char *config_path,
   cyclone_handle->user_arg   = user_arg;
   
   boost::property_tree::read_ini(config_path, cyclone_handle->pt);
+  boost::property_tree::read_ini(config_path_client, cyclone_handle->pt_client);
   std::string path_raft           = cyclone_handle->pt.get<std::string>("storage.raftpath");
   char me_str[100];
   sprintf(me_str, "%d", me);
@@ -696,7 +699,9 @@ void* cyclone_boot(const char *config_path,
   cyclone_handle->router = new raft_switch(cyclone_handle->zmq_context,
 					   &cyclone_handle->pt,
 					   cyclone_handle->me,
-					   cyclone_handle->replicas);
+					   cyclone_handle->replicas,
+					   clients,
+					   &cyclone_handle->pt_client);
 
   bool i_am_active = false;
   for(int i=0;i<cyclone_handle->pt.get<int>("active.replicas");i++) {

### 2016-06-22 core/cyclone.hpp

@@ -40,12 +40,14 @@ typedef void (*cyclone_build_image_t)(void *socket);
 					    
 // Returns a cyclone handle
 extern void* cyclone_boot(const char *config_path,
+			  const char *client_path,
 			  cyclone_callback_t cyclone_rep_callback,
 			  cyclone_callback_t cyclone_pop_callback,
 			  cyclone_callback_t cyclone_commit_callback,
 			  cyclone_build_image_t cyclone_build_image_callback,
 			  int me,
 			  int replicas,
+			  int clients,
 			  void *user_arg);
 
 extern void cyclone_shutdown(void *cyclone_handle);

### 2016-06-22 core/cyclone_comm.hpp

@@ -283,11 +283,14 @@ class raft_switch {
   void *request_socket_out;
   void *control_socket_in;
   int replicas;
+  client_paths cpaths;
 public:
   raft_switch(void *context,
 	      boost::property_tree::ptree *pt,
 	      int me,
-	      int replicas_in)
+	      int replicas_in,
+	      int clients_in,
+	      boost::property_tree::ptree *pt_client)
     :replicas(replicas_in)
   {
     std::stringstream key;
@@ -348,6 +351,12 @@ public:
       addr << ":" << port;
       cyclone_connect_endpoint(control_sockets_out[i], addr.str().c_str());
     }
+
+    cpaths.client_machines =  pt_client->get<int>(key.str().c_str());
+    cpaths.clients  = clients_in;
+    cpaths.saved_pt_client = pt_client;
+    cpaths.saved_context = context;
+    cpaths.init();
   }
 
   void * output_socket(int machine)
@@ -380,6 +389,16 @@ public:
     return control_sockets_out[machine];
   }
 
+  void client_doorbell(int mc, int client, int port)
+  {
+    cpaths.ring_doorbell(mc, client, port);
+  }
+
+  void* client_socket(int mc, int client)
+  {
+    return cpaths.socket(mc, client);
+  }
+  
   ~raft_switch()
   {
     for(int i=0;i<=replicas;i++) {

### 2016-06-22 core/cyclone_context.hpp

@@ -64,6 +64,7 @@ typedef struct
 struct cyclone_monitor;
 typedef struct cyclone_st {
   boost::property_tree::ptree pt;
+  boost::property_tree::ptree pt_client;
   void *zmq_context;
   raft_switch *router;
   int replicas;

### 2016-06-22 core/dispatcher.cpp

@@ -1153,12 +1153,14 @@ void dispatcher_start(const char* config_server_path,
   pending_rpc_head = pending_rpc_tail = NULL;
   // Boot cyclone -- this can lead to rep cbs on recovery
   cyclone_handle = cyclone_boot(config_server_path,
+				config_client_path,
 				&cyclone_rep_cb,
 				&cyclone_pop_cb,
 				&cyclone_commit_cb,
 				&checkpoint_callback,
 				me,
 				replicas,
+				clients,
 				NULL);
   // Listen on port
   void *zmq_context = zmq_init(zmq_threads);

### 2016-06-22 core/cyclone_comm.hpp
Fix.

@@ -352,6 +352,8 @@ public:
       cyclone_connect_endpoint(control_sockets_out[i], addr.str().c_str());
     }
 
+    key.str("");key.clear();
+    key << "machines.machines";
     cpaths.client_machines =  pt_client->get<int>(key.str().c_str());
     cpaths.clients  = clients_in;
     cpaths.saved_pt_client = pt_client;

### 2016-06-26 core/cyclone_context.hpp
Updates.

@@ -480,7 +480,9 @@ typedef struct cyclone_st {
       break;
 
     case MSG_ASSISTED_QUORUM_OK:
-      // Drop
+      raft_recv_assisted_quorum(raft_handle,
+				&msg->rep,
+				msg->quorum);
       break;
     case MSG_CLIENT_REQ:
       client_req.id = rand();

### 2016-06-26 paper/paper.tex
Updates to paper.

--- 2016-05-10 paper/paper.tex
@@
 
+\section{Concurrency}
+\subsection{Network}
+The single leader in RAFT is responsible for sending every log entry to every
+replica. Sending log entries one by one or even in batches causes inefficient
+use of the network link between the leader replica and followers when the leader
+must wait for an acknowledgment from the follower before sending the next batch
+of log entries. The solution is to send log entries from the leader as they are
+appended to its log without waiting for an acknowledgment. The result however is
+that entries can arrive out of order at follower replicas due to an unreliable
+network.
+
+We note that RAFT as specified already tolerates out of order arrival of log
+entries at follower replicas (a newly elected leader assumed replicas are up to
+date when it starts sending out log entries). However follower replicas can only
+apply log entries \emph{in order}- dropping log entries that cannot be applied
+to extend the log at a follower replica. To bridge this gap, with no changes to
+the protocol, we added an in-memory cache of log entries (that we dub the log
+cache) at follower replicas to cache log fragments that arrive out of
+order. These are then applied opportunistically as soon as all preceding entries
+have arrived.
+
+Pseudocode for the log cache is shown in Figure~\ref{fig:logcache}.
+
+\begin{figure}
+{ \scriptsize
+\begin{verbatim}
+on ReceiveEntry(leaderTerm, entry) {
+  if (leaderTerm < logcache_term) {
+    drop entry // Stale
+  }
+  else if(leaderTerm > logcache_term){
+    drop all entries from logcache
+    logcache_term = leaderTerm
+  }
+  while(entry = canAdd(raftLog, logcache)) {
+    AddEntry(raftLog, entry);
+    RemoveEntry(logcache, entry)
+    Send RAFT response for entry;
+  }
+}
+\end{verbatim}
+}
+\caption{Logcache}
+\label{fig:logcache}
+\end{figure}
+
+The log cache is built around the key property of RAFT that there is exactly one
+leader in any term and the log at the leader of a term can only be appended to. We
+therefore need only support adding entries to the log cache and flush the entire
+log cache on detecting a change in leadership due to a term larger than that of
+the leader whose entries are cached. Application of log entries follows the
+normal RAFT protocol where the last entry in the log must match the desired
+previous log index and previous log entry term specified in the new entry to be
+appended.
+
+The log cache enables leaders to send batches of log entries as soon as they are
+available rather than waiting for an acknowledgment from followers for the
+previous batch - enabling better utilization of the network link. We note
+however that the leader continues to track the last log entry \emph{known} to be
+replicated at any follower (the {\tt match index} in the RAFT protocol
+specification) as usual and resends log fragments on timeouts - tolerating
+unreliable networks that drop log fragments.
+
+In addition to better utilization of the network link out of the leader replica,
+the log cache also enables the leader replica to effectively offload the work of
+replication. To do so it simply informs the client originating the log entry of
+the list of replicas to which the entry must be sent. The client sends the
+entries, collects the responses and responds to the leader replica as soon as a
+majority quorum has responded. From the perspective of the RAFT protocol the
+client simply acts as an unreliable router. The log cache allows follower
+replicas to utilize chunks that might arrive out of order from different
+assisting clients. The benefit of such a scheme is that the replication load on
+the leader (in terms of network bandwidth) is distributed across the clients
+with a larger load (more clients) permitting more offload. As we show in the
+evaluation this is an effective technique to keeping tail latency under control
+in Cyclone deployments that can tolerate the loss in performance due to the
+extra hop but are willing to pay that price for better tail latencies.
+
 \section{Example Application}

### 2016-08-15 paper/paper.tex
First batch of changes for multi-log.

--- 2016-06-18 paper/paper.tex
@@
 faults that take one more machines offline by combining the persistence offered by
-durable storage with availability through replication. Building such fault
-tolerant applications is hard and users are forced to either build
-them from scratch or modify their applications to fit the interfaces of solutions
-such as zookeeper. In addition to a loss of flexibility such applications today
-are unable to take advantage of the speed of directly attached non-volatile
-memory as opposed to older slower storage in the form of disks and SSDs.
+durable storage with availability through replication. Unfortunately, building
+fault tolerant applications is known to be hard - users must reason about and
+deal with faults by adding fault recovery to their applications. In addition,
+replication comes with performance overheads that current research contends can
+only be overcome with sophisticated network hardware and intrusive application
+changes such as RDMA.
 
-This paper presents Cyclone - fault tolerance middleware that exposes a
-replicated and durable heap. This heap can be directly manipulated using loads and
-stores. Cyclone provides both failure atomicity as well as
-strongly consistent replication when updates to the heap are enclosed in
-software transactions thereby giving strong guarantees to programmers
-with minimal effort on their part. Under the hood Cyclone decouples modification
-of local persistent memory from replication to maximize performance. Unlike
-traditional systems that provide strong replication, the ``in-memory'' image is
-persistent and there is no need for checkpoints - except when new replicas are
-added. In addition linearizable reads can be directly executed without the need to
-combine pending updates from a log.
+This paper presents Cyclone - fault tolerance middleware for adding reliability
+via replication to applications written for NVM. Cyclone aims for API
+transparency - replicating RPC calls across replicas for client-server programs
+to cause minimum disruption to NVM programming models. In addition, Cyclone
+provides at most once RPC execution across the set of replicas with the
+  ability to recall the result of the last executed RPC call. As we demonstrate
+in the paper - this is sufficient to execute distributed transactions without
+the need for users to explicitly handle fault recovery. Cyclone also includes a
+high performance but light weight zero copy batching implementation of the RAFT
+consensus protocol built around the DPDK network stack. This allows a single
+machine to efficiently support multiple RAFT logs at the same time, making full
+use of available network resources, replicating tens of millions of RPC calls a
+second. This allows Cyclone to provide transparent performance - that is -
+minimal performance overheads for concurrent applications written for NVM that
+also wish to use Cyclone for fault tolerance without limiting themselves to
+distributed transactions on RDMA capable network hardware.
 \end{abstract}
 
-%\section{Introduction}
-
-\section{Single Node Programming Model}
-Cyclone provides users with a replicated and durable heap that can be
-manipulated via a set of library routines. Cyclone builds on top of Intel's Non
-Volatile Memory Library (NVML) and provides transparent replication to programs
-written to use NVML for persistence on a single node. This section therefore
-describes NVML's programming model.
+\section{Introduction}
+TBD
+\section{Background: Single Node Programming Model}
+Cyclone builds on top of Intel's Non Volatile Memory Library (NVML) and provides
+transparent replication to programs written to use NVML for persistence on a
+single node. We describe NVML's programming model in this section to provide
+adequate background for the rest of this paper.
 
@@
 
-\section{Failure Atomicity}
+\subsection{Failure Atomicity}
 Programmers typically find it hard to reason about failures when writing
@@
 calls to mark the objects being updated \emph{before} the actual updates to
-them. Figure~\ref{fig:undo_log} shows how Cyclone maintains the undo-log of
-objects.
+them.
 
@@
 
-\begin{figure}
-{ \scriptsize
-\begin{verbatim}
-TX_BEGIN()
-{
-  Initialize undo log
-}
-
-TX_ADD(o)
-{
-  if(o not already in undo log) {
-    Append <o, value of o> to undo log
-  }
-}
-
-TX_END()
-{
-  for each entry <o, v> in undo log {
-    flush o to persistent storage
-  }
-  truncate undo log
-}
-
-Recover()
-{
-  for each entry <o, v> in undo log {
-    set o to v
-    flush o to persistent storage
-  }   
-}
-
-\end{verbatim}
-}
-\caption{Undo Log in Cyclone}
-\label{fig:undo_log}
-\end{figure}
-
 A crucial feature of NVML is that it abstracts the details of the persistence
 mechanism away from programmers. This allows the same code to be deployed on
-clusters with a range of persistence mechanisms - mapping pages from secondary
+across a range of persistence mechanisms - mapping pages from secondary
 storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
@@
 \section{Replication}
-Cyclone replicates the state of the NVML heap across multiple machines in a
-cluster. Cyclone provides strong consistency - a majority of machines in the
-quorum of replicas always agrees on the current state of the NVM heap. Unlike
-traditional approaches to strong replication however, Cyclone is designed to
-decouple replication from execution of user code modifying NVM. It does so by
-encapsulating changes to the NVM heap in server side code driven by client RPC
-requests. We chose this model since it closely reflects the use cases most
-suitable for NVM-based clusters.
+Cyclone is designed to add fault tolerance to NVML client server applications
+via replication - multiple machines maintain \emph{equivalent} state. Server
+state is queried and manipulated via RPC calls from clients. Cyclone's approach
+to replication is simple - we replicate the RPC call itself across replicas
+rather than replicating every access made during execution of the RPC call,
+which would be prohibitively expensive.
 
-Cyclone replicates state \emph{by executing the same sequence of RPC calls on
-  all members of the quorum}. By executing the same sequence of RPC calls on all
-machines we end up with the same state\footnote{We discuss how to handle
-  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
-  raft paper] under the hood for consensus on the log of RPC calls - although
-any other similar state machine replication protocol such as
-Paxos is also suitable. Cyclone provides the means to clients to achieve exactly
-once semantics for the RPC calls they make by numbering RPC calls and
-remembering the number and result of the last executed RPC call at the server end.
+Cyclone provides strongly consistent replication - every machines in the quorum of
+replicas maintains a log of RPC calls and all machines agree on the sequnce of
+RPC calls in the log. We achieve this by running an instance of the RAFT [raft]
+consensus protocol to keep the log of RPC calls on different machines in
+synchronization. The standard replicated state machine approach requires that
+commands (entries in the log) be committed before they can be applied to the
+state machine i.e. in our case - the RPC call be executed. However, the fact
+that RPC calls that modify state must themselves be executed in a crash
+consistent transaction allows us to overlap execution with replication as we
+describe next.
 
-\subsection{RPC Call Numbering}
-Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
-from a client are numbered and Cyclone remembers (in a persistent manner) the
-number and result of the last RPC call made by a client. A particular client can
-exist on any machine - cyclone treats all instances of the client as a single
-client. Cyclone rejects RPC calls that are not in sequence. Any RPC call is
-executed exactly once across the quorum - regardless of failure and
-reconfigurations. This provides the means for clients to easily achieve exactly
-once semantics, provided the client does not crash. Cyclone can supply the last
-executed RPC call number and its result to assist recovering and stateless
-clients.
+\subsection{Decoupled Replication}
+Cyclone decouples execution of the RPC call on any replica from its replication
+by performing them in parallel to save time. Normally, this is a problem with
+any consenus protocol including RAFT as replication might failt for a number of
+reasons primarily due to a change of leader and subsequent of rollback of
+previously appended (but not committed) log entries. In order to achieve
+decoupled replication therefore Cyclone depends on the ability of NVML to roll
+back crash consistent transactions. The ability to roll back a crash consistent
+transaction is important to tolerate power failure as incomplete transactions
+must be rolled back after recovery from a power failure to preserve the
+integrity of the heap. NVML also allows the user to roll back a transaction if
+needed. Cyclone exploits this feature by delaying the commit of the failure
+atomic transaction after execution till replication is complete. In the event
+that replication fails, the transaction is aborted and any changes to the non
+volatile heap are rolled back.
 
-\subsection{Asynchronous Replication}
-By default, Cyclone decouples execution of the RPC call on any replica from its
-replication. This is particularly beneficial to performance since it allows
-overlapping the update of non-volatile state with the network I/O associated
-with replication. Making the call to the application and executing the it in a
-failure atomic transaction takes a
-significant amount of time and therefore it makes sense to overlap this with the
-round-trip delay on the network for replication. This is possible in Cyclone
-since we use failure atomic transactions that can be rolled back. Replication
-can fail due to various reasons such as failure of nodes, a network partition or
-a new leader getting elected in the underlying replication protocol
-(we refer to this as a view change). Since failure atomic transactions can be
-rolled back, we are able to speculatively apply the changes to NVM without
-waiting for replication to finish.
-
-The runtime steps for asynchronous replication are shown in
-Figure~\ref{fig:async_rep}.
+A pseudocode description of how the runtime handles decoupled replication is
+shown in Figure~\ref{fig:async_rep}.
 
@@
 
-\subsection{Synchronous Replication}
-\label{sec:synch}
-Asynchronous replication provides good performance but lacks guarantees
-occasionally necessary for distributed systems paradigms such as output commit
-and the ability to integrate code that has non-deterministic execution paths
-(such as being dependent on the current time) or side effects - such as making
-further requests. Cyclone therefore provides programmers the ability to closely
-synchronize execution across the replicas using synchronous
-replication. Synchronous replication extends the programming model provided by
-cyclone - but neccessitates some awareness on the part of the user about the
-difference between the leader and follower replicas. We show later
-(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
-fault tolerant applications with little effort on the part of the programmer.
+Asynchronous replication provides good performance and we believe would be the
+common case for RPC calls replicated via Cyclone. However the program might on
+occasion require successful replication to be a precondition to exection. A case
+where we find this necessary is output commit - where the RPC call has external
+side effects, such as communicating with other machines in the cluster and
+therefore crash consistency is insufficient to roll back the execution of the
+RPC call in the event that replication fails. We therefore provide users the
+ability to declare at the client side that an RPC call should execute
+synchronously at the server side. This results in Cyclone ensuring that the RPC
+call is successfully replicated before it attempts execution on any of the
+replicas.
 
-Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
-of the RPC call \emph{waits until the call and its arguments have been
-  replicated.} This crucial difference from asynchronous RPC execution ensures
-that any externally visible side effects of the RPC such as calls to other
-machines only happen after it is guaranteed that the effects of the RPC call
-will be committed to the replicas. This provides the output commit property
-necessary in some applications.
+\subsection{Zero Copy Batching}
+Replication protocols are expensive to operate - fact that has been underlined in
+multiple pieces of research [FARM, in a box]. It would therefore appear that
+replication would become the bottleneck in Cyclone. We observed however that
+previous work did not implement replication around a high performance network
+stack possibly explaining why they chose both high performance hardware (40
+GigE+ NICS, custom FPGA designs) as well as specific high performance but
+restrictive software APIs (one sided RDMA with metadata co-location, application
+compiled down to the FPGA). As a counterpoint, we designed Cyclone around DPDK
+[dpdk] a high performance user-space network stack. DPDK provides convenient
+access to packets arriving at an ethernet NIC by moving them into packet buffers
+with the contained cache lines also directly placed into the last level cache of
+the CPU during DMA, a first step to a high performance implementation of RAFT.
 
-Next, the RPC call is executed on the leader and the result of the RPC call is
-replicated to the followers. The followers then proceed to execute the RPC call
-locally taking into account the result from the leader. This result can include
-data to synchronize non-deterministic steps such as reading the time on the
-replica.
+A second step is the observation that replication is by definition a zero copy
+activity. The exact same RPC call must be sent to all replicas - the protocol
+state machine by itself has no role to play in the contents of the log
+entries. We therefore designed our implementation of RAFT in Cyclone to avoid
+copying incoming packets containing RPC calls from clients when replicating
+them. The only addition from the protocol is a RAFT specific header containing
+the current term and index of the RPC call in the RAFT log and term of the
+leader, which is prepended to the packet before sending it out for replication.
 
-This process is made conditional on the view (current configuration of
-replicas and choice of leader) staying unchanged. To successfully complete
-execution, the leader must be able to execute the RPC call and replicate the
-result within the current view. This ensures that in the event of a network
-partition and later repair, an old execution of a previous leader cannot commit a
-log record with results from a synchronously executed RPC call.
+The third and final step is batching. Although simple, RAFT's consensus protocol
+requires a non-zero number of CPU cycles to process each entry in log. To
+amortize this cost, we make extensive use of batching. DPDK by itself can
+efficiently batch incoming packets under load - allowing the user to easily
+receive a burst (currently limited to 32) packets rather than a single one for a
+polling call. DPDKs separation of packet data from metadata means that we can
+chain these packets together and treat them as a single log entry for the
+purposes of replication.
 
-\begin{figure*}
-\centering
-\begin{minipage}{0.7\textwidth}  
-{ \scriptsize
-\begin{verbatim}
-Initiate replication of RPC call and arguments
-Block till replication of RPC call returns
-if(replication success) {  
-  while (true) { // Must complete 
-    view = current RAFT view (term)
-    TX_BEGIN {
-      if(is_leader && !have_execution_result) {
-        Execute RPC call and modify local NVM
-        replicate result in view;
-        Block till result replication returns
-        if(result replication failed) {
-          TX_ABORT;
-        }
-      }
-      else {
-        Block till committed result replication or view changes
-        if(!committed result replication) {
-          TX_ABORT;
-        }
-        Execute RPC call to modify local NVM with result from leader
-      }
-    } TX_ONABORT {
-      continue; // Try again
-    } TX_END
-    break; // All done
-  }
-}
-\end{verbatim}
-}
-\end{minipage}
-\caption{Synchronous Replication}
-\label{fig:sync_rep}
-\end{figure*}
+TBD - put it together diagram.
 
-\section{Concurrency}
-\subsection{Network}
-The single leader in RAFT must send a message to every replica and receive a
-response from every replica to replicate a log entry. The number of
-messages the leader must handle grows therefore with the number of replicas
-while follower replicas only handle a fixed number of
-messages for each log entry. This imbalance causes the leader to become a
-bottleneck at high throughputs for even a small number of replicas.
+\subsection{Multiple RAFT Logs}
+Say we maintain multiple RAFT logs on the same machine with dedicated NIC queue pairs.
 
-In order to ease this bottleneck, Cyclone allows clients to assist in
-replication. To do so with minimum changes to the RAFT protocol, the leader
-continues to replicate entries to clients on a timeout, with the clients
-providing best effort replication. The flow for assisted replication is
-simple. For any RPC call added to the log, the leader sends back a message to
-the client with a list of replicas. The client then sends the RPC call to
-follower replicas on behalf of the leader. It then then waits for a timeout,
-gathers and sends back a message to the leader with the list of replicas that
-have responded within the timeout. Client assisted replication is therefore best
-effort. The leader also continues to run the normal RAFT protocol - in the event
-that a log entry remains unreplicated at a follower it sends it the log entry
-for replication as usual.
+\subsection{RPC Call Execution}
+Call steering
+Exactly once
 
-The only change necessary for client assisted replication from the perspective
-of RAFT is for follower replicas to be able to handle out of order arrival of
-log entries. This is accomplished by an in memory cache of replicated log
-entries. When an arriving entry completes a sequence of log entries starting at
-the next desired index, the cache attempts to append the log entries to the
-actual RAFT log. For each entry the appropriate RAFT response (accept/reject) is
-sent back to the client that had sent the entry on behalf of the leader.
-
-\section{Example Application}
+\section{Putting it Together - A linearizable hash table}
 \label{sec:examples}

### 2016-08-15 core/cyclone.cpp

@@ -9,7 +9,6 @@
 extern dpdk_context_t * global_dpdk_context;
 struct rte_ring ** to_cores;
 struct rte_ring *from_cores;
-extern volatile int sending_checkpoints;
 
 /** Raft callback for sending request vote message */
 static int __send_requestvote(raft_server_t* raft,
@@ -233,7 +232,7 @@ static int __applylog(raft_server_t* raft,
     delta_node_id = cfg->node;
     BOOST_LOG_TRIVIAL(info) << "COMPLETE ADD nonvoting node " << delta_node_id;
     // TBD: This should be a signal from the application
-    sending_checkpoints = 0;
+    cyclone_handle->sending_checkpoints = 0;
   }
   else if(ety->type == RAFT_LOGTYPE_ADD_NODE) {
     cfg_change_t *cfg = (cfg_change_t *)(chunk + sizeof(rpc_t));
@@ -256,7 +255,7 @@ static void handle_cfg_change(cyclone_t * cyclone_handle,
 			     (void *)(unsigned long)cyclone_handle->router->replica_mc(delta_node_id),
 			     delta_node_id,
 			     delta_node_id == cyclone_handle->me ? 1:0);
-    sending_checkpoints = 1; // Start sending checkpoints
+    cyclone_handle->sending_checkpoints = 1; // Start sending checkpoints
   }
   else if(ety->type == RAFT_LOGTYPE_ADD_NODE) {
     cfg_change_t *cfg = (cfg_change_t *)((char *)chunk + sizeof(rpc_t));
@@ -355,7 +354,7 @@ static int __raft_logentry_offer_batch(raft_server_t* raft,
 	  void *pair[2];
 	  pair[0] = m;
 	  pair[1] = rpc;
-	  if(rte_ring_sp_enqueue_bulk(to_cores[core], pair, 2) == -ENOBUFS) {
+	  if(rte_ring_mp_enqueue_bulk(to_cores[core], pair, 2) == -ENOBUFS) {
 	    BOOST_LOG_TRIVIAL(fatal) << "raft->core comm ring is full";
 	    exit(-1);
 	  }
@@ -459,7 +458,7 @@ int __raft_has_sufficient_logs(raft_server_t *raft,
 {
   msg_entry_t completion_entry;
   cyclone_t* cyclone_handle = (cyclone_t *)user_data;
-  if(sending_checkpoints)
+  if(cyclone_handle->sending_checkpoints)
     return -1;
   completion_entry.id = 1;
[…]

### 2016-08-15 core/cyclone.hpp

@@ -25,6 +25,7 @@ typedef void (*cyclone_build_image_t)(void *socket);
 // Returns a cyclone handle
 extern void* cyclone_boot(const char *config_quorum_path,
 			  void *router,
+			  int quorum_id,
 			  int me,
 			  int clients,
 			  void *user_arg);

### 2016-08-15 core/cyclone_comm.hpp

@@ -15,7 +15,8 @@ const int MSG_MAXSIZE  = 3500; // Maximum user data in pkt
 class quorum_switch {
   int* replicas;
 public:
-  quorum_switch(boost::property_tree::ptree *cluster, boost::property_tree::ptree *quorum)
+  quorum_switch(boost::property_tree::ptree *cluster,
+		boost::property_tree::ptree *quorum)
   {
     std::stringstream key;
     std::stringstream addr;

### 2016-08-15 core/cyclone_context.hpp

@@ -128,6 +128,7 @@ typedef struct cyclone_st {
   quorum_switch *router;
   int replicas;
   int me;
+  int me_quorum;
   boost::thread *checkpoint_thread;
   int RAFT_LOGENTRIES;
   PMEMobjpool *pop_raft_state;
@@ -137,24 +138,30 @@ typedef struct cyclone_st {
   unsigned char* cyclone_buffer_out;
   unsigned char* cyclone_buffer_in;
   cyclone_monitor *monitor_obj;
+  volatile int sending_checkpoints;
 
   msg_t ae_responses[PKT_BURST];
   int ae_response_sources[PKT_BURST];
   int ae_response_cnt;
 
+  int my_q(int q)
+  {
+    return num_queues*me_quorum + q;
+  }
+  
   void send_msg(msg_t *msg, int dst_replica)
   {
-    rte_mbuf *m = rte_pktmbuf_alloc(global_dpdk_context->mempools[q_raft]);
+    rte_mbuf *m = rte_pktmbuf_alloc(global_dpdk_context->mempools[my_q(q_raft)]);
     if(m == NULL) {
       BOOST_LOG_TRIVIAL(fatal) << "Out of mbufs for send mesg";
     }
     cyclone_prep_mbuf(global_dpdk_context, 
 		      router->replica_mc(dst_replica), 
-		      q_raft, 
+		      my_q(q_raft), 
 		      m, 
 		      msg, 
 		      sizeof(msg_t));
-    cyclone_tx(global_dpdk_context, m, q_raft);
+    cyclone_tx(global_dpdk_context, m, my_q(q_raft));
   }
   
   void send_ae_responses()
@@ -335,7 +342,7 @@ struct cyclone_monitor {
     while(!terminate) {
       // Handle any outstanding requests
       available = rte_eth_rx_burst(global_dpdk_context->port_id,
-				   q_raft,
+				   my_q(q_raft),
 				   &pkt_array[0],
 				   PKT_BURST);
       cyclone_handle->ae_response_cnt = 0;
@@ -353,7 +360,7 @@ struct cyclone_monitor {
       memset(chain_size, 0, 2*PKT_BURST);
       while(accepted <= PKT_BURST) {
 	available = rte_eth_rx_burst(global_dpdk_context->port_id,
-				     q_dispatcher,
+				     my_q(q_dispatcher),
[…]

### 2016-08-15 core/dispatch_client.cpp

@@ -24,6 +24,11 @@ typedef struct rpc_client_st {
   int replicas;
   unsigned long channel_seq;
   dpdk_rx_buffer_t *buf;
+
+  int quorum_q(int quorum_id, int q)
+  {
+    return num_queues*quorum_id + q;
+  }
   
   void update_server(const char *context)
   {
@@ -69,7 +74,7 @@ typedef struct rpc_client_st {
     return resp_sz;
   }
 
-  void send_to_server(int sz)
+  void send_to_server(int sz, int quorum_id)
   {
     rte_mbuf *mb = rte_pktmbuf_alloc(global_dpdk_context->mempools[me_queue]);
     if(mb == NULL) {
@@ -77,14 +82,14 @@ typedef struct rpc_client_st {
     }
     cyclone_prep_mbuf(global_dpdk_context,
 		      router->replica_mc(server),
-		      q_dispatcher,
+		      quorum_q(quorum_id, q_dispatcher),
 		      mb,
 		      packet_out,
 		      sz);
     cyclone_tx(global_dpdk_context, mb, me_queue);
   }
 
-  int get_last_txid()
+  int get_last_txid(int quorum_id)
   {
     int retcode;
     int resp_sz;
@@ -97,7 +102,7 @@ typedef struct rpc_client_st {
       packet_out->channel_seq = channel_seq++;
       packet_out->requestor   = me_mc;
       packet_out->payload_sz  = 0;
-      send_to_server(sizeof(rpc_t));
+      send_to_server(sizeof(rpc_t), quorum_id);
       while(true) {
 	resp_sz = cyclone_rx_timeout(global_dpdk_context,
 				     me_queue,
@@ -128,7 +133,7 @@ typedef struct rpc_client_st {
     return packet_in->last_client_txid;
   }
 
-  int delete_node(int txid, int nodeid)
+  int delete_node(int txid, int quorum_id, int nodeid)
   {
     int retcode;
     int resp_sz;
@@ -143,7 +148,7 @@ typedef struct rpc_client_st {
       packet_out->payload_sz  = sizeof(cfg_change_t);
       cfg_change_t *cfg = (cfg_change_t *)(packet_out + 1);
       cfg->node = nodeid;
-      send_to_server(sizeof(rpc_t) + sizeof(cfg_change_t));
+      send_to_server(sizeof(rpc_t) + sizeof(cfg_change_t), quorum_id);
[…]

### 2016-08-15 core/dispatcher.cpp

@@ -20,14 +20,11 @@
 #include "checkpoint.hpp"
 #include "cyclone_context.hpp"
 
-static void *cyclone_handle;
 dpdk_context_t * global_dpdk_context = NULL;
 extern struct rte_ring ** to_cores;
 extern struct rte_ring *from_cores;
-static quorum_switch *router;
 static PMEMobjpool *state;
 static rpc_callbacks_t app_callbacks;
-volatile int sending_checkpoints = 0;
 static void client_reply(rpc_t *req, 
 			 rpc_t *rep,
 			 void *payload,
@@ -309,11 +306,30 @@ void dispatcher_start(const char* config_cluster_path,
   // Load/Setup state
   std::string file_path = pt_quorum.get<std::string>("dispatch.filepath");
   unsigned long heapsize = pt_quorum.get<unsigned long>("dispatch.heapsize");
+  int quorums = pt_quorum.get<int>("quorums");
   char me_str[100];
   sprintf(me_str,"%d", me);
   file_path.append(me_str);
   app_callbacks = *rpc_callbacks;
   bool i_am_active = false;
+
+  // Initialize comm rings
+  
+  to_cores = (struct rte_ring **)malloc(executor_threads*sizeof(struct rte_ring *));
+
+  sprintf(ringname, "FROM_CORES");
+  from_cores =  rte_ring_create(ringname, 
+				65536,
+				rte_socket_id(), 
+				RING_F_SC_DEQ);
+  for(int i=0;i<executor_threads;i++) {
+    sprintf(ringname, "TO_CORE%d", i);
+    to_cores[i] =  rte_ring_create(ringname, 
+				   65536,
+				   rte_socket_id(), 
+				   RING_F_SC_DEQ); 
+  }
+  
   for(int i=0;i<pt_quorum.get<int>("active.replicas");i++) {
     char nodeidxkey[100];
     sprintf(nodeidxkey, "active.entry%d",i);
@@ -370,16 +386,19 @@ void dispatcher_start(const char* config_cluster_path,
     BOOST_LOG_TRIVIAL(info) << "DISPATCHER: Recovered state";
   }
 
-  router = new quorum_switch(&pt_cluster, &pt_quorum);
-  cyclone_handle = cyclone_boot(config_quorum_path,
-				router,
-				me,
-				clients,
-				NULL);
+  for(int i=0;i<quorums;i++) {
[…]

### 2016-08-15 core/libcyclone.hpp

@@ -94,23 +94,26 @@ int make_rpc(void *handle,
 	     int sz,
 	     void **response,
 	     int txid,
+	     int quorum_id,
 	     int rpc_flags);
 // Noop RPC -- do nothing at the other end
 // client txid is essentially ignored
 int make_noop_rpc(void *handle,
 		  int txid,
+		  int quorum_id,
 		  int rpc_flags);
 
 // Get last accepred txid
-int get_last_txid(void *handle);
+int get_last_txid(void *handle, int quorum_id);
 // Get the last response
 int get_response(void *handle,
 		 void **response,
-		 int txid);
+		 int txid,
+		 int quorum_id);
 
-int delete_node(void *handle, int txid, int node);
+int delete_node(void *handle, int txid, int quorum_id, int node);
 
-int add_node(void *handle, int txid, int node);
+int add_node(void *handle, int txid, int quorum_id, int node);
 
 
 // Possible flags 

### 2016-08-15 core/tuning.hpp

@@ -25,9 +25,8 @@ static const int zmq_threads = 4;
 static const int q_junk       = 0;
 static const int q_raft       = 1;
 static const int q_dispatcher = 2;
-static const int q_control    = 3;
-static const int q_client     = 4;
-static const int num_queues   = 5;
+static const int num_queues   = 3;
+static const int num_quorums  = 5;
 static const int Q_BUFS = 8191;
 
 

### 2016-08-15 test/counter_add_node.cpp

@@ -42,6 +42,7 @@
 #include "counter.hpp"
 #include "../core/clock.hpp"
 #include "../core/logging.hpp"
+#include "../core/tuning.hpp"
 #include <libcyclone.hpp>
 
 
@@ -56,9 +57,11 @@ int main(int argc, const char *argv[]) {
 				     0,
 				     argv[4],
 				     argv[5]);
-  
-  int ctr = get_last_txid(handle) + 1;
-  int sz  = add_node(handle, ctr, atoi(argv[6])); 
+  int ctr = get_last_txid(handle, i) + 1;
+  for(int i=0;i<num_quorums;i++) {
+    int sz  = add_node(handle, ctr, i, atoi(argv[6]));
+    ctr++;
+  }
   BOOST_LOG_TRIVIAL(info) << "Done";
   return 0;
 }

### 2016-08-15 test/counter_delete_node.cpp

@@ -56,9 +56,11 @@ int main(int argc, const char *argv[]) {
 				     0,
 				     argv[4],
 				     argv[5]);
-  
   int ctr = get_last_txid(handle) + 1;
-  int sz  = delete_node(handle, ctr, atoi(argv[6])); 
+  for(int i=0;i<num_quorums;i++) {
+    int sz  = delete_node(handle, ctr, atoi(argv[6]));
+    ctr++;
+  }
   BOOST_LOG_TRIVIAL(info) << "Done";
   return 0;
 }

### 2016-08-15 test/counter_driver_noop_mt.cpp

@@ -42,6 +42,7 @@
 #include "counter.hpp"
 #include "../core/clock.hpp"
 #include "../core/logging.hpp"
+#include "../core/tuning.hpp"
 #include <libcyclone.hpp>
 #include <rte_launch.h>
 
@@ -114,11 +115,15 @@ int driver(void *arg)
 		  sizeof(struct proposal),
 		  (void **)&resp,
 		  ctr[partition],
+		  rand()%num_quorums,
 		  rpc_flags);
     */
     int rpc_flags = 0;
     // int rpc_flags = RPC_FLAG_RO;
-    sz = make_noop_rpc(handles[partition], ctr[partition], rpc_flags);
+    sz = make_noop_rpc(handles[partition],
+		       ctr[partition],
+		       rand() % num_quorums,
+		       rpc_flags);
     ctr[partition]++;
     tx_block_cnt++;
     /*

### 2016-08-15 test/counter_server.cpp

@@ -276,7 +276,9 @@ int main(int argc, char *argv[])
   else {
     sleep_time = 0;
   }
-  cyclone_network_init(argv[4], atoi(argv[2]), num_queues + executor_threads);
+  cyclone_network_init(argv[4],
+		       atoi(argv[2]),
+		       num_queues*num_quorums + executor_threads);
   dispatcher_start(argv[4], 
 		   argv[5], 
 		   &rpc_callbacks,

### 2016-08-15 core/cyclone.cpp
Compile fixes.

@@ -564,7 +564,6 @@ void* cyclone_boot(const char *config_quorum_path,
   cyclone_t *cyclone_handle;
   std::stringstream key;
   std::stringstream addr;
-  char ringname[50];
   
   cyclone_handle = new cyclone_t();
   cyclone_handle->user_arg   = user_arg;

### 2016-08-15 core/cyclone_context.hpp

@@ -342,7 +342,7 @@ struct cyclone_monitor {
     while(!terminate) {
       // Handle any outstanding requests
       available = rte_eth_rx_burst(global_dpdk_context->port_id,
-				   my_q(q_raft),
+				   cyclone_handle->my_q(q_raft),
 				   &pkt_array[0],
 				   PKT_BURST);
       cyclone_handle->ae_response_cnt = 0;
@@ -360,7 +360,7 @@ struct cyclone_monitor {
       memset(chain_size, 0, 2*PKT_BURST);
       while(accepted <= PKT_BURST) {
 	available = rte_eth_rx_burst(global_dpdk_context->port_id,
-				     my_q(q_dispatcher),
+				     cyclone_handle->my_q(q_dispatcher),
 				     &pkt_array[0],
 				     PKT_BURST);
 	if(available == 0) {

### 2016-08-15 core/dispatcher.cpp

@@ -312,6 +312,7 @@ void dispatcher_start(const char* config_cluster_path,
   file_path.append(me_str);
   app_callbacks = *rpc_callbacks;
   bool i_am_active = false;
+  char ringname[50];
 
   // Initialize comm rings
   
@@ -387,7 +388,7 @@ void dispatcher_start(const char* config_cluster_path,
   }
 
   for(int i=0;i<quorums;i++) {
-    router = new quorum_switch(&pt_cluster, &pt_quorum);
+    quorum_switch *router = new quorum_switch(&pt_cluster, &pt_quorum);
     cyclone_boot(config_quorum_path,
 		 router,
 		 i,

### 2016-08-15 test/counter_add_node.cpp

@@ -57,7 +57,7 @@ int main(int argc, const char *argv[]) {
 				     0,
 				     argv[4],
 				     argv[5]);
-  int ctr = get_last_txid(handle, i) + 1;
+  int ctr = get_last_txid(handle, 0) + 1;
   for(int i=0;i<num_quorums;i++) {
     int sz  = add_node(handle, ctr, i, atoi(argv[6]));
     ctr++;

### 2016-08-15 test/counter_delete_node.cpp

@@ -42,6 +42,7 @@
 #include "counter.hpp"
 #include "../core/clock.hpp"
 #include "../core/logging.hpp"
+#include "../core/tuning.hpp"
 #include <libcyclone.hpp>
 
 
@@ -56,9 +57,9 @@ int main(int argc, const char *argv[]) {
 				     0,
 				     argv[4],
 				     argv[5]);
-  int ctr = get_last_txid(handle) + 1;
+  int ctr = get_last_txid(handle, 0) + 1;
   for(int i=0;i<num_quorums;i++) {
-    int sz  = delete_node(handle, ctr, atoi(argv[6]));
+    int sz  = delete_node(handle, ctr, i, atoi(argv[6]));
     ctr++;
   }
   BOOST_LOG_TRIVIAL(info) << "Done";

### 2016-08-15 test/counter_driver_noop_mt.cpp

@@ -83,7 +83,7 @@ int driver(void *arg)
   int ctr[partitions];
   for(int i=0;i<partitions;i++) {
     BOOST_LOG_TRIVIAL(info) << "Connecting to quorum " << i;
-    ctr[i] = get_last_txid(handles[i]) + 1;
+    ctr[i] = get_last_txid(handles[i], 0) + 1;
     BOOST_LOG_TRIVIAL(info) << "Done";
   }
 

### 2016-08-15 core/cyclone.cpp
Fixes.

@@ -683,7 +683,7 @@ void* cyclone_boot(const char *config_quorum_path,
   }
   /* Launch cyclone service */
   __sync_synchronize(); // Going to give the thread control over the socket
-  int e = rte_eal_remote_launch(dpdk_raft_monitor, (void *)cyclone_handle->monitor_obj, 1);
+  int e = rte_eal_remote_launch(dpdk_raft_monitor, (void *)cyclone_handle->monitor_obj, 1 + quorum_id);
   if(e != 0) {
     BOOST_LOG_TRIVIAL(fatal) << "Failed to launch raft monitor on remote lcore";
     exit(-1);

### 2016-08-15 core/dispatcher.cpp

@@ -306,7 +306,6 @@ void dispatcher_start(const char* config_cluster_path,
   // Load/Setup state
   std::string file_path = pt_quorum.get<std::string>("dispatch.filepath");
   unsigned long heapsize = pt_quorum.get<unsigned long>("dispatch.heapsize");
-  int quorums = pt_quorum.get<int>("quorums");
   char me_str[100];
   sprintf(me_str,"%d", me);
   file_path.append(me_str);
@@ -340,7 +339,7 @@ void dispatcher_start(const char* config_cluster_path,
     }
   }
 
-  if(i_am_active) {
+  if(!i_am_active) {
     BOOST_LOG_TRIVIAL(info) << "Starting inactive server";
   }
   if(access(file_path.c_str(), F_OK)) {
@@ -387,7 +386,7 @@ void dispatcher_start(const char* config_cluster_path,
     BOOST_LOG_TRIVIAL(info) << "DISPATCHER: Recovered state";
   }
 
-  for(int i=0;i<quorums;i++) {
+  for(int i=0;i<num_quorums;i++) {
     quorum_switch *router = new quorum_switch(&pt_cluster, &pt_quorum);
     cyclone_boot(config_quorum_path,
 		 router,

### 2016-08-15 core/cyclone.cpp
Fixes.

@@ -115,7 +115,7 @@ static int __send_appendentries_opt(raft_server_t* raft,
     msg->ae.term          = m->term;
     msg->source        = cyclone_handle->me;
     // Bump refcnt, add ethernet header and handoff for transmission
-    rte_mbuf *e = rte_pktmbuf_alloc(global_dpdk_context->extra_pool);
+    rte_mbuf *e = rte_pktmbuf_alloc(global_dpdk_context->extra_pools[cyclone_handle->me_quorum]);
     if(e == NULL) {
       BOOST_LOG_TRIVIAL(info) << "Unable to allocate header ";
       break;

### 2016-08-15 core/cyclone_comm_dpdk.hpp

@@ -107,7 +107,7 @@ static void init_fdir_conf()
 typedef struct {
   struct ether_addr *mc_addresses;
   struct rte_mempool **mempools;
-  struct rte_mempool *extra_pool;
+  struct rte_mempool **extra_pools;
   struct rte_eth_dev_tx_buffer **buffers;
   int me;
   int port_id;
@@ -405,6 +405,7 @@ static void dpdk_context_init(dpdk_context_t *context,
   context->mempools = (rte_mempool **)malloc(queues*sizeof(rte_mempool *));
   context->buffers   = (rte_eth_dev_tx_buffer **)malloc
     (queues*sizeof(rte_eth_dev_tx_buffer *));
+  context->extra_pools = (rte_mempool **)malloc(num_quorums*sizeof(rte_mempool *));
   // Assume port 0, core 1 ....
 
   for(int i=0;i<queues;i++) {
@@ -425,8 +426,10 @@ static void dpdk_context_init(dpdk_context_t *context,
     
     
     BOOST_LOG_TRIVIAL(info) << "Init mempool max pktsize = " << max_pktsize;
+    bool is_disp_pool = (i < num_queues*num_quorums) && (i % num_queues == q_dispatcher);
+    bool is_raft_pool = (i < num_queues*num_quorums) && (i % num_queues == q_raft);
     context->mempools[i] = rte_pktmbuf_pool_create(pool_name,
-						   i != q_dispatcher ? Q_BUFS:Q_BUFS*pack_ratio,
+						   !is_disp_pool ? Q_BUFS:Q_BUFS*pack_ratio,
 						   32,
 						   0,
 						   RTE_PKTMBUF_HEADROOM + max_pktsize,
@@ -434,16 +437,16 @@ static void dpdk_context_init(dpdk_context_t *context,
     if (context->mempools[i] == NULL)
       rte_exit(EXIT_FAILURE, "Cannot init mbuf pool\n");
 
-    if(i == q_raft) {
-      strcat(pool_name, "extra");
-      context->extra_pool = rte_pktmbuf_pool_create(pool_name,
-						    Q_BUFS,
-						    4*PKT_BURST,
-						    0,
-						    RTE_PKTMBUF_HEADROOM + sizeof(struct ether_hdr),
-						    rte_socket_id());
-
-      if (context->extra_pool == NULL)
+    if(is_raft_pool) {
+      sprintf(pool_name, "extra_%d", i);
[…]

### 2016-08-15 utils-global/setup_dpdk.sh
Add script to setup dpdk.

@@ -1,0 +1,4 @@
+#!/bin/sh
+cat inputs | ./dpdk-16.04/tools/setup.sh
+ifdown p1p1
+./dpdk-16.04/tools/dpdk_nic_bind.py --bind igb_uio 0000:41:00.0

### 2016-08-15 core/cyclone.cpp
Fixes.

@@ -22,18 +22,19 @@ static int __send_requestvote(raft_server_t* raft,
   msg.source      = cyclone_handle->me;
   msg.msg_type    = MSG_REQUESTVOTE;
   msg.rv          = *m;
-  rte_mbuf *mb = rte_pktmbuf_alloc(global_dpdk_context->mempools[q_raft]);
+  int my_raft_q   = cyclone_handle->my_q(q_raft);
+  rte_mbuf *mb = rte_pktmbuf_alloc(global_dpdk_context->mempools[my_raft_q]);
   if(mb == NULL) {
     BOOST_LOG_TRIVIAL(fatal) << "Out of mbufs for send requestvote";
     exit(-1);
   }
   cyclone_prep_mbuf(global_dpdk_context,
 		    (int)(unsigned long)socket,
-		    q_raft,
+		    my_raft_q,
 		    mb,
 		    &msg,
 		    sizeof(msg_t));
-  cyclone_tx(global_dpdk_context, mb, q_raft);
+  cyclone_tx(global_dpdk_context, mb, my_raft_q);
   return 0;
 }
 
@@ -48,18 +49,19 @@ static void __send_appendentries_response(void *udata,
   resp.msg_type = MSG_APPENDENTRIES_RESPONSE;
   memcpy(&resp.aer, r, sizeof(msg_appendentries_response_t));
   resp.source = cyclone_handle->me;
-  rte_mbuf *m = rte_pktmbuf_alloc(global_dpdk_context->mempools[q_raft]);
+  int my_raft_q   = cyclone_handle->my_q(q_raft);
+  rte_mbuf *m = rte_pktmbuf_alloc(global_dpdk_context->mempools[my_raft_q]);
   if(m == NULL) {
     BOOST_LOG_TRIVIAL(fatal) << "Out of mbufs for send requestvote";
     exit(-1);
   }
   cyclone_prep_mbuf(global_dpdk_context,
 		    (int)(unsigned long)socket,
-		    q_raft,
+		    my_raft_q,
 		    m,
 		    &resp,
 		    sizeof(msg_t));
-  cyclone_tx(global_dpdk_context, m, q_raft);
+  cyclone_tx(global_dpdk_context, m, my_raft_q);
 }
 
 /** Raft callback for sending appendentries message */
@@ -79,17 +81,18 @@ static int __send_appendentries(raft_server_t* raft,
[…]

### 2016-08-15 core/cyclone_comm_dpdk.hpp

@@ -216,7 +216,6 @@ static void cyclone_prep_mbuf(dpdk_context_t *context,
 
 static void cyclone_prep_eth(dpdk_context_t *context,
 			     int dst,
-			     int dst_q,
 			     struct ether_hdr *eth)
 {
   memset(eth, 0, sizeof(struct ether_hdr));

### 2016-08-15 core/cyclone_context.hpp

@@ -107,15 +107,6 @@ const int  MSG_REQUESTVOTE              = 1;
 const int  MSG_REQUESTVOTE_RESPONSE     = 2;
 const int  MSG_APPENDENTRIES            = 3;
 const int  MSG_APPENDENTRIES_RESPONSE   = 4;
-const int  MSG_CLIENT_REQ               = 5;
-const int  MSG_CLIENT_REQ_BATCH         = 6;
-const int  MSG_CLIENT_REQ_TERM          = 7;
-const int  MSG_CLIENT_STATUS            = 8;
-const int  MSG_CLIENT_REQ_CFG           = 9;
-const int  MSG_CLIENT_REQ_SET_IMGBUILD  = 10;
-const int  MSG_CLIENT_REQ_UNSET_IMGBUILD= 11;
-const int  MSG_ASSISTED_APPENDENTRIES   = 12;
-const int  MSG_ASSISTED_QUORUM_OK       = 13;
 
 extern struct rte_ring ** to_cores;
 extern struct rte_ring *from_cores;

### 2016-08-15 core/dispatcher.cpp

@@ -104,7 +104,7 @@ typedef struct executor_st {
 		   resp_buffer, 
 		   NULL, 
 		   0,
-		   num_queues + tid);
+		   num_queues*num_quorums + tid);
     }
     else if(client_buffer->code == RPC_REQ_STATUS) {
       cookie.client_id = client_buffer->client_id;
@@ -116,7 +116,7 @@ typedef struct executor_st {
 		     resp_buffer, 
 		     NULL, 
 		     0,
-		     num_queues + tid);
+		     num_queues*num_quorums + tid);
       }
       else if(cookie.client_txid == client_buffer->client_txid) {
 	resp_buffer->code = RPC_REP_COMPLETE;
@@ -124,7 +124,7 @@ typedef struct executor_st {
 		     resp_buffer, 
 		     cookie.ret_value, 
 		     cookie.ret_size,
-		     num_queues + tid);
+		     num_queues*num_quorums + tid);
       }
       else {
 	resp_buffer->code = RPC_REP_OLD;
@@ -132,7 +132,7 @@ typedef struct executor_st {
 		     resp_buffer, 
 		     NULL,
 		     0,
-		     num_queues + tid);
+		     num_queues*num_quorums + tid);
       }
     }
     else if(client_buffer->code == RPC_REQ_NOOP) {
@@ -141,18 +141,18 @@ typedef struct executor_st {
 	if(client_buffer->wal.leader) {
 	  if(client_buffer->wal.rep == REP_SUCCESS) {
 	    resp_buffer->code = RPC_REP_COMPLETE;
-	    client_reply(client_buffer, resp_buffer, NULL, 0, num_queues + tid);
+	    client_reply(client_buffer, resp_buffer, NULL, 0, num_queues*num_quorums + tid);
 	  }
 	  else {
 	    resp_buffer->code = RPC_REP_UNKNOWN;
-	    client_reply(client_buffer, resp_buffer, NULL, 0, num_queues + tid);
+	    client_reply(client_buffer, resp_buffer, NULL, 0, num_queues*num_quorums + tid);
 	  }
 	}
       }
       else {
 	if(client_buffer->wal.leader) {
 	  resp_buffer->code = RPC_REP_COMPLETE;
-	  client_reply(client_buffer, resp_buffer, NULL, 0, num_queues + tid);
+	  client_reply(client_buffer, resp_buffer, NULL, 0, num_queues*num_quorums + tid);
 	}
       }
     }
[…]

### 2016-08-15 core/dispatch_client.cpp
Fixes.

@@ -20,7 +20,7 @@ typedef struct rpc_client_st {
   rpc_t *packet_out;
   msg_t *packet_rep;
   rpc_t *packet_in;
-  int server;
+  int *servers;
   int replicas;
   unsigned long channel_seq;
   dpdk_rx_buffer_t *buf;
@@ -30,20 +30,23 @@ typedef struct rpc_client_st {
     return num_queues*quorum_id + q;
   }
   
-  void update_server(const char *context)
+  void update_server(const char *context, int quorum)
   {
     BOOST_LOG_TRIVIAL(info) 
       << "CLIENT DETECTED POSSIBLE FAILED LEADER: "
-      << server
+      << servers[quorum]
+      << " of quorum "
+      << quorum
       << " Reason " 
       << context;
-    server = (server + 1)%replicas;
-    BOOST_LOG_TRIVIAL(info) << "CLIENT SET NEW LEADER " << server;
+    servers[quorum] = (servers[quorum] + 1)%replicas;
+    BOOST_LOG_TRIVIAL(info) << "CLIENT SET NEW LEADER " << servers[quorum];
   }
 
-  void set_server()
+  void set_server(int quorum)
   {
-    BOOST_LOG_TRIVIAL(info) << "CLIENT SETTING LEADER " << server;
+    BOOST_LOG_TRIVIAL(info) << "CLIENT SETTING LEADER for quorum " 
+			    << quorum << " = " << servers[quorum];
   }
 
 
@@ -81,7 +84,7 @@ typedef struct rpc_client_st {
       BOOST_LOG_TRIVIAL(fatal) << "Out of mbufs for send requestvote";
     }
     cyclone_prep_mbuf(global_dpdk_context,
-		      router->replica_mc(server),
+		      router->replica_mc(servers[quorum_id]),
 		      quorum_q(quorum_id, q_dispatcher),
 		      mb,
 		      packet_out,
@@ -121,11 +124,11 @@ typedef struct rpc_client_st {
 	break;
       }
       if(resp_sz == -1) {
-	update_server("rx timeout, get txid");
+	update_server("rx timeout, get txid", quorum_id);
 	continue;
       }
       if(packet_in->code == RPC_REP_INVSRV) {
-	update_server("Server not leader");
+	update_server("Server not leader", quorum_id);
 	continue;
       }
       break;
@@ -151,11 +154,11 @@ typedef struct rpc_client_st {
       send_to_server(sizeof(rpc_t) + sizeof(cfg_change_t), quorum_id);
[…]

### 2016-08-16 core/cyclone_context.hpp
Fix.

@@ -371,11 +371,10 @@ struct cyclone_monitor {
 	     rpc->code == RPC_REQ_STATUS ||  
 	     rpc->flags & RPC_FLAG_RO) {
 	    if(cyclone_is_leader(cyclone_handle)) {
-	      if(rte_ring_mp_enqueue(to_cores[core], m) == -ENOBUFS) {
-		BOOST_LOG_TRIVIAL(fatal) << "raft->core comm ring is full";
-		exit(-1);
-	      }
-	      if(rte_ring_mp_enqueue(to_cores[core], rpc) == -ENOBUFS) {
+	      void *pair[2];
+	      pair[0] = m;
+	      pair[1] = rpc;
+	      if(rte_ring_mp_enqueue_bulk(to_cores[core], pair, 2) == -ENOBUFS) {
 		BOOST_LOG_TRIVIAL(fatal) << "raft->core comm ring is full";
 		exit(-1);
 	      }

### 2016-08-27 paper/paper.tex
Additions to paper.

@@
 \begin{document}
-\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
+\title{Cyclone: Replication Middleware for NVM Clusters}
 \author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
@@
 Memory technology in the datacenter is due to undergo a paradigm shift with the
-introduction of directly addressable non-volatile memory both in the form of
+increased use of directly addressable non-volatile memory both in the form of
 battery backed non-volatile DIMMs as well as newer memory types such as
-3D XPoint. Datacenter applications typically use storage devices to survive
-faults that take one more machines offline by combining the persistence offered by
-durable storage with availability through replication. Unfortunately, building
-fault tolerant applications is known to be hard - users must reason about and
-deal with faults by adding fault recovery to their applications. In addition,
-replication comes with performance overheads that current research contends can
-only be overcome with sophisticated network hardware and intrusive application
-changes such as RDMA.
+3D XPoint. This will allow a broad range of new applications that exploit the
+combination of durability with performance far superior to traditional durable
+storage media such as NAND flash. However, durability does not guaratee
+availability and without a means to survive machine or storage medium failures
+these applications cannot match the survivability offered by sophisticated
+distributed file systems, databases and key value stores.
 
-This paper presents Cyclone - fault tolerance middleware for adding reliability
-via replication to applications written for NVM. Cyclone aims for API
-transparency - replicating RPC calls across replicas for client-server programs
-to cause minimum disruption to NVM programming models. In addition, Cyclone
-provides at most once RPC execution across the set of replicas with the
-  ability to recall the result of the last executed RPC call. As we demonstrate
-in the paper - this is sufficient to execute distributed transactions without
-the need for users to explicitly handle fault recovery. Cyclone also includes a
-high performance but light weight zero copy batching implementation of the RAFT
-consensus protocol built around the DPDK network stack. This allows a single
-machine to efficiently support multiple RAFT logs at the same time, making full
-use of available network resources, replicating tens of millions of RPC calls a
-second. This allows Cyclone to provide transparent performance - that is -
-minimal performance overheads for concurrent applications written for NVM that
-also wish to use Cyclone for fault tolerance without limiting themselves to
-distributed transactions on RDMA capable network hardware.
+Cyclone is replication middleware aimed at helping application developers
+targeting new memory types to easily add fault tolerance to their
+applications. Cyclone eschews external co-ordination services like zookeeper and
+instead replicates RPC calls from clients being servered out of durable memory
+across a quorum of clusters using the RAFT consensus protocol. Cyclone also
+provides a client side library to enable transparent failover. Finally, Cyclone
+is designed to provide at most once semantics to RPC calls. We demonstrate in
+this paper that this is sufficient to design applications that support
+distributed transactions without any need for the developer to be concerned with
+fault recovery - a challenging aspect of designing systems that support
+distributed transaction and at the same time are fault tolerant.
+
+Cyclone is also designed to provide maximum performance without being dependent
+on high end networking stacks - such as RDMA/Infiniband combinations.  To enable
+the lowest possible latency we use DPDK - a userspace networking stack and
+designed our implementation of RAFT to use zero copy for the replication
+process. To enable maximum throughput under load we extensively batch
+replication operations and support multiple concurrent instances of RAFT to
+maximize throughput. As a result, Cyclone can replicate as many as 4 million 100
+byte RPC messages every second on a very modest commodity 10 GigE network. Under
+low load, the latency seen by a client for an RPC call the requires replication
+is of the order of 20 us.
 \end{abstract}
@@
 NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
-instructions. NVDIMMs would conceivably require only a CLFLUSH to flush the
+instructions. NVDIMMs require only a CLFLUSH to flush the
 cacheline and ensure that the line enters the asynchronous data refresh
@@
 Cyclone is designed to add fault tolerance to NVML client server applications
-via replication - multiple machines maintain \emph{equivalent} state. Server
+via replication - a quorum of machines maintains \emph{equivalent} state. Server
 state is queried and manipulated via RPC calls from clients. Cyclone's approach
@@
 
-Cyclone provides strongly consistent replication - every machines in the quorum of
-replicas maintains a log of RPC calls and all machines agree on the sequnce of
-RPC calls in the log. We achieve this by running an instance of the RAFT [raft]
-consensus protocol to keep the log of RPC calls on different machines in
-synchronization. The standard replicated state machine approach requires that
-commands (entries in the log) be committed before they can be applied to the
-state machine i.e. in our case - the RPC call be executed. However, the fact
-that RPC calls that modify state must themselves be executed in a crash
-consistent transaction allows us to overlap execution with replication as we
-describe next.
+Cyclone provides strongly consistent replication - every machines in the quorum
+of replicas maintains a log of RPC calls. All RPC calls (including read-only
+calls that do not change state and therefore do not need replication) are always
+executed by the leader replica. Cyclone treats the RPC call itself as a variable
+sized opaque blob of data, leaving the application free to choose how it wishes
+to marshall call related details and argument. All machines agree on the
+committed sequence of RPC calls in the log. We achieve this by running an
+instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
+different machines in synchronization. The standard replicated state machine
+approach requires that commands (entries in the log) be committed before they
+can be applied to the state machine i.e. in our case - the RPC call be
+executed. However, the fact that RPC calls that modify state must themselves be
+executed in a crash consistent transaction allows us to overlap execution with
+replication (described in the next section).
 
 \subsection{Decoupled Replication}
+\label{sec:decouple}
 Cyclone decouples execution of the RPC call on any replica from its replication
@@
 
-\subsection{Zero Copy Batching}
+\subsection{Network Stack}
 Replication protocols are expensive to operate - fact that has been underlined in
@@
 with the contained cache lines also directly placed into the last level cache of
-the CPU during DMA, a first step to a high performance implementation of RAFT.
+the CPU during DMA.
 
-A second step is the observation that replication is by definition a zero copy
-activity. The exact same RPC call must be sent to all replicas - the protocol
-state machine by itself has no role to play in the contents of the log
-entries. We therefore designed our implementation of RAFT in Cyclone to avoid
-copying incoming packets containing RPC calls from clients when replicating
-them. The only addition from the protocol is a RAFT specific header containing
-the current term and index of the RPC call in the RAFT log and term of the
-leader, which is prepended to the packet before sending it out for replication.
+Cyclone separates the available cores on the machine into cores dedicated to
+work related to the consensus protocol and cores decidated to running
+application code. The division is configurable but Fig [] shows the division for
+our testbed. Each RAFT core runs an \emph{independent} instance of the RAFT
+protocol (Section~\ref{sec:ind}). RAFT cores pass work to the application cores
+through lock-free double ended queues. A RAFT core dispatches an incoming RPC
+call for execution simultaneously with sending it out for replication. As
+explained in Section~\ref{sec:decouple} this is beneficial for reducing latency
+to completion.
 
-The third and final step is batching. Although simple, RAFT's consensus protocol
-requires a non-zero number of CPU cycles to process each entry in log. To
-amortize this cost, we make extensive use of batching. DPDK by itself can
-efficiently batch incoming packets under load - allowing the user to easily
-receive a burst (currently limited to 32) packets rather than a single one for a
-polling call. DPDKs separation of packet data from metadata means that we can
-chain these packets together and treat them as a single log entry for the
-purposes of replication.
+Each RAFT core has a dedicated NIC queue pair. The input queue receives RPC
+calls from clients as well as RAFT protocol related traffic. The output queue
+only carries RAFT protocol related traffic. Each application core has a
+dedicated \emph{output} queue on the NIC. When completed, the response to the
+RPC call is sent from the application core via its dedicated output queue. This
+division of NIC queues eliminates any need for synchronization between RAFT and
+application cores when accessing the NIC.
 
-TBD - put it together diagram.
+\subsection{Zero Copy Replication}
+Consensus protcols such as RAFT usually have a simple common case path. Once a
+quorum leader is elected, the leader's primary job is to transmit new entries
+appended to the log to its followers. The primary job of the followers is to
+receive these entries and send acknowledgments to the leader. An entry is
+considered committed when the leader receives an acknowledgment for an entry
+from a majority of the quorum. Followers receive intimation of committed entries
+as piggybacked information on subsequent replication messages.
 
-\subsection{Multiple RAFT Logs}
-Say we maintain multiple RAFT logs on the same machine with dedicated NIC queue pairs.
+Our aim when designing Cyclone was to be relatively insensitive to the size of
+the RPC call - i.e. the size of the RAFT log entry being replicated. Since
+Cyclone is targeted towards general purpose applications we intended to support
+a range of RPC call sizes without a degradation in performance caused by copying
+the RPC call contents for transmission to each follower replica. Hence, we built
+a zero copy mechanism for common case of a leader replica sending the same log
+entry to a set of follower replicas. To achieve our zero copy goal we explot the
+fact that DPDKs packet buffers allow prepending a data without physically moving
+the data in the packet buffer. Fig [] shows how the layout of a packet buffer
+changes at the leader during replication. The received packets contains only the
+RPC call from the client. We prepend a RAFT related header (containing the
+previous log index, previous log term, leader's term and leader's commit
+point) to packet buffer itself. To this RAFT header we then prepend an IPv4
+header. Although we use the raw packet interface from DPDK and do not use IP
+based routing, we set IP addresses in a dummy header to use 5-tuple steering in
+order to place the packet in the right queue at the receiving NIC. This
+composite packet buffer is then sent to every replica. In order to properly
+route the packet through switches, we allocate a small buffer containing an
+ethernet header for each follower replica and \emph{chain} it in front of the
+prepared packet - ensuring we do not need to change the contents for every
+replica.
+
+\subsection{Batching}
+Although the common case path in RAFT is straightforward, it still adds many
+hundreds of cycles for each log entry to be processed. This adds up to a few
+microseconds, a significant amount in the face of sub 10 us latency between
+nodes with DPDK. We amortize this overhead when operating under load by batching
+multiple RPC calls together and treating the batch as a single RAFT log entry.
+DPDK by itself can efficiently batch incoming packets under load - allowing the
+user to easily receive a burst (currently limited to 32) packets rather than a
+single one for a polling call. DPDK further allows chaining packet buffers
+together and treating the chain as a single packet. This is convenient for our
+zero copy approach and we chain packet mbufs together to form a single packet
+as shown in Fig []. Since the entire chain is sent as a single packet when being
+replicated, the length of a chain is limited by the size of the coalesced
+packet, that must fit into an ethernet jumbo frame of approximately 9K bytes.
+
+\subsection{Multiple RAFT Instances}
+\label{sec:ind}
+We scale replication performance in Cyclone by using multiple cores on the same
+machine to each run an independent instance of the RAFT protocol. For a quorum
+of homogeneous machines, we dedicate the same core and NIC queue pair on each
+machine to the same instance of RAFT. Each instance maintains an
+\emph{independent} log. All the RAFT instances on a machine share the common set
+of application cores, dispatching work to them via the (shared) lock-free double
+ended queues (implemented as circular buffers in DPDK). It is important to note
+that there is no fixed ordering between RPC calls dispatched for execution from
+different RAFT instances. We discuss in Sections
+~\ref{sec:exec},\ref{sec:example} how a Cyclone user can still impose
+determinism for linearizable data structures, while continuing to scale across
+the multiple RAFT instances.
+
+Other than the lock free queues for dispatching work to application cores, the
+multiple RAFT instances share nothing, being assigned dedicated NIC queue pairs
+and hence also dedicated packet buffer pools associated with those queue
+pairs. This removes any synchronization bottlenecks on memory allocation. We
+note that DPDK uses reference counting to decide when to free a packet
+buffer. RAFT cores continue to hold a reference to a packet containing an RPC
+call even after dispatching it for execution until it has completed replicating
+the contents of the packet and is ready to compact away that entry from its RAFT
+log.
 
 \subsection{RPC Call Execution}
-Call steering
-Exactly once
+\label{sec:exec}
+We now describe how RPC calls are executed and results returned to the
+client. There are two important aspects of RPC call execution in Cyclone -
+steering, and at most once semantics.
 
-\section{Putting it Together - A linearizable hash table}
-\label{sec:examples}
+An RPC call via Cyclone must traverse a specific RAFT instance (RAFT log) and
+specific execution core. These details are provided by the user when making the
+RPC call to the client side library provided by Cyclone. The user therefore is
+aware of the number of execution cores and RAFT instances and given complete
+control of \emph{RPC call steering}.
+
+Cyclone provides at most once semantics. It supports a large but fixed (compile
+time configured) number of clients. Each RPC call from a client bears a sequence
+number. \emph{Each execution core independently enforces execution of RPC calls
+  in increasing sequence numbers}. This is done by storing the last executed
+sequence number in a per-core durable datastructure as part of the crash
+consistent transaction wrapping the execution of the RPC call itself. Even in
+the presence of fail-overs, this durable memory of sequence numbers seen from a
+client ensures at most once semantics. In addition to guarding agains repeated
+execution of the same call, Cyclone also remembers the \emph{results} of the
+last executed RPC call. In order to assist clients (particularly stateless
+clients), Cyclone allows the client to query the last seen sequence number and
+result for that RPC call from the quorum of servers.
+
+\section{Putting it together - A Durable Replicated Hash Table}
+\label{sec:example}
 We now demonstrate how Cyclone can be used in practise by putting together an
-example of a linearizable set of counters. The problem is to construct a sparse
-set of counters indexed by a 64 bit key and support \emph{linearizable}
-increment, decrement and an atomic set increment. The atomic set increment takes
-as input a set of counter indexes and their current value and atomically
-increments the counters from the the given values.
+example of a linearizable hash table - a useful and common data structure
+[memcached]. The baseline is a durable hash table written in NVML that runs on a
+single machine in client-server mode. The server side maintains the durable
+open-addressed hash table and services client RPC calls to insert and delete
+key-value pairs and lookup keys. Such a hash table is simple to derive from a
+volatile hash table and readily available in NVML [nvml.io] as an example [src
+  code link]. To use that (non client-server) example with Cyclone we add a
+simple RPC call format, manually marshall and demarshall the (at most two)
+arguments to the RPC call in an RPC call handler we wrote (a few tens of lines
+of code) and wrote a simple client driver program (a few tens of additional
+lines of code).
 
-Figure~\ref{fig:counter_set} shows a linearizable implementation executing on
-servers. Odd counter values are considered unstable and even values stable.
-Clients make RPC calls to servers the trigger the execution of the
-appropriate function. Any client can make calls for the get and increment
-operations shown on the left hand side. A client retries an operation until it
-finds a stable counter value.
+In order to use the hash table with Cyclone, we simply need to to declare the RPC
+call handler as the point of entry to Cyclone at the server end - shown in
+Figure []. The client end issues RPC calls via Cyclone's client side library as
+shown in Figure []. Crucially neither side involves fault awareness or fault
+recovery code - a key goal for Cyclone.
 
-A single distinguished client called the co-ordinator runs any atomic increment
-set operations on behalf of other clients - shown on the right hand side of
-Figure~\ref{fig:counter_set}. We chose to have a single
-distinguished co-ordinator to avoid making the example complex by introducting
-concurrency.
+There are two challenges in scaling the performance and improving the utility of
+the linearizable and durable hash table. The first is to ensure linearizability
+while scaling performance by adding RAFT cores and application cores. The second
+is to support distributed transactions - that manipulate a set of key-value
+pairs atomically in the face of failure.
 
-The operations in Figure~\ref{fig:counter_set} do not comprehend failures.
-Cyclone includes a client library to automatically sequence RPC calls at the client
-side, retry them, connect to a new quorum leader on a view change and finally
-take care of unrealiable networks. At the server side the results of increment,
-prepare and commit RPC calls are always persisted and replicated before the
-response to the client. For individual get, increment, prepare and commit
-operations therefore persistence, replication and failover is transparently
-handled.
+\subsection{Scaling}
+Scaling the performance of the hash table by using multiple RAFT cores and
+execution cores presents determinism related challenges. Two RPC calls being
+steered through different RAFT cores can end up executing in different orders on
+different replicas even if steered to the same execution core. At the same time,
+two RPC calls steered through the same RAFT core but dispatched to different
+execution cores can end up completing in different orders in different replicas.
 
-The AtomicIncrementSet operation presents a special case however. Clients can
-fail and therefore a client executing the steps of the operation can fail midway
-leaving some counters in an unstable state forever. The solution is the
-replicate the distinguished co-ordinator client too. The co-ordinator therefore
-becomes both a server, receiving AtomicIncrementSet operations from clients as
-well as a client, making individual Prepare and Commit calls to the quorum
-leader. The implementation of the AtomicIncrementSet however requires guarantees
-beyond that of simple asynchronous replication
+Our solution to this problem in Cyclone is one that applies to most durable
+data structures - not just the hash table we consider here. We observed that
+replicas of the durable data structure require \emph{per-key determinism}
+and not determinism across \emph{all} operations. We therefore steer
+operations to the same key to the same RAFT core and the same execution core, by
+hashing the key to determine the RAFT core and execution core. The actual
+configuration of the hash table therefore can be different across different
+replicas - for example due to the insertion of two keys mapping to the same hash
+bucket completing in different orders across different replicas. However, since the
+results of insert, delete and lookup operations for a key depend only on
+previous operations to the same key the replicas provide identical results for
+committed operations.
 
-\begin{itemize}
-  \item Output Commit: AtomicIncrementSet has externally visible side effects at
-    other servers when execting. It should therefore be guaranteed to finish
-    executing once it starts
-  \item Exactly once: Regardless of restarts AtomicIncrementSet should be
-    preceived to have executed exactly once by an external observer.
-\end{itemize}
+The RPC call steering mechanism described above leaves the programmer free to
+use any synchronization mechanism they wish between concurrent accesses from the
+application cores to the data structure, since the ordering between accesses
+from different cores is allowed to be non-deterministic. For our open addressed
+hash table we chose the simple scheme of a per-bucket lock.
 
-The solution is to simply run AtomicIncrementSet in synchronous replication
-mode. This guarantees the first property. It also guarantees that exactly once
-execution will succced in ``leader'' mode with followers simply remembering the
-success status of a transaction. To ensure that due to a failover we do not
-prepare or commit a counter more than once, we use exactly once RPCs that
-prevent RPC calls from executing more than once. Finally, since only
-one AtomicIncrementSet is executed at a time, we include the current step (the
-place in the for loop) in RPC calls to the quorum leader and have it reflect the
-information back in the response (a cookie of sorts). By querying the results of
-the last made call, the co-ordinator can simply fast forward to the correct
-index in the for loop during recovery and use the value of the ``success'' field
-in the response to update its own local success variable.
+\subsection{Distributed Transactions}
+A useful primitive for many distributed systems is the ability to run
+distributed transactions - an atomic sequence of conditional updates spanning
+multiple machines. Distributed transactions have two distinct facets:
+atomicity and fault tolerance. Distributed transactions appear atomic with
+respect to other atomic transactions and simple updates or reads by virtue of
+standard concurrency control techniques. The most common technique is two phase
+locking: cquire locks, verify that previously read objects are at the desired
+values, update the locked objects and finally release the locks. A survey of
+various distributed systems and informal discussions with programmers led us to
+the conclusion that this aspect of distributed transactions was fairly easy to
+understand and implement. The hard part for most programmers is the second
+aspect: fault tolerance. A distributed transaction spans many machines and has a
+large footprint to clean up in the event of failure. Failure handling is made
+more complex when considering cascading failures - further failures while
+recovering from a failure. Cyclone was designed to allow primitives such as
+distributed transctions to be added without the programmer needing to worry
+about faults or write fault recovery code - as we demonstrate with the hash
+table.
 
-\begin{figure*}
-\begin{tabular}{cc}    
-\begin{minipage}{0.5\textwidth}  
-{ \scriptsize
-\begin{verbatim}
-bool is_stable(uint64_t value)
-{
-  if(value is even) {
-    return true
-  }
-  else {
-    return false
-  }
-}
+We support distributed transactions in our hash table by allowing the user to
+specify a group of insert/delete/lookup operations that must occur
+atomically. We use the existing bucket locks to implement a simple two phase
+locking protocol to implement the transction. A client makes an RPC call to the
+server specifying the transaction and the server executes it on the client's
+behalf. Figure [] shows pseudocode for the transaction. Execution of the
+transaction has four distinct phases: acquire locks, verify values, apply
+updates and finally release locks.
 
-Get(uint64_t index)
-{
-  if(not is_stable(counters[index]) {
-    return EUNSTABLE
-  }
-  else {
-    return counters[index]
-  }
-}
+We guard against fault tolerance by executing the RPC call on a quorum of
+machines via Cyclone. Since execution of the RPC call leaves \emph{externally
+  visible side effects} by virtue of the RPC calls it makes we require output
+commit [] - the property that the execution is guaranteed to complete once it
+starts. This is simply achieved by a flag in the RPC call that turns off
+decoupled execution (Section~\ref{sec:decouple}). The flag causes execution to
+start only after the RPC call has been successfully replicated.
 
-Increment(uint64_t index)
-{
-  if(not is_stable(counters[index]) {
-    return EUNSTABLE
-  }
-  else {
-    counters[index] = counters[index] + 2
-    return SUCCESS
-  }
-}
-\end{verbatim}
-}
-\end{minipage} &
-\begin{minipage}{0.5\textwidth}  
-{ \scriptsize
-\begin{verbatim}
-Prepare(uint64_t index, 
-        uint64_t value, 
-        bool success)
-{
-  if(counters[index] == value) {
-    success = success AND TRUE
-  }
-  else {
-    success = success AND FALSE
-  }
-  // Make unstable
-  counters[index] = counters[index] + 1
-}
+Next, we need to ensure that although every machine in the quorum makes all
+necessary RPC calls, the call itself is executed only once. This is achieved by
+using Cyclone's at most once semantics. All machines in the quorum executing the
+transaction use the same client identifier and the same montonically increasing
+sequence of RPC call numbers. Finally, we need to deal with the fact that
+Cyclone only remembers the last executed RPC call for a client. This means that
+if different machines in the quorum move at different speeds, some machines will
+not receive a response to the RPC call - they will instead receive an error that
+tells them that they are too far behind. To ensure that the transaction is still
+guaranteed to complete, we use a \emph{different} client identifier for each
+phase. Thus, until the transaction completes and a new one is started the
+response for any RPC call continues to be available. Figure [] details in
+pseudocode the execution of a transaction with these added details. Crucially
+there is no need for awareness of faults nor any need to write fault recovery
+code. Finally, we need to ensure that the client can reliably query the result
+of the transaction. If the leader of the quorum executing the transaction falls
+behind it cannot supply the result. We therefore immediately return and
+acknowledment to the client as soon as the RPC call for the transaction has been
+replicated. We store the result of the transaction in a special data structure
+updated through an additional RPC call made at the end of the transaction. The
+client queries this data structure to determine the fate of the transaction.
 
-Commit(uint64_t index, bool success)
-{
-  if(success == TRUE) {
-    // Update
-    counters[index] = counters[index] + 1
-  }
-  else {
-    // Rollback
-    counters[index] = counters[index] - 1
-  }
-}
+\subsection{Checkpoints}
 
-AtomicIncrementSet(uint64_t indices[],
-                   uint64_t pre_values[],
-                   int count)
-{
-  bool success = TRUE
-  for i = 1 to count
-    success =
-      success AND
-        Prepare(indices[i],
-                pre_values[i],
-                success)
-  for i = 1 to count
-    Commit(indices[i], success)       
-  return success
-}
-\end{verbatim}
-}
-\end{minipage}
-\end{tabular}
-\caption{Counter Set}
-\label{fig:counter_set}
-\end{figure*}
+\section{Evaluation}
 

### 2016-08-28 core/cyclone.cpp
Print raft header size.

@@ -584,7 +584,7 @@ void* cyclone_boot(const char *config_quorum_path,
   cyclone_handle->me_quorum       = quorum_id;
   cyclone_handle->raft_handle = raft_new();
   raft_set_multi_inflight(cyclone_handle->raft_handle);
-
+  BOOST_LOG_TRIVIAL(info) << "RAFT start. sizeof(msg_t) is :" << sizeof(msg_t);
   /* Setup raft state */
   if(access(path_raft.c_str(), F_OK)) {
     // TBD: figure out how to make this atomic

### 2016-09-08 paper/paper.tex
Additions to paper.

@@
 \subsection{Network Stack}
-Replication protocols are expensive to operate - fact that has been underlined in
-multiple pieces of research [FARM, in a box]. It would therefore appear that
+Replication protocols are expensive to operate - fact that has been underlined
+in multiple pieces of research [FARM, in a box]. It would therefore appear that
 replication would become the bottleneck in Cyclone. We observed however that
@@
 stack possibly explaining why they chose both high performance hardware (40
-GigE+ NICS, custom FPGA designs) as well as specific high performance but
-restrictive software APIs (one sided RDMA with metadata co-location, application
-compiled down to the FPGA). As a counterpoint, we designed Cyclone around DPDK
+GigE+ NICS, custom FPGA designs) as well as specific software APIs (one sided
+RDMA with metadata co-location, application compiled down to the FPGA) that
+bypass the network stack. As a counterpoint, we designed Cyclone around DPDK
 [dpdk] a high performance user-space network stack. DPDK provides convenient
@@
 
-Cyclone separates the available cores on the machine into cores dedicated to
-work related to the consensus protocol and cores decidated to running
-application code. The division is configurable but Fig [] shows the division for
-our testbed. Each RAFT core runs an \emph{independent} instance of the RAFT
-protocol (Section~\ref{sec:ind}). RAFT cores pass work to the application cores
-through lock-free double ended queues. A RAFT core dispatches an incoming RPC
-call for execution simultaneously with sending it out for replication. As
-explained in Section~\ref{sec:decouple} this is beneficial for reducing latency
-to completion.
+Cyclone separates the available cores on a machine into cores dedicated to work
+related to the consensus protocol and cores decidated to running application
+code. The division is configurable. Each RAFT core runs an \emph{independent}
+instance of the RAFT protocol (Section~\ref{sec:ind}). RAFT cores pass work to
+the application cores through lock-free double ended queues. A RAFT core
+dispatches an incoming RPC call for execution simultaneously with sending it out
+for replication. As explained in Section~\ref{sec:decouple} this is beneficial
+for reducing latency to completion.
 
@@
 RPC call is sent from the application core via its dedicated output queue. This
-division of NIC queues eliminates any need for synchronization between RAFT and
-application cores when accessing the NIC.
+division of NIC queues eliminates any need for synchronization between RAFT
+cores and application cores when accessing the NIC.
+
+In the event that the machine has multiple NICs Cyclone binds each RAFT or
+application core to a specific NIC and picks the necessary queues from that
+NIC. Spreading the network demands from the cores across the NICs in this manner
+allows Cyclone to scale with multiple NICs when available.
 
@@
 quorum leader is elected, the leader's primary job is to transmit new entries
-appended to the log to its followers. The primary job of the followers is to
-receive these entries and send acknowledgments to the leader. An entry is
-considered committed when the leader receives an acknowledgment for an entry
-from a majority of the quorum. Followers receive intimation of committed entries
-as piggybacked information on subsequent replication messages.
+appended to the log to its followers. The primary job of a follower is to
+receive these entries, append them to its log and send acknowledgments to the
+leader. An entry is considered committed when the leader receives an
+acknowledgment for an entry from a majority of the quorum. Followers receive
+intimation of committed entries as piggybacked information on subsequent
+replication messages.
 
@@
 the RPC call contents for transmission to each follower replica. Hence, we built
-a zero copy mechanism for common case of a leader replica sending the same log
-entry to a set of follower replicas. To achieve our zero copy goal we explot the
-fact that DPDKs packet buffers allow prepending a data without physically moving
-the data in the packet buffer. Fig [] shows how the layout of a packet buffer
-changes at the leader during replication. The received packets contains only the
-RPC call from the client. We prepend a RAFT related header (containing the
-previous log index, previous log term, leader's term and leader's commit
-point) to packet buffer itself. To this RAFT header we then prepend an IPv4
-header. Although we use the raw packet interface from DPDK and do not use IP
-based routing, we set IP addresses in a dummy header to use 5-tuple steering in
-order to place the packet in the right queue at the receiving NIC. This
-composite packet buffer is then sent to every replica. In order to properly
-route the packet through switches, we allocate a small buffer containing an
-ethernet header for each follower replica and \emph{chain} it in front of the
-prepared packet - ensuring we do not need to change the contents for every
-replica.
+a zero copy mechanism for the common case of a leader replica sending the same
+log entry to a set of follower replicas. To achieve our zero copy goal, we
+exploit the fact that DPDKs packet buffers allow prepending data without
+physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
+how the layout of a packet buffer changes at the leader during replication. The
+received packets contains only the RPC call from the client. We prepend a RAFT
+related header (containing the previous log index, previous log term, leader's
+term and leader's commit point) to packet buffer itself. To this RAFT header we
+then prepend an IPv4 header. Although we use the raw packet interface from DPDK
+and do not use IP based routing, we set IP addresses in a dummy header to use
+5-tuple steering in order to place the packet in the right queue at the
+receiving NIC. This composite packet buffer is then sent to every replica. In
+order to properly route the packet through switches, we allocate a small buffer
+containing an ethernet header for each follower replica and \emph{chain} it in
+front of the prepared packet - ensuring we do not need to change the contents
+for every replica.
 
@@
 hundreds of cycles for each log entry to be processed. This adds up to a few
-microseconds, a significant amount in the face of sub 10 us latency between
-nodes with DPDK. We amortize this overhead when operating under load by batching
-multiple RPC calls together and treating the batch as a single RAFT log entry.
-DPDK by itself can efficiently batch incoming packets under load - allowing the
-user to easily receive a burst (currently limited to 32) packets rather than a
-single one for a polling call. DPDK further allows chaining packet buffers
-together and treating the chain as a single packet. This is convenient for our
-zero copy approach and we chain packet mbufs together to form a single packet
-as shown in Fig []. Since the entire chain is sent as a single packet when being
-replicated, the length of a chain is limited by the size of the coalesced
-packet, that must fit into an ethernet jumbo frame of approximately 9K bytes.
+microseconds, a significant amount in the face of sub 10 us network latency
+between nodes with DPDK. We amortize this overhead when operating under load by
+batching multiple RPC calls together and treating the batch as a single RAFT log
+entry.  DPDK by itself can efficiently batch incoming packets under load -
+allowing the user to easily receive a burst (currently limited to 32) of packets
+rather than a single one for a polling call. DPDK further allows chaining packet
+buffers together and treating the chain as a single packet. This is convenient
+for our zero copy approach and we chain packet mbufs together to form a single
+packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
+single packet when being replicated, the length of a chain is limited by the
+size of the coalesced packet, that must fit into an ethernet jumbo frame of
+approximately 9K bytes.
 
@@
 \label{sec:example}
-We now demonstrate how Cyclone can be used in practise by putting together an
+We now demonstrate how Cyclone can be used in practice by putting together an
 example of a linearizable hash table - a useful and common data structure
@@
 
-In order to use the hash table with Cyclone, we simply need to to declare the RPC
-call handler as the point of entry to Cyclone at the server end - shown in
-Figure []. The client end issues RPC calls via Cyclone's client side library as
-shown in Figure []. Crucially neither side involves fault awareness or fault
-recovery code - a key goal for Cyclone.
+In order to use the hash table with Cyclone, we simply need to to declare the
+RPC call handler as the point of entry to Cyclone at the server end - shown in
+Figure~\ref{fig:server_api}. The client end issues RPC calls via Cyclone's
+client side library as shown in Figure~\ref{fig:client_api}. Crucially neither
+side involves fault awareness or fault recovery code - a key goal for Cyclone.
 
@@
 pairs atomically in the face of failure.
+
+\begin{figure}
+{ \scriptsize
+\begin{verbatim}
+int exec_rpc_call(void * input, int input_size, void *output)
+{ 
+  ...
+}
+
+struct rpc_callbacks_t callbacks = {
+.exec_call = exec_rpc_call,
+...
+};
+
+main()
+{
+  ...
+  cyclone_start(callbacks, ...);
+  ...
+}
+\end{verbatim}
+}
+\caption{Server side API}
+\label{fig:server_api}
+\end{figure}
+
+\begin{figure}
+{ \scriptsize
+\begin{verbatim}
+int make_rpc_call(..., void *data, int size, int rpc_number);
+int get_last_rpc_number(...);
+main()
+{
+  ...
+  int rpc_number = get_last_rpc_number(...);
+  while(true) {
+    ...
+    make_rpc_call(..., data, size, rpc_number++);
+
+  }
+  ...
+}
+\end{verbatim}
+}
+\caption{Client side API}
+\label{fig:client_api}
+\end{figure}
 

### 2016-09-11 paper/paper.tex
More text in paper.

@@
 quorum leader is elected, the leader's primary job is to transmit new entries
-appended to the log to its followers. The primary job of a follower is to
-receive these entries, append them to its log and send acknowledgments to the
-leader. An entry is considered committed when the leader receives an
-acknowledgment for an entry from a majority of the quorum. Followers receive
-intimation of committed entries as piggybacked information on subsequent
+appended to its persistent log to its followers. The primary job of a follower
+is to receive these entries, append them to its persistent log and send
+acknowledgments to the leader. An entry is considered committed when the leader
+receives an acknowledgment for an entry from a majority of the quorum. Followers
+receive intimation of committed entries as piggybacked information on subsequent
 replication messages.
+
+In order to avoid copying the incoming RPC call to a persistent memory range,
+Cyclone uses persistent memory on the system itself to allocate space for packet
+buffers. However although a packet is accessible at the core the most up to date
+data for the memory region occupied by the packet is in the CPU cache and not
+necessarily in memory. To ensure that packet data has truly been made
+persistent, Cyclone issues a series of cacheline flushes. This ensures that the
+packet data is persistent. Cyclone maintains its log as a circular sequence of
+pointers to packet buffers and therefore the append operation is completed by
+adding a pointer to the received packet and flushing the cacheline containing
+the pointer.
 

### 2016-09-14 paper/paper.tex
Added list of graphs.

@@
 
-\subsection{Checkpoints}
+\section{Evaluation}
+\begin{figure}
+  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  ports = 4. Fix payload = 0. One line per number of replicas (1,2,3).
+  \caption{Replicaton Performance - Varying replica count}
+  \label{fig:perf_rep}
+\end{figure}
 
-\section{Evaluation}
+\begin{figure}
+  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  replicas = 3. Fix payload = 0. One line per number of ports.
+  \caption{Replicaton Performance - Varying ethernet ports}
+  \label{fig:perf_ports}
+\end{figure}
+
+\begin{figure}
+  Source: results_raw.csv. Fix replicas = 3. Set of 4 graphs. One graph for each
+  port count (1,2,3,4). X-axis throughput (Gbps). Y-axis latency. One line per
+  payload size.
+  \caption{Replicaton Performance - Varying payload size}
+  \label{fig:perf_payload}
+\end{figure}
+
+\begin{figure}
+  Source: results_clwb.csv. Bar graph with 3 groups of bars - one group for
+  payload = 0, one group for payload = 190, one for 512. Each group has 3 bars -
+  one for window =1, one for window = 10, one for window = 10000 (infty)
+  \caption{Replicaton Performance - CLFLUSH window}
+  \label{fig:perf_clflush}
+\end{figure}
+
+\begin{figure}
+  Source: results_quorum.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  replicas = 3. Fix payload = 0. One line for each of ports = 1,2,3,4.
+  \caption{Replicaton Performance - quorum count}
+  \label{fig:perf_quorums}
+\end{figure}
+
+\begin{figure}
+  Source: results_cc.csv.Fix replicas = 3. X-axis is throughput
+  (Mops/sec). Y-axis is Latency. One line for each of ports = 1,2,3,4.
+  \caption{Crash consistency overheads}
+  \label{fig:perf_cc}
+\end{figure}
+
+\begin{figure}
+  Source: results_app.csv. X-axis is throughput
+  (Mops/sec). Y-axis is Latency. One line for each of read_fraction = 0, 50, 100
+  \caption{Hash Table Performance}
+  \label{fig:perf_app}
+\end{figure}
 

### 2016-09-18 paper/paper.tex
One more graph.
Prepare for short paper.

@@
-\documentclass[twocolumn]{article}
-\usepackage{times}
-\usepackage{parskip}
+\documentclass[letterpaper,twocolumn,10pt]{article}
+\usepackage{usenix,epsfig,endnotes}
 \begin{document}
-\title{Cyclone: Replication Middleware for NVM Clusters}
-\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
+
+%don't want date printed
+\date{}
+
+%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
+\title{\Large \bf Cyclone: Replication Middleware for NVM Applications}
+
+%for single author (just remove % characters)
+%\author{
+%  {\rm Paper #: XXX }\\
+%  XXX
+%  \and
+%  {\rm YYY}\\
+%  YYY
+%} % end author
+
 \maketitle
-\begin{abstract}
+
+% Use the following at camera-ready time to suppress page numbers.
+% Comment it out when you first submit the paper for review.
+\thispagestyle{empty}
+
+\subsection*{Abstract}
 Memory technology in the datacenter is due to undergo a paradigm shift with the
@@
 battery backed non-volatile DIMMs as well as newer memory types such as
-3D XPoint. This will allow a broad range of new applications that exploit the
-combination of durability with performance far superior to traditional durable
-storage media such as NAND flash. However, durability does not guaratee
-availability and without a means to survive machine or storage medium failures
-these applications cannot match the survivability offered by sophisticated
-distributed file systems, databases and key value stores.
+3D XPoint. In anticipation, libraries that allow programmers to build durable
+data structures on non-volatile heaps have become available. A missing piece
+however is a component to make such durable data structures available using
+replication across a commodity network such as ethernet. Existing solutions that
+offer good performance are biased towards pure transactions across RDMA with a
+dependence on external services and application specific code for fault recovery.
 
-Cyclone is replication middleware aimed at helping application developers
-targeting new memory types to easily add fault tolerance to their
-applications. Cyclone eschews external co-ordination services like zookeeper and
-instead replicates RPC calls from clients being servered out of durable memory
-across a quorum of clusters using the RAFT consensus protocol. Cyclone also
-provides a client side library to enable transparent failover. Finally, Cyclone
-is designed to provide at most once semantics to RPC calls. We demonstrate in
-this paper that this is sufficient to design applications that support
-distributed transactions without any need for the developer to be concerned with
-fault recovery - a challenging aspect of designing systems that support
-distributed transaction and at the same time are fault tolerant.
+Cyclone is replication middleware that aims to add fault tolerance as
+transparently as possible. A design for Cyclone is that NVM application
+developers should not need to add fault recovery code. A second design goal for
+Cyclone is to strive for good performance on commodity ethernet - with
+innovations in the software stack rather than a requirement for RDMA capable
+networking hardware.
 
-Cyclone is also designed to provide maximum performance without being dependent
-on high end networking stacks - such as RDMA/Infiniband combinations.  To enable
-the lowest possible latency we use DPDK - a userspace networking stack and
-designed our implementation of RAFT to use zero copy for the replication
-process. To enable maximum throughput under load we extensively batch
-replication operations and support multiple concurrent instances of RAFT to
-maximize throughput. As a result, Cyclone can replicate as many as 4 million 100
-byte RPC messages every second on a very modest commodity 10 GigE network. Under
-low load, the latency seen by a client for an RPC call the requires replication
-is of the order of 20 us.
-\end{abstract}
+Cyclone achieves these goals by providing a simple API for client server
+applications that allows server state to be replicated by executing the same
+sequence of RPC calls across all replicas. It provides at most once semantics
+with automatic failover to hide fault recovery from programmers. Finally it uses
+DPDK - a high performance, low latency network stack coupled with batching,
+zero-copy and the ability to scale replication performance across
+multiple instances of a consensus protocol to replicate millions of small
+payload RPC messages a second on commodity ethernet.
 
@@
 TBD
-\section{Background: Single Node Programming Model}
-Cyclone builds on top of Intel's Non Volatile Memory Library (NVML) and provides
-transparent replication to programs written to use NVML for persistence on a
-single node. We describe NVML's programming model in this section to provide
-adequate background for the rest of this paper.
-
-NVML provides special functions to allocate and free persistent memory as well
-as fat pointers to persistent memory to ensure that referential integrity in the
-heap is maintained across restarts - since the virtual memory ranges into which
-persistent memory is mapped may change.
-
-\begin{figure}
-{\scriptsize
-\begin{verbatim}
-
-struct ll_node {
-  int value;
-  TOID(ll_node) next;
-};
-
-void insert_after(TOID(ll_node) prev, 
-                  TOID(ll_node) new_node)
-{
-  D_RW(new_node)->next = D_RO(prev)->next;
-  D_RW(prev)->next = new_node;
-}
-
-\end{verbatim}
-}
-\caption{A Persistent Linked List of Integers}
-\label{fig:example}
-\end{figure}
-
-Figure~\ref{fig:example} shows the example of a linked list node and a function
-for inserting a new node after a given one. The {\tt TOID} declaration wraps the
-linked list node type to produce a fat pointer to it. A {\tt TOID(ll\_node)}
-type therefore is a pointer to a linked list node stored in persistent memory.
-The {\tt insert\_after} function uses the {\tt D\_RO} and {\tt D\_RW} calls to
-convert fat pointers into directly accessible read-only and read-write pointers
-respectively. Other than the mechanics of using fat pointers the code entirely
-resembles that which a programmer might write for a standard linked list in
-DRAM.
-
-\subsection{Failure Atomicity}
-Programmers typically find it hard to reason about failures when writing
-code. This is a problem when directly manipulating data structures in persistent
-memory. Depending on the specifics of the system architecture, updates to
-persistent memory can be persisted out of order thereby leaving the data
-structure in an inconsistent state. To illustrate this consider a failure during
-execution of the code to add a node to a linked list in
-Figure~\ref{fig:example}. If either of the two updates to the next fields of the
-previous and new node fail to be persisted, the linked list is left in a
-disconnected state with one or more connected nodes inaccessible.
-
-NVML provides tools for programmers to write code that is failure
-atomic by allowing a block of code to be delimited as a transaction. A
-transaction is either completely executed in terms of its effects on
-persistent memory or not executed at all. Under the hood, NVML tracks all
-updates to persistent memory in a persistent undo log - the log entries are
-persisted before the update to the actual memory is allowed. On a failure, the
-undo log is used to roll back incomplete transactions. A transaction is
-considered complete when it completes execution and the updated objects are
-flushed (automatically by NVML) to persistent memory. The log entries for the
-transaction are then immediately deleted. The only burden placed on a programmer
-for failure atomicity is to delimit transactions and that they must inform NVML
-about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
-convert the function for adding a node to a linked list into a failure atomic
-one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
-calls to mark the objects being updated \emph{before} the actual updates to
-them.
+\section{Single Node Programming Model}
+Cyclone starts with the assumption that the programmer has already built a
+concurrent durable application on a non-volatile heap. A number of libraries and
+filesystems make this possible today [DAX, NVML, others ?]. As a starting point,
+we support applications built on top of Intel's Non Volatile Memory Library
+(NVML) [nvml.io]. We describe NVML's programming model in this section to provide
+adequate background for the rest of this paper. The primary feature of interest
+is the ability to write code that is crash consistent. In other words code that
+makes changes to the non-volatile heap is enclosed in a crash consistent
+transaction such that either all of the transaction executes or none of it
+executes in the face of power failures. This is critical to maintaining
+consistency as the data structure, being durable, must survive across power
+failures. To illustrate this idea, Figure~\ref{fig:example} shows how one might
+build a persistent linked list using NVML - the key idea is to enclose the
+entire process of updating the linked list within a crash consistent
+transaction. This ensure that the linked list is not corrupted due to a power
+failure. A second point worth noting is that NVM libraries such as NVML usually
+provide a way to wrap pointers to non-volatile memory in order that such memory
+may be mapped to different locations in virtual memory across restarts without
+impact to the accessing application.
 
@@
 \begin{verbatim}
-
 struct ll_node {
@@
   }
-\caption{Failure Atomic Peristent Linked List}
-\label{fig:example_fatomic}
+\caption{A Persistent Linked List}
+\label{fig:example}
 \end{figure}
 
-A crucial feature of NVML is that it abstracts the details of the persistence
-mechanism away from programmers. This allows the same code to be deployed on
-across a range of persistence mechanisms - mapping pages from secondary
-storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
-save DRAM state to SSD on failure [FARM] or using directly accessed persistent
-memory [PMFS]. Each of these options has different requirements. For example the
-NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
-instructions. NVDIMMs require only a CLFLUSH to flush the
-cacheline and ensure that the line enters the asynchronous data refresh
-domain. Using secondary storage is also an option with faster 3DXPoint storage
-behind an NVMe interface - this would require an {\tt msync} and OS pagecache
-intervention to persist data to secondary storage. NVML seamlessly uses the
-right persistence mechanism without requiring code using the library to be aware
-of these details.
+It is easy to turn the linked list example in NVML into a concurrent linked list
+by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
+so desired for performance. More sophisticated fine-grained or optimistic
+concurrency control mechanisms for NVM are also possible [mnemosyne].
 
@@
 via replication - a quorum of machines maintains \emph{equivalent} state. Server
-state is queried and manipulated via RPC calls from clients. Cyclone's approach
-to replication is simple - we replicate the RPC call itself across replicas
-rather than replicating every access made during execution of the RPC call,
-which would be prohibitively expensive.
+state is queried and manipulated via RPC calls from clients. Cyclone replicates
+the RPC call itself across replicas rather than replicating every access made
+during execution of the RPC call.
 
-Cyclone provides strongly consistent replication - every machines in the quorum
-of replicas maintains a log of RPC calls. All RPC calls (including read-only
-calls that do not change state and therefore do not need replication) are always
-executed by the leader replica. Cyclone treats the RPC call itself as a variable
+Cyclone provides strongly consistent replication - every machine in the quorum
+of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as a variable
 sized opaque blob of data, leaving the application free to choose how it wishes
@@
 instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
-different machines in synchronization. The standard replicated state machine
-approach requires that commands (entries in the log) be committed before they
-can be applied to the state machine i.e. in our case - the RPC call be
-executed. However, the fact that RPC calls that modify state must themselves be
-executed in a crash consistent transaction allows us to overlap execution with
-replication (described in the next section).
+different machines in synchronization. Specially flagged `read-only` RPC calls
+are executed on the leader replica and are not replicated.
 
@@
 \label{sec:decouple}
-Cyclone decouples execution of the RPC call on any replica from its replication
-by performing them in parallel to save time. Normally, this is a problem with
-any consenus protocol including RAFT as replication might failt for a number of
-reasons primarily due to a change of leader and subsequent of rollback of
-previously appended (but not committed) log entries. In order to achieve
-decoupled replication therefore Cyclone depends on the ability of NVML to roll
-back crash consistent transactions. The ability to roll back a crash consistent
-transaction is important to tolerate power failure as incomplete transactions
-must be rolled back after recovery from a power failure to preserve the
-integrity of the heap. NVML also allows the user to roll back a transaction if
-needed. Cyclone exploits this feature by delaying the commit of the failure
-atomic transaction after execution till replication is complete. In the event
-that replication fails, the transaction is aborted and any changes to the non
-volatile heap are rolled back.
+The standard replicated state machine approach requires that commands (entries
+in the log) be committed before they can be applied to the state machine i.e. in
+our case - the RPC call be executed. In contrast, Cyclone decouples execution of
+the RPC call on any replica from its replication.
 
-A pseudocode description of how the runtime handles decoupled replication is
-shown in Figure~\ref{fig:async_rep}.
+However, the fact that RPC calls that modify state must themselves be executed
+in a crash consistent transaction allows us to overlap execution with
+replication. Normally, this would be a problem as replication might failt for a
+number of reasons but primarily due to a change of leader and subsequent of
+rollback of previously appended (but not committed) log entries. To solve this
+problem, Cyclone delays the commit of the failure atomic transaction after
+execution till replication is complete. In the event that replication fails, the
+transaction is aborted and any changes to the non volatile heap are rolled back.
+
+A pseudocode description of how the runtime handles \emph{decoupled replication}
+is shown in Figure~\ref{fig:async_rep}.
 
@@
   if(replication failed) {
-    TX_ABORT;
+     TX_ABORT 
   } 
@@
 }
-\caption{Asynchronous Replication}
+\caption{Decoupled Replication}
 \label{fig:async_rep}
@@
 
-Asynchronous replication provides good performance and we believe would be the
+Asynchronous replication provides good performance and we believe it would be the
 common case for RPC calls replicated via Cyclone. However the program might on
@@
 \subsection{Network Stack}
-Replication protocols are expensive to operate - fact that has been underlined
-in multiple pieces of research [FARM, in a box]. It would therefore appear that
-replication would become the bottleneck in Cyclone. We observed however that
-previous work did not implement replication around a high performance network
-stack possibly explaining why they chose both high performance hardware (40
-GigE+ NICS, custom FPGA designs) as well as specific software APIs (one sided
-RDMA with metadata co-location, application compiled down to the FPGA) that
-bypass the network stack. As a counterpoint, we designed Cyclone around DPDK
-[dpdk] a high performance user-space network stack. DPDK provides convenient
-access to packets arriving at an ethernet NIC by moving them into packet buffers
-with the contained cache lines also directly placed into the last level cache of
-the CPU during DMA.
+Cyclone is designed around DPDK [dpdk] a high performance network stack. DPDK
+provides convenient access to packets arriving at an ethernet NIC by moving them
+into packet buffers with the contained cache lines also directly placed into the
+last level cache of the CPU during DMA [ddio].
 
-Cyclone separates the available cores on a machine into cores dedicated to work
-related to the consensus protocol and cores decidated to running application
-code. The division is configurable. Each RAFT core runs an \emph{independent}
-instance of the RAFT protocol (Section~\ref{sec:ind}). RAFT cores pass work to
-the application cores through lock-free double ended queues. A RAFT core
-dispatches an incoming RPC call for execution simultaneously with sending it out
-for replication. As explained in Section~\ref{sec:decouple} this is beneficial
-for reducing latency to completion.
+Cyclone separates the available CPU cores on a machine, in a configurable way,
+into cores dedicated to work related to the consensus protocol and cores
+decidated to running application code. Each RAFT core runs an \emph{independent}
+instance of the RAFT protocol. RAFT cores pass work to the application cores
+through lock-free double ended queues. A RAFT core dispatches an incoming RPC
+call for execution simultaneously with sending it out for replication.
 
@@
 receive intimation of committed entries as piggybacked information on subsequent
-replication messages.
+replication messages. The common case path at the leader lends itself to zero
+copy since the same set of log entries is sent to all replicas.
 
-In order to avoid copying the incoming RPC call to a persistent memory range,
-Cyclone uses persistent memory on the system itself to allocate space for packet
-buffers. However although a packet is accessible at the core the most up to date
-data for the memory region occupied by the packet is in the CPU cache and not
-necessarily in memory. To ensure that packet data has truly been made
+In order to avoid copying the incoming RPC call contents to a persistent memory
+range, Cyclone uses persistent memory on the system itself to allocate space for
+packet buffers. However although a packet is accessible at the core the most up
+to date data for the memory region occupied by the packet is in the CPU cache
+and not necessarily in memory. To ensure that packet data has truly been made
 persistent, Cyclone issues a series of cacheline flushes. This ensures that the
@@
 
-Our aim when designing Cyclone was to be relatively insensitive to the size of
-the RPC call - i.e. the size of the RAFT log entry being replicated. Since
-Cyclone is targeted towards general purpose applications we intended to support
-a range of RPC call sizes without a degradation in performance caused by copying
-the RPC call contents for transmission to each follower replica. Hence, we built
-a zero copy mechanism for the common case of a leader replica sending the same
-log entry to a set of follower replicas. To achieve our zero copy goal, we
-exploit the fact that DPDKs packet buffers allow prepending data without
+We exploit the fact that DPDKs packet buffers allow prepending data without
 physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
@@
 term and leader's commit point) to packet buffer itself. To this RAFT header we
-then prepend an IPv4 header. Although we use the raw packet interface from DPDK
-and do not use IP based routing, we set IP addresses in a dummy header to use
+then prepend an IPv4 header. We set IP addresses in the header to use
 5-tuple steering in order to place the packet in the right queue at the
@@
 front of the prepared packet - ensuring we do not need to change the contents
-for every replica.
+of the packet itself for every replica.
 
@@
 size of the coalesced packet, that must fit into an ethernet jumbo frame of
-approximately 9K bytes.
-
-\subsection{Multiple RAFT Instances}
-\label{sec:ind}
-We scale replication performance in Cyclone by using multiple cores on the same
-machine to each run an independent instance of the RAFT protocol. For a quorum
-of homogeneous machines, we dedicate the same core and NIC queue pair on each
-machine to the same instance of RAFT. Each instance maintains an
-\emph{independent} log. All the RAFT instances on a machine share the common set
-of application cores, dispatching work to them via the (shared) lock-free double
-ended queues (implemented as circular buffers in DPDK). It is important to note
-that there is no fixed ordering between RPC calls dispatched for execution from
-different RAFT instances. We discuss in Sections
-~\ref{sec:exec},\ref{sec:example} how a Cyclone user can still impose
-determinism for linearizable data structures, while continuing to scale across
-the multiple RAFT instances.
-
-Other than the lock free queues for dispatching work to application cores, the
-multiple RAFT instances share nothing, being assigned dedicated NIC queue pairs
-and hence also dedicated packet buffer pools associated with those queue
-pairs. This removes any synchronization bottlenecks on memory allocation. We
-note that DPDK uses reference counting to decide when to free a packet
-buffer. RAFT cores continue to hold a reference to a packet containing an RPC
-call even after dispatching it for execution until it has completed replicating
-the contents of the packet and is ready to compact away that entry from its RAFT
-log.
+approximately 9KB.
 
@@
 client. There are two important aspects of RPC call execution in Cyclone -
-steering, and at most once semantics.
+steering and at most once semantics.
 
-An RPC call via Cyclone must traverse a specific RAFT instance (RAFT log) and
-specific execution core. These details are provided by the user when making the
-RPC call to the client side library provided by Cyclone. The user therefore is
-aware of the number of execution cores and RAFT instances and given complete
-control of \emph{RPC call steering}.
+An RPC call via Cyclone must traverse a specific RAFT core (and associated
+consensus protocol instance) and specific application core. These details are
+provided by the user when making the RPC call to the client side library
+provided by Cyclone. The user therefore is aware of the number of execution
+cores and RAFT instances and given complete control of \emph{RPC call steering}.
+For most applications we envisage a simple modulo hashing scheme should suffice
+for load balancing.
 
-Cyclone provides at most once semantics. It supports a large but fixed (compile
-time configured) number of clients. Each RPC call from a client bears a sequence
-number. \emph{Each execution core independently enforces execution of RPC calls
-  in increasing sequence numbers}. This is done by storing the last executed
-sequence number in a per-core durable datastructure as part of the crash
-consistent transaction wrapping the execution of the RPC call itself. Even in
-the presence of fail-overs, this durable memory of sequence numbers seen from a
-client ensures at most once semantics. In addition to guarding agains repeated
-execution of the same call, Cyclone also remembers the \emph{results} of the
-last executed RPC call. In order to assist clients (particularly stateless
-clients), Cyclone allows the client to query the last seen sequence number and
-result for that RPC call from the quorum of servers.
+Cyclone provides at most once semantics. It supports a large but fixed (at
+startup time) number of clients. Each RPC call from a client bears the client
+number and a sequence number. \emph{Each execution core independently enforces
+  execution of RPC calls in increasing sequence numbers}. This is done by
+storing the last executed sequence number in a per-core durable datastructure as
+part of the crash consistent transaction wrapping the execution of the RPC call
+itself. Even in the presence of fail-overs, this durable memory of sequence
+numbers seen from a client ensures at most once semantics. In addition to
+guarding agains repeated execution of the same call, Cyclone also remembers the
+\emph{results} of the last executed RPC call. This is done to assist clients
+(particularly stateless clients). Cyclone allows the client to query the last
+seen sequence number and result for that RPC call from the quorum of servers.
 
-\section{Putting it together - A Durable Replicated Hash Table}
+\section{A Durable Replicated Hash Table}
 \label{sec:example}
 We now demonstrate how Cyclone can be used in practice by putting together an
-example of a linearizable hash table - a useful and common data structure
+example of a linearizable hash table - a useful and common data structure.
 [memcached]. The baseline is a durable hash table written in NVML that runs on a
@@
 volatile hash table and readily available in NVML [nvml.io] as an example [src
-  code link]. To use that (non client-server) example with Cyclone we add a
-simple RPC call format, manually marshall and demarshall the (at most two)
-arguments to the RPC call in an RPC call handler we wrote (a few tens of lines
-of code) and wrote a simple client driver program (a few tens of additional
-lines of code).
+  code link]. To use that (non client-server) example with Cyclone we added a
+simple RPC call format, manually marshalled and demarshalled the few
+arguments to the RPC call in an RPC call handler we wrote - a few tens of lines
+of code. Finally, we added a simple client driver program (a few tens of
+additional lines of code).
 
@@
 
-There are two challenges in scaling the performance and improving the utility of
-the linearizable and durable hash table. The first is to ensure linearizability
-while scaling performance by adding RAFT cores and application cores. The second
-is to support distributed transactions - that manipulate a set of key-value
-pairs atomically in the face of failure.
+There are two interesting challenges in scaling the performance and improving
+the utility of the linearizable and durable hash table. The first is to ensure
+linearizability while scaling performance by adding RAFT cores and application
+cores. The second is to support distributed transactions - that manipulate a set
+of key-value pairs atomically in the face of failure.
 
@@
 Scaling the performance of the hash table by using multiple RAFT cores and
-execution cores presents determinism related challenges. Two RPC calls being
+application cores presents determinism related challenges. Two RPC calls being
 steered through different RAFT cores can end up executing in different orders on
-different replicas even if steered to the same execution core. At the same time,
+different replicas even if steered to the same application core. At the same time,
 two RPC calls steered through the same RAFT core but dispatched to different
-execution cores can end up completing in different orders in different replicas.
+application cores can end up completing in different orders on different replicas.
 
-Our solution to this problem in Cyclone is one that applies to most durable
-data structures - not just the hash table we consider here. We observed that
-replicas of the durable data structure require \emph{per-key determinism}
-and not determinism across \emph{all} operations. We therefore steer
-operations to the same key to the same RAFT core and the same execution core, by
-hashing the key to determine the RAFT core and execution core. The actual
-configuration of the hash table therefore can be different across different
-replicas - for example due to the insertion of two keys mapping to the same hash
-bucket completing in different orders across different replicas. However, since the
-results of insert, delete and lookup operations for a key depend only on
-previous operations to the same key the replicas provide identical results for
-committed operations.
+Our solution to this problem in Cyclone is to observe that replicas of the hash
+table require \emph{per-key determinism} and not determinism across \emph{all}
+operations. We therefore steer operations to the same key to the same RAFT core
+and the same application core, by hashing the key to determine the RAFT core and
+execution core. The actual configuration of the hash table therefore can be
+different across different replicas - for example due to the insertion of two
+keys mapping to the same hash bucket completing in different orders across
+different replicas. However, since the results of insert, delete and lookup
+operations for a key depend only on previous operations to the same key the
+replicas provide identical results for committed operations.
 
@@
 distributed transactions - an atomic sequence of conditional updates spanning
-multiple machines. Distributed transactions have two distinct facets:
-atomicity and fault tolerance. Distributed transactions appear atomic with
-respect to other atomic transactions and simple updates or reads by virtue of
-standard concurrency control techniques. The most common technique is two phase
-locking: cquire locks, verify that previously read objects are at the desired
-values, update the locked objects and finally release the locks. A survey of
-various distributed systems and informal discussions with programmers led us to
-the conclusion that this aspect of distributed transactions was fairly easy to
-understand and implement. The hard part for most programmers is the second
-aspect: fault tolerance. A distributed transaction spans many machines and has a
-large footprint to clean up in the event of failure. Failure handling is made
-more complex when considering cascading failures - further failures while
-recovering from a failure. Cyclone was designed to allow primitives such as
-distributed transctions to be added without the programmer needing to worry
-about faults or write fault recovery code - as we demonstrate with the hash
-table.
+multiple machines. Distributed transactions have two distinct facets: atomicity
+and fault tolerance. Distributed transactions appear atomic with respect to
+other atomic transactions and simple updates or reads by virtue of standard
+concurrency control techniques. The most common technique is two phase locking:
+cquire locks, verify that previously read objects are at the desired values,
+update the locked objects and finally release the locks. We believe that NVM
+application programmers - particularly those having experience with concurrent
+code should not find it hard to deal with distributed concurrency. The hard part
+for most programmers is likely to be the second aspect: fault tolerance. A
+distributed transaction spans many machines and has a large footprint to clean
+up in the event of failure. Failure handling is made more complex when
+considering cascading failures - further failures while recovering from a
+failure. Cyclone was designed to allow primitives such as distributed
+transctions to be added without the programmer needing to worry about faults or
+write fault recovery code. As an exercise we show how to add support for
+distributed transactions to our hash table.
 
@@
 server specifying the transaction and the server executes it on the client's
-behalf. Figure [] shows pseudocode for the transaction. Execution of the
-transaction has four distinct phases: acquire locks, verify values, apply
-updates and finally release locks.
+behalf. Execution of the transaction has four distinct phases: acquire locks,
+verify values, apply updates and finally release locks.
 
@@
 not receive a response to the RPC call - they will instead receive an error that
-tells them that they are too far behind. To ensure that the transaction is still
-guaranteed to complete, we use a \emph{different} client identifier for each
-phase. Thus, until the transaction completes and a new one is started the
-response for any RPC call continues to be available. Figure [] details in
-pseudocode the execution of a transaction with these added details. Crucially
-there is no need for awareness of faults nor any need to write fault recovery
-code. Finally, we need to ensure that the client can reliably query the result
-of the transaction. If the leader of the quorum executing the transaction falls
-behind it cannot supply the result. We therefore immediately return and
-acknowledment to the client as soon as the RPC call for the transaction has been
-replicated. We store the result of the transaction in a special data structure
-updated through an additional RPC call made at the end of the transaction. The
-client queries this data structure to determine the fate of the transaction.
+tells them that they are too far behind.
+
+To ensure that the transaction is still guaranteed to complete, we use a
+\emph{different} client identifier for each phase. Thus, until the transaction
+completes and a new one is started the response for any RPC call continues to be
+available. Figure [] shows in pseudocode the server side details of a
+transaction. It is worth comparing the lack of any need for explicit fault
+recovery in Cyclone with fault recovery in systems systems that explicity
+require such cleanup [FARM, also refer to pages].  We believe that making
+availability easy for NVM applications is key to increasing the adoption rate of
+NVM by data center operators.
 
@@
 \begin{figure}
-  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
   ports = 4. Fix payload = 0. One line per number of replicas (1,2,3).
@@
 \begin{figure}
-  Source: results_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
   replicas = 3. Fix payload = 0. One line per number of ports.
@@
 \begin{figure}
-  Source: results_raw.csv. Fix replicas = 3. Set of 4 graphs. One graph for each
+  Source: results\_raw.csv. Fix replicas = 3. Set of 4 graphs. One graph for each
   port count (1,2,3,4). X-axis throughput (Gbps). Y-axis latency. One line per
@@
 \begin{figure}
-  Source: results_clwb.csv. Bar graph with 3 groups of bars - one group for
+  Source: results\_clwb.csv. Bar graph with 3 groups of bars - one group for
   payload = 0, one group for payload = 190, one for 512. Each group has 3 bars -
@@
 \begin{figure}
-  Source: results_quorum.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
+  Source: results\_quorum.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
   replicas = 3. Fix payload = 0. One line for each of ports = 1,2,3,4.
@@
 \begin{figure}
-  Source: results_cc.csv.Fix replicas = 3. X-axis is throughput
+  Source: results\_cc.csv.Fix replicas = 3. X-axis is throughput
   (Mops/sec). Y-axis is Latency. One line for each of ports = 1,2,3,4.
   \caption{Crash consistency overheads}
+  \label{fig:perf_cc_overheads}
+\end{figure}
+
+\begin{figure}
+  Source: results\_quorum\_cc.csv.X-axis throughput (Mops/sec). Y-axis latency.
+  Fix replicas = 3. Fix payload = 0. One line for each of ports = 1,2,3,4.
+  \caption{Crash consistency scaling}
   \label{fig:perf_cc}
@@
 \begin{figure}
-  Source: results_app.csv. X-axis is throughput
-  (Mops/sec). Y-axis is Latency. One line for each of read_fraction = 0, 50, 100
+  Source: results\_app.csv. X-axis is throughput
+  (Mops/sec). Y-axis is Latency. One line for each of read\_fraction = 0, 50, 100
   \caption{Hash Table Performance}
@@
 
+\section{Future Work}
+
 \end{document}

### 2016-09-18 paper/paper.tex
Updates.

@@
 we support applications built on top of Intel's Non Volatile Memory Library
-(NVML) [nvml.io]. We describe NVML's programming model in this section to provide
-adequate background for the rest of this paper. The primary feature of interest
-is the ability to write code that is crash consistent. In other words code that
-makes changes to the non-volatile heap is enclosed in a crash consistent
-transaction such that either all of the transaction executes or none of it
-executes in the face of power failures. This is critical to maintaining
+(NVML) [nvml.io]. We describe NVML's programming model in this section to
+provide adequate background for the rest of this paper. The primary feature of
+interest is the ability to write code that is crash consistent. In other words
+code that makes changes to the non-volatile heap is enclosed in a crash
+consistent transaction such that either all of the transaction executes or none
+of it executes in the face of power failures. This is critical to maintaining
 consistency as the data structure, being durable, must survive across power
@@
 failure. A second point worth noting is that NVM libraries such as NVML usually
-provide a way to wrap pointers to non-volatile memory in order that such memory
-may be mapped to different locations in virtual memory across restarts without
-impact to the accessing application.
+provide a way to wrap pointers to non-volatile memory (using {\tt TOID} macros)
+in order that such memory may be mapped to different locations in virtual memory
+across restarts without impact to the accessing application.
 
@@
 Cyclone provides strongly consistent replication - every machine in the quorum
-of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as a variable
-sized opaque blob of data, leaving the application free to choose how it wishes
-to marshall call related details and argument. All machines agree on the
-committed sequence of RPC calls in the log. We achieve this by running an
+of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as
+a variable sized opaque blob of data, leaving the application free to choose how
+it wishes to marshall call related details and argument. All machines agree on
+the committed sequence of RPC calls in the log. We achieve this by running an
 instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
@@
 
-We exploit the fact that DPDKs packet buffers allow prepending data without
-physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
-how the layout of a packet buffer changes at the leader during replication. The
-received packets contains only the RPC call from the client. We prepend a RAFT
-related header (containing the previous log index, previous log term, leader's
-term and leader's commit point) to packet buffer itself. To this RAFT header we
-then prepend an IPv4 header. We set IP addresses in the header to use
-5-tuple steering in order to place the packet in the right queue at the
-receiving NIC. This composite packet buffer is then sent to every replica. In
-order to properly route the packet through switches, we allocate a small buffer
-containing an ethernet header for each follower replica and \emph{chain} it in
-front of the prepared packet - ensuring we do not need to change the contents
-of the packet itself for every replica.
+We exploit the fact that DPDKs packet buffers allow prepending data to a packet
+without physically moving the data in the packet buffer. Fig
+[figures/packets.bmp] shows how the layout of a packet buffer changes at the
+leader during replication. The received packets contains only the RPC call from
+the client. We prepend a RAFT related header (containing the previous log index,
+previous log term, leader's term and leader's commit point) to packet buffer
+itself. To this RAFT header we then prepend an IPv4 header. We set IP addresses
+in the header to use 5-tuple steering in order to place the packet in the right
+queue at the receiving NIC. This composite packet buffer is then sent to every
+replica. In order to properly route the packet through switches, we allocate a
+small buffer containing an ethernet header for each follower replica and
+\emph{chain} it in front of the prepared packet - ensuring we do not need to
+change the contents of the packet itself for every replica.
 
@@
 In order to use the hash table with Cyclone, we simply need to to declare the
-RPC call handler as the point of entry to Cyclone at the server end - shown in
-Figure~\ref{fig:server_api}. The client end issues RPC calls via Cyclone's
-client side library as shown in Figure~\ref{fig:client_api}. Crucially neither
-side involves fault awareness or fault recovery code - a key goal for Cyclone.
+RPC call handler as the point of entry to Cyclone at the server end.  The client
+end issues RPC calls via Cyclone's client side library. Crucially neither side
+involves fault awareness or fault recovery code - a key goal for Cyclone.
 
@@
 of key-value pairs atomically in the face of failure.
-
-\begin{figure}
-{ \scriptsize
-\begin{verbatim}
-int exec_rpc_call(void * input, int input_size, void *output)
-{ 
-  ...
-}
-
-struct rpc_callbacks_t callbacks = {
-.exec_call = exec_rpc_call,
-...
-};
-
-main()
-{
-  ...
-  cyclone_start(callbacks, ...);
-  ...
-}
-\end{verbatim}
-}
-\caption{Server side API}
-\label{fig:server_api}
-\end{figure}
-
-\begin{figure}
-{ \scriptsize
-\begin{verbatim}
-int make_rpc_call(..., void *data, int size, int rpc_number);
-int get_last_rpc_number(...);
-main()
-{
-  ...
-  int rpc_number = get_last_rpc_number(...);
-  while(true) {
-    ...
-    make_rpc_call(..., data, size, rpc_number++);
-
-  }
-  ...
-}
-\end{verbatim}
-}
-\caption{Client side API}
-\label{fig:client_api}
-\end{figure}
 
@@
 other atomic transactions and simple updates or reads by virtue of standard
-concurrency control techniques. The most common technique is two phase locking:
-cquire locks, verify that previously read objects are at the desired values,
-update the locked objects and finally release the locks. We believe that NVM
-application programmers - particularly those having experience with concurrent
-code should not find it hard to deal with distributed concurrency. The hard part
-for most programmers is likely to be the second aspect: fault tolerance. A
-distributed transaction spans many machines and has a large footprint to clean
-up in the event of failure. Failure handling is made more complex when
-considering cascading failures - further failures while recovering from a
-failure. Cyclone was designed to allow primitives such as distributed
-transctions to be added without the programmer needing to worry about faults or
-write fault recovery code. As an exercise we show how to add support for
-distributed transactions to our hash table.
+concurrency control techniques. A common technique is strict two phase
+locking: acquire all locks, update the locked objects and finally release all
+the locks. We believe that NVM application programmers - particularly those
+having experience with concurrent code should not find it hard to deal with
+distributed concurrency. The hard part for most programmers is likely to be the
+second aspect: fault tolerance. A distributed transaction spans many machines
+and has a large footprint to clean up in the event of failure. Failure handling
+is made more complex when considering cascading failures - further failures
+while recovering from a failure. Cyclone was designed to allow primitives such
+as distributed transctions to be added without the programmer needing to worry
+about faults or write fault recovery code. As an exercise we show how to add
+support for distributed transactions to our hash table.
 
@@
 
-We guard against fault tolerance by executing the RPC call on a quorum of
+We guard against faults by executing the RPC call on a quorum of
 machines via Cyclone. Since execution of the RPC call leaves \emph{externally
@@
 completes and a new one is started the response for any RPC call continues to be
-available. Figure [] shows in pseudocode the server side details of a
-transaction. It is worth comparing the lack of any need for explicit fault
-recovery in Cyclone with fault recovery in systems systems that explicity
-require such cleanup [FARM, also refer to pages].  We believe that making
-availability easy for NVM applications is key to increasing the adoption rate of
-NVM by data center operators.
+available. Figure~\ref{fig:dist_tx} shows in pseudocode the server side details
+of a transaction. One initiated, the distributed transaction is guaranteed to
+complete given quorum availabilty. However, if the client requires intimation of
+completion, the final transaction state can be written to a distinguished
+partition as part of the transaction itself.
+
+We note that this design is not intended for performance, under the assumption
+that distributed transactions will be infrequent compared to single key
+operations.  On the other hand, it is worth comparing the lack of any need for
+explicit fault recovery in Cyclone with fault recovery in systems systems that
+explicity require such cleanup [FARM, also refer to pages].  We believe that
+such designs, which make availability easy for NVM applications are key to
+increasing the adoption rate of NVM by data center operators.
+
+\begin{figure}
+{ \scriptsize
+\begin{verbatim}
+wait until RPC call is replicated
+respond to client saying tx accepted for execution
+acquire:
+ set client identifier to 1
+ for each machine
+   Acquire write or read locks on the machine
+   If machine reponds RPC sequence too old
+     Transaction completed by someone else
+
+update:
+ set client identifier to 2
+ for each machine
+   Update object
+   If machine reponds RPC sequence too old
+     Transaction completed by someone else
+
+release:
+ set client identifier to 3
+ for each machine
+   Update object
+   If machine reponds RPC sequence too old
+     Transaction completed by someone else
+
+optional:
+ write transaction result to <quorum_tx_result>
+\end{verbatim}
+}
+\caption{Distributed Transactions}
+\label{fig:dist_tx}
+\end{figure}
 
@@
 
-\section{Related Work}
-
-\section{Future Work}
+\section{Conclusion}
 

### 2016-09-20 paper/paper.tex
add packet layout figure to paper.
Additions to paper.

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{usenix,epsfig,endnotes}
+\usepackage{usenix,epsfig,endnotes,graphicx}
 \begin{document}
@@
 
-\subsection{Zero Copy Replication}
+\subsection{Zero Copy Batching}
 Consensus protcols such as RAFT usually have a simple common case path. Once a
@@
 in the header to use 5-tuple steering in order to place the packet in the right
-queue at the receiving NIC. This composite packet buffer is then sent to every
-replica. In order to properly route the packet through switches, we allocate a
-small buffer containing an ethernet header for each follower replica and
-\emph{chain} it in front of the prepared packet - ensuring we do not need to
-change the contents of the packet itself for every replica.
+queue at the receiving NIC. We do not use different ip addresses for different
+replicas, rather the ip addresses are only set in conjunction with NIC
+programming to enable steering of the received packets to the correct NIC queue.
+This composite packet buffer is then sent \emph{unchanged} to every replica. In
+order to properly route the packet through switches, we allocate a small buffer
+containing an ethernet header for each follower replica and \emph{chain} it in
+front of the prepared packet - ensuring we do not need to change the contents of
+the packet itself for every replica.
 
-\subsection{Batching}
 Although the common case path in RAFT is straightforward, it still adds many
@@
 approximately 9KB.
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth,height=6cm]{figures/chain.pdf}
+\caption{Zero Copy Batching}
+\label{fig:zc_batch}
+\end{figure}
+
+This set of transformations to incoming request packets from clients at the
+leader replica is summarized in Figure~\ref{fig:zc_batch}.
 

### 2016-09-21 paper/paper.tex
Added pure replication results.

@@
 \section{Evaluation}
-\begin{figure}
-  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
-  ports = 4. Fix payload = 0. One line per number of replicas (1,2,3).
-  \caption{Replicaton Performance - Varying replica count}
-  \label{fig:perf_rep}
-\end{figure}
 
-\begin{figure}
-  Source: results\_raw.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
-  replicas = 3. Fix payload = 0. One line per number of ports.
-  \caption{Replicaton Performance - Varying ethernet ports}
-  \label{fig:perf_ports}
-\end{figure}
-
-\begin{figure}
-  Source: results\_raw.csv. Fix replicas = 3. Set of 4 graphs. One graph for each
-  port count (1,2,3,4). X-axis throughput (Gbps). Y-axis latency. One line per
-  payload size.
-  \caption{Replicaton Performance - Varying payload size}
-  \label{fig:perf_payload}
-\end{figure}
-
-\begin{figure}
-  Source: results\_clwb.csv. Bar graph with 3 groups of bars - one group for
-  payload = 0, one group for payload = 190, one for 512. Each group has 3 bars -
-  one for window =1, one for window = 10, one for window = 10000 (infty)
-  \caption{Replicaton Performance - CLFLUSH window}
-  \label{fig:perf_clflush}
-\end{figure}
-
-\begin{figure}
-  Source: results\_quorum.csv. X-axis throughput (Mops/sec). Y-axis latency. Fix
-  replicas = 3. Fix payload = 0. One line for each of ports = 1,2,3,4.
-  \caption{Replicaton Performance - quorum count}
-  \label{fig:perf_quorums}
-\end{figure}
-
-\begin{figure}
-  Source: results\_cc.csv.Fix replicas = 3. X-axis is throughput
-  (Mops/sec). Y-axis is Latency. One line for each of ports = 1,2,3,4.
-  \caption{Crash consistency overheads}
-  \label{fig:perf_cc_overheads}
-\end{figure}
-
-\begin{figure}
-  Source: results\_quorum\_cc.csv.X-axis throughput (Mops/sec). Y-axis latency.
-  Fix replicas = 3. Fix payload = 0. One line for each of ports = 1,2,3,4.
-  \caption{Crash consistency scaling}
-  \label{fig:perf_cc}
-\end{figure}
-
-\begin{figure}
-  Source: results\_app.csv. X-axis is throughput
-  (Mops/sec). Y-axis is Latency. One line for each of read\_fraction = 0, 50, 100
-  \caption{Hash Table Performance}
-  \label{fig:perf_app}
-\end{figure}
+\begin{figure*}
+\begin{tabular}{cc}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_0.pdf}
+\end{minipage}&
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_0.pdf}
+\end{minipage}\\
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_512.pdf}
+\end{minipage}&
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_512.pdf}
+\end{minipage}
+\end{tabular}
+\caption{Pure Replication}
+\label{fig:pure_rep}
+\end{figure*}
 

### 2016-09-22 paper/paper.tex
Added cc results.
Added app results.
Edits.

@@
 \documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{usenix,epsfig,endnotes,graphicx}
+\usepackage{usenix,epsfig,endnotes,graphicx,url}
 \begin{document}
@@
 % Comment it out when you first submit the paper for review.
-\thispagestyle{empty}
+%\thispagestyle{empty}
 
 \subsection*{Abstract}
-Memory technology in the datacenter is due to undergo a paradigm shift with the
-increased use of directly addressable non-volatile memory both in the form of
-battery backed non-volatile DIMMs as well as newer memory types such as
-3D XPoint. In anticipation, libraries that allow programmers to build durable
-data structures on non-volatile heaps have become available. A missing piece
-however is a component to make such durable data structures available using
-replication across a commodity network such as ethernet. Existing solutions that
-offer good performance are biased towards pure transactions across RDMA with a
-dependence on external services and application specific code for fault recovery.
-
-Cyclone is replication middleware that aims to add fault tolerance as
-transparently as possible. A design for Cyclone is that NVM application
-developers should not need to add fault recovery code. A second design goal for
-Cyclone is to strive for good performance on commodity ethernet - with
-innovations in the software stack rather than a requirement for RDMA capable
-networking hardware.
-
-Cyclone achieves these goals by providing a simple API for client server
-applications that allows server state to be replicated by executing the same
-sequence of RPC calls across all replicas. It provides at most once semantics
-with automatic failover to hide fault recovery from programmers. Finally it uses
-DPDK - a high performance, low latency network stack coupled with batching,
-zero-copy and the ability to scale replication performance across
-multiple instances of a consensus protocol to replicate millions of small
-payload RPC messages a second on commodity ethernet.
+The introduction of non-volatile directly attached memory in datacenter - in
+the form of NVDIMMs and new memory technologies allows efficient data structures
+to be built on non-volatile heaps that provide good performance coupled with
+durability. However high availability remains a problem for these primarily
+single machine solutions. While transactional interfaces built over RDMA
+capable networking hardware offer a viable alternative, they are often
+unavailable in datacenters. Cyclone solves this problem by providing transparent
+replication middleware that to be used with Intel's Non Volatile Memory Library 
+(NVML) to construct durable data-structures that map keys to values.
+Cyclone is designed around an optimized userspace networking stack to reduce
+latency and can scale its consensus protocol across cores and network ports to
+provide good throughput under load. We show that Cyclone can replicate as many
+as 8 million updates a second using just 4 commodity 10 GigE ports on a machine.
 
 \section{Introduction}
-TBD
+Memory technology in datacenters is due to undergo a paradigm shift with the
+increased use of directly addressable non-volatile memory both in the form of
+battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types
+such as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
+programmers to build durable data structures on non-volatile heaps have become
+available~\cite{mnemosyne, nvheaps, cdds}  - an example of a such a
+library that we have used in this paper is Intel's non volatile memory library
+(NVML)~\cite{nvml}. A missing piece however is a component to make such durable 
+data structures highly available using replication across a commodity network
+such as ethernet. Existing solutions that offer good performance are biased
+towards pure transactions across RDMA with a dependence on external services
+(such as zookeeper~\cite{zookeeper}) and application specific code for
+fault recovery~\cite{farm, htm}.
+This is a problem for programmers using NVML to construct durable and available 
+data structures where one cannot apriori guarantee the availability of
+RDMA (for e.g., when running on externally hosted environments or public
+clouds).
+
+Cyclone is a prototype replication middleware that aims to demonstrate that one
+can add fault tolerance transparently to a specific class of client server
+applications written on top of libraries such as NVML. We specifically target
+key-value maps in this paper and in that context make two key
+contributions. First, we propose a simple API that enables server state to be
+replicated without the programmer needing to worry about fault recovery. This
+allows one to start by building a concurrent durable data structure using
+libraries such as NVML and add availability at a later stage using Cyclone. The
+second contribution is to show that one can achieve reasonably good replication 
+performance using commodity ethernet via innovations in the networking stack and
+by scaling replication with the application. This result should be a useful
+design point for developers looking to deploy applications on non-volatile heaps
+in commodity ethernet-based environments such as public clouds.
+
 \section{Single Node Programming Model}
-Cyclone starts with the assumption that the programmer has already built a
-concurrent durable application on a non-volatile heap. A number of libraries and
-filesystems make this possible today [DAX, NVML, others ?]. As a starting point,
-we support applications built on top of Intel's Non Volatile Memory Library
-(NVML) [nvml.io]. We describe NVML's programming model in this section to
-provide adequate background for the rest of this paper. The primary feature of
-interest is the ability to write code that is crash consistent. In other words
-code that makes changes to the non-volatile heap is enclosed in a crash
+Cyclone assumes a concurrent durable application on non-volatile heap as the
+starting point. A number of libraries and filesystems make this possible
+today~\cite{nvml, dax, pmfs, mnemosyne, nvheaps, cdds}. Cyclone currently
+supports applications built on top of Intel's Non Volatile Memory Library
+(NVML~\cite{nvml}). We describe NVML's programming model in this section to
+provide adequate background for the rest of this paper. NVML requires 
+that code that makes changes to non-volatile heap is enclosed in a crash
 consistent transaction such that either all of the transaction executes or none
-of it executes in the face of power failures. This is critical to maintaining
-consistency as the data structure, being durable, must survive across power
-failures. To illustrate this idea, Figure~\ref{fig:example} shows how one might
-build a persistent linked list using NVML - the key idea is to enclose the
-entire process of updating the linked list within a crash consistent
-transaction. This ensure that the linked list is not corrupted due to a power
-failure. A second point worth noting is that NVM libraries such as NVML usually
-provide a way to wrap pointers to non-volatile memory (using {\tt TOID} macros)
-in order that such memory may be mapped to different locations in virtual memory
-across restarts without impact to the accessing application.
+of it executes in the event of power failures.
+%This is critical to maintaining consistency as the data structure, being
+%durable, must survive across power failures.
+Figure~\ref{fig:example} illustrates the use of NVML to build a persistent
+linked list. Updates to the linked list are enclosed in a NVML transaction to
+avoid corruption due to power failure. It is also worth noting that NVM
+libraries such as NVML usually allow remapping non-volatile heaps to a different
+virtual memory address range across restarts (e.g., using {\tt TOID} macros to
+implement fat pointers in NVML).
 
@@
 so desired for performance. More sophisticated fine-grained or optimistic
-concurrency control mechanisms for NVM are also possible [mnemosyne].
+concurrency control mechanisms for NVM are also
+possible~\cite{mnemosyne, cdds, nvheaps}.
 
@@
 Cyclone is designed to add fault tolerance to NVML client server applications
-via replication - a quorum of machines maintains \emph{equivalent} state. Server
-state is queried and manipulated via RPC calls from clients. Cyclone replicates
-the RPC call itself across replicas rather than replicating every access made
-during execution of the RPC call.
+via replication -- a quorum of machines maintains \emph{equivalent} state.
+Server state is queried and manipulated via RPC calls from clients.
+Cyclone replicates the RPC call itself across replicas rather than replicating
+every access (to the server state) made during execution of the RPC call.
 
@@
 a variable sized opaque blob of data, leaving the application free to choose how
-it wishes to marshall call related details and argument. All machines agree on
+it wishes to marshall call related details and arguments. All machines agree on
 the committed sequence of RPC calls in the log. We achieve this by running an
-instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
-different machines in synchronization. Specially flagged `read-only` RPC calls
-are executed on the leader replica and are not replicated.
+instance of the RAFT~\cite{raft} consensus protocol to keep the log of RPC calls
+on different machines in synchronization. Specially flagged `read-only` RPC
+calls are executed only on the leader replica and are not replicated.
 
@@
 The standard replicated state machine approach requires that commands (entries
-in the log) be committed before they can be applied to the state machine i.e. in
-our case - the RPC call be executed. In contrast, Cyclone decouples execution of
-the RPC call on any replica from its replication.
-
-However, the fact that RPC calls that modify state must themselves be executed
+in the log) be committed before they can be applied to the state machine --
+i.e. in our case before the RPC call can be executed. In contrast, Cyclone
+decouples execution of an RPC call on any replica from its replication.
+The fact that RPC calls that modify state must themselves be executed
 in a crash consistent transaction allows us to overlap execution with
-replication. Normally, this would be a problem as replication might failt for a
-number of reasons but primarily due to a change of leader and subsequent of
+replication. Normally, this would be a problem as replication might fail for a
+number of reasons but primarily due to change of leadership and subsequent
 rollback of previously appended (but not committed) log entries. To solve this
-problem, Cyclone delays the commit of the failure atomic transaction after
-execution till replication is complete. In the event that replication fails, the
-transaction is aborted and any changes to the non volatile heap are rolled back.
-
-A pseudocode description of how the runtime handles \emph{decoupled replication}
-is shown in Figure~\ref{fig:async_rep}.
+problem, Cyclone delays the actual commit of the failure atomic transaction
+until after the replication is complete. In the event that replication fails,
+the transaction is aborted and any changes to the non volatile heap are rolled
+back. Figure~\ref{fig:async_rep} shows a pseudocode description of how the
+runtime handles \emph{decoupled replication}.
 
@@
 
-Asynchronous replication provides good performance and we believe it would be the
-common case for RPC calls replicated via Cyclone. However the program might on
-occasion require successful replication to be a precondition to exection. A case
-where we find this necessary is output commit - where the RPC call has external
-side effects, such as communicating with other machines in the cluster and
-therefore crash consistency is insufficient to roll back the execution of the
-RPC call in the event that replication fails. We therefore provide users the
+Asynchronous replication provides good performance and we believe it would be
+the common case for RPC calls replicated via Cyclone. However the program might 
+on occasion require successful replication to be a precondition to exection.
+A case where we find this necessary is
+\textit{output commit}~\cite{output_commit} - where the RPC call has
+external side effects (e.g., communication with other machines in the cluster)
+and therefore crash consistency is insufficient to roll back the execution of
+the RPC call in the event that replication fails. We therefore provide users the
 ability to declare at the client side that an RPC call should execute
-synchronously at the server side. This results in Cyclone ensuring that the RPC
+synchronously at the server side. Cyclone would then ensure that the RPC
 call is successfully replicated before it attempts execution on any of the
@@
 \subsection{Network Stack}
-Cyclone is designed around DPDK [dpdk] a high performance network stack. DPDK
-provides convenient access to packets arriving at an ethernet NIC by moving them
-into packet buffers with the contained cache lines also directly placed into the
-last level cache of the CPU during DMA [ddio].
+Cyclone is designed around DPDK~\cite{dpdk} -- a high performance network stack
+optimized for Intel processors. DPDK provides convenient access to packets
+arriving at an ethernet NIC by moving them into packet buffers and directly
+placing those buffers into the last level cache of the CPU during DMA~\cite{ddio}.
 
 Cyclone separates the available CPU cores on a machine, in a configurable way,
-into cores dedicated to work related to the consensus protocol and cores
+into cores dedicated to work related to the (RAFT) consensus protocol and cores
 decidated to running application code. Each RAFT core runs an \emph{independent}
@@
 RPC call is sent from the application core via its dedicated output queue. This
-division of NIC queues eliminates any need for synchronization between RAFT
+division of NIC queues eliminates the need for synchronization between RAFT
 cores and application cores when accessing the NIC.
 
-In the event that the machine has multiple NICs Cyclone binds each RAFT or
-application core to a specific NIC and picks the necessary queues from that
-NIC. Spreading the network demands from the cores across the NICs in this manner
-allows Cyclone to scale with multiple NICs when available.
+In the event that the machine has multiple NICs, Cyclone spreads the network
+demands and scales by binding each RAFT or application core to a specific NIC
+and picking the necessary queues from that NIC.
 
@@
 
-In order to avoid copying the incoming RPC call contents to a persistent memory
-range, Cyclone uses persistent memory on the system itself to allocate space for
-packet buffers. However although a packet is accessible at the core the most up
-to date data for the memory region occupied by the packet is in the CPU cache
-and not necessarily in memory. To ensure that packet data has truly been made
-persistent, Cyclone issues a series of cacheline flushes. This ensures that the
-packet data is persistent. Cyclone maintains its log as a circular sequence of
+In order to avoid an additional copy of the incoming RPC call contents from DRAM
+to persistent memory, Cyclone allocates space for packet buffers on non-volatile
+heap. However, since packet data is cached, Cyclone ensures durability
+(or persistence) of this data by explicitly flushing the dirty
+cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
 pointers to packet buffers and therefore the append operation is completed by
@@
 
-We exploit the fact that DPDKs packet buffers allow prepending data to a packet
-without physically moving the data in the packet buffer. Fig
-[figures/packets.bmp] shows how the layout of a packet buffer changes at the
-leader during replication. The received packets contains only the RPC call from
-the client. We prepend a RAFT related header (containing the previous log index,
-previous log term, leader's term and leader's commit point) to packet buffer
-itself. To this RAFT header we then prepend an IPv4 header. We set IP addresses
-in the header to use 5-tuple steering in order to place the packet in the right
-queue at the receiving NIC. We do not use different ip addresses for different
-replicas, rather the ip addresses are only set in conjunction with NIC
-programming to enable steering of the received packets to the correct NIC queue.
-This composite packet buffer is then sent \emph{unchanged} to every replica. In
-order to properly route the packet through switches, we allocate a small buffer
-containing an ethernet header for each follower replica and \emph{chain} it in
-front of the prepared packet - ensuring we do not need to change the contents of
-the packet itself for every replica.
+We exploit the fact that DPDK allows prepending data to a packet
+without physically moving the data in a packet buffer. The received packet
+contains only the RPC call from the client. We prepend a RAFT related header
+(containing the previous log index, previous log term, leader's term and
+leader's commit point~\cite{raft}) to packet buffer itself. To this RAFT header 
+we then prepend an IPv4 header. We assign pre-determined values to the
+IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
+the placement of the incoming packet in the right queue.
+This composite packet buffer is then sent \emph{unchanged} to every replica.
+To properly route the packet through switches, we allocate a small buffer
+containing an ethernet header for each follower replica and \emph{chain} the
+buffer in front of the prepared packet - thereby ensuring that we do not need
+to change the contents of the packet itself for every replica.
 
-Although the common case path in RAFT is straightforward, it still adds many
-hundreds of cycles for each log entry to be processed. This adds up to a few
-microseconds, a significant amount in the face of sub 10 us network latency
-between nodes with DPDK. We amortize this overhead when operating under load by
+Although the common case path in RAFT is straightforward, it still requires
+hundreds of cycles (or a few microseconds) to process each log entry, which is
+a significant amount in the face of sub 10 us network latency between nodes with
+DPDK. We amortize this overhead when operating under load by
 batching multiple RPC calls together and treating the batch as a single RAFT log
-entry.  DPDK by itself can efficiently batch incoming packets under load -
-allowing the user to easily receive a burst (currently limited to 32) of packets
-rather than a single one for a polling call. DPDK further allows chaining packet
+entry.  DPDK by itself can efficiently batch incoming packets under load,
+therefore allowing the user to easily receive a burst of packets
+(currently limited to 32) rather than a single packet for a polling call.
+DPDK further allows chaining packet
 buffers together and treating the chain as a single packet. This is convenient
 for our zero copy approach and we chain packet mbufs together to form a single
-packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
-single packet when being replicated, the length of a chain is limited by the
-size of the coalesced packet, that must fit into an ethernet jumbo frame of
-approximately 9KB.
+packet. Since the entire chain is sent as a single packet when being replicated,
+the length of a chain is limited by the size of the coalesced packet, that must
+fit into an ethernet jumbo frame of approximately 9KB.
 
@@
 distributed transactions - an atomic sequence of conditional updates spanning
-multiple machines. Distributed transactions have two distinct facets: atomicity
-and fault tolerance. Distributed transactions appear atomic with respect to
-other atomic transactions and simple updates or reads by virtue of standard
-concurrency control techniques. A common technique is strict two phase
+multiple groups of replicas. Distributed transactions have two distinct facets:
+atomicity and fault tolerance. Distributed transactions appear atomic with
+respect to other atomic transactions and simple updates or reads by virtue of
+standard concurrency control techniques. A common technique is strict two phase
 locking: acquire all locks, update the locked objects and finally release all
@@
 
-We guard against faults by executing the RPC call on a quorum of
-machines via Cyclone. Since execution of the RPC call leaves \emph{externally
-  visible side effects} by virtue of the RPC calls it makes we require output
-commit [] - the property that the execution is guaranteed to complete once it
-starts. This is simply achieved by a flag in the RPC call that turns off
-decoupled execution (Section~\ref{sec:decouple}). The flag causes execution to
-start only after the RPC call has been successfully replicated.
+We guard against faults by executing the RPC call on a quorum of machines via
+Cyclone. Since execution of the RPC call leaves \emph{externally visible side
+  effects} by virtue of the RPC calls it makes we require output
+commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
+complete once it starts. This is simply achieved by a flag in the RPC call that
+turns off decoupled execution (Section~\ref{sec:decouple}). The flag causes
+execution to start only after the RPC call has been successfully replicated.
 
@@
 explicit fault recovery in Cyclone with fault recovery in systems systems that
-explicity require such cleanup [FARM, also refer to pages].  We believe that
+explicity require such cleanup (~\cite{farm}, pp 59--63).  We believe that
 such designs, which make availability easy for NVM applications are key to
@@
 \section{Evaluation}
+We evaluate Cyclone against its design goals by using it for up to 3-way
+replication in a cluster of twelve 48 core Xeon servers. Three servers run
+Cyclone's server side code, while the remaining machines run multithreaded
+clients to load the system for our studies. The server side machines are each
+equipped with four commodity 82599 10 Gbit ethernet controllers, while each
+client machine has one 82599 ethernet controller. All the machines are connected
+through the same commodity 10Gbit switch. We use 8 cores on the server machines
+for running 8 instance of RAFT, leaving 39 cores to run application code. The
+last remaining core is reserved for management code. We use NVML with the non
+volatile pool held in DRAM, which proxies for the performance we would expect
+from NVDIMMs. We evaluate Cyclone in three stages.
 
-\begin{figure*}
+First, we evaluate pure replication performance by replicating RPC calls. There
+is crash consistent transaction that remembers the RPC call sequence number -
+every RPC call is accepted and once replicated a response is sent to the
+client. Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one
+where the clients send an RPC call with no payload and one where they add a 512
+byte dummy payload. For the no payload case the bottleneck is the message
+handling rate. We note that Cyclone can handle small message replication rates
+as high as \emph{8 million messages a second}. In the case of RPC calls with 512
+byte payloads, the bottleneck is the available network bandwidth. We plot
+transmit throughput derived from the message rate as the rate at which data is
+sent to replicas from all the 8 quorum leaders and back to the clients.
+It is interesting to note that with 3 replicas Cyclone achieves \emph{more than
+  40 Gbps} of bandwidth. The reason for this is that the quorum leaders can end
+up residing on different machines. Although replication traffic is still limited
+by the receive bandwidth the responses to the clients, originating from
+different machines have access to independent 4x10GigE NICs, allowing total
+transmit throughput to exceed 40 Gbps. Although not discussed in this paper,
+Cyclone's control plane is designed to exploit this by trying to spread quorum
+leaders across machines.
+
+Also worth noting is that thanks to DPDK's low latency access to the NICs, three
+way replication costs as little as 25 us of latency in the unloaded case. This
+latency is measured at the client and therefore is four commodity 10 GigE
+network hops through the switch, with each hop taking under 7 us including
+network stack delays. These results should serve to illustrate that simple
+commodity ethernet can indeed provide reasonable replication performance with a
+carefully designed software stack that can exploit it.
+
+\begin{figure}
 \begin{tabular}{cc}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_0.pdf}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
 \end{minipage}&
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_0.pdf}
+\begin{minipage}{0.25\textwidth}
+  \hspace{-0.2in}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
 \end{minipage}\\
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_512.pdf}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
 \end{minipage}&
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_512.pdf}
+\begin{minipage}{0.25\textwidth}
+  \hspace{-0.2in}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
 \end{minipage}
@@
 \label{fig:pure_rep}
-\end{figure*}
+\end{figure}
+
+Next, we consider the effect of adding NVMLs crash consistent
+transactions. Figure~\ref{fig:cc_rep} shows the effect of adding a crash
+consistent transaction where the application still does nothing but Cyclone
+remembers the last sequence number and RPC call result (the application simply
+returns the received payload). Throughput is now bottlenecked by NVML allocator
+performance at just under a million messages a second. The final result is
+adding the hash table in the picture with the clients updating the values of
+keys already present in the hash table - a stable workload for easy
+comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
+under 1M calls a second mark for pure updates. For comparison we also show a
+pure read only case where the clients only query the value of a key - there is
+no replication \emph{nor} any crash consistent transaction applied. It is
+interesting to note that the pure read only case is bottlenecked at the same
+point as the pure replication case of 8 million operations a second.
+
+These results lead us to conclude that Cyclone's provides adequate performance
+headroom for replication to NVML applications. In fact our current focus in the
+light of these results is towards improving application performance on top of
+NVML. NVML is a general purpose library written for a variety of substrates
+including traditional SSD based storage and there is adequate room for
+improvement of performance for applications running on directly attached
+persistent memory with close to DRAM performance.
+
+\begin{figure}
+\begin{tabular}{cc}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
+  \caption{Crash Consistency + Replication}
+  \label{fig:cc_rep}
+\end{minipage} &
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
+  \caption{Crash Consistent Replicated Hash Table}
+  \label{fig:app_rep}
+\end{minipage}
+\end{tabular}
+\end{figure}
 
 \section{Conclusion}
+Cyclone allows developers to add availability to applications built on top of
+libraries such as NVML and achieve good performance on commodity networking
+hardware. A key question is whether the ideas in Cyclone extend to more than
+just key-value maps - perhaps to application such as filesystems ? We believe
+that the requirement for scaling across multiple replication quorums can be
+satisfied by applications where the scalable commutativity
+rule~\cite{scalable_commutativity} applies to common case primitives. Although
+our intitial release of Cyclone will be targeted towards key value style
+applications, our future research is targeted towards investigating this
+hypothesis.
+\newcommand\myurl[2]{\url{#1}}
+\bibliographystyle{acm}
+\bibliography{paper}
 

### 2016-09-22 paper/paper.tex
More stuff in the paper.

--- 2016-09-21 paper/paper.tex
@@
 % Comment it out when you first submit the paper for review.
-\thispagestyle{empty}
+%\thispagestyle{empty}
 
 \subsection*{Abstract}
+The introduction of non-volatile directly attached memory in the datacenter - in
+the form of NVDIMMs and new memory technologies allows efficient datastructures
+to be built on non-volatile heaps that provide good performance coupled with
+durability. However availability remains a problem for these single machine
+solutions with proposed solutions counting on transactional interfaces built
+over RDMA capable networking hardware that is often unavailable in datacenters.
+Cyclone is designed to help solve this problem. It works as a transparent layer
+with Intel's Non Volaile Memory Library (NVML) and allows the construction of
+durable data-structures targeted towards mapping keys to values. We show that
+Cyclone can replicate as many as 8 million updates a second using just 4
+commodity 10 GigE ports on a machine. Cyclone is designed around a userspace
+networking stack to reduce latency and can scale its consensus protocol across
+cores and network ports to provide good throughput under load.
+
+\section{Introduction}
 Memory technology in the datacenter is due to undergo a paradigm shift with the
 increased use of directly addressable non-volatile memory both in the form of
-battery backed non-volatile DIMMs as well as newer memory types such as
-3D XPoint. In anticipation, libraries that allow programmers to build durable
-data structures on non-volatile heaps have become available. A missing piece
-however is a component to make such durable data structures available using
-replication across a commodity network such as ethernet. Existing solutions that
-offer good performance are biased towards pure transactions across RDMA with a
-dependence on external services and application specific code for fault recovery.
+battery backed non-volatile DIMMs as well as newer memory types such as 3D
+XPoint. In anticipation, libraries that allow programmers to build durable data
+structures on non-volatile heaps have become available - an example of a such a
+library that we have used in this paper is Intel's non volatile memory library
+(NVML). A missing piece however is a component to make such durable data
+structures available using replication across a commodity network such as
+ethernet. Existing solutions that offer good performance are biased towards pure
+transactions across RDMA with a dependence on external services and application
+specific code for fault recovery [FARM, etc]. This is a problem for programmers
+using NVML to construct durable and available datastructures where one cannot
+apriori guarantee the availability of RDMA in the target environment especially
+when running on externally hosted environments such as the cloud.
 
-Cyclone is replication middleware that aims to add fault tolerance as
-transparently as possible. A design for Cyclone is that NVM application
-developers should not need to add fault recovery code. A second design goal for
-Cyclone is to strive for good performance on commodity ethernet - with
-innovations in the software stack rather than a requirement for RDMA capable
-networking hardware.
+Cyclone is a prototype replication middleware that aims to demonstrate that one
+can add fault tolerance transparently to a specific class of client server
+applications written on top of libraries such as NVML. We specifically target
+key-value maps in this paper and in that context make two key
+contributions. First, we show that a simple API allows server state to be
+replicated without the programmer needing to worry about fault recovery. This
+allows one to start by building a concurrent durable data structure using
+libraries such as NVML and add availability as a later step using Cyclone. The
+second contribution we make is to show that one can achieve reasonably good
+replication performance using commodity ethernet via innovations in the
+networking stack and by scaling replication with the application. This result
+should be a useful design point for developers looking to deploy applications on
+non-volatile heaps in commodity environments such as public clouds.
 
-Cyclone achieves these goals by providing a simple API for client server
-applications that allows server state to be replicated by executing the same
-sequence of RPC calls across all replicas. It provides at most once semantics
-with automatic failover to hide fault recovery from programmers. Finally it uses
-DPDK - a high performance, low latency network stack coupled with batching,
-zero-copy and the ability to scale replication performance across
-multiple instances of a consensus protocol to replicate millions of small
-payload RPC messages a second on commodity ethernet.
-
-\section{Introduction}
-TBD
 \section{Single Node Programming Model}
@@
 \section{Evaluation}
+We evaluate Cyclone against its design goals by using it for up to 3-way
+replication in a cluster of twelve 48 core Xeon servers. Three servers run
+Cyclone's server side code, while the remaining machines run multithreaded
+clients to load the system for our studies. The server side machines are each
+equipped with four commodity 82599 10 Gbit ethernet controllers, while each
+client machine has one 82599 ethernet controller. All the machines are connected
+through the same commodity 10Gbit switch. We use 8 cores on the server machines
+for running 8 instance of RAFT, leaving 39 cores to run application code. The
+last remaining core is reserved for management code. We use NVML with the non
+volatile pool held in DRAM, which proxies for the performance we would expect
+from NVDIMMs. We evaluate Cyclone in three stages.
+
+First, we evaluate pure replication performance by replicating RPC calls. There
+is crash consistent transaction that remembers the RPC call sequence number -
+every RPC call is accepted and once replicated a response is sent to the
+client. Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one
+where the clients send an RPC call with no payload and one where they add a 512
+byte dummy payload. For the no payload case the bottleneck is the message
+handling rate. We note that Cyclone can handle small message replication rates
+as high as \emph{8 million messages a second}. In the case of RPC calls with 512
+byte payloads, the bottleneck is the available network bandwidth. We plot
+transmit throughput derived from the message rate as the rate at which data is
+sent to replicas from all the 8 quorum leaders and back to the clients.
+It is interesting to note that with 3 replicas Cyclone achieves \emph{more than
+  40 Gbps} of bandwidth. The reason for this is that the quorum leaders can end
+up residing on different machines. Although replication traffic is still limited
+by the receive bandwidth the responses to the clients, originating from
+different machines have access to independent 4x10GigE NICs, allowing total
+transmit throughput to exceed 40 Gbps. Although not discussed in this paper,
+Cyclone's control plane is designed to exploit this by trying to spread quorum
+leaders across machines.
+
+Also worth noting is that thanks to DPDK's low latency access to the NICs, three
+way replication costs as little as 25 us of latency in the unloaded case. This
+latency is measured at the client and therefore is four commodity 10 GigE
+network hops through the switch, with each hop taking under 7 us including
+network stack delays. These results should serve to illustrate that simple
+commodity ethernet can indeed provide reasonable replication performance with a
+carefully designed software stack that can exploit it.
 
@@
 
+Next, we consider the effect of adding NVMLs crash consistent
+transactions. Figure~\ref{fig:cc_rep} shows the effect of adding a crash
+consistent transaction where the application still does nothing but Cyclone
+remembers the last sequence number and RPC call result (the application simply
+returns the received payload). Throughput is now bottlenecked by NVML allocator
+performance at just under a million messages a second. The final result is
+adding the hash table in the picture with the clients updating the values of
+keys already present in the hash table - a stable workload for easy
+comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
+under 1M calls a second mark for pure updates. For comparison we also show a
+pure read only case where the clients only query the value of a key - there is
+no replication \emph{nor} any crash consistent transaction applied. It is
+interesting to note that the pure read only case is bottlenecked at the same
+point as the pure replication case of 8 million operations a second.
+
+These results lead us to conclude that Cyclone's provides adequate performance
+headroom for replication to NVML applications. In fact our current focus in the
+light of these results is towards improving application performance on top of
+NVML. NVML is a general purpose library written for a variety of substrates
+including traditional SSD based storage and there is adequate room for
+improvement of performance for applications running on directly attached
+persistent memory with close to DRAM performance.
+
+\begin{figure}
+  \includegraphics[width=0.5\textwidth,height=6cm]{results/cc_mops.pdf}
+  \caption{Crash Consistency + Replication}
+  \label{fig:cc_rep}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=0.5\textwidth,height=6cm]{results/app_mops.pdf}
+  \caption{Crash Consistent Replicated Hash Table}
+  \label{fig:app_rep}
+\end{figure}
+
 \section{Conclusion}
-
+Cyclone allows developers to add availability to applications built on top of
+libraries such as NVML and achieve good performance on commodity networking
+hardware. A key question is whether the ideas in Cyclone extend to more than
+just key-value maps - perhaps to application such as filesystems. We believe
+that the requirement for scaling across multiple replication quorums is
+can be satisfied by applications where the scalable commutativity rule [scalable
+commutativity] applies to common case primitives. Although our intitial release
+of Cyclone will be targeted towards key value style applications, our future
+research is targeted towards investigating this hypothesis.
 \end{document}

### 2016-09-22 paper/paper.tex
Updates to paper.

@@
 ethernet. Existing solutions that offer good performance are biased towards pure
-transactions across RDMA with a dependence on external services and application
-specific code for fault recovery [FARM, etc]. This is a problem for programmers
-using NVML to construct durable and available datastructures where one cannot
-apriori guarantee the availability of RDMA in the target environment especially
-when running on externally hosted environments such as the cloud.
+transactions across RDMA with a dependence on external services (such as
+zookeeper~\cite{zookeeper}) and application specific code for fault recovery~\cite{farm,
+  htm}. This is a problem for programmers using NVML to construct durable and
+available datastructures where one cannot apriori guarantee the availability of
+RDMA in the target environment especially when running on externally hosted
+environments such as the cloud.
 
@@
 should be a useful design point for developers looking to deploy applications on
-non-volatile heaps in commodity environments such as public clouds.
+non-volatile heaps in commodity environments such as public clouds as well as
+library builders that aim to support them.
 
@@
 concurrent durable application on a non-volatile heap. A number of libraries and
-filesystems make this possible today [DAX, NVML, others ?]. As a starting point,
-we support applications built on top of Intel's Non Volatile Memory Library
-(NVML) [nvml.io]. We describe NVML's programming model in this section to
+filesystems make this possible today~\cite{nvml, dax}. As a starting point, we
+support applications built on top of Intel's Non Volatile Memory Library
+(NVML~\cite{nvml}). We describe NVML's programming model in this section to
 provide adequate background for the rest of this paper. The primary feature of
@@
 so desired for performance. More sophisticated fine-grained or optimistic
-concurrency control mechanisms for NVM are also possible [mnemosyne].
+concurrency control mechanisms for NVM are also possible~\cite{mnemosyne}.
 
@@
 the committed sequence of RPC calls in the log. We achieve this by running an
-instance of the RAFT [raft] consensus protocol to keep the log of RPC calls on
+instance of the RAFT~\cite{raft} consensus protocol to keep the log of RPC calls on
 different machines in synchronization. Specially flagged `read-only` RPC calls
@@
 \subsection{Network Stack}
-Cyclone is designed around DPDK [dpdk] a high performance network stack. DPDK
+Cyclone is designed around DPDK~\cite{dpdk} a high performance network stack. DPDK
 provides convenient access to packets arriving at an ethernet NIC by moving them
 into packet buffers with the contained cache lines also directly placed into the
-last level cache of the CPU during DMA [ddio].
+last level cache of the CPU during DMA~\cite{ddio}.
 
@@
 We exploit the fact that DPDKs packet buffers allow prepending data to a packet
-without physically moving the data in the packet buffer. Fig
-[figures/packets.bmp] shows how the layout of a packet buffer changes at the
-leader during replication. The received packets contains only the RPC call from
-the client. We prepend a RAFT related header (containing the previous log index,
-previous log term, leader's term and leader's commit point) to packet buffer
-itself. To this RAFT header we then prepend an IPv4 header. We set IP addresses
-in the header to use 5-tuple steering in order to place the packet in the right
-queue at the receiving NIC. We do not use different ip addresses for different
-replicas, rather the ip addresses are only set in conjunction with NIC
-programming to enable steering of the received packets to the correct NIC queue.
-This composite packet buffer is then sent \emph{unchanged} to every replica. In
-order to properly route the packet through switches, we allocate a small buffer
-containing an ethernet header for each follower replica and \emph{chain} it in
-front of the prepared packet - ensuring we do not need to change the contents of
-the packet itself for every replica.
+without physically moving the data in the packet buffer. The received packets
+contains only the RPC call from the client. We prepend a RAFT related header
+(containing the previous log index, previous log term, leader's term and
+leader's commit point) to packet buffer itself. To this RAFT header we then
+prepend an IPv4 header. We set IP addresses in the header to use 5-tuple
+steering in order to place the packet in the right queue at the receiving
+NIC. We do not use different ip addresses for different replicas, rather the ip
+addresses are only set in conjunction with NIC programming to enable steering of
+the received packets to the correct NIC queue.  This composite packet buffer is
+then sent \emph{unchanged} to every replica. In order to properly route the
+packet through switches, we allocate a small buffer containing an ethernet
+header for each follower replica and \emph{chain} it in front of the prepared
+packet - ensuring we do not need to change the contents of the packet itself for
+every replica.
 
@@
 for our zero copy approach and we chain packet mbufs together to form a single
-packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
-single packet when being replicated, the length of a chain is limited by the
-size of the coalesced packet, that must fit into an ethernet jumbo frame of
-approximately 9KB.
+packet. Since the entire chain is sent as a single packet when being replicated,
+the length of a chain is limited by the size of the coalesced packet, that must
+fit into an ethernet jumbo frame of approximately 9KB.
 
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{tabular}{cc}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_0.pdf}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
 \end{minipage}&
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_0.pdf}
+\begin{minipage}{0.25\textwidth}
+  \hspace{-0.2in}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
 \end{minipage}\\
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_mops_512.pdf}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
 \end{minipage}&
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\textwidth,height=6cm]{results/replication_gbps_512.pdf}
+\begin{minipage}{0.25\textwidth}
+  \hspace{-0.2in}
+  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
 \end{minipage}
@@
 \label{fig:pure_rep}
-\end{figure*}
+\end{figure}
 
@@
 \begin{figure}
-  \includegraphics[width=0.5\textwidth,height=6cm]{results/cc_mops.pdf}
+\begin{tabular}{cc}
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
   \caption{Crash Consistency + Replication}
   \label{fig:cc_rep}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=0.5\textwidth,height=6cm]{results/app_mops.pdf}
+\end{minipage} &
+\begin{minipage}{0.25\textwidth}
+  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
   \caption{Crash Consistent Replicated Hash Table}
   \label{fig:app_rep}
+\end{minipage}
+\end{tabular}
 \end{figure}
@@
 hardware. A key question is whether the ideas in Cyclone extend to more than
-just key-value maps - perhaps to application such as filesystems. We believe
-that the requirement for scaling across multiple replication quorums is
-can be satisfied by applications where the scalable commutativity rule [scalable
-commutativity] applies to common case primitives. Although our intitial release
-of Cyclone will be targeted towards key value style applications, our future
-research is targeted towards investigating this hypothesis.
+just key-value maps - perhaps to application such as filesystems ? We believe
+that the requirement for scaling across multiple replication quorums can be
+satisfied by applications where the scalable commutativity
+rule~\cite{scalable_commutativity} applies to common case primitives. Although
+our intitial release of Cyclone will be targeted towards key value style
+applications, our future research is targeted towards investigating this
+hypothesis.
+\newcommand\myurl[2]{\url{#1}}
+\bibliographystyle{acm}
+\bibliography{paper}
+
 \end{document}

### 2016-09-23 paper/paper.tex
Tweak graph placement.
Submission related updates.

--- 2016-09-22 paper/paper.tex
@@
 \subsection*{Abstract}
-The introduction of non-volatile directly attached memory in datacenter - in
+The introduction of non-volatile directly attached memory in datacenters - in
 the form of NVDIMMs and new memory technologies allows efficient data structures
@@
 durability. However high availability remains a problem for these primarily
-single machine solutions. While transactional interfaces built over RDMA
-capable networking hardware offer a viable alternative, they are often
-unavailable in datacenters. Cyclone solves this problem by providing transparent
-replication middleware that to be used with Intel's Non Volatile Memory Library 
-(NVML) to construct durable data-structures that map keys to values.
-Cyclone is designed around an optimized userspace networking stack to reduce
-latency and can scale its consensus protocol across cores and network ports to
-provide good throughput under load. We show that Cyclone can replicate as many
-as 8 million updates a second using just 4 commodity 10 GigE ports on a machine.
+single machine solutions. While transactional interfaces built over RDMA capable
+networking hardware offer a viable alternative, such hardware is often
+unavailable in datacenters. Cyclone offers a solution to this problem by
+providing transparent replication middleware that can be used with Intel's Non
+Volatile Memory Library (NVML) to construct durable data-structures that map
+keys to values.  Cyclone is designed around an optimized userspace networking
+stack to reduce latency and can scale its consensus protocol across cores and
+network ports to provide good throughput under load. We show that Cyclone can
+replicate as many as 8 million updates a second using just 4 commodity 10 GigE
+ports on a machine.
 
@@
 programmers to build durable data structures on non-volatile heaps have become
-available~\cite{mnemosyne, nvheaps, cdds}  - an example of a such a
+available~\cite{mnemosyne, nvheaps, cdds}  - an example of such a
 library that we have used in this paper is Intel's non volatile memory library
@@
 
-Cyclone is a prototype replication middleware that aims to demonstrate that one
+Cyclone is prototype replication middleware that aims to demonstrate that one
 can add fault tolerance transparently to a specific class of client server
@@
 \section{Single Node Programming Model}
-Cyclone assumes a concurrent durable application on non-volatile heap as the
+Cyclone assumes a concurrent durable application on a non-volatile heap as the
 starting point. A number of libraries and filesystems make this possible
@@
 In order to avoid an additional copy of the incoming RPC call contents from DRAM
-to persistent memory, Cyclone allocates space for packet buffers on non-volatile
+to persistent memory, Cyclone allocates space for packet buffers on the non-volatile
 heap. However, since packet data is cached, Cyclone ensures durability
@@
 (containing the previous log index, previous log term, leader's term and
-leader's commit point~\cite{raft}) to packet buffer itself. To this RAFT header 
+leader's commit point~\cite{raft}) to the packet buffer itself. To this RAFT header 
 we then prepend an IPv4 header. We assign pre-determined values to the
 IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
-the placement of the incoming packet in the right queue.
+the placement of the incoming packet in the correct queue.
 This composite packet buffer is then sent \emph{unchanged} to every replica.
@@
 example of a linearizable hash table - a useful and common data structure.
-[memcached]. The baseline is a durable hash table written in NVML that runs on a
+The baseline is a durable hash table written in NVML that runs on a
 single machine in client-server mode. The server side maintains the durable
 open-addressed hash table and services client RPC calls to insert and delete
-key-value pairs and lookup keys. Such a hash table is simple to derive from a
-volatile hash table and readily available in NVML [nvml.io] as an example [src
-  code link]. To use that (non client-server) example with Cyclone we added a
-simple RPC call format, manually marshalled and demarshalled the few
-arguments to the RPC call in an RPC call handler we wrote - a few tens of lines
-of code. Finally, we added a simple client driver program (a few tens of
-additional lines of code).
+key-value pairs and lookup keys.
+
+A durable hash table is simple to derive from a volatile hash table and readily
+available in NVML as an example~\cite{nvml_hash}. To use that (non client-server)
+example with Cyclone we added a simple RPC call format, manually marshalled and
+demarshalled the few arguments to the RPC call in an RPC call handler we wrote -
+a few tens of lines of code. Finally, we added a simple client driver program (a
+few tens of additional lines of code).
 
@@
 specify a group of insert/delete/lookup operations that must occur
-atomically. We use the existing bucket locks to implement a simple two phase
+atomically. We use the existing bucket locks to support a simple two phase
 locking protocol to implement the transction. A client makes an RPC call to the
@@
 
-We guard against faults by executing the RPC call on a quorum of machines via
-Cyclone. Since execution of the RPC call leaves \emph{externally visible side
-  effects} by virtue of the RPC calls it makes we require output
+We guard against faults by executing the the entire transaction on a quorum of
+machines via Cyclone. All machines in the quorum execute the transaction, making
+calls to other quorums to execute steps in the two phase locking protocol. Since
+execution of the RPC call leaves \emph{externally visible side effects} by
+virtue of the RPC calls it makes we require output
 commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
@@
 clients to load the system for our studies. The server side machines are each
-equipped with four commodity 82599 10 Gbit ethernet controllers, while each
-client machine has one 82599 ethernet controller. All the machines are connected
-through the same commodity 10Gbit switch. We use 8 cores on the server machines
+equipped with four commodity 82599-based 10 Gbit ethernet ports, while each
+client machine has one 82599-based ethernet port. All the machines are connected
+through the same commodity 10 Gbit switch. We use 8 cores on the server machines
 for running 8 instance of RAFT, leaving 39 cores to run application code. The
@@
 
-First, we evaluate pure replication performance by replicating RPC calls. There
-is crash consistent transaction that remembers the RPC call sequence number -
-every RPC call is accepted and once replicated a response is sent to the
-client. Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one
-where the clients send an RPC call with no payload and one where they add a 512
-byte dummy payload. For the no payload case the bottleneck is the message
-handling rate. We note that Cyclone can handle small message replication rates
-as high as \emph{8 million messages a second}. In the case of RPC calls with 512
-byte payloads, the bottleneck is the available network bandwidth. We plot
-transmit throughput derived from the message rate as the rate at which data is
-sent to replicas from all the 8 quorum leaders and back to the clients.
-It is interesting to note that with 3 replicas Cyclone achieves \emph{more than
-  40 Gbps} of bandwidth. The reason for this is that the quorum leaders can end
-up residing on different machines. Although replication traffic is still limited
-by the receive bandwidth the responses to the clients, originating from
-different machines have access to independent 4x10GigE NICs, allowing total
-transmit throughput to exceed 40 Gbps. Although not discussed in this paper,
-Cyclone's control plane is designed to exploit this by trying to spread quorum
-leaders across machines.
+First, we evaluate pure replication performance by replicating RPC
+calls. Cyclone simply replicates the contents of the RPC call without executing
+it and returns a response to the client when replication is complete.
+Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one where
+the clients send an RPC call with no payload and one where they add a 512 byte
+dummy payload. For the no payload case the bottleneck is the message handling
+rate. We note that Cyclone can handle small message replication rates as high as
+\emph{8 million messages a second}. In the case of RPC calls with 512 byte
+payloads, the bottleneck is the available network bandwidth. To show this, we
+also plot transmit throughput derived from the message by adding the rate at
+which data is sent to replicas from all the 8 quorum leaders to the rate at
+which responses are sent back to the clients.  It is interesting to note that
+with 3 replicas Cyclone achieves \emph{more than 40 Gbps} of bandwidth. The
+reason for this is that the quorum leaders can end up residing on different
+machines. Although replication traffic is still limited by the receive bandwidth
+the responses to the clients, originating from different machines have access to
+independent 4x10GigE ports, allowing total transmit throughput to exceed 40
+Gbps. Although not discussed in this paper, Cyclone's control plane is designed
+to exploit this by trying to spread quorum leaders across machines.
 
@@
 \begin{minipage}{0.25\textwidth}
+  \hspace{-0.17in}
   \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
@@
 \begin{minipage}{0.25\textwidth}
-  \hspace{-0.2in}
+  \hspace{-0.37in}
   \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
@@
 \begin{minipage}{0.25\textwidth}
+  \hspace{-0.17in}
   \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
@@
 \begin{minipage}{0.25\textwidth}
-  \hspace{-0.2in}
+  \hspace{-0.37in}
   \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
@@
 consistent transaction where the application still does nothing but Cyclone
-remembers the last sequence number and RPC call result (the application simply
-returns the received payload). Throughput is now bottlenecked by NVML allocator
-performance at just under a million messages a second. The final result is
-adding the hash table in the picture with the clients updating the values of
-keys already present in the hash table - a stable workload for easy
+durably remembers the last sequence number and RPC call result (the application
+simply returns the received payload). Throughput is now bottlenecked by NVML
+allocator performance at just under a million messages a second. The final
+result is adding the hash table in the picture with the clients updating the
+values of keys already present in the hash table - a stable workload for easy
 comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
@@
 
-These results lead us to conclude that Cyclone's provides adequate performance
-headroom for replication to NVML applications. In fact our current focus in the
-light of these results is towards improving application performance on top of
-NVML. NVML is a general purpose library written for a variety of substrates
-including traditional SSD based storage and there is adequate room for
-improvement of performance for applications running on directly attached
+These results lead us to conclude that Cyclone provides adequate performance
+headroom for transparent replication in NVML applications. In fact our current
+focus in the light of these results is towards improving application performance
+on top of NVML. NVML is a general purpose library written for a variety of
+substrates including traditional SSD based storage and there is adequate room
+for improvement of performance for applications running on directly attached
 persistent memory with close to DRAM performance.
@@
 \begin{minipage}{0.25\textwidth}
+  \hspace{-0.17in}
   \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
-  \caption{Crash Consistency + Replication}
+  \caption{Crash Consistency}
   \label{fig:cc_rep}
 \end{minipage} &
+  \hspace{-0.37in}
 \begin{minipage}{0.25\textwidth}
   \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
-  \caption{Crash Consistent Replicated Hash Table}
+  \caption{Hash Table}
   \label{fig:app_rep}
@@
 libraries such as NVML and achieve good performance on commodity networking
-hardware. A key question is whether the ideas in Cyclone extend to more than
-just key-value maps - perhaps to application such as filesystems ? We believe
-that the requirement for scaling across multiple replication quorums can be
-satisfied by applications where the scalable commutativity
-rule~\cite{scalable_commutativity} applies to common case primitives. Although
-our intitial release of Cyclone will be targeted towards key value style
-applications, our future research is targeted towards investigating this
-hypothesis.
+hardware. A key question is whether Cyclone can be used for other applications
+such as filesystems ? We believe that the requirement for scaling across
+multiple replication quorums can be satisfied by applications where the scalable
+commutativity rule~\cite{scalable_commutativity} applies to common case
+primitives. Although our intitial release of Cyclone will be for
+key value style applications, our future research is based on this hypothesis.
 \newcommand\myurl[2]{\url{#1}}

### 2016-09-26 paper/paper.tex
More tweaks to paper.

@@
 network ports to provide good throughput under load. We show that Cyclone can
-replicate as many as 8 million updates a second using just 4 commodity 10 GigE
+replicate as many as 7 million updates a second using just 4 commodity 10 GigE
 ports on a machine.
@@
 provide adequate background for the rest of this paper. NVML requires 
-that code that makes changes to non-volatile heap is enclosed in a crash
+that code that makes changes to the non-volatile heap is enclosed in a crash
 consistent transaction such that either all of the transaction executes or none
@@
 
-Asynchronous replication provides good performance and we believe it would be
+Decoupled replication provides good performance and we believe it would be
 the common case for RPC calls replicated via Cyclone. However the program might 
@@
 ability to declare at the client side that an RPC call should execute
-synchronously at the server side. Cyclone would then ensure that the RPC
+in a coupled manner at the server side. Cyclone would then ensure that the RPC
 call is successfully replicated before it attempts execution on any of the
@@
 This composite packet buffer is then sent \emph{unchanged} to every replica.
-To properly route the packet through switches, we allocate a small buffer
-containing an ethernet header for each follower replica and \emph{chain} the
-buffer in front of the prepared packet - thereby ensuring that we do not need
-to change the contents of the packet itself for every replica.
+To properly route the packet through switches, for each follower replica we
+allocate a small buffer containing an ethernet header with appropriate addresses
+and \emph{chain} the buffer in front of the prepared packet - thereby ensuring
+that we do not need to change the contents of the packet itself.
 
@@
 number and a sequence number. \emph{Each execution core independently enforces
-  execution of RPC calls in increasing sequence numbers}. This is done by
-storing the last executed sequence number in a per-core durable datastructure as
-part of the crash consistent transaction wrapping the execution of the RPC call
-itself. Even in the presence of fail-overs, this durable memory of sequence
-numbers seen from a client ensures at most once semantics. In addition to
-guarding agains repeated execution of the same call, Cyclone also remembers the
-\emph{results} of the last executed RPC call. This is done to assist clients
-(particularly stateless clients). Cyclone allows the client to query the last
-seen sequence number and result for that RPC call from the quorum of servers.
+  that accepted RPC calls from the same client bear increasing sequence
+  numbers}. This is done by storing the last executed sequence number for every
+client in a per-core durable datastructure as part of the crash consistent
+transaction wrapping the execution of the RPC call. Even in the presence of
+fail-overs, this durable memory of sequence numbers seen from a client ensures
+at most once semantics. In addition to guarding against repeated execution of
+the same call, Cyclone also remembers the \emph{result} of the last executed RPC
+call from every client. This is done to assist clients (particularly stateless
+clients). Cyclone allows the client to query the last seen sequence number and
+result for that RPC call from the quorum of servers.
 
@@
 such designs, which make availability easy for NVM applications are key to
-increasing the adoption rate of NVM by data center operators.
+increasing the adoption rate of NVM by datacenter operators.
 
@@
 Cyclone's server side code, while the remaining machines run multithreaded
-clients to load the system for our studies. The server side machines are each
+clients. The server side machines are each
 equipped with four commodity 82599-based 10 Gbit ethernet ports, while each
-client machine has one 82599-based ethernet port. All the machines are connected
-through the same commodity 10 Gbit switch. We use 8 cores on the server machines
-for running 8 instance of RAFT, leaving 39 cores to run application code. The
-last remaining core is reserved for management code. We use NVML with the non
-volatile pool held in DRAM, which proxies for the performance we would expect
-from NVDIMMs. We evaluate Cyclone in three stages.
+client machine has one 82599-based 10 Gbit ethernet port. All the machines are
+connected through the same 10 Gbit switch. We use 8 cores on the
+server machines for running 8 instance of RAFT, leaving 39 cores to run
+application code. The last remaining core is reserved for management code. We
+use NVML with the non volatile pool held in DRAM, which proxies for the
+performance we would expect from NVDIMMs. We evaluate Cyclone in three stages.
 
@@
 rate. We note that Cyclone can handle small message replication rates as high as
-\emph{8 million messages a second}. In the case of RPC calls with 512 byte
+\emph{7 million messages a second}. In the case of RPC calls with 512 byte
 payloads, the bottleneck is the available network bandwidth. To show this, we
@@
 machines. Although replication traffic is still limited by the receive bandwidth
-the responses to the clients, originating from different machines have access to
-independent 4x10GigE ports, allowing total transmit throughput to exceed 40
-Gbps. Although not discussed in this paper, Cyclone's control plane is designed
-to exploit this by trying to spread quorum leaders across machines.
+at follower replicas, the responses to the clients, originating from different
+machines have access to independent 4x10GigE ports, allowing total transmit
+throughput to exceed 40 Gbps. Although not discussed in this paper, Cyclone's
+control plane is designed to exploit this by trying to spread quorum leaders
+across machines.
 
@@
 
-Next, we consider the effect of adding NVMLs crash consistent
-transactions. Figure~\ref{fig:cc_rep} shows the effect of adding a crash
+Next, Figure~\ref{fig:cc_rep} shows the effect of adding a crash
 consistent transaction where the application still does nothing but Cyclone
@@
 comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
-under 1M calls a second mark for pure updates. For comparison we also show a
+under 1M calls a second mark for pure updates. We also show a
 pure read only case where the clients only query the value of a key - there is
 no replication \emph{nor} any crash consistent transaction applied. It is
-interesting to note that the pure read only case is bottlenecked at the same
-point as the pure replication case of 8 million operations a second.
+interesting to note that the pure read only case is bottlenecked almost at the same
+point as the pure replication case of 7 million operations a second.
 
@@
 focus in the light of these results is towards improving application performance
-on top of NVML. NVML is a general purpose library written for a variety of
-substrates including traditional SSD based storage and there is adequate room
-for improvement of performance for applications running on directly attached
-persistent memory with close to DRAM performance.
+on top of NVML rather than improving the performane of replication in Cyclone.
+%NVML is a general purpose library written for a variety of
+%substrates including traditional SSD based storage and there is adequate room
+%for improvement of performance for applications running on directly attached
+%persistent memory with close to DRAM performance.
 

### 2016-09-27 paper/paper.tex
updates.

@@
 keys to values.  Cyclone is designed around an optimized userspace networking
-stack to reduce latency and can scale its consensus protocol across cores and
-network ports to provide good throughput under load. We show that Cyclone can
-replicate as many as 7 million updates a second using just 4 commodity 10 GigE
-ports on a machine.
+stack to reduce latency. Further, it can scale its consensus protocol across
+cores and network ports to provide good throughput under load. We show that
+Cyclone can replicate as many as 7 million updates a second using just 8 cores
+(for the consensus protocol) and 4 commodity 10 GigE ports on each replica.
 
@@
 increased use of directly addressable non-volatile memory both in the form of
-battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types
-such as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
+battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
+as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
 programmers to build durable data structures on non-volatile heaps have become
-available~\cite{mnemosyne, nvheaps, cdds}  - an example of such a
-library that we have used in this paper is Intel's non volatile memory library
+available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
+have used in this paper is Intel's non volatile memory library
 (NVML)~\cite{nvml}. A missing piece however is a component to make such durable
 data structures highly available using replication across a commodity network
-such as ethernet. Existing solutions that offer good performance are biased
-towards pure transactions across RDMA with a dependence on external services
-(such as zookeeper~\cite{zookeeper}) and application specific code for
-fault recovery~\cite{farm, htm}.
-This is a problem for programmers using NVML to construct durable and available 
-data structures where one cannot apriori guarantee the availability of
-RDMA (for e.g., when running on externally hosted environments or public
+such as ethernet. Existing solutions that offer good performance are dependent
+on high performance RDMA networking hardware, with a dependence on external
+services (such as zookeeper~\cite{zookeeper}) and application specific code for
+fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
+where one cannot apriori guarantee the availability of such RDMA networking
+hardware (for e.g., when running on externally hosted environments or public
 clouds).
 
-Cyclone is prototype replication middleware that aims to demonstrate that one
+Cyclone is prototype replication middleware aiming to demonstrate that one
 can add fault tolerance transparently to a specific class of client server
@@
 via replication -- a quorum of machines maintains \emph{equivalent} state.
-Server state is queried and manipulated via RPC calls from clients.
-Cyclone replicates the RPC call itself across replicas rather than replicating
+Server state is queried and manipulated via RPC calls from clients.  Cyclone
+treats the RPC call itself as a variable sized opaque blob of data, leaving the
+application free to choose how it wishes to marshall call related details and
+arguments.
+
+Cyclone replicates the RPC call across replicas rather than replicating
 every access (to the server state) made during execution of the RPC call.
-
 Cyclone provides strongly consistent replication - every machine in the quorum
-of replicas maintains a log of RPC calls. Cyclone treats the RPC call itself as
-a variable sized opaque blob of data, leaving the application free to choose how
-it wishes to marshall call related details and arguments. All machines agree on
-the committed sequence of RPC calls in the log. We achieve this by running an
-instance of the RAFT~\cite{raft} consensus protocol to keep the log of RPC calls
-on different machines in synchronization. Specially flagged `read-only` RPC
-calls are executed only on the leader replica and are not replicated.
+of replicas maintains a log of RPC calls.  All machines agree on the committed
+sequence of RPC calls in the log. We achieve this by running an instance of the
+RAFT~\cite{raft} consensus protocol to keep the log of RPC calls on different
+machines in synchronization. Specially flagged `read-only` RPC calls are
+executed only on the leader replica and are not replicated.
 
@@
 in the log) be committed before they can be applied to the state machine --
-i.e. in our case before the RPC call can be executed. In contrast, Cyclone
-decouples execution of an RPC call on any replica from its replication.
-The fact that RPC calls that modify state must themselves be executed
-in a crash consistent transaction allows us to overlap execution with
-replication. Normally, this would be a problem as replication might fail for a
-number of reasons but primarily due to change of leadership and subsequent
-rollback of previously appended (but not committed) log entries. To solve this
-problem, Cyclone delays the actual commit of the failure atomic transaction
-until after the replication is complete. In the event that replication fails,
-the transaction is aborted and any changes to the non volatile heap are rolled
-back. Figure~\ref{fig:async_rep} shows a pseudocode description of how the
-runtime handles \emph{decoupled replication}.
+i.e. in our case before the RPC call can be executed. This is because
+replication of an accepted log entry can fail for a
+number of reasons such as change of leadership and subsequent
+rollback of previously appended (but not committed) log entries.
+
+In contrast to other systems, Cyclone decouples execution of an RPC call on any
+replica from its replication by executing the RPC call as soon as it is received
+on any replica. We exploit the fact that RPC calls that modify state must
+themselves be executed in a crash consistent transaction. Cyclone delays the
+actual commit of the crash consistent transaction until after the replication is
+complete. In the event that replication fails, the transaction is aborted and
+any changes to the non volatile heap are rolled back. Figure~\ref{fig:async_rep}
+shows a pseudocode description of how the runtime handles \emph{decoupled
+  replication}.
 
@@
 
-Decoupled replication provides good performance and we believe it would be
-the common case for RPC calls replicated via Cyclone. However the program might 
-on occasion require successful replication to be a precondition to exection.
-A case where we find this necessary is
-\textit{output commit}~\cite{output_commit} - where the RPC call has
-external side effects (e.g., communication with other machines in the cluster)
-and therefore crash consistency is insufficient to roll back the execution of
-the RPC call in the event that replication fails. We therefore provide users the
-ability to declare at the client side that an RPC call should execute
-in a coupled manner at the server side. Cyclone would then ensure that the RPC
-call is successfully replicated before it attempts execution on any of the
-replicas.
+Decoupled replication provides good performance and we believe it would be the
+common case for RPC calls replicated via Cyclone. However the program might on
+occasion require successful replication to be a precondition to exection. A case
+where we find this necessary is \textit{output commit}~\cite{output_commit} -
+where the RPC call has external side effects (e.g., communication with other
+machines in the cluster) and therefore crash consistency is insufficient to roll
+back the execution of the RPC call in the event that replication fails. We
+therefore provide users the ability to declare that an RPC call should execute
+in a coupled manner. Cyclone would then ensure that the RPC call is successfully
+replicated before it attempts execution on any of the replicas.
 
@@
 decidated to running application code. Each RAFT core runs an \emph{independent}
-instance of the RAFT protocol. RAFT cores pass work to the application cores
-through lock-free double ended queues. A RAFT core dispatches an incoming RPC
-call for execution simultaneously with sending it out for replication.
+instance of the RAFT protocol - in conjunction with a similar instance on every
+other replica. RAFT cores pass work to the application cores through lock-free
+double ended queues. A RAFT core dispatches an incoming RPC call for execution
+simultaneously with sending it out for replication.
 
@@
 
-In the event that the machine has multiple NICs, Cyclone spreads the network
-demands and scales by binding each RAFT or application core to a specific NIC
-and picking the necessary queues from that NIC.
+Cyclone can exploit multiple NICs on a machine by binding each RAFT or
+application core to a specific NIC and picking the necessary queues from that
+NIC.
 
@@
 receive intimation of committed entries as piggybacked information on subsequent
-replication messages. The common case path at the leader lends itself to zero
-copy since the same set of log entries is sent to all replicas.
+replication messages. The common case replication path at the leader lends
+itself to \emph{zero copy} transmission to the replicas, since the same set of
+log entries is sent to all replicas.
 
-In order to avoid an additional copy of the incoming RPC call contents from DRAM
-to persistent memory, Cyclone allocates space for packet buffers on the non-volatile
-heap. However, since packet data is cached, Cyclone ensures durability
-(or persistence) of this data by explicitly flushing the dirty
+In the receive path, we avoid an additional copy of the incoming RPC call
+contents from DRAM to persistent memory by allocating space for packet buffers
+on the non-volatile heap. However, since packet data is cached, Cyclone must
+ensure durability (or persistence) of this data by explicitly flushing the dirty
 cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
@@
 and \emph{chain} the buffer in front of the prepared packet - thereby ensuring
-that we do not need to change the contents of the packet itself.
+that we do not need to copy the contents of the packet itself.
 
 Although the common case path in RAFT is straightforward, it still requires
-hundreds of cycles (or a few microseconds) to process each log entry, which is
-a significant amount in the face of sub 10 us network latency between nodes with
-DPDK. We amortize this overhead when operating under load by
-batching multiple RPC calls together and treating the batch as a single RAFT log
-entry.  DPDK by itself can efficiently batch incoming packets under load,
-therefore allowing the user to easily receive a burst of packets
-(currently limited to 32) rather than a single packet for a polling call.
-DPDK further allows chaining packet
-buffers together and treating the chain as a single packet. This is convenient
-for our zero copy approach and we chain packet mbufs together to form a single
-packet. Since the entire chain is sent as a single packet when being replicated,
-the length of a chain is limited by the size of the coalesced packet, that must
-fit into an ethernet jumbo frame of approximately 9KB.
+hundreds of cycles (or a few microseconds) to process each log entry, which is a
+significant amount in the face of sub 10 us network latency between nodes with
+DPDK. We amortize this overhead when operating under load by batching multiple
+RPC calls together and treating the batch as a single RAFT log entry.  DPDK by
+itself can efficiently batch incoming packets under load, therefore allowing the
+user to easily receive a burst of packets (currently limited to 32) rather than
+a single packet for a polling call.  DPDK further allows chaining packet buffers
+together and treating the chain as a single packet. We chain these packet
+buffers together to form a single packet. Since the entire chain is sent as a
+single packet when being replicated, the length of a chain is limited by the
+size of the coalesced packet, that must fit into an ethernet jumbo frame of
+approximately 9KB.
 
@@
 consensus protocol instance) and specific application core. These details are
-provided by the user when making the RPC call to the client side library
-provided by Cyclone. The user therefore is aware of the number of execution
-cores and RAFT instances and given complete control of \emph{RPC call steering}.
-For most applications we envisage a simple modulo hashing scheme should suffice
-for load balancing.
+provided by the user when making the RPC call to the Cyclone client side
+library. The user therefore is aware of the number of execution cores and RAFT
+instances and given complete control of \emph{RPC call steering}. For most
+applications we envisage a simple modulo hashing scheme should suffice for load
+balancing, something that we plan to later automate in Cyclone.
 
@@
   numbers}. This is done by storing the last executed sequence number for every
-client in a per-core durable datastructure as part of the crash consistent
+client in a per-core durable data structure as part of the crash consistent
 transaction wrapping the execution of the RPC call. Even in the presence of
@@
 the same call, Cyclone also remembers the \emph{result} of the last executed RPC
-call from every client. This is done to assist clients (particularly stateless
-clients). Cyclone allows the client to query the last seen sequence number and
-result for that RPC call from the quorum of servers.
+call from every client. This is done to assist clients - particularly stateless
+clients. Cyclone allows the client to query the last seen sequence number and
+result for that RPC call from any application core on the quorum of servers.
 
@@
 A durable hash table is simple to derive from a volatile hash table and readily
-available in NVML as an example~\cite{nvml_hash}. To use that (non client-server)
-example with Cyclone we added a simple RPC call format, manually marshalled and
-demarshalled the few arguments to the RPC call in an RPC call handler we wrote -
-a few tens of lines of code. Finally, we added a simple client driver program (a
-few tens of additional lines of code).
+available in NVML as an example~\cite{nvml_hash}. To use that (non
+client-server) example with Cyclone we added a simple RPC call format, manually
+marshalled and demarshalled the few arguments to the RPC call in an RPC call
+handler - a few tens of lines of code. Finally, we added a simple client driver
+program (a few tens of additional lines of code).
 
@@
 table require \emph{per-key determinism} and not determinism across \emph{all}
-operations. We therefore steer operations to the same key to the same RAFT core
-and the same application core, by hashing the key to determine the RAFT core and
-execution core. The actual configuration of the hash table therefore can be
-different across different replicas - for example due to the insertion of two
-keys mapping to the same hash bucket completing in different orders across
-different replicas. However, since the results of insert, delete and lookup
-operations for a key depend only on previous operations to the same key the
-replicas provide identical results for committed operations.
+operations. We therefore steer operations to a particular key to the same RAFT
+core and the same application core (using a hash on the key).  The actual
+configuration of the hash table therefore can be different across different
+replicas - for example due to the insertion of two keys mapping to the same hash
+bucket completing in different orders across different replicas. However, since
+the results of insert, delete and lookup operations for a key depend only on
+previous operations to the same key the replicas provide identical results for
+committed operations. \emph{In other words, we exploit the fact that operations to
+different keys in a map are commutative.}
 
@@
 the locks. We believe that NVM application programmers - particularly those
-having experience with concurrent code should not find it hard to deal with
+with experience in concurrency should not find it hard to deal with
 distributed concurrency. The hard part for most programmers is likely to be the
@@
 as distributed transctions to be added without the programmer needing to worry
-about faults or write fault recovery code. As an exercise we show how to add
+about faults or writing fault recovery code. As an exercise we show how to add
 support for distributed transactions to our hash table.
@@
 execution of the RPC call leaves \emph{externally visible side effects} by
-virtue of the RPC calls it makes we require output
+virtue of the further RPC calls it makes, we require output
 commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
-complete once it starts. This is simply achieved by a flag in the RPC call that
-turns off decoupled execution (Section~\ref{sec:decouple}). The flag causes
-execution to start only after the RPC call has been successfully replicated.
+complete once it starts. Hence, for distributed transactions, we turn off
+decoupled execution (Section~\ref{sec:decouple}).
 
@@
 To ensure that the transaction is still guaranteed to complete, we use a
-\emph{different} client identifier for each phase. Thus, until the transaction
-completes and a new one is started the response for any RPC call continues to be
-available. Figure~\ref{fig:dist_tx} shows in pseudocode the server side details
-of a transaction. One initiated, the distributed transaction is guaranteed to
-complete given quorum availabilty. However, if the client requires intimation of
-completion, the final transaction state can be written to a distinguished
-partition as part of the transaction itself.
+\emph{different} client identifier for each \emph{phase}. Thus, until the
+transaction completes and a new one is started the response for any RPC call
+continues to be available. Figure~\ref{fig:dist_tx} shows in pseudocode the
+server side details of a transaction. Once initiated, the distributed transaction
+is guaranteed to complete given availability of a quorum.
+%However, if the client
+%requires intimation of completion, the final transaction state can be written to
+%a distinguished partition as part of the transaction itself.
 
-We note that this design is not intended for performance, under the assumption
-that distributed transactions will be infrequent compared to single key
-operations.  On the other hand, it is worth comparing the lack of any need for
-explicit fault recovery in Cyclone with fault recovery in systems systems that
-explicity require such cleanup (~\cite{farm}, pp 59--63).  We believe that
-such designs, which make availability easy for NVM applications are key to
-increasing the adoption rate of NVM by datacenter operators.
+Distributed transactions built around at most once semantics are not intended
+for performance. On the other hand, at most once semantics allows users of
+Cyclone to avoid writing explicit fault recovery code as compared to fault
+recovery in systems that explicity require such cleanup (~\cite{farm}, pp
+59--63). Under the assumption that distributed transactions will be infrequent
+compared to single key operations, such simplicity is key to increasing the
+adoption of NVM among datacenter operators concerned about availability.
 
@@
      Transaction completed by someone else
-
-optional:
- write transaction result to <quorum_tx_result>
 \end{verbatim}
@@
 Cyclone's server side code, while the remaining machines run multithreaded
-clients. The server side machines are each
-equipped with four commodity 82599-based 10 Gbit ethernet ports, while each
-client machine has one 82599-based 10 Gbit ethernet port. All the machines are
-connected through the same 10 Gbit switch. We use 8 cores on the
-server machines for running 8 instance of RAFT, leaving 39 cores to run
-application code. The last remaining core is reserved for management code. We
-use NVML with the non volatile pool held in DRAM, which proxies for the
-performance we would expect from NVDIMMs. We evaluate Cyclone in three stages.
+clients. The server side machines are each equipped with four commodity
+82599-based 10 Gbit ethernet ports, while each client machine has one
+82599-based 10 Gbit ethernet port. All the machines are connected to the same 10
+Gbit switch. We use 8 cores on the server machines for running 8 instances of
+RAFT and 39 cores to run application code. The last remaining core is reserved
+for management code. We use NVML with the non volatile pool held in DRAM, which
+proxies for the performance we would expect from NVDIMMs. We evaluate Cyclone in
+three stages.
 
-First, we evaluate pure replication performance by replicating RPC
-calls. Cyclone simply replicates the contents of the RPC call without executing
-it and returns a response to the client when replication is complete.
-Figure~\ref{fig:pure_rep} shows the results. We consider two cases, one where
-the clients send an RPC call with no payload and one where they add a 512 byte
-dummy payload. For the no payload case the bottleneck is the message handling
-rate. We note that Cyclone can handle small message replication rates as high as
-\emph{7 million messages a second}. In the case of RPC calls with 512 byte
-payloads, the bottleneck is the available network bandwidth. To show this, we
-also plot transmit throughput derived from the message by adding the rate at
-which data is sent to replicas from all the 8 quorum leaders to the rate at
-which responses are sent back to the clients.  It is interesting to note that
-with 3 replicas Cyclone achieves \emph{more than 40 Gbps} of bandwidth. The
-reason for this is that the quorum leaders can end up residing on different
-machines. Although replication traffic is still limited by the receive bandwidth
-at follower replicas, the responses to the clients, originating from different
-machines have access to independent 4x10GigE ports, allowing total transmit
-throughput to exceed 40 Gbps. Although not discussed in this paper, Cyclone's
-control plane is designed to exploit this by trying to spread quorum leaders
-across machines.
+First, we evaluate pure replication by having Cyclone simply replicate the
+contents of the RPC call without executing it and return a response to the
+client when replication is complete.  Figure~\ref{fig:pure_rep} shows the
+results. We consider two cases, one where the clients send an RPC call with no
+payload and one where they add a 512 byte dummy payload. For the no payload case
+the bottleneck is the message handling rate. We note that Cyclone can handle
+small message replication rates as high as \emph{7 million messages a
+  second}. In the case of RPC calls with 512 byte payloads, the bottleneck is
+the available network bandwidth. To show this, we also plot aggregate transmit
+throughput derived from these results by adding the rate at which data is sent
+to replicas from all the 8 quorum leaders to the rate at which responses are
+sent back to the clients.  With 3 replicas Cyclone shows \emph{slightly more
+  than 40 Gbps} of aggregate transmit bandwidth. The reason for this is that the
+quorum leaders can end up residing on different machines. Although replication
+traffic is still limited by the receive bandwidth at follower replicas, the
+responses to the clients, originating from different machines have access to
+independent 4x10GigE ports, allowing total transmit throughput to exceed 40
+Gbps. Although not discussed in this paper, Cyclone's control plane is designed
+to exploit this by trying to spread quorum leaders across machines.
 
@@
 way replication costs as little as 25 us of latency in the unloaded case. This
-latency is measured at the client and therefore is four commodity 10 GigE
+latency is measured at the client and is therefore four commodity 10 GigE
 network hops through the switch, with each hop taking under 7 us including
@@
 hardware. A key question is whether Cyclone can be used for other applications
-such as filesystems ? We believe that the requirement for scaling across
+such as filesystems ? \emph{We believe that the requirement for scaling across
 multiple replication quorums can be satisfied by applications where the scalable
 commutativity rule~\cite{scalable_commutativity} applies to common case
-primitives. Although our intitial release of Cyclone will be for
+primitives.} Although our intitial release of Cyclone will be for
 key value style applications, our future research is based on this hypothesis.

### 2016-09-27 paper/paper.tex
final version.

@@
 
-Next, Figure~\ref{fig:cc_rep} shows the effect of adding a crash
-consistent transaction where the application still does nothing but Cyclone
-durably remembers the last sequence number and RPC call result (the application
-simply returns the received payload). Throughput is now bottlenecked by NVML
-allocator performance at just under a million messages a second. The final
-result is adding the hash table in the picture with the clients updating the
-values of keys already present in the hash table - a stable workload for easy
-comparison. As Figure~\ref{fig:app_rep} shows performance stays at the just
-under 1M calls a second mark for pure updates. We also show a
-pure read only case where the clients only query the value of a key - there is
-no replication \emph{nor} any crash consistent transaction applied. It is
-interesting to note that the pure read only case is bottlenecked almost at the same
-point as the pure replication case of 7 million operations a second.
+Next, Figure~\ref{fig:cc_rep} shows the effect of adding a crash consistent
+transaction where the application still does nothing but Cyclone durably
+remembers the last sequence number and RPC call result (the application simply
+returns the received payload). Throughput is now bottlenecked by NVML allocator
+performance at just under a million messages a second. The final result is
+adding the hash table in the picture with the clients updating the values of
+keys already present in the hash table - a stable workload for easy
+comparison. As Figure~\ref{fig:app_rep} shows, performance stays at just under
+1M calls a second for pure updates. We also show a pure read only case where the
+clients only query the value of a key - there is no replication \emph{nor} any
+crash consistent transaction applied. It is interesting to note that the pure
+read only case is bottlenecked almost at the same point as the pure replication
+case of 7 million operations a second, demonstrating the efficiency of Cyclone's
+replication under load.
 
-These results lead us to conclude that Cyclone provides adequate performance
-headroom for transparent replication in NVML applications. In fact our current
-focus in the light of these results is towards improving application performance
-on top of NVML rather than improving the performane of replication in Cyclone.
+Contrary to common wisdom that replication over commodity networks is expensive,
+these results lead us to conclude that Cyclone provides adequate performance
+headroom for transparent replication in NVML-based key value maps.
+%In fact our current focus in the light of
+%these results is towards improving application performance on top of NVML rather
+%than improving the performane of replication in Cyclone.
 %NVML is a general purpose library written for a variety of
@@
 \section{Conclusion}
-Cyclone allows developers to add availability to applications built on top of
+Cyclone allows developers to add availability to key value maps built on top of
 libraries such as NVML and achieve good performance on commodity networking
 hardware. A key question is whether Cyclone can be used for other applications
-such as filesystems ? \emph{We believe that the requirement for scaling across
-multiple replication quorums can be satisfied by applications where the scalable
-commutativity rule~\cite{scalable_commutativity} applies to common case
-primitives.} Although our intitial release of Cyclone will be for
-key value style applications, our future research is based on this hypothesis.
+such as filesystems? \emph{We believe that the requirement for scaling across
+  multiple replication quorums can be satisfied by applications where the
+  scalable commutativity rule~\cite{scalable_commutativity} applies to common
+  case primitives.} Although our intitial release of Cyclone will be for key
+value style applications, our on-going research is based on the scalable
+commutativity hypothesis.
 \newcommand\myurl[2]{\url{#1}}

### 2016-09-27 paper/paper.tex
Fixing captions.

@@
   }
+\vspace{-0.2in}
 \caption{A Persistent Linked List}
@@
 size of the coalesced packet, that must fit into an ethernet jumbo frame of
-approximately 9KB.
+approximately 9KB. Figure~\ref{fig:zc_batch} summarizes this set of
+transformations to incoming request packets from clients at the leader replica.
 
 \begin{figure}
-\includegraphics[width=0.7\textwidth,height=6cm]{figures/chain.pdf}
+\includegraphics[width=0.9\columnwidth]{figures/chain}
+\vspace{-0.15in}
 \caption{Zero Copy Batching}
@@
 \end{figure}
-
-This set of transformations to incoming request packets from clients at the
-leader replica is summarized in Figure~\ref{fig:zc_batch}.
 
@@
 }
+\vspace{-0.22in}
 \caption{Distributed Transactions}
@@
 \end{tabular}
-\caption{Pure Replication}
+\caption{Evaluation of pure replication performance}
 \label{fig:pure_rep}
+\end{figure}
+
+\begin{figure}[htb]
+%\vspace{-0.2in}
+\centering
+\begin{tabular}{c}
+\includegraphics[height=3cm,width=0.8\columnwidth]{results/cc_mops.pdf}
+\end{tabular}
+\vspace{-0.22in}
+\caption{Evaluation of replication performance with crash consistency}
+\label{fig:cc_rep}
+\vspace{-0.08in}
+\end{figure}
+
+\begin{figure}[htb]
+%\vspace{-0.2in}
+\centering
+\begin{tabular}{c}
+\includegraphics[height=3cm,width=0.8\columnwidth]{results/app_mops.pdf}
+\end{tabular}
+\vspace{-0.22in}
+\caption{Evaluation of in-memory hash table performance with 3-way crash consistent replication}
+\label{fig:app_rep}
+\vspace{-0.08in}
 \end{figure}
@@
 
-\begin{figure}
-\begin{tabular}{cc}
-\begin{minipage}{0.25\textwidth}
-  \hspace{-0.17in}
-  \includegraphics[width=\textwidth,height=3cm]{results/cc_mops.pdf}
-  \caption{Crash Consistency}
-  \label{fig:cc_rep}
-\end{minipage} &
-  \hspace{-0.37in}
-\begin{minipage}{0.25\textwidth}
-  \includegraphics[width=\textwidth,height=3cm]{results/app_mops.pdf}
-  \caption{Hash Table}
-  \label{fig:app_rep}
-\end{minipage}
-\end{tabular}
-\end{figure}
-
 \section{Conclusion}

### 2017-03-03 paper/paper.tex
bring back author names for circulation.
Switch to sigplanconf format.

@@
-\documentclass[letterpaper,twocolumn,10pt]{article}
-\usepackage{usenix,epsfig,endnotes,graphicx,url}
+\documentclass[10pt, preprint]{sigplanconf}
+\usepackage{times,epsfig,endnotes,graphicx}
+\usepackage{datetime, url, hyperref}
+
+\conferenceinfo{SOSP'17}{October 29--31, 2017, Shanghai, China}
+\copyrightyear{2017}
+
+% These only appear when the 'preprint' option is specified.
+% Enabling these will cause the first page of the document to fail the 
+% format check on HotCRP :-(
+%\titlebanner{Under submission to SOSP 2017 - do not cite or distribute}
+%\preprintfooter{Draft of {\currenttime}, \today{}}
+
+% No date in title area.
+\date{}
+
+% Paper number and no. of pages as author
+\authorinfo{Paper \textbf{\#XX}}{NN pages}
+
 \begin{document}
 
-%don't want date printed
-\date{}
-
 %make title bold and 14 pt font (Latex default is non-bold, 16 pt)
-\title{\Large \bf Cyclone: Replication Middleware for NVM Applications}
-
-%for single author (just remove % characters)
-%\author{
-%  {\rm Paper #: XXX }\\
-%  XXX
-%  \and
-%  {\rm YYY}\\
-%  YYY
-%} % end author
+\title{\Large \bf Cyclone: Warp Speed Replication for Key Value Stores}
 
@@
 
-\subsection*{Abstract}
+\begin{abstract}
 The introduction of non-volatile directly attached memory in datacenters - in
@@
 (for the consensus protocol) and 4 commodity 10 GigE ports on each replica.
+\end{abstract}
 
@@
 \newcommand\myurl[2]{\url{#1}}
-\bibliographystyle{acm}
+\bibliographystyle{plainnat}
 \bibliography{paper}

### 2017-04-18 paper/paper.tex
Updates to paper.

@@
 \usepackage{times,epsfig,endnotes,graphicx}
-\usepackage{datetime, url, hyperref}
+\usepackage{datetime, url, hyperref, pdfpages}
 
@@
 % Paper number and no. of pages as author
-\authorinfo{Paper \textbf{\#XX}}{NN pages}
+\authorinfo{Paper \textbf{\#224}}{NN pages}
 
@@
 \begin{abstract}
-The introduction of non-volatile directly attached memory in datacenters - in
-the form of NVDIMMs and new memory technologies allows efficient data structures
-to be built on non-volatile heaps that provide good performance coupled with
-durability. However high availability remains a problem for these primarily
-single machine solutions. While transactional interfaces built over RDMA capable
-networking hardware offer a viable alternative, such hardware is often
-unavailable in datacenters. Cyclone offers a solution to this problem by
-providing transparent replication middleware that can be used with Intel's Non
-Volatile Memory Library (NVML) to construct durable data-structures that map
-keys to values.  Cyclone is designed around an optimized userspace networking
-stack to reduce latency. Further, it can scale its consensus protocol across
-cores and network ports to provide good throughput under load. We show that
-Cyclone can replicate as many as 7 million updates a second using just 8 cores
-(for the consensus protocol) and 4 commodity 10 GigE ports on each replica.
+Persistent key value stores are rapidly becoming an important component of many
+distributed data serving solutions - with innovations targeted at taking
+advantage of growing flash speeds. Unfortunately their performance is hampered
+by the need to maintain and replicate a write ahead log to guarantee
+availability in the face of machine and storage failures. Cyclone offers a
+solution to this problem by leveraging a small amount of directly attached
+non-volatile memory such as NVDIMMs to maintain a two level log - the upper
+level in a small amount of NVDIMM memory draining into a larger lower level log
+maintained on flash. Replicating the smaller log kept in NVDIMMs transforms the
+replication problem into a packet switching problem - the leader sends the exact
+same packet to all follower replicas and we solve the problem as such by using
+the RAFT protocol implemented as a software multicast module on the Data Plane
+Development Kit (DPDK). Second, we levergage the observation that key value store
+interfaces are commutative for operations to different keys to scale our two
+level log horizontally into a number of physical two level logs - with a hash of
+the key being used to choose the right log. Each log is replicated by an
+independent instance of RAFT, thereby leveraging the scalability afforded by
+classic software packet switching when using multiple CPU cores. Cyclone is able
+to replicate millions of small updates a second using only commodity 10 gigabit
+ethernet adapters and improves the performance of RocksDB - a popular persistent
+key value store used in industry by orders of magnitude when compared to its own
+write ahead log without replication.
 \end{abstract}
@@
 \section{Evaluation}
-We evaluate Cyclone against its design goals by using it for up to 3-way
-replication in a cluster of twelve 48 core Xeon servers. Three servers run
-Cyclone's server side code, while the remaining machines run multithreaded
-clients. The server side machines are each equipped with four commodity
-82599-based 10 Gbit ethernet ports, while each client machine has one
-82599-based 10 Gbit ethernet port. All the machines are connected to the same 10
-Gbit switch. We use 8 cores on the server machines for running 8 instances of
-RAFT and 39 cores to run application code. The last remaining core is reserved
-for management code. We use NVML with the non volatile pool held in DRAM, which
-proxies for the performance we would expect from NVDIMMs. We evaluate Cyclone in
-three stages.
+We evaluate Cyclone on a 12 node cluster connected via a 10 GigE switch. Three
+of the machines are equipped with 1.6TB Intel DC P3600 SSDs and 4*10 GigE
+ports. The remaining nine machines do not have SSDs and have only one 10 GigE
+port, serving as clients for most of the experiments. As with other
+work~\cite{faast}, we use DRAM on the machines to proxy for NVDIMMs where
+necessary - the persistent memory needed never exceeds 64 MB regardless of the
+size of the key value store or second level log on flash. We divide the
+evaluation into three parts. First, we evaluate Cyclone's performance with a
+single level log as a pure software packet switch. Next, we evaluate performance
+when adding a second level of log on flash. Finally, we evaluate performance
+when integrated with Rocksdb as an alternative to Rocksdb's write ahead
+log. Unless otherwise mentioned, we use a 60 byte header followed by a payload
+for experiments. We log both the header and payload. In all cases the server
+echoes the received entry (header and payload) back to the client. We fix the
+number of application threads at 32 using at most 8 threads for running RAFT
+instances.
 
-First, we evaluate pure replication by having Cyclone simply replicate the
-contents of the RPC call without executing it and return a response to the
-client when replication is complete.  Figure~\ref{fig:pure_rep} shows the
-results. We consider two cases, one where the clients send an RPC call with no
-payload and one where they add a 512 byte dummy payload. For the no payload case
-the bottleneck is the message handling rate. We note that Cyclone can handle
-small message replication rates as high as \emph{7 million messages a
-  second}. In the case of RPC calls with 512 byte payloads, the bottleneck is
-the available network bandwidth. To show this, we also plot aggregate transmit
-throughput derived from these results by adding the rate at which data is sent
-to replicas from all the 8 quorum leaders to the rate at which responses are
-sent back to the clients.  With 3 replicas Cyclone shows \emph{slightly more
-  than 40 Gbps} of aggregate transmit bandwidth. The reason for this is that the
-quorum leaders can end up residing on different machines. Although replication
-traffic is still limited by the receive bandwidth at follower replicas, the
-responses to the clients, originating from different machines have access to
-independent 4x10GigE ports, allowing total transmit throughput to exceed 40
-Gbps. Although not discussed in this paper, Cyclone's control plane is designed
-to exploit this by trying to spread quorum leaders across machines.
-
-Also worth noting is that thanks to DPDK's low latency access to the NICs, three
-way replication costs as little as 25 us of latency in the unloaded case. This
-latency is measured at the client and is therefore four commodity 10 GigE
-network hops through the switch, with each hop taking under 7 us including
-network stack delays. These results should serve to illustrate that simple
-commodity ethernet can indeed provide reasonable replication performance with a
-carefully designed software stack that can exploit it.
+Cyclone is built on top of DPDK to apply software packet switching techniques to
+the log replication problem. We therefore begin by systematically evaluating
+optimizations applied in Cyclone to replicate the top level log in
+Figure~\ref{fig:network_opts} - with no payload. The y-axis reports latency seen
+at the client (which means two network round trips with replication). Using
+TCP/IP to replicate a RAFT log tops out at around 30K entries/s. Applying
+optimizations improves performance by a large enough amount that we use a log
+scale on the x-axis. Switching to DPDK (the line marked +DPDK) improves the
+throughput by an order of magnitude to around 50K entries/s. Using batching
+(line marked +batching) improves the performance further bringing us close to a
+million entries/s. Scaling horizontally to 8 physical logs (+8 phy logs)
+improves performance to close to 2M entries/s. Finally using all 4 ports on the
+machine to replicate entries improves performance considerably to 6M
+entries/s. In all performance improves by 200X over the TCP/IP single log
+baseline. Cyclone also considerably improves the latency for replication, from
+close to 100us with TCP/IP to around 30us at peak throughput.
 
 \begin{figure}
-\begin{tabular}{cc}
-\begin{minipage}{0.25\textwidth}
-  \hspace{-0.17in}
-  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_0.pdf}
-\end{minipage}&
-\begin{minipage}{0.25\textwidth}
-  \hspace{-0.37in}
-  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_0.pdf}
-\end{minipage}\\
-\begin{minipage}{0.25\textwidth}
-  \hspace{-0.17in}
-  \includegraphics[width=\textwidth,height=3cm]{results/replication_mops_512.pdf}
-\end{minipage}&
-\begin{minipage}{0.25\textwidth}
-  \hspace{-0.37in}
-  \includegraphics[width=\textwidth,height=3cm]{results/replication_gbps_512.pdf}
-\end{minipage}
-\end{tabular}
-\caption{Evaluation of pure replication performance}
-\label{fig:pure_rep}
+\includegraphics[scale=0.6]{results2/network_opts.pdf}
+\caption{Network optimizations for top level log}
+\label{fig:network_opts}
 \end{figure}
 
-\begin{figure}[htb]
-%\vspace{-0.2in}
-\centering
-\begin{tabular}{c}
-\includegraphics[height=3cm,width=0.8\columnwidth]{results/cc_mops.pdf}
-\end{tabular}
-\vspace{-0.22in}
-\caption{Evaluation of replication performance with crash consistency}
-\label{fig:cc_rep}
-\vspace{-0.08in}
+There are two factors that can have significant impact on Cyclone's
+performance. First, the number of replicas dictates the outgoing message rate
+from the leader replica and therefore increasing the replication factor can
+decrease Cyclone's performance. Figure~\ref{fig:replicas} shows the impact of
+varying replica count. Using only a single replica cuts out a network round trip
+and shows the best unloaded latency (10 us) and peak throughput (near 10M
+entries/s). Adding replicas decreases the peak throughput down to around 2M
+entries/s. We note that a number of previous pieces of work~\cite{faast, farm}
+use three replicas and therefore we focus on three replicas for the replicated
+cases we consider below. The second factor that dictates Cyclone's performance
+is the size of the log entry being replicated. Figure~\ref{fig:payload} shows
+the effect of increasing the payload size from zero to 512 bytes. Peak
+throughput drops from 6M entries/s to approximately 2M entries/s. At this
+replication rate, the leader replica needs to transmit data at approximately 30
+Gbit/s. Coupled with the cost of network headers all four 10 GigE ports are now
+saturated and therefore Cyclone hits the network line rate bottleneck at this
+point.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/replicas.pdf}
+\caption{Impact of replica count}
+\label{fig:replicas}
 \end{figure}
 
-\begin{figure}[htb]
-%\vspace{-0.2in}
-\centering
-\begin{tabular}{c}
-\includegraphics[height=3cm,width=0.8\columnwidth]{results/app_mops.pdf}
-\end{tabular}
-\vspace{-0.22in}
-\caption{Evaluation of in-memory hash table performance with 3-way crash consistent replication}
-\label{fig:app_rep}
-\vspace{-0.08in}
+\begin{figure}
+\includegraphics[scale=0.6]{results2/512.pdf}
+\caption{Impact of payload size}
+\label{fig:payload}
 \end{figure}
 
-Next, Figure~\ref{fig:cc_rep} shows the effect of adding a crash consistent
-transaction where the application still does nothing but Cyclone durably
-remembers the last sequence number and RPC call result (the application simply
-returns the received payload). Throughput is now bottlenecked by NVML allocator
-performance at just under a million messages a second. The final result is
-adding the hash table in the picture with the clients updating the values of
-keys already present in the hash table - a stable workload for easy
-comparison. As Figure~\ref{fig:app_rep} shows, performance stays at just under
-1M calls a second for pure updates. We also show a pure read only case where the
-clients only query the value of a key - there is no replication \emph{nor} any
-crash consistent transaction applied. It is interesting to note that the pure
-read only case is bottlenecked almost at the same point as the pure replication
-case of 7 million operations a second, demonstrating the efficiency of Cyclone's
-replication under load.
+We now turn our attention from the network component of Cyclone to the storage
+one by adding the second level flashlog. We evaluate the impact of adding
+storage to our hitherto pure packet switching scenario in
+Figure~\ref{fig:flashlog}. Batching entries from the top level NVDIMM log to the
+second level flashlog is clearly beneficial as adding flash storage at the
+second level has almost no impact on peak performance in Cyclone for small
+entries. The situation however changes for larger
+entries. Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
+significant impact on peak throughput - it drops to approximately 350K
+ops/sec. This corresponds to around 50K 4KB IOPS to the SSD to write out the
+flashlog pages. The peak for the drive is 160K IOPS using a queue depth
+(concurrency) of 128. With 32 application threads we expect a lower peak
+throughput around 40K IOPS explaining our bottleneck at 50K IOPS.  It is
+possible to tune our observed performance further by aligning the flush boundary
+to increase the number of outstanding requests - we do not do so in this paper,
+keeping Cyclone agnostic to the exact size of entry being replicated. A final
+point about Figure~\ref{fig:flashlog_512} is that once we are past the storage
+bottleneck the latency spike is dramatic and large enough to trigger Cyclone's
+failure detector and repeated retries from the clients. There are - therefore -
+no points on the ``knee'' of the curve as in the pure packet switched one level
+log case.
 
-Contrary to common wisdom that replication over commodity networks is expensive,
-these results lead us to conclude that Cyclone provides adequate performance
-headroom for transparent replication in NVML-based key value maps.
-%In fact our current focus in the light of
-%these results is towards improving application performance on top of NVML rather
-%than improving the performane of replication in Cyclone.
-%NVML is a general purpose library written for a variety of
-%substrates including traditional SSD based storage and there is adequate room
-%for improvement of performance for applications running on directly attached
-%persistent memory with close to DRAM performance.
+\begin{figure}
+\includegraphics[scale=0.6]{results2/flashlog.pdf}
+\caption{Impact of adding second level log}
+\label{fig:flashlog}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/flashlog_512.pdf}
+\caption{Impact of adding second level log (512 bytes payload)}
+\label{fig:flashlog_512}
+\end{figure}
+
+The final dimension we evaluate is of ganged operations. The primary purpose of
+ganged operations is to avoid the need for distributed transactions to
+manipulate what is a single shared memory image and therefore we were most
+concerned about unloaded latency given the complexity of synchronizing different
+replication quorums as well executing our rendezvous protocol on multiple
+cores. We therefore setup an experiment where a single client reflecting the
+unloaded case made ganged requests to the replicas. We varied the number of
+cores participating in the request from one to the full complement of 32
+application cores. Figure~\ref{fig:ganged} shows the results both using a single
+level log as well as a two level log. The primary takeaway is that unloaded
+latency increases slowly as we increase the number of active cores - to around
+40 us from the baseline of 20 us. There are two causes of this. First, there is
+a rapid increase from 1 to 8 active cores, the reason being that this
+corresponds to an increase in the number of quorums that must be synchronized
+on. There is some amount of synchronization involved in the userspace DPDK
+driver for access to common resources for accessing the NIC and this affects the
+ability to simultaneously issues replication messages on all quorums. The
+smaller contribution to increasing latency that continues past the 8 core case
+is due to cacheline pingponging when executing the rendevous. Both these sources
+of latency drift could be corrected using replication quorums mapped to
+dedicated NICs and more scalable rendesvous designs (such as with machine aware
+broadcast trees~\cite{broadcast_tree}). However we deemed the complexity of such
+optimizations unnecessary. The added latency for most cases is well under the
+extra round trip delay, which would the minimum needed for a solution using
+distributed transactions via two phase commit.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/multi.pdf}
+\caption{Ganged Operations}
+\label{fig:ganged}
+\end{figure}
 

### 2017-04-18 test/rocksdb_client_multicore.cpp
Fixes to rocksdb client.

@@ -70,6 +70,7 @@ int gen_multi(char * buffer, int keys, unsigned long max_key, unsigned long *cor
   rock_kv_pair_t *tail = (rock_kv_pair_t *)(buffer + sizeof(rock_kv_t));
   for(int i=1;i<keys;i++,tail++) {
     tail->key = rand() % max_key;
+    (*core_mask) |= (1UL << (tail->key % executor_threads));
     size += sizeof(rock_kv_pair_t);
   }
   return size;
@@ -133,6 +134,7 @@ int driver(void *arg)
       rpc_flags = 0;
       kv->op    = OP_PUT;
       unsigned long cmask = 0;
+      //int random_core = rand() % executor_threads;
       int req_sz = gen_multi(buffer, active, keys, &cmask);
       sz = make_rpc(handles[0],
 		    buffer,

### 2017-04-18 paper/paper.tex
updates to paper.

@@
 of latency drift could be corrected using replication quorums mapped to
-dedicated NICs and more scalable rendesvous designs (such as with machine aware
+dedicated NICs and more scalable rendezvous designs (such as with machine aware
 broadcast trees~\cite{broadcast_tree}). However we deemed the complexity of such
@@
 \label{fig:ganged}
+\end{figure}
+
+We now evaluate Cyclone integrated with the Rocksdb persistent key value
+store. Rocksdb is a complex industrial strength persistent key value store and
+this means that it is accompanied by a complex array of performance tuning knobs
+to get the best performance from flash. We were also aware that SSD performance
+is slated for dramatic increases in the coming years with the introduction of
+new memory technology such as 3DXPoint based Optane SSDs. To demonstrate that
+Cyclone is future proof we eliminated flash media performance from the picture
+as far as the key value store is concerned by placing all files for the key
+value store (SSTables) on a RAMdisk, which presumably represents the limit in
+performance for flash in the near future. All log files on secondary storage
+however - both Rocksdb's own write ahead log and the alternative of Cyclone's
+second level flashlog - are placed on the SSDs.
+
+We evaluate two different request sizes: 8 byte keys with 8 byte values and 8
+bytes keys with 256 byte values. We run 8 physical logs on 8 cores and devote 32
+cores to making requests to Rocksdb. Since we are interested in performance of
+the log, which is only used for update requests, our workload is 100\% writes.
+
+Before evaluating with Rocksdb we measure the baseline performance of
+replicating the log with Cycloine for the given request sizes - Rocksdb performs
+a no-op. We note that in addition to the key and value, we are also logging
+RocksDB specific request data such as operation type and the request header.
+Figure~\ref{fig:kv_baseline} shows the baseline performance for the chosen
+request sizes. With the smaller request size, Cyclone can conservatively sustain
+close to a million requests a second at a latency of just under 25us. With the
+larger request size, Cyclone can sustain around 350K requests a second, again at
+a latency of just under 25 us. Armed with these baseline numbers we now examine
+how well Cyclone performs with Rocksdb.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/kv_baseline.pdf}
+\caption{Baseline KV replication performance}
+\label{fig:kv_baseline}
+\end{figure}
+
+The performance of Rocksdb with Cyclone for the small update workload is shown
+in Figure~\ref{fig:rocksdb}. We consider four different settings. The line
+labeled Rocksdb is the key value store running with no logging whatsover - a
+system crash would lead to data loss. The line labeled Rocksdb/WAL is for
+Rocksdb running with its write ahead logging turned on. The large gap between
+these two is the overhead of the existing Rocksdb WAL solution. The line labeled
+Rocksdb/Cyclone 1 way is a two level Cyclone log but without any
+replication. The line almost exactly tracks the performance of Rockdb. As
+suggested by the baseline replication performance, Cyclone is able to provide a
+write ahead log with no overhead to Rocksdb. The line labeled Rocksdb/Cyclone 3
+way is with 3-way replication turned on. Other than a 20us delta due to the
+extra network round trip, the line almost exactly tracks the Rocksdb no logging
+case. Cyclone therefore provides high availability to Rocksdb at a fraction of
+the cost of its existing single machine write ahead log. We also repeat the
+experiment for the larger update size in Figure~\ref{fig:rocksdb_256}. The
+conclusions are identical: Cyclone solves Rocksdb's write ahead logging
+problem.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/rocksdb.pdf}
+\caption{Rocksdb - small updates}
+\label{fig:rocksdb}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/rocksdb_256.pdf}
+\caption{Rocksdb - large updates}
+\label{fig:rocksdb_256}
+\end{figure}
+
+Next, we consider the problem of supporting Rocksdb's write-batch operation that
+atomically writes a set of key-value pairs into the KV store. In order to
+perform the operation with Cyclone managing the log, the client must issue a
+ganged operation across cores owning the keys in the write batch. This is in
+contrast to baseline Rocksdb where the entire write batch can be sent to any
+core. A key concern here was whether Cyclone would add any latency to the
+operation due to the extra synchronization needed across replication quorums and
+participating application cores. We examine the problem for the unloaded case
+and small updates in Figure~\ref{fig:rocksdb_multi} for a single client with
+increasing number of keys in the batch - till 32 keys that covers all
+application cores. The line labeled Rocksdb is with no write-ahead logging. We
+note an increasing latency for this baseline indicating Rocksdb takes longer
+with larger key batches. For the other options - Rocksdb/wal has considerably
+larger latency. Cyclone does an effective job of cutting down on this latency
+even as it needs to pay a price for synchronizing multiple quorums and
+application cores making it somewhat slower than running Rocksdb with no logging
+for batched writes.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/rocksdb_multi.pdf}
+\caption{Rocksdb - batched writes}
+\label{fig:rocksdb_multi}
+\end{figure}
+
+Finally we showcase the benefit of using Cyclone beyond pure performance as
+compared to the existing single machine Rocksdb write ahead log. Cyclone brings
+multi-machine availability with the ability to automatically failover. We
+demonstrate this in Figure ~\ref{fig:timeline} that shows the timeline of a run
+where we kill the server process on the leader replica. Cyclone is configured to
+with a 30ms failure detection timeout after which the client library tries
+replicas in turn to locate the leader - in this case it fails over in about
+60ms.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/failover.pdf}
+\caption{Rocksdb - failover}
+\label{fig:timeline}
 \end{figure}

### 2017-04-19 paper/paper.tex
Updates to paper.

@@
-\documentclass[10pt, preprint]{sigplanconf}
-\usepackage{times,epsfig,endnotes,graphicx}
+\documentclass[10pt, preprint, nonatbib]{sigplanconf}
+\usepackage{times,epsfig,endnotes,graphicx, verbatim}
 \usepackage{datetime, url, hyperref, pdfpages}
@@
 solution to this problem by leveraging a small amount of directly attached
-non-volatile memory such as NVDIMMs to maintain a two level log - the upper
-level in a small amount of NVDIMM memory draining into a larger lower level log
-maintained on flash. Replicating the smaller log kept in NVDIMMs transforms the
+non-volatile memory (NVM) such as NVDIMMs to maintain a two level log - the upper
+level in a small amount of NVM draining into a larger lower level log
+maintained on flash. Replicating the smaller log kept in NVMs transforms the
 replication problem into a packet switching problem - the leader sends the exact
@@
 ethernet adapters and improves the performance of RocksDB - a popular persistent
-key value store used in industry by orders of magnitude when compared to its own
-write ahead log without replication.
+key value store by orders of magnitude when compared to its own write ahead log
+without replication.
 \end{abstract}
@@
 \section{Introduction}
-Memory technology in datacenters is due to undergo a paradigm shift with the
-increased use of directly addressable non-volatile memory both in the form of
-battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
-as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
-programmers to build durable data structures on non-volatile heaps have become
-available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
-have used in this paper is Intel's non volatile memory library
-(NVML)~\cite{nvml}. A missing piece however is a component to make such durable
-data structures highly available using replication across a commodity network
-such as ethernet. Existing solutions that offer good performance are dependent
-on high performance RDMA networking hardware, with a dependence on external
-services (such as zookeeper~\cite{zookeeper}) and application specific code for
-fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
-where one cannot apriori guarantee the availability of such RDMA networking
-hardware (for e.g., when running on externally hosted environments or public
-clouds).
 
-Cyclone is prototype replication middleware aiming to demonstrate that one
-can add fault tolerance transparently to a specific class of client server
-applications written on top of libraries such as NVML. We specifically target
-key-value maps in this paper and in that context make two key
-contributions. First, we propose a simple API that enables server state to be
-replicated without the programmer needing to worry about fault recovery. This
-allows one to start by building a concurrent durable data structure using
-libraries such as NVML and add availability at a later stage using Cyclone. The
-second contribution is to show that one can achieve reasonably good replication 
-performance using commodity ethernet via innovations in the networking stack and
-by scaling replication with the application. This result should be a useful
-design point for developers looking to deploy applications on non-volatile heaps
-in commodity ethernet-based environments such as public clouds.
+\section{Two Level Log}
+A fundamental challenge in building an efficient replicated log is dealing with
+network overheads. Cyclone'e construction is predicated on the observation that
+in the common case protocols such as RAFT simply multicast the same entry
+received from a client, from the leader replica to all follower replicas. This
+suggests that an efficient way to implement a replicated log is to treat it as a
+software packet switching problem. However peristing logs on flash represents an
+impediment in that block storage devices are not good candidates to hold packets
+being software switched. Our solution is to split the log into two levels. The
+upper level is held in persistent directly attached non-volatile memory (NVM)
+such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
+level log as the flash log. We periodically drain entries in batches from the
+NVM log to the backing flashlog. An important and beneficial side effect of this
+two level approach is that we need only limited amounts of NVM with the bulk of
+the log residing on flash - an arrangement that maps well to the relative cost
+of the two types of memory. We begin by describing the NVM log and how we
+implement the RAFT consensus protocol in the packet switching paradigm around
+it. We then describe the flashlog and how we drain entries in batches from the
+NVM level to the flash level.
 
-\section{Single Node Programming Model}
-Cyclone assumes a concurrent durable application on a non-volatile heap as the
-starting point. A number of libraries and filesystems make this possible
-today~\cite{nvml, dax, pmfs, mnemosyne, nvheaps, cdds}. Cyclone currently
-supports applications built on top of Intel's Non Volatile Memory Library
-(NVML~\cite{nvml}). We describe NVML's programming model in this section to
-provide adequate background for the rest of this paper. NVML requires 
-that code that makes changes to the non-volatile heap is enclosed in a crash
-consistent transaction such that either all of the transaction executes or none
-of it executes in the event of power failures.
-%This is critical to maintaining consistency as the data structure, being
-%durable, must survive across power failures.
-Figure~\ref{fig:example} illustrates the use of NVML to build a persistent
-linked list. Updates to the linked list are enclosed in a NVML transaction to
-avoid corruption due to power failure. It is also worth noting that NVM
-libraries such as NVML usually allow remapping non-volatile heaps to a different
-virtual memory address range across restarts (e.g., using {\tt TOID} macros to
-implement fat pointers in NVML).
+\subsection{NVM Log}
+The NVM log is maintained as a circular log of fixed sized pointers (a pointer
+log) to memory buffers - each of which holds a client request. This arrangement
+is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of
+such an arragement. Adding a level of indirection in the NVM log allows us to
+isolate the circular log being manipulated by the RAFT protocol from the
+problems of variable sized client requests that can be scattered across
+non-volatile memory depending on the internals of the memory allocator of the
+packets switching library we use. Another advantage is that it makes recovery
+from NVM easy - appends to the circular pointer log are atomic and we use the
+pointer log to recover allocator state i.e. what pieces of NVM are currently in
+use by the log.
 
 \begin{figure}
-  { \scriptsize
+\centering
+\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+\hspace{-0.2in}
+\caption{NVM Log Structure}
+\label{fig:nvm_log}
+\end{figure}
+
+We now describe how we implement the RAFT consensus protocol to replicate the
+NVM log across a set of replicas. The software packet switching idea we apply is
+to separate the control plane from the dataplane using the Data Plane
+Development Kit (DPDK~\cite{dpdk}) library for direct userspace access to the
+network interface.
+
+\subsubsection{Dataplane Separation}
+We use RAFT~\cite{raft} as the underlying protocol for log replication. RAFT
+replicates a log across a set of replicas. Its operations are leader-based and
+RAFT incorporates a leader election algorithm that elects a leader with the most
+upto date log. The leader receives log entries from clients, appends them to the
+persistent log (NVM log in this case) and multicasts the new entry to follower
+replicas. On a response from enough follower replicas (constituting a majority
+quorum) the leader replica can then execute the command in the log entry. In
+our case this involves accessing the key value store and sending the response
+back to the client.
+
+We implement the common case flow in RAFT as a software packet switch using
+DPDK. The dataplane now encompasses the NIC and the CPU caches with the NIC
+placing packet data directly in the CPU cache via DDIO~\cite{ddio} - a standard
+feature of most NICs today. The dataplane for the RAFT protocol is therefore
+separated from the control plane. The control plane now runs on a CPU core and
+takes care of checking adherence to the protocol: checking for correct term, log
+index and that the replica receiving the packet from the client is indeed the
+leader. In addition, the control plane is responsible for non common-case
+activities such as dealing with timeout and leader elections.
+
+The pseudocode in Figure~\ref{fig:control_plane} describes the control plane in
+Cyclone organized as event handlers triggered on receiving a packet. We focus on
+two events for illustration (and brevity): the event where a request is received
+from a client and the event where a leader election related message is
+received. A key concern when we built Cyclone was adhering as far as possible to
+the software packet switching principle of separating the control plane from the
+dataplane. This means that the CPU should avoid examining every byte of the
+packet as far as possible.
+
+If an election message is received all the code related to handling
+the election related message is run on the CPU core and all contents of the
+packet are examined to decide what to do in accordance with the RAFT protocol
+state machine. This is an exception to the software packet switching principle
+but one that we found acceptable since leader elections are rare compared to the
+common case of replicating client packets.
+
+The sequence of operations when a packet is received from the client however
+needs careful design for efficient software packet switching. To illustrate how
+this is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates
+packet layouts across the two steps of prepending a RAFT header and transmitting
+to follower replicas. DPDK describes packets using an ``mbuf'' data
+structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
+actually containing the byte and a fixed size piece of external metadata that
+describes various aspects of the packet, most crucially a pointer to the start
+and end of the packet in the byte array. DPDK's userspace drivers receive
+packets from the NIC such that they are offset in the byte array by a
+configurable amount referred to as ``headroom''. We prepend RAFT related
+information in the headroom and shift the start pointer appropriately - these
+operations are standard enough for software packet switches that DPDK provides
+convenient library calls to do it. For the final step, we need to prepare the
+packet for transmission to the various follower replicas. To do this we prepare
+a different packet containing an ethernet header for the targeted replica and
+``chain'' the data packet to each of these headers. Each header is then
+separately handed off to the driver for transmission via the NIC.
+
+\begin{figure}
 \begin{verbatim}
-struct ll_node {
-  int value;
-  TOID(ll_node) next;
-};
-
-void insert_after(TOID(ll_node) prev, 
-                  TOID(ll_node) new_node)
+    event_receive_client_req()
     {
-  TX_BEGIN {
-    TX_ADD(new_node);
-    D_RW(new_node)->next = D_RO(prev)->next;
-    TX_ADD(prev);
-    D_RW(prev)->next = new_node;
-  } TX_END
+      if(!check_is_leader()) {
+        drop_msg
+        return
+      }
+      Prepend raft log term and index
+      Persist to local log
+      Transmit to follower replicas
     }
 
+    event_receive_election_msg()
+    {
+      Handle_election_msg()
+    }
 \end{verbatim}
-  }
-\vspace{-0.2in}
-\caption{A Persistent Linked List}
-\label{fig:example}
+\caption{Cyclone Control Plane Fragment}
+\label{fig:control_plane}
 \end{figure}
 
-It is easy to turn the linked list example in NVML into a concurrent linked list
-by simply adding a lock to it. Such a lock can be maintined in volatile DRAM if
-so desired for performance. More sophisticated fine-grained or optimistic
-concurrency control mechanisms for NVM are also
-possible~\cite{mnemosyne, cdds, nvheaps}.
-
-\section{Replication}
-Cyclone is designed to add fault tolerance to NVML client server applications
-via replication -- a quorum of machines maintains \emph{equivalent} state.
-Server state is queried and manipulated via RPC calls from clients.  Cyclone
-treats the RPC call itself as a variable sized opaque blob of data, leaving the
-application free to choose how it wishes to marshall call related details and
-arguments.
-
-Cyclone replicates the RPC call across replicas rather than replicating
-every access (to the server state) made during execution of the RPC call.
-Cyclone provides strongly consistent replication - every machine in the quorum
-of replicas maintains a log of RPC calls.  All machines agree on the committed
-sequence of RPC calls in the log. We achieve this by running an instance of the
-RAFT~\cite{raft} consensus protocol to keep the log of RPC calls on different
-machines in synchronization. Specially flagged `read-only` RPC calls are
-executed only on the leader replica and are not replicated.
-
-\subsection{Decoupled Replication}
-\label{sec:decouple}
-The standard replicated state machine approach requires that commands (entries
-in the log) be committed before they can be applied to the state machine --
-i.e. in our case before the RPC call can be executed. This is because
-replication of an accepted log entry can fail for a
-number of reasons such as change of leadership and subsequent
-rollback of previously appended (but not committed) log entries.
-
-In contrast to other systems, Cyclone decouples execution of an RPC call on any
-replica from its replication by executing the RPC call as soon as it is received
-on any replica. We exploit the fact that RPC calls that modify state must
-themselves be executed in a crash consistent transaction. Cyclone delays the
-actual commit of the crash consistent transaction until after the replication is
-complete. In the event that replication fails, the transaction is aborted and
-any changes to the non volatile heap are rolled back. Figure~\ref{fig:async_rep}
-shows a pseudocode description of how the runtime handles \emph{decoupled
-  replication}.
-
 \begin{figure}
-{ \scriptsize
-\begin{verbatim}
-Initiate replication of RPC call and arguments 
-TX_BEGIN { 
-  Execute RPC call and modify local NVM 
-  Block till replication result known 
-  if(replication failed) {
-     TX_ABORT 
-  } 
-} TX_END
-\end{verbatim}
-}
-\caption{Decoupled Replication}
-\label{fig:async_rep}
+  \caption{Cyclone Network Packet Layout}
+  \label{fig:packet_layout}
 \end{figure}
 
-Decoupled replication provides good performance and we believe it would be the
-common case for RPC calls replicated via Cyclone. However the program might on
-occasion require successful replication to be a precondition to exection. A case
-where we find this necessary is \textit{output commit}~\cite{output_commit} -
-where the RPC call has external side effects (e.g., communication with other
-machines in the cluster) and therefore crash consistency is insufficient to roll
-back the execution of the RPC call in the event that replication fails. We
-therefore provide users the ability to declare that an RPC call should execute
-in a coupled manner. Cyclone would then ensure that the RPC call is successfully
-replicated before it attempts execution on any of the replicas.
+TBD: talk about how we do not need a reliable network layer like TCP/IP.
+\subsubsection{Batching}
 
-\subsection{Network Stack}
-Cyclone is designed around DPDK~\cite{dpdk} -- a high performance network stack
-optimized for Intel processors. DPDK provides convenient access to packets
-arriving at an ethernet NIC by moving them into packet buffers and directly
-placing those buffers into the last level cache of the CPU during DMA~\cite{ddio}.
+\subsection{Flash Log}
+TBD: talk about why we don't need group commit.
+\section{Horizontal Scaling}
 
-Cyclone separates the available CPU cores on a machine, in a configurable way,
-into cores dedicated to work related to the (RAFT) consensus protocol and cores
-decidated to running application code. Each RAFT core runs an \emph{independent}
-instance of the RAFT protocol - in conjunction with a similar instance on every
-other replica. RAFT cores pass work to the application cores through lock-free
-double ended queues. A RAFT core dispatches an incoming RPC call for execution
-simultaneously with sending it out for replication.
+\subsection{API Commutativity}
 
-Each RAFT core has a dedicated NIC queue pair. The input queue receives RPC
-calls from clients as well as RAFT protocol related traffic. The output queue
-only carries RAFT protocol related traffic. Each application core has a
-dedicated \emph{output} queue on the NIC. When completed, the response to the
-RPC call is sent from the application core via its dedicated output queue. This
-division of NIC queues eliminates the need for synchronization between RAFT
-cores and application cores when accessing the NIC.
+\subsection{Core Assignment}
 
-Cyclone can exploit multiple NICs on a machine by binding each RAFT or
-application core to a specific NIC and picking the necessary queues from that
-NIC.
+\subsection{NIC Queue Assignmnt}
 
-\subsection{Zero Copy Batching}
-Consensus protcols such as RAFT usually have a simple common case path. Once a
-quorum leader is elected, the leader's primary job is to transmit new entries
-appended to its persistent log to its followers. The primary job of a follower
-is to receive these entries, append them to its persistent log and send
-acknowledgments to the leader. An entry is considered committed when the leader
-receives an acknowledgment for an entry from a majority of the quorum. Followers
-receive intimation of committed entries as piggybacked information on subsequent
-replication messages. The common case replication path at the leader lends
-itself to \emph{zero copy} transmission to the replicas, since the same set of
-log entries is sent to all replicas.
-
-In the receive path, we avoid an additional copy of the incoming RPC call
-contents from DRAM to persistent memory by allocating space for packet buffers
-on the non-volatile heap. However, since packet data is cached, Cyclone must
-ensure durability (or persistence) of this data by explicitly flushing the dirty
-cachelines~\cite{pmfs}. Cyclone maintains its log as a circular sequence of
-pointers to packet buffers and therefore the append operation is completed by
-adding a pointer to the received packet and flushing the cacheline containing
-the pointer.
-
-We exploit the fact that DPDK allows prepending data to a packet
-without physically moving the data in a packet buffer. The received packet
-contains only the RPC call from the client. We prepend a RAFT related header
-(containing the previous log index, previous log term, leader's term and
-leader's commit point~\cite{raft}) to the packet buffer itself. To this RAFT header 
-we then prepend an IPv4 header. We assign pre-determined values to the
-IP address fields in order to exploit 5-tuple steering at the receiving NIC for 
-the placement of the incoming packet in the correct queue.
-This composite packet buffer is then sent \emph{unchanged} to every replica.
-To properly route the packet through switches, for each follower replica we
-allocate a small buffer containing an ethernet header with appropriate addresses
-and \emph{chain} the buffer in front of the prepared packet - thereby ensuring
-that we do not need to copy the contents of the packet itself.
-
-Although the common case path in RAFT is straightforward, it still requires
-hundreds of cycles (or a few microseconds) to process each log entry, which is a
-significant amount in the face of sub 10 us network latency between nodes with
-DPDK. We amortize this overhead when operating under load by batching multiple
-RPC calls together and treating the batch as a single RAFT log entry.  DPDK by
-itself can efficiently batch incoming packets under load, therefore allowing the
-user to easily receive a burst of packets (currently limited to 32) rather than
-a single packet for a polling call.  DPDK further allows chaining packet buffers
-together and treating the chain as a single packet. We chain these packet
-buffers together to form a single packet. Since the entire chain is sent as a
-single packet when being replicated, the length of a chain is limited by the
-size of the coalesced packet, that must fit into an ethernet jumbo frame of
-approximately 9KB. Figure~\ref{fig:zc_batch} summarizes this set of
-transformations to incoming request packets from clients at the leader replica.
-
-\begin{figure}
-\includegraphics[width=0.9\columnwidth]{figures/chain}
-\vspace{-0.15in}
-\caption{Zero Copy Batching}
-\label{fig:zc_batch}
-\end{figure}
-
-\subsection{RPC Call Execution}
-\label{sec:exec}
-We now describe how RPC calls are executed and results returned to the
-client. There are two important aspects of RPC call execution in Cyclone -
-steering and at most once semantics.
-
-An RPC call via Cyclone must traverse a specific RAFT core (and associated
-consensus protocol instance) and specific application core. These details are
-provided by the user when making the RPC call to the Cyclone client side
-library. The user therefore is aware of the number of execution cores and RAFT
-instances and given complete control of \emph{RPC call steering}. For most
-applications we envisage a simple modulo hashing scheme should suffice for load
-balancing, something that we plan to later automate in Cyclone.
-
-Cyclone provides at most once semantics. It supports a large but fixed (at
-startup time) number of clients. Each RPC call from a client bears the client
-number and a sequence number. \emph{Each execution core independently enforces
-  that accepted RPC calls from the same client bear increasing sequence
-  numbers}. This is done by storing the last executed sequence number for every
-client in a per-core durable data structure as part of the crash consistent
-transaction wrapping the execution of the RPC call. Even in the presence of
-fail-overs, this durable memory of sequence numbers seen from a client ensures
-at most once semantics. In addition to guarding against repeated execution of
-the same call, Cyclone also remembers the \emph{result} of the last executed RPC
-call from every client. This is done to assist clients - particularly stateless
-clients. Cyclone allows the client to query the last seen sequence number and
-result for that RPC call from any application core on the quorum of servers.
-
-\section{A Durable Replicated Hash Table}
-\label{sec:example}
-We now demonstrate how Cyclone can be used in practice by putting together an
-example of a linearizable hash table - a useful and common data structure.
-The baseline is a durable hash table written in NVML that runs on a
-single machine in client-server mode. The server side maintains the durable
-open-addressed hash table and services client RPC calls to insert and delete
-key-value pairs and lookup keys.
-
-A durable hash table is simple to derive from a volatile hash table and readily
-available in NVML as an example~\cite{nvml_hash}. To use that (non
-client-server) example with Cyclone we added a simple RPC call format, manually
-marshalled and demarshalled the few arguments to the RPC call in an RPC call
-handler - a few tens of lines of code. Finally, we added a simple client driver
-program (a few tens of additional lines of code).
-
-In order to use the hash table with Cyclone, we simply need to to declare the
-RPC call handler as the point of entry to Cyclone at the server end.  The client
-end issues RPC calls via Cyclone's client side library. Crucially neither side
-involves fault awareness or fault recovery code - a key goal for Cyclone.
-
-There are two interesting challenges in scaling the performance and improving
-the utility of the linearizable and durable hash table. The first is to ensure
-linearizability while scaling performance by adding RAFT cores and application
-cores. The second is to support distributed transactions - that manipulate a set
-of key-value pairs atomically in the face of failure.
-
-\subsection{Scaling}
-Scaling the performance of the hash table by using multiple RAFT cores and
-application cores presents determinism related challenges. Two RPC calls being
-steered through different RAFT cores can end up executing in different orders on
-different replicas even if steered to the same application core. At the same time,
-two RPC calls steered through the same RAFT core but dispatched to different
-application cores can end up completing in different orders on different replicas.
-
-Our solution to this problem in Cyclone is to observe that replicas of the hash
-table require \emph{per-key determinism} and not determinism across \emph{all}
-operations. We therefore steer operations to a particular key to the same RAFT
-core and the same application core (using a hash on the key).  The actual
-configuration of the hash table therefore can be different across different
-replicas - for example due to the insertion of two keys mapping to the same hash
-bucket completing in different orders across different replicas. However, since
-the results of insert, delete and lookup operations for a key depend only on
-previous operations to the same key the replicas provide identical results for
-committed operations. \emph{In other words, we exploit the fact that operations to
-different keys in a map are commutative.}
-
-The RPC call steering mechanism described above leaves the programmer free to
-use any synchronization mechanism they wish between concurrent accesses from the
-application cores to the data structure, since the ordering between accesses
-from different cores is allowed to be non-deterministic. For our open addressed
-hash table we chose the simple scheme of a per-bucket lock.
-
-\subsection{Distributed Transactions}
-A useful primitive for many distributed systems is the ability to run
-distributed transactions - an atomic sequence of conditional updates spanning
-multiple groups of replicas. Distributed transactions have two distinct facets:
-atomicity and fault tolerance. Distributed transactions appear atomic with
-respect to other atomic transactions and simple updates or reads by virtue of
-standard concurrency control techniques. A common technique is strict two phase
-locking: acquire all locks, update the locked objects and finally release all
-the locks. We believe that NVM application programmers - particularly those
-with experience in concurrency should not find it hard to deal with
-distributed concurrency. The hard part for most programmers is likely to be the
-second aspect: fault tolerance. A distributed transaction spans many machines
-and has a large footprint to clean up in the event of failure. Failure handling
-is made more complex when considering cascading failures - further failures
-while recovering from a failure. Cyclone was designed to allow primitives such
-as distributed transctions to be added without the programmer needing to worry
-about faults or writing fault recovery code. As an exercise we show how to add
-support for distributed transactions to our hash table.
-
-We support distributed transactions in our hash table by allowing the user to
-specify a group of insert/delete/lookup operations that must occur
-atomically. We use the existing bucket locks to support a simple two phase
-locking protocol to implement the transction. A client makes an RPC call to the
-server specifying the transaction and the server executes it on the client's
-behalf. Execution of the transaction has four distinct phases: acquire locks,
-verify values, apply updates and finally release locks.
-
-We guard against faults by executing the the entire transaction on a quorum of
-machines via Cyclone. All machines in the quorum execute the transaction, making
-calls to other quorums to execute steps in the two phase locking protocol. Since
-execution of the RPC call leaves \emph{externally visible side effects} by
-virtue of the further RPC calls it makes, we require output
-commit~\cite{output_commit}. This is achieved if the execution is guaranteed to
-complete once it starts. Hence, for distributed transactions, we turn off
-decoupled execution (Section~\ref{sec:decouple}).
-
-Next, we need to ensure that although every machine in the quorum makes all
-necessary RPC calls, the call itself is executed only once. This is achieved by
-using Cyclone's at most once semantics. All machines in the quorum executing the
-transaction use the same client identifier and the same montonically increasing
-sequence of RPC call numbers. Finally, we need to deal with the fact that
-Cyclone only remembers the last executed RPC call for a client. This means that
-if different machines in the quorum move at different speeds, some machines will
-not receive a response to the RPC call - they will instead receive an error that
-tells them that they are too far behind.
-
-To ensure that the transaction is still guaranteed to complete, we use a
-\emph{different} client identifier for each \emph{phase}. Thus, until the
-transaction completes and a new one is started the response for any RPC call
-continues to be available. Figure~\ref{fig:dist_tx} shows in pseudocode the
-server side details of a transaction. Once initiated, the distributed transaction
-is guaranteed to complete given availability of a quorum.
-%However, if the client
-%requires intimation of completion, the final transaction state can be written to
-%a distinguished partition as part of the transaction itself.
-
-Distributed transactions built around at most once semantics are not intended
-for performance. On the other hand, at most once semantics allows users of
-Cyclone to avoid writing explicit fault recovery code as compared to fault
-recovery in systems that explicity require such cleanup (~\cite{farm}, pp
-59--63). Under the assumption that distributed transactions will be infrequent
-compared to single key operations, such simplicity is key to increasing the
-adoption of NVM among datacenter operators concerned about availability.
-
-\begin{figure}
-{ \scriptsize
-\begin{verbatim}
-wait until RPC call is replicated
-respond to client saying tx accepted for execution
-acquire:
- set client identifier to 1
- for each machine
-   Acquire write or read locks on the machine
-   If machine reponds RPC sequence too old
-     Transaction completed by someone else
-
-update:
- set client identifier to 2
- for each machine
-   Update object
-   If machine reponds RPC sequence too old
-     Transaction completed by someone else
-
-release:
- set client identifier to 3
- for each machine
-   Update object
-   If machine reponds RPC sequence too old
-     Transaction completed by someone else
-\end{verbatim}
-}
-\vspace{-0.22in}
-\caption{Distributed Transactions}
-\label{fig:dist_tx}
-\end{figure}
+\subsection{Ganged Operations}
 
@@
 
+\section{Related Work}
+Memory technology in datacenters is due to undergo a paradigm shift with the
+increased use of directly addressable non-volatile memory both in the form of
+battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
+as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
+programmers to build durable data structures on non-volatile heaps have become
+available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
+have used in this paper is Intel's non volatile memory library
+(NVML)~\cite{nvml}. A missing piece however is a component to make such durable
+data structures highly available using replication across a commodity network
+such as ethernet. Existing solutions that offer good performance are dependent
+on high performance RDMA networking hardware, with a dependence on external
+services (such as zookeeper~\cite{zookeeper}) and application specific code for
+fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
+where one cannot apriori guarantee the availability of such RDMA networking
+hardware (for e.g., when running on externally hosted environments or public
+clouds).
+
 \section{Conclusion}
@@
 \newcommand\myurl[2]{\url{#1}}
-\bibliographystyle{plainnat}
+\bibliographystyle{plain}
 \bibliography{paper}

### 2017-04-19 paper/paper.tex
Updates to paper.

@@
 \usepackage{times,epsfig,endnotes,graphicx, verbatim}
-\usepackage{datetime, url, hyperref, pdfpages}
+\usepackage{datetime, url, pdfpages}
 
@@
 \includegraphics[scale=0.3]{figures2/nvm_log.pdf}
-\hspace{-0.2in}
 \caption{NVM Log Structure}
@@
 configurable amount referred to as ``headroom''. We prepend RAFT related
-information in the headroom and shift the start pointer appropriately - these
-operations are standard enough for software packet switches that DPDK provides
-convenient library calls to do it. For the final step, we need to prepare the
-packet for transmission to the various follower replicas. To do this we prepare
-a different packet containing an ethernet header for the targeted replica and
-``chain'' the data packet to each of these headers. Each header is then
-separately handed off to the driver for transmission via the NIC.
+information specific to each log entry in the headroom and shift the start
+pointer appropriately - these operations are standard enough for software packet
+switches that DPDK provides convenient library calls to do it. For the final
+step, we need to prepare the packet for transmission to the various follower
+replicas. To do this we prepare a different packet containing an ethernet header
+for the targeted replica and ``chain'' the data packet to each of these
+headers. Each header is then separately handed off to the driver for
+transmission via the NIC, carrying the data packet with it by association.
 
@@
 \begin{figure}
+  \centering
+  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
   \caption{Cyclone Network Packet Layout}
@@
 
-TBD: talk about how we do not need a reliable network layer like TCP/IP.
+We note that by discrding stream oriented protocols the network layer is not
+longer reliable or ordered. This is not a problem for RAFT that is designed to
+tolerate network failures resulting in drops or reordering of network
+packets. However this rules out consensus protocols such as Zookeeper Atomic
+Broadcast (ZAB~\cite{zab}) that depend on a reliable stream-oriented network
+layer to function correctly.
+
+Finally, we turn our attention to the persistence step in
+Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
+before it is multicast out to follower replicas. Since the NVM is directly
+attached we do this by executing a cacheline flush ({\tt clflush}) instruction
+for every cacheline in the packet and the pointer in the pointer buffer to
+persist these via the memory bus. This is the one exception to our otherwise
+clean separation between the control and data plane - necessitated by the fact
+that at the moment NICs do not support persistent memory. This is not too
+onerous a burden because we can use the newly introduced {\tt
+  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
+efficiently flush to persistent memory without the overhead of the serialization
+normally introduced by {\tt clflush}. This allows one to hit full memory
+bandwidth on present generation platforms, a quantity in excess of 200 Gb/s per
+core, well above the near term speeds of network interface cards.
+
 \subsubsection{Batching}

### 2017-04-19 paper/paper.tex
Updates to paper.

@@
 packets from the NIC such that they are offset in the byte array by a
-configurable amount referred to as ``headroom''. We prepend RAFT related
-information specific to each log entry in the headroom and shift the start
-pointer appropriately - these operations are standard enough for software packet
-switches that DPDK provides convenient library calls to do it. For the final
-step, we need to prepare the packet for transmission to the various follower
-replicas. To do this we prepare a different packet containing an ethernet header
-for the targeted replica and ``chain'' the data packet to each of these
-headers. Each header is then separately handed off to the driver for
-transmission via the NIC, carrying the data packet with it by association.
+configurable amount referred to as ``headroom''. We strip off the existing
+network headers in the packet and prepend RAFT related information specific to
+each log entry in the headroom by shifting the start pointer appropriately -
+these operations are standard enough for software packet switches that DPDK
+provides convenient library calls to do it. For the final step, we need to
+prepare the packet for transmission to the various follower replicas. To do this
+we prepare a different packet containing an ethernet header for the targeted
+replica and ``chain'' the data packet to each of these headers. Each header is
+then separately handed off to the driver for transmission via the NIC, carrying
+the data packet with it by association.
 
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \caption{Batching}
+  \label{fig:batching}
+\end{figure}
+
 \subsubsection{Batching}
+Although the control plane code only runs operations related to the RAFT state
+machine it is still slow enough relative to the dataplane to add significant
+overhead for each packet. We tackle this problem by applying another classic
+software packet switching technique: batching - illustrated in
+Figure~\ref{fig:batching}. We use a burst receive call available in the DPDK
+userspace driver to receive a burst of client packets at a time. We then chain
+these packets together and treat them as a single log entry from the perspective
+of RAFT, amortizing the control plane overheads over the packets (at most 32 at
+a time due to current driver limitations).
+
+We note there that batching in Cyclone does not involve a latency-throughput
+tradeoff like in many other systems~\cite{ix-dataplane}. The batch receive call
+we use in DPDK returns immediately with whatever number of packets is available,
+including zero. We always flush the transmit buffer after every call to DPDK to
+transmit packets to replicas. Therefore, we never tradeoff latency for
+throughput when batching.
 

### 2017-04-20 core/flash_log.cpp
Don't cross 4KB boundaries.

@@ -141,6 +141,17 @@ int log_append(void *log_,
   else if(log->entries_on_active_page >= flashlog_hwm) {
     log_switch_page(log);
   }
+  unsigned long stop_byte = log->bytes_on_active_page + size - 1;
+  // Skip to next page if necessary
+  if(stop_byte/4096 != log->bytes_on_active_page/4096) {
+    if(bytes_left < 4096) {
+      log_switch_page(log);
+    }
+    else {
+      log->bytes_on_active_page = 
+	((log->bytes_on_active_page + 4096)/4096)*4096;
+    }
+  }
   char *buffer = log->log_pages[log->active_page].page;
   buffer = buffer + log->bytes_on_active_page;
   *(unsigned long *)buffer = size;

### 2017-04-20 paper/paper.tex
Updates to paper.

@@
 \subsection{Flash Log}
-TBD: talk about why we don't need group commit.
+Cyclone is designed to use only limited amounts of NVM and drains log entries
+down into a log placed on a flash SSD - that we term the flashlog. The flashlog
+is written out in segments of configurable size (we use 128KB segments). A
+segment buffer is prepared in memory (volatile DRAM) using the layout shown in
+Figure~\ref{fig:flashlog_page}. We do not allow objects in the flashlog to cross
+a 4KB boundary - linking multiple objects together with a special flag encoded
+into the size if necessary. We flush log segments out to the log file using
+asynchronous direct IO, and therefore we fill one log segment buffer while
+keeping IO to another one outstanding to the flash drive. To avoid having to do
+a synchronous metadata flush, we preallocate (using the posix\_fallocate call) a
+gigabyte worth of zero filled disk pages at the end of a file, at a time.
+
+In order to recover from a crash, we make two important assumptions about the
+underlying SSD. First, we assume that 4KB is the \emph{minimum} atomic unit for
+updating pages on the SSD even under power failure i.e. there are no shorn
+writes on a 4KB page~\cite{shorn_writes}. We also assume that the SSD has power
+loss data protection meaning that writes cached in the drive's volatile cache
+are written to the SSD using a backup capacitor in the event of power failure -
+a property of many data center class SSDs today, including the Intel DC P3600
+SSD we use in our evaluation. Together these two assumptions mean that we can
+recover a consistent prefix of the log on a power failure.
+
+We move log entries from the head of the NVM log to the flashlog buffers in FIFO
+order. THe NVM log entry is only actually removed then the IO for the
+corresponding flashlog page is complete. This means that during recovery we can
+have the same log entry both in the NVM log and the flashlog, a condition that
+can be detected by examining the RAFT related information that is embedded as
+part of the logged packet.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.3]{figures2/flashlog_page.pdf}
+  \caption{Flashlog segment}
+  \label{fig:flashlog_page}
+\end{figure}
+
+The fact that we can immedately absorb updates into an NVM log and only move
+them to the flashlog in suitable units for block IO solves a common problem with
+logging to secondary storage - the need to do group commit. Logging systems
+often wait to collect a batch of entries to write to secondary storage trading
+latency for throughput. This is unnecessary in Cyclone as it
+effectively makes use of directly attached NVM with a two level log that removes
+the need to make this tradeoff.
+
 \section{Horizontal Scaling}

### 2017-04-20 paper/paper.tex
Updates to paper.

@@
 a property of many data center class SSDs today, including the Intel DC P3600
-SSD we use in our evaluation. Together these two assumptions mean that we can
-recover a consistent prefix of the log on a power failure.
+SSD~\cite{ssd_spec} we use in our evaluation. Together these two assumptions
+mean that we can recover a consistent prefix of the log on a power failure.
 
@@
 The fact that we can immedately absorb updates into an NVM log and only move
-them to the flashlog in suitable units for block IO solves a common problem with
-logging to secondary storage - the need to do group commit. Logging systems
+them to the flashlog in suitable units for block IO, sidesteps a common problem
+with logging to secondary storage - the need to do group commit. Logging systems
 often wait to collect a batch of entries to write to secondary storage trading
-latency for throughput. This is unnecessary in Cyclone as it
-effectively makes use of directly attached NVM with a two level log that removes
-the need to make this tradeoff.
+latency for throughput. This is unnecessary in Cyclone as it effectively makes
+use of directly attached NVM with a two level log that removes the need to make
+this tradeoff.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.3]{figures2/race.pdf}
+  \caption{Race with multiple physical logs}
+  \label{fig:race}
+\end{figure}
 
 \section{Horizontal Scaling}
+The basic principle in Cyclone is to turn log replication into a software packet
+switching problem. A key ingredient in software packet switching is applying
+multiple CPU cores to parallelize the control plane, scaling message processing
+throughput. In this section we describe how we can horizontally scale Cyclone's
+single \emph{logical} log across multiple \emph{physical} logs. For each
+physical log, we run an independent version of RAFT with the corresponding
+control plane mapped to a dedicated core. Further, as we shortly describe we
+assume a concurrent key value store where a set of CPU cores independently run
+operations from the physical logs on the shared memory key-value store data
+structure, synchronizing as necessary.
 
-\subsection{API Commutativity}
+We begin by describing how a key-value store API can be mapped on these
+independent physical logs without divergence in replica state. We then discuss
+how different pieces of work related to replication and the key value store
+itself are distributed across available CPU cores and available NIC queues.
+
+\subsection{Leveraging API Commutativity}
+Update requests to a key-value store consist of a key and an associated operation -
+such as ``put'' that set the value for the key or ``merge'' that calls a
+user-defined operator to merge a new value for the key into the existing value.
+A naive approach that maps a key to a random physical log and subsequently to a
+random core on the system for execution runs into the determinism problem -
+replicas could diverge with respect to the final value assigned to a key on
+different replicas. We illustrate this problem with an example in
+Figure~\ref{fig:race} where two updates to the same key race with each other on
+different physical logs and execution cores to end up being applied in different
+orders on different replicas, leading to a divergence of state.
+
+The solution to this problem is observe that the key-value store is a concurrent
+data structure that ultimately admits a serialization of operations to
+it. Further, consecutive operations to different keys in the serial history can
+be interchanged, because operations to different keys are
+\emph{commutative}. The correct way to use different physical logs is therefore
+to simply hash the key to select the physical log \emph{and} the core on which
+the operation executes. This guarantees determinism as the serial history of
+two different replicas can be transformed into the same serial history by
+reordering operations to different keys.
+
+Further relaxations to the mapping scheme are possible when operations to the
+same key are also commutative. A good example is when key value stores are used
+to maintain counters. Different increment operations to the same counter are
+commutative if one is prepared to tolerate non-determinism in the sequence of
+updates seen by a reader with the counter reaching the same eventual value. This
+specific use-case, in fact, lead to the addition of the merge operation in
+Rocksdb, a key value store that we use for evaluation. Merge operations
+therefore could be dispatched to a random physical log with no harm. For this
+paper however, we stick on the stricter requirement that operations to the same
+key are not commutative.
 
 \subsection{Core Assignment}
+
+TBD: mention no linearizability for reads at the moment
 

### 2017-04-20 paper/paper.tex
Updates to paper.

@@
 independent instance of RAFT, thereby leveraging the scalability afforded by
-classic software packet switching when using multiple CPU cores. Cyclone is able
+classic software packet switching when using multiple threads. Cyclone is able
 to replicate millions of small updates a second using only commodity 10 gigabit
@@
 feature of most NICs today. The dataplane for the RAFT protocol is therefore
-separated from the control plane. The control plane now runs on a CPU core and
-takes care of checking adherence to the protocol: checking for correct term, log
-index and that the replica receiving the packet from the client is indeed the
-leader. In addition, the control plane is responsible for non common-case
-activities such as dealing with timeout and leader elections.
+separated from the control plane. The control plane now runs in a thread pinned
+to a dedicated CPU core and takes care of checking adherence to the protocol:
+checking for correct term, log index and that the replica receiving the packet
+from the client is indeed the leader. In addition, the control plane is
+responsible for non common-case activities such as dealing with timeout and
+leader elections.
 
@@
 physical log, we run an independent version of RAFT with the corresponding
-control plane mapped to a dedicated core. Further, as we shortly describe we
-assume a concurrent key value store where a set of CPU cores independently run
-operations from the physical logs on the shared memory key-value store data
-structure, synchronizing as necessary.
+control plane mapped to a dedicated CPU core. Further, we assume a concurrent
+key value store where a set of CPU cores independently run operations from the
+physical logs on the shared memory key-value store data structure, synchronizing
+as necessary.
 
-We begin by describing how a key-value store API can be mapped on these
-independent physical logs without divergence in replica state. We then discuss
-how different pieces of work related to replication and the key value store
-itself are distributed across available CPU cores and available NIC queues.
+We begin by showing how we divide CPU cores and NIC resources on the plaform across
+replication and key value store work, and show how a KV store request flows
+across them.  We then describe how a key-value store API can be mapped to
+independent physical logs without divergence in replica state. Finally, we
+describe special handling in Cyclone for multi-key operations.
+
+\subsection{Core Allocation}
+A key-value store integrated with Cyclone runs two types of threads. One type of
+thread does key-value store work. This includes servicing lookup and update
+requests and background threads that flush in-memory data structures to
+secondary storage and do maintainence work on secondary storage data structures
+such as compacting log-structured merge trees~\cite{lsmtree}. We use a dedicated
+thread per-physical log to run the corresponding control plane including the
+associated RAFT state machine. We use static assignment of threads to cores as
+recommended for packet switching libraries such as DPDK for the best possible
+performance. Therefore we use the term threads and cores interchangeably in this
+paper. Client requests are always received by the control plane core for
+the physical log the request is directed to. A response for the core is always
+sent by the KV store core that finally handles the request.
+
+An important resource for Cyclone is the NIC that must be shared among the
+control plane cores that receive and replicate requests and the KV store cores
+that send back responses. Most NICs provide independent transmit/receive queue
+pairs and we aim for synchronization free access to these to aid horizontal
+scalability with respect to the network resource. DPDK already provides fairly
+scalable access to NIC queues, with the queue number being a parameter to send
+and receive calls. All that remains for us is to map the queues to the control
+plane and KV store cores, keeping in mind that there may be multiple NICs on the
+system. We do this by iterating over all control plane cores followed by all KV
+store cores. We assign two queue pairs to each control plane core, one to
+receive requests from the client and the other transmit and receive replication
+protocol related messages. We assign one queue pair to each KV store core to
+send back responses to clients. We are careful to spread the different queue
+pairs for a core across the available NICs as far as possible. This is important
+due to the asymmetry between the different types of traffic. Queue pairs
+carrying replication traffic will likely be more busy that those carrying client
+request responses due to the amplification to multicast the same message to
+different replicas.
+
+The flow for an update operation in Cyclone is as follows. The network packet is
+received by the control plane for the targeted physical log. The packet is then
+immediately sent out for replication as described in the previous section. When
+responses from a majority quorum is received, the request is passed to the
+targeted KV store core for execution. The KV store core process the request and
+sends back the response to the client. We note that replication is pipelined and
+decoupled from execution and thus multiple updates requests are usually in
+flight for replication at a time.
+
+The flow for a read operation is simpler. It is still targeted to a physical log
+but instead to replicating it the control plane core simply passes it on
+straight away to the KV store core for execution and response.
+
+The control plane cores and KV store cores communicate through a one-way
+lock-free FIFO queue passing pointers to packets. We leverage the convenient
+ring library in DPDK for this.
+
+Finally an asynchronous source of work in the system is maintenance of the
+two-level log. We depart from a 1:1 mapping between the first level NVM log and
+the second level flashlog here. Each physical log consists of one NVM log
+managed by the corresponding control plane core and multiple flashlogs each of
+which is managed as an additional responsibility by a KV store core. We
+therefore enforce a many to one mapping between key value store cores and
+control plane cores - thereby partitioning KV store cores across control plane
+cores. A request picks a control plane core and any one of the associated KV
+store cores for execution. The request is logged in the NVM log of the control
+plane core and garbage collected once it is drained to the flashlog of the
+selected KV store core.
 
@@
 therefore could be dispatched to a random physical log with no harm. For this
-paper however, we stick on the stricter requirement that operations to the same
-key are not commutative.
+paper however, we stick to the stricter requirement that operations to the same
+key are not commutative. However, Cyclone's client side libraries leave it to
+the user to specify the exact physical log and key-value thread to use to enable
+more relaxed mappings when necessary.
 
-\subsection{Core Assignment}
+We note that commutativity as assumed by Cyclone is an \emph{API level
+  property}. There are no assumption as to how application cores synchronize to
+access the shared key-value state or the underlying data structures for holding
+key-value pairs.
 
-TBD: mention no linearizability for reads at the moment
-
-\subsection{NIC Queue Assignmnt}
+Finally, Cyclone provides seraializable reads but does not - at the moment -
+guarantee the ability to read one's own writes, repeatable reads or more
+generally, linearizable reads. The reason is that on a failover the new leader
+replica might still have update operations in its NVM logs that are yet to be
+applied to the key value store. However the original failed leader replica might
+have applied these updates and a client might have already observed them. Users
+who desire linearizable reads can issue a special API call to ensure that reads
+are also routed through the logs. Flushing the NVM logs on failover is a better
+solution that we are currently in the process of engineering into Cyclone.
 

### 2017-04-20 paper/paper.tex
Updates to paper.

@@
 \subsection{Ganged Operations}
+Key value stores (such as Rocksdb) usually provide a batched write operation
+that atomically updates a set of key-value pairs. A batched write is often a key
+primitive for building more sophisticated operations such as
+transactions. Batched writes require special handling in Cyclone due to the
+requirement that we route a write request by hashing the key to select both the
+control plane core (physical log) as well as KV store core. A batched write has
+multiple keys and therefore we need to synchronize the progress of the operation
+across multiple physical logs (and associated replication quorums) as well as KV
+store cores. In effect the serial history contains a grouped set of operation on
+multiple keys spanning multiple cores as a single atomic operation. All serial
+histories across replicas therefore continue to map to the same equivalent
+serial history modulo reordering of operations to different keys.
+
+A simple solution to this problem is to use a distributed transaction, using
+two-phase commit to couple the otherwise independent replication quorums and KV
+store cores. However this method has a number of problems. First, it introduces
+an additional network round trip that is normally not necessary for a key value
+store with a single shared memory image holding all the keys. Second, handling
+failures with distributed transactions would have cause an explosion in
+complexity in Cyclone, something we wished to avoid in the interests of
+practical deployment to independent end users who need to understand and 
+maintain Cyclone as part of their KV store infrastructure.
+
+Our solution instead is to use a technique we termed ganged operations, based on
+the observation that in the failure free case, executing a batched write across
+multiple control plane cores and KV store cores does not look very much
+different from a synchronous shared memory barrier. The challenge is to
+effectively manage failures in replication without causing participating cores
+to stop making progress. The first step is to reconfigure RAFTs leader election
+protocol to ensure that the leaders for all the physical logs are co-located on
+the same host. Clients always dispatch batched writes to a fixed control plane
+core (called the injection core) which is then responsible for serially
+forwarding the requests to the participating control plane cores (for
+corresponding physical logs). The injection core uses DPDKs packet cloning
+primitives to avoid dataplane work while forwarding copies of the same batched
+write packet to multiple control plane cores. The injection core also adds a
+``nonce'' - a unique timestamp to the packet. In addition the injection core
+adds a unique view number to the packet corresponding to the term numbers of all
+the participating RAFT instances (read from shared memory of the co-located leaders).
+We describe the remaining flow for ganged operations in terms of the pseudocode
+in Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce
+to later in this section. We also assume that a unique barrier is allocated in
+shared memory for each ganged operation. We discuss later how this is done
+without using dynamic allocation.
+
+In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. One
+replication is complete on all participating physical logs, the corresponding KV
+store cores execute {\tt event\_replicated\_ganged\_op}. The use a bitmask for
+the barrier setting the appropriate bit for themselves. Once all KV store cores
+have arrived at the barrier the operation is executed. A distinguished leader
+core (we use use the minumum numbered participating KV store core) is
+responsible for sending back the response to the client. Two types of failure
+need to be considered. First, one of the non-leader cores could fail. The
+remaning cores would never progress past the barrier. To solve this the leader
+core montors the remaning cores for their current view - which is the RAFT term
+of the last executed log entry. If the view has moved past the view in the
+packet, replication failed on that physical log. The leader then sets the bit
+for the core that will no longer participate in the barrier to aid progress (a
+technique we borrowed from lock-free data structures~\cite{lock-free}). It also
+sets a failed flag to signal failure to the other participating cores. The
+second failure case is that the leader core might not receive the operation if
+it fails during replication from corresponding physical log. This case is
+detected in a similar manner by the remaining cores by monitoring the published
+view from the leader core in shared memory.
+
+\begin{figure*}
+  \centering
+\begin{verbatim}
+    // Control plane core
+    event_recv_ganged_op(packet)
+    {
+      Begin replication of packet
+    }
+
+    // KV store core
+    event_replicated_ganged_op (packet, barrier)
+    {
+      if(leader KV core)
+        atomic set bit me in barrier.mask         
+        do
+         for each KV store core in packet
+          if core_public_data[core].view > packet.view
+           barrier.failed = true 
+           atomic set bit for core in barrier.mask
+        while barrrer.mask != mask of participating KV cores
+        if barrier.failed
+         Ganged replication failed. Send retry to client. return.
+        else       
+         execute operation
+         send response
+      else
+        wait until 
+         core_public_data[leader core].view > packet.view
+         OR core_public_data[me].barrier == mask of participating KV cores
+        if core_public_data[leader core].view > packet.view
+         Ganged replication failed. Send retry to client. return.
+        else
+         execute operation
+    }
+\end{verbatim}
+\caption{Ganged Operation}
+\label{fig:ganged_ops}
+\end{figure*}
 

### 2017-04-21 paper/paper.tex
More updates to paper.

@@
 \section{Introduction}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.6]{results2/problem.pdf}
+\caption{Rocksdb Write Ahead Logging Impact}
+\label{fig:wal_impact}
+\end{figure}
 
@@
 
+The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
+mapped to a unique barrier. We achieve this by using a fixed piece memory owned
+by the leader core to hold the barrier and write the nonce to it in order to
+indicate that the barrer is active. Non-leader cores watch for the nonce to know
+when to execute the ganged operation barrier in Figure~\ref{fig:ganged_ops},
+while also monitoring the leader's published view to detect the case where the
+ganged operation fails to replicate on the leader's physical log.
+
+Finally, we describe how we generate the nonce. The nonce is generated on the
+injector core by concatenating the ethernet MAC ID of the first NIC on the
+system with a 64 bit value that is the number of CPU timestamp counter cycles
+since epoch time (read from the real time clock at startup plus the number of
+cycles from the CPU {\tt rdtsc} instruction). The nonce can only be repeated if
+the same machine manages to fail and come back up in less time than the real
+time clock drift (controlled with NTP), a possiblity that we discount.
+
+Ganged operations possibly consititute the most complex part of Cyclone but the
+code weighs in at well under a couple of hundred lines of code. We believe that
+this complexity is worth paying to avoid distributed transactions.
+
 \section{Evaluation}

### 2017-04-21 paper/paper.tex
Updates to paper.

@@
 \section{Introduction}
+Persistent key value stores are an increasingly important component of
+datacenter scale storage services. Key value stores such as
+Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
+large amounts of effort on both the engineering and deployment fronts. These key
+value stores represent sophisticated in-memory data structures built around Log
+Structured Merge trees~\cite{lsmtree} and are heavily tuned to extract maximum
+performance from present and future generations of flash solid state drives
+(SSDs).
+
+These key value stores however have a hitherto ignored an important component -
+the write ahead log. A machine or storage failure leading to total loss or
+temporary unavailability of data is unacceptable in many internet scale services
+where high availabilty and revenue are often interconnected. Key value stores
+therefore usually incorporate support for a write-ahead log that if replicated
+and made durable for every appended update provides the necessary high
+availability. Unfortunately the write ahead is a performance achilles heel for
+these systems - rendering much of the work on improving performance of the LSM
+component moot. To illustrate the impact of the write ahead log, consider
+Figure~\ref{fig:problem}. The line marked Rocksdb shows the performance of
+Rocksdb without the write ahead log. Both persisting the write ahead log without
+replicating it - the line marked Rockdb/WAL as well simply replicating the log
+three ways using the RAFT~\cite{raft} protocol over TCP/IP without persisting it
+are at least an order of magnitude slower (observe the log scale on the
+x-axis). It is therefore no suprise that deployments of Rocksdb often turn off
+the write ahead log~\cite{samza}. On the other side of the spectrum, key value
+store researchers such as the FloDB~\cite{flodb} authors turn off the write
+ahead log to be able to showcase benefits of sophisticated extensions to LSM
+datastructures.
 
@@
 \caption{Rocksdb Write Ahead Logging Impact}
-\label{fig:wal_impact}
+\label{fig:problem}
 \end{figure}
 
+Cyclone is a high speed strongly consistent replicated write ahead logging
+service specialized for key value stores such as Rocksdb. Cyclone provides the
+order of magnitude better performance required to close the performance gap in
+Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
+\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the
+server - such as in the form of NVDIMMs~\cite{farm} or directly attached
+persistent memory~\cite{pmfs}.
+
+Cyclone makes use of a limited amount of attached NVM using two simple
+ideas. First it leverages the observation that with persistence being a first
+class property of a subset of directly addressable DRAM, replicating the same
+log entry from the leader to follower replicas is no different from multicast
+using software packet switching techniques. It directly leverages this
+observation to implement a version of RAFT the replicates the NVM log using
+classic packet switching techniques such as separation of control and
+dataplanes. Second, Cyclone uses a two level log structure where it drains
+entries from the NVM log in IO friendly batches to a log place on a flash
+drive. This allows Cyclone to use only a limited amount of NVM as current forms
+of NVM such as NVDIMMs are more expensive than DRAM. Our experiments use an NVM
+footprint of a mere 64 MB to achieve performance comparable to placing the
+entire log in NVM. This allows Cyclone to be used in deployments where the key
+value store in maintanied entirely in main memory with the log serving as the
+source for recovery in the event of failure, rather than a multi-level LSM tree
+structure.
+
+We discuss the two level structure focusing on the packet switching components
+in Section~\ref{sec:2level}. We then describe how we can horizontally scale this
+basic idea across multiple cores by leveraging the commutativity of key value
+store APIs across different keys in Section~\ref{sec:horizontal}. We then
+demonstrate in the evaluation in Section~\ref{sec:evaluation} that Cyclone can
+replicate millions of updates a second using only commodity 10 Gigabit ethernet
+thereby closing the performance gap between key value store performance and
+write ahead logging performance using only a small amount of directly attached
+NVM. A discussion of related work follows before we conclude.
+
 \section{Two Level Log}
+\label{sec:2level}
 A fundamental challenge in building an efficient replicated log is dealing with
@@
 \section{Horizontal Scaling}
+\label{sec:horizontal}
 The basic principle in Cyclone is to turn log replication into a software packet
@@
 \section{Evaluation}
+\label{sec:evaluation}
 We evaluate Cyclone on a 12 node cluster connected via a 10 GigE switch. Three

### 2017-04-21 paper/paper.tex
updates to paper.

@@
          core_public_data[leader core].view > packet.view
-         OR core_public_data[me].barrier == mask of participating KV cores
+         OR barrier.mask == mask of participating KV cores
         if core_public_data[leader core].view > packet.view
+         OR barrier.failed
          Ganged replication failed. Send retry to client. return.
@@
 \section{Related Work}
-Memory technology in datacenters is due to undergo a paradigm shift with the
-increased use of directly addressable non-volatile memory both in the form of
-battery backed non-volatile DIMMs~\cite{farm} as well as newer memory types such
-as 3D XPoint~\cite{pmfs, bpfs}. In anticipation, libraries that allow
-programmers to build durable data structures on non-volatile heaps have become
-available~\cite{mnemosyne, nvheaps, cdds} - an example of such a library that we
-have used in this paper is Intel's non volatile memory library
-(NVML)~\cite{nvml}. A missing piece however is a component to make such durable
-data structures highly available using replication across a commodity network
-such as ethernet. Existing solutions that offer good performance are dependent
-on high performance RDMA networking hardware, with a dependence on external
-services (such as zookeeper~\cite{zookeeper}) and application specific code for
-fault recovery~\cite{farm, htm}.  This is a problem for programmers using NVML
-where one cannot apriori guarantee the availability of such RDMA networking
-hardware (for e.g., when running on externally hosted environments or public
-clouds).
+
+Cyclone takes a software only approach to improving replication performance over
+commodity network hardware. At the other end of the spectrum, Consensus in a
+box~\cite{consensus_box} implements the zookeeper atomic broadcast protocol
+together with a NIC on an FPGA. That work ignored durability focusing purely on
+replication performance. Latency numbers are impressive from the FPGA that cuts
+out the path from CPU caches to the NIC are as low as 3us for a round trip from
+leader to follower replica compared to the 7us we observe with DPDK. Notably
+however they reported 7us when using TCP/IP from the FPGA rather than their own
+connection oriented network protocol. On the other hand peak replication
+throughput appears to be 6M replications/sec on 66 Gbps of aggregate network
+bandwith, comparable to the mark obtained by Cyclone on an aggregate 40 Gbps of
+network bandwidth. We believe the general applicability afforded by our software
+only packes switching solution makes it attractive in comparison to an FPGA
+based solution even given the larger latency.
+
+Networked Ordered Paxos~\cite{nopaxos} reprograms top of rack switches to order
+client requests and multicast them to replicas - attacking replication latency
+by removing the need to direct requests through a leader replica. It is hard to
+compare performance due to the 1Gbps links used to leaf nodes in NoPaxos but we
+note that avoiding changes to switching infrastructure was a specific constraint
+in the design of Cyclone due to the difficulty we encountered in convincing
+public cloud customers to allow such changes to be made.
+
+{CRANE}~\cite{crane} is a system for replicating mutithreaded programs using
+Paxos. It uses deterministic multi-threading to ensure replicas converge to the
+same state as opposed to semantic equivalence in Cyclone using the commutativity
+of the key value store interface. Although CRANE is more generally applicable in
+that it makes no assumptions about the application API, deterministic
+multithreading comes with high overheads for applications that store large
+amounts of data - CRANE reports around a 2X slowdown for MySQL likely rendering
+it unsuitable as a replacement for the WAL in key value stores.
+
+A number of researchers have proposed using high performance networking hardware
+for improving the performance of various distributed system stacks such as
+distributed transactions~\cite{farm} and replication~\cite{dare, mojim}. We note
+that the ideas in Cyclone are not specific to ethernet. The same ideas of packet
+switching and horizontal scaling can be equally applied to RDMA. However, we
+chose commodity ethernet due to the short term deployment plans that are a
+priority design target.
+
+The scalable commutativity rule~\cite{scalable_commutativity} generalized the
+idea of commutativity in key value store interfaces that we have used for
+horizontal scaling in Cyclone. Conveniently it provides a tool for checking when
+API calls can commute in a history. It is a simple exercise to express a key
+value store in COMMUTER and check that operations to different keys
+commute. This provides a theoretical basis for horizontal scaling, something we
+have omitted for brevity from this paper. The scalable commutativity rule also
+suggests that scalable key value store implementations would exploit
+commutativity something we have observed in the case of in-memory layers for key
+value stores such as Flodb and indeed in the way we distributed keys across
+physical logs in Cyclone.
+
+Finally, we note the in progress transformation of durable memory in datacenters
+with the advents of 3DXPoint. Researchers have speculated on the impact of cheap
+and high volume directly attached 3DXPoint~\cite{tiering}. Memory vendors have
+also begun shipping software libraries such as NVML~\cite{nvml} for building durable
+in-memory data structures using undo logging. A durable key value store using
+3DXPoint would be an idea target for Cyclone because each NVM log entry can be
+discarded as soon as it is applied to the key value store, since the results are
+persistent. This would allow Cyclone to operate at pure packet switching speeds,
+a goal we hope to achieve in the future with a single level log for NVML
+applications.
 
 \section{Conclusion}
-Cyclone allows developers to add availability to key value maps built on top of
-libraries such as NVML and achieve good performance on commodity networking
-hardware. A key question is whether Cyclone can be used for other applications
-such as filesystems? \emph{We believe that the requirement for scaling across
-  multiple replication quorums can be satisfied by applications where the
-  scalable commutativity rule~\cite{scalable_commutativity} applies to common
-  case primitives.} Although our intitial release of Cyclone will be for key
-value style applications, our on-going research is based on the scalable
-commutativity hypothesis.
+
 \newcommand\myurl[2]{\url{#1}}

### 2017-04-21 paper/paper.tex
Updates to paper.

@@
 \usepackage{times,epsfig,endnotes,graphicx, verbatim}
-\usepackage{datetime, url, pdfpages}
+\usepackage{datetime, url, pdfpages, parskip}
 
@@
 Persistent key value stores are rapidly becoming an important component of many
-distributed data serving solutions - with innovations targeted at taking
-advantage of growing flash speeds. Unfortunately their performance is hampered
-by the need to maintain and replicate a write ahead log to guarantee
-availability in the face of machine and storage failures. Cyclone offers a
-solution to this problem by leveraging a small amount of directly attached
-non-volatile memory (NVM) such as NVDIMMs to maintain a two level log - the upper
-level in a small amount of NVM draining into a larger lower level log
-maintained on flash. Replicating the smaller log kept in NVMs transforms the
-replication problem into a packet switching problem - the leader sends the exact
-same packet to all follower replicas and we solve the problem as such by using
-the RAFT protocol implemented as a software multicast module on the Data Plane
-Development Kit (DPDK). Second, we levergage the observation that key value store
-interfaces are commutative for operations to different keys to scale our two
-level log horizontally into a number of physical two level logs - with a hash of
-the key being used to choose the right log. Each log is replicated by an
-independent instance of RAFT, thereby leveraging the scalability afforded by
-classic software packet switching when using multiple threads. Cyclone is able
-to replicate millions of small updates a second using only commodity 10 gigabit
-ethernet adapters and improves the performance of RocksDB - a popular persistent
-key value store by orders of magnitude when compared to its own write ahead log
-without replication.
+distributed data serving solutions with innovations targeted at taking advantage
+of growing flash speeds. Unfortunately their performance is hampered by the need
+to maintain and replicate a write ahead log to guarantee availability in the
+face of machine and storage failures. Cyclone offers a solution to this problem
+by leveraging a small amount of directly attached non-volatile memory (NVM) such
+as NVDIMMs to maintain a two level log: the upper level in a small amount of NVM
+draining into a larger lower level log maintained on flash. Replicating the
+smaller log kept in NVMs transforms the replication problem into a packet
+switching problem as the leader sends the exact same packet to all follower
+replicas and we solve the problem as such by using the RAFT protocol implemented
+as a software multicast module on the Data Plane Development Kit (DPDK). Second,
+we leverage the observation that key value store interfaces are commutative for
+operations to different keys to scale our two level log horizontally into a
+number of physical two level logs - with a hash of the key being used to choose
+the physical log. Each log is replicated by an independent instance of RAFT,
+thereby leveraging the scalability afforded by classic software packet switching
+when using multiple CPU cores. Cyclone is able to replicate millions of small
+updates a second using only commodity 10 gigabit ethernet adapters and improves
+the performance and availability of RocksDB, a popular persistent key value
+store by an order of magnitude when compared to its own write ahead log without
+replication.
 \end{abstract}
@@
 Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
-large amounts of effort on both the engineering and deployment fronts. These key
-value stores represent sophisticated in-memory data structures built around Log
+large amounts of effort on both the engineering and research fronts. These key
+value stores include sophisticated in-memory data structures built around Log
 Structured Merge trees~\cite{lsmtree} and are heavily tuned to extract maximum
-performance from present and future generations of flash solid state drives
-(SSDs).
+performance from flash-based solid state drives (SSDs).
 
-These key value stores however have a hitherto ignored an important component -
-the write ahead log. A machine or storage failure leading to total loss or
-temporary unavailability of data is unacceptable in many internet scale services
-where high availabilty and revenue are often interconnected. Key value stores
-therefore usually incorporate support for a write-ahead log that if replicated
-and made durable for every appended update provides the necessary high
+These key value stores however tend to ignore an important component: the write
+ahead log. A machine or storage failure leading to total loss or temporary
+unavailability of data is unacceptable in many internet scale services where
+high availabilty and revenue are often interconnected. Key value stores
+therefore usually incorporate support for a write ahead log that if replicated
+and kept durable for every appended update provides the necessary high
 availability. Unfortunately the write ahead is a performance achilles heel for
-these systems - rendering much of the work on improving performance of the LSM
-component moot. To illustrate the impact of the write ahead log, consider
+these systems - eclipsing much of the work on improving performance of the LSM
+component. To illustrate the impact of the write ahead log, consider
 Figure~\ref{fig:problem}. The line marked Rocksdb shows the performance of
-Rocksdb without the write ahead log. Both persisting the write ahead log without
-replicating it - the line marked Rockdb/WAL as well simply replicating the log
-three ways using the RAFT~\cite{raft} protocol over TCP/IP without persisting it
-are at least an order of magnitude slower (observe the log scale on the
+Rocksdb without the write ahead log. The performance when persisting the write
+ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
+line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
+without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
+persisting every update to the log or replicating it using TCP/IP causes
+performance to drop by an order of magnitude (observe the log scale on the
 x-axis). It is therefore no suprise that deployments of Rocksdb often turn off
 the write ahead log~\cite{samza}. On the other side of the spectrum, key value
-store researchers such as the FloDB~\cite{flodb} authors turn off the write
-ahead log to be able to showcase benefits of sophisticated extensions to LSM
-datastructures.
+store research such as FloDB~\cite{flodb} turns off the write ahead log to be
+able to showcase benefits of sophisticated extensions to LSM datastructures.
 
@@
 \emph{a small amount of} directly attached Non Volatile Memory (NVM) on the
-server - such as in the form of NVDIMMs~\cite{farm} or directly attached
+host - such as in the form of NVDIMMs~\cite{farm} or directly attached
 persistent memory~\cite{pmfs}.
 
-Cyclone makes use of a limited amount of attached NVM using two simple
-ideas. First it leverages the observation that with persistence being a first
-class property of a subset of directly addressable DRAM, replicating the same
-log entry from the leader to follower replicas is no different from multicast
-using software packet switching techniques. It directly leverages this
-observation to implement a version of RAFT the replicates the NVM log using
-classic packet switching techniques such as separation of control and
-dataplanes. Second, Cyclone uses a two level log structure where it drains
-entries from the NVM log in IO friendly batches to a log place on a flash
-drive. This allows Cyclone to use only a limited amount of NVM as current forms
-of NVM such as NVDIMMs are more expensive than DRAM. Our experiments use an NVM
-footprint of a mere 64 MB to achieve performance comparable to placing the
-entire log in NVM. This allows Cyclone to be used in deployments where the key
-value store in maintanied entirely in main memory with the log serving as the
-source for recovery in the event of failure, rather than a multi-level LSM tree
-structure.
+Cyclone makes use of a limited amount of attached NVM using two ideas. First, it
+leverages the observation that with persistence being a first class property of
+a subset of directly addressable DRAM, replicating the same log entry from the
+leader to follower replicas is no different from multicast using software packet
+switching techniques. It directly leverages this observation to implement a
+version of RAFT that replicates the NVM log using classic packet switching
+techniques such as separation of control and data planes. Second, Cyclone uses a
+two level log structure where it drains entries from the NVM log in IO friendly
+batches to a log placed on a flash drive. This allows Cyclone to use only a
+limited amount of NVM reasoning that current forms of NVM such as NVDIMMs are
+more expensive than DRAM. Our experiments use an NVM footprint of a mere 64 MB
+to achieve performance comparable to placing the entire log in NVM. This allows
+Cyclone to be used in deployments where the key value store is maintained
+entirely in main memory rather than as a multi-level LSM tree structure, with
+the log serving as the source for recovery in the event of failure.
 
@@
 replicate millions of updates a second using only commodity 10 Gigabit ethernet
-thereby closing the performance gap between key value store performance and
-write ahead logging performance using only a small amount of directly attached
-NVM. A discussion of related work follows before we conclude.
+adapters thereby closing the performance gap between key value store performance
+and write ahead logging performance using only a small amount of directly
+attached NVM. A discussion of related work follows before we conclude.
 
@@
 A fundamental challenge in building an efficient replicated log is dealing with
-network overheads. Cyclone'e construction is predicated on the observation that
-in the common case protocols such as RAFT simply multicast the same entry
+network overheads. Cyclone's construction is predicated on the observation that
+in the common case, protocols such as RAFT simply multicast the same entry
 received from a client, from the leader replica to all follower replicas. This
@@
 such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
-level log as the flash log. We periodically drain entries in batches from the
-NVM log to the backing flashlog. An important and beneficial side effect of this
-two level approach is that we need only limited amounts of NVM with the bulk of
-the log residing on flash - an arrangement that maps well to the relative cost
-of the two types of memory. We begin by describing the NVM log and how we
-implement the RAFT consensus protocol in the packet switching paradigm around
-it. We then describe the flashlog and how we drain entries in batches from the
-NVM level to the flash level.
+level log as the flashlog. We periodically drain entries in batches from the NVM
+log to the backing flashlog. An important consequence of this two level approach
+is that we need only a limited amounts of NVM with the bulk of the log residing
+on flash - an arrangement that maps well to the current relative costs of the
+two types of memory. We begin by describing the NVM log and how we implement the
+RAFT consensus protocol in the packet switching paradigm around it. We then
+describe the flashlog and how we drain entries in batches from the NVM level to
+the flash level.
 
@@
 log) to memory buffers - each of which holds a client request. This arrangement
-is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of
-such an arragement. Adding a level of indirection in the NVM log allows us to
-isolate the circular log being manipulated by the RAFT protocol from the
-problems of variable sized client requests that can be scattered across
-non-volatile memory depending on the internals of the memory allocator of the
-packets switching library we use. Another advantage is that it makes recovery
-from NVM easy - appends to the circular pointer log are atomic and we use the
-pointer log to recover allocator state i.e. what pieces of NVM are currently in
-use by the log.
+is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of such
+an arragement. Adding a level of indirection in the NVM log allows us to
+separate the FIFO ordered circular log being manipulated by the RAFT from
+variable sized client being managed by the memory allocator of the packet
+switching library we use. Another advantage is that it makes recovery from NVM
+easy - appends to the circular pointer log are atomic and we can use the pointer
+log to recover allocator state i.e. what pieces of NVM are currently in use by
+the log.
 
@@
 NVM log across a set of replicas. The software packet switching idea we apply is
-to separate the control plane from the dataplane using the Data Plane
+to separate the control plane from the data plane using the Data Plane
 Development Kit (DPDK~\cite{dpdk}) library for direct userspace access to the
@@
 RAFT incorporates a leader election algorithm that elects a leader with the most
-upto date log. The leader receives log entries from clients, appends them to the
+up to date log. The leader receives log entries from clients, appends them to the
 persistent log (NVM log in this case) and multicasts the new entry to follower
@@
 We implement the common case flow in RAFT as a software packet switch using
-DPDK. The dataplane now encompasses the NIC and the CPU caches with the NIC
-placing packet data directly in the CPU cache via DDIO~\cite{ddio} - a standard
-feature of most NICs today. The dataplane for the RAFT protocol is therefore
-separated from the control plane. The control plane now runs in a thread pinned
-to a dedicated CPU core and takes care of checking adherence to the protocol:
-checking for correct term, log index and that the replica receiving the packet
-from the client is indeed the leader. In addition, the control plane is
-responsible for non common-case activities such as dealing with timeout and
-leader elections.
+DPDK. The dataplane now encompasses the Network Interface Card (NIC) and the CPU
+caches with the NIC placing packet data directly in the CPU cache via
+DDIO~\cite{ddio} - a standard feature of most NICs today. The dataplane for the
+RAFT protocol is therefore separated from the control plane. The control plane
+runs in a thread pinned to a dedicated CPU core and takes care of the protocol
+state machine: checking for correct term, log index and that the replica
+receiving the packet from the client is indeed the leader. On the follower side
+it is responsible for checking that the leader is in the correct term before
+accepting the packet by appending it to the NVM log and initiating a
+response. In addition, the control plane is responsible for uncommon case
+activities such as dealing with timeouts and leader elections.
 
-The pseudocode in Figure~\ref{fig:control_plane} describes the control plane in
-Cyclone organized as event handlers triggered on receiving a packet. We focus on
-two events for illustration (and brevity): the event where a request is received
-from a client and the event where a leader election related message is
-received. A key concern when we built Cyclone was adhering as far as possible to
-the software packet switching principle of separating the control plane from the
-dataplane. This means that the CPU should avoid examining every byte of the
-packet as far as possible.
+The pseudocode in Figure~\ref{fig:control_plane} describes part of the control
+plane in Cyclone organized as event handlers triggered on receiving a packet at
+the leader. We focus on one event for illustration (and brevity): the event
+where a request is received from a client. A key concern when we built Cyclone
+was adhering as far as possible to the software packet switching principle of
+separating the control plane from the dataplane. This means that the CPU should
+avoid examining every byte of the packet as far as possible.
 
-If an election message is received all the code related to handling
-the election related message is run on the CPU core and all contents of the
-packet are examined to decide what to do in accordance with the RAFT protocol
-state machine. This is an exception to the software packet switching principle
-but one that we found acceptable since leader elections are rare compared to the
-common case of replicating client packets.
-
-The sequence of operations when a packet is received from the client however
-needs careful design for efficient software packet switching. To illustrate how
-this is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates
-packet layouts across the two steps of prepending a RAFT header and transmitting
-to follower replicas. DPDK describes packets using an ``mbuf'' data
+The sequence of operations when a packet is received from the client needs
+careful design for efficient software packet switching. To illustrate how this
+is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates packet
+layouts across the two steps of prepending a RAFT header and transmitting to
+follower replicas. DPDK describes packets using an ``mbuf'' data
 structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
-actually containing the byte and a fixed size piece of external metadata that
+actually containing the packet and a fixed size piece of external metadata that
 describes various aspects of the packet, most crucially a pointer to the start
@@
 network headers in the packet and prepend RAFT related information specific to
-each log entry in the headroom by shifting the start pointer appropriately -
-these operations are standard enough for software packet switches that DPDK
-provides convenient library calls to do it. For the final step, we need to
-prepare the packet for transmission to the various follower replicas. To do this
-we prepare a different packet containing an ethernet header for the targeted
-replica and ``chain'' the data packet to each of these headers. Each header is
-then separately handed off to the driver for transmission via the NIC, carrying
-the data packet with it by association.
+each log entry in the headroom by shifting the start pointer appropriately.
+These operations are standard enough for software packet switches that DPDK
+provides convenient library calls for it. For the final step, we need to prepare
+the packet for transmission to the various follower replicas. To do this we
+prepare a different packet containing an ethernet header for \emph{each}
+targeted replica and ``chain'' the data packet to each of these headers. Each
+header is then separately handed off to the driver for transmission via the NIC,
+carrying the data packet with it by association.
 
@@
     }
-
-    event_receive_election_msg()
-    {
-      Handle_election_msg()
-    }
 \end{verbatim}
@@
 
-We note that by discrding stream oriented protocols the network layer is not
-longer reliable or ordered. This is not a problem for RAFT that is designed to
-tolerate network failures resulting in drops or reordering of network
-packets. However this rules out consensus protocols such as Zookeeper Atomic
-Broadcast (ZAB~\cite{zab}) that depend on a reliable stream-oriented network
-layer to function correctly.
+We note that by discrding stream oriented protocols such as TCP, the network
+layer is not longer reliable or ordered. This is not a problem for RAFT that is
+designed to tolerate network failures resulting in drops or reordering of
+network packets. However, this rules out consensus protocols such as Zookeeper
+Atomic Broadcast (ZAB~\cite{zab}) that depend on a reliable stream-oriented
+network layer to function correctly.
 
@@
 
-We note there that batching in Cyclone does not involve a latency-throughput
+We note that batching in Cyclone does not involve a latency-throughput
 tradeoff like in many other systems~\cite{ix-dataplane}. The batch receive call
@@
 a synchronous metadata flush, we preallocate (using the posix\_fallocate call) a
-gigabyte worth of zero filled disk pages at the end of a file, at a time.
+gigabyte worth of zero filled disk pages at the end of the log file whenever we
+hit its end.
 
@@
 We move log entries from the head of the NVM log to the flashlog buffers in FIFO
-order. THe NVM log entry is only actually removed then the IO for the
+order. The NVM log entry is only actually removed when the IO for the
 corresponding flashlog page is complete. This means that during recovery we can
@@
 The fact that we can immedately absorb updates into an NVM log and only move
-them to the flashlog in suitable units for block IO, sidesteps a common problem
+them to the flashlog in suitable units for block IO sidesteps a common problem
 with logging to secondary storage - the need to do group commit. Logging systems
 often wait to collect a batch of entries to write to secondary storage trading
-latency for throughput. This is unnecessary in Cyclone as it effectively makes
-use of directly attached NVM with a two level log that removes the need to make
-this tradeoff.
+latency for throughput. This is unnecessary in Cyclone as the use of directly
+attached NVM with a two level log removes the need to make this tradeoff.
 
@@
 multiple CPU cores to parallelize the control plane, scaling message processing
-throughput. In this section we describe how we can horizontally scale Cyclone's
+throughput. In this section, we describe how we can horizontally scale Cyclone's
 single \emph{logical} log across multiple \emph{physical} logs. For each
@@
 physical logs on the shared memory key-value store data structure, synchronizing
-as necessary.
+among themselves as necessary.
 

### 2017-04-21 paper/paper.tex
Updates to paper.

@@
 \subsection{Core Allocation}
-A key-value store integrated with Cyclone runs two types of threads. One type of
-thread does key-value store work. This includes servicing lookup and update
-requests and background threads that flush in-memory data structures to
+A key-value store integrated with Cyclone runs two types of threads. The first
+type of thread does key-value store work. This includes servicing lookup and
+update requests and background threads that flush in-memory data structures to
 secondary storage and do maintainence work on secondary storage data structures
-such as compacting log-structured merge trees~\cite{lsmtree}. We use a dedicated
-thread per-physical log to run the corresponding control plane including the
-associated RAFT state machine. We use static assignment of threads to cores as
-recommended for packet switching libraries such as DPDK for the best possible
-performance. Therefore we use the term threads and cores interchangeably in this
-paper. Client requests are always received by the control plane core for
-the physical log the request is directed to. A response for the core is always
-sent by the KV store core that finally handles the request.
+such as compacting log-structured merge trees~\cite{lsmtree}. The second type of
+thread is a dedicated thread per-physical log to run the corresponding control
+plane including the associated RAFT state machine. We use static assignment of
+threads to cores as recommended for packet switching libraries such as DPDK for
+the best possible performance. Therefore we use the terms thread and core
+interchangeably in this paper. Client requests are always received by the
+control plane core for the physical log the request is directed to. A response
+for the core is always sent by the KV store core that finally handles the
+request.
 
@@
 store cores. We assign two queue pairs to each control plane core, one to
-receive requests from the client and the other transmit and receive replication
-protocol related messages. We assign one queue pair to each KV store core to
-send back responses to clients. We are careful to spread the different queue
-pairs for a core across the available NICs as far as possible. This is important
-due to the asymmetry between the different types of traffic. Queue pairs
-carrying replication traffic will likely be more busy that those carrying client
-request responses due to the amplification to multicast the same message to
-different replicas.
+receive requests from the client and the other to transmit and receive
+replication protocol related messages. We assign one queue pair to each KV store
+core to send back responses to clients. We are careful to spread the different
+queue pairs for a core across the available NICs as far as possible. This is
+important due to the asymmetry between the different types of traffic. Queue
+pairs carrying replication traffic will likely be more busy than those carrying
+client request responses due to the amplification in message count to multicast
+the same message to different replicas.
 
@@
 sends back the response to the client. We note that replication is pipelined and
-decoupled from execution and thus multiple updates requests are usually in
-flight for replication at a time.
+decoupled from execution and thus multiple updates are usually in
+flight on the network for replication at the same time.
 
 The flow for a read operation is simpler. It is still targeted to a physical log
-but instead to replicating it the control plane core simply passes it on
+but instead of replicating it, the control plane core simply passes it on
 straight away to the KV store core for execution and response.
 
-The control plane cores and KV store cores communicate through a one-way
+The control plane cores communicate with the KV store cores through a one-way
 lock-free FIFO queue passing pointers to packets. We leverage the convenient
@@
 
-Finally an asynchronous source of work in the system is maintenance of the
+Finally, an asynchronous source of work in the system is maintenance of the
 two-level log. We depart from a 1:1 mapping between the first level NVM log and
@@
 which is managed as an additional responsibility by a KV store core. We
-therefore enforce a many to one mapping between key value store cores and
-control plane cores - thereby partitioning KV store cores across control plane
-cores. A request picks a control plane core and any one of the associated KV
-store cores for execution. The request is logged in the NVM log of the control
-plane core and garbage collected once it is drained to the flashlog of the
-selected KV store core.
+therefore enforce a many to one mapping between key value store cores and a
+front end control plane core - thereby partitioning KV store cores across
+control plane cores. A request picks a control plane core and any one of the
+associated KV store cores for execution. The request is logged in the NVM log of
+the control plane core and garbage collected once it is drained to the flashlog
+of the selected KV store core.
 
 \subsection{Leveraging API Commutativity}
-Update requests to a key-value store consist of a key and an associated operation -
-such as ``put'' that set the value for the key or ``merge'' that calls a
+Update requests to a key-value store include a key and an associated operation -
+such as ``put'' that sets the value for the key or ``merge'' that calls a
 user-defined operator to merge a new value for the key into the existing value.
@@
 two different replicas can be transformed into the same serial history by
-reordering operations to different keys.
+reordering operations to different keys. Operations to the same key are already
+in the same order in the serial histories by virtue of being replicated on the
+same physical log and being executed on the same KV store core.
 
@@
 therefore could be dispatched to a random physical log with no harm. For this
-paper however, we stick to the stricter requirement that operations to the same
+paper however, we stick to the stricter case that operations to the same
 key are not commutative. However, Cyclone's client side libraries leave it to
-the user to specify the exact physical log and key-value thread to use to enable
-more relaxed mappings when necessary.
+the user to specify the exact physical log and key-value store core to use in
+order to enable more relaxed mappings when necessary.
 
@@
 access the shared key-value state or the underlying data structures for holding
-key-value pairs.
+key-value pairs. We emphasize that this means replicas ultimately have the same
+content but not necessarily the same memory state.
 
@@
 across multiple physical logs (and associated replication quorums) as well as KV
-store cores. In effect the serial history contains a grouped set of operation on
-multiple keys spanning multiple cores as a single atomic operation. All serial
-histories across replicas therefore continue to map to the same equivalent
-serial history modulo reordering of operations to different keys.
+store cores. In effect, the serial history contains a grouped set of operation
+on multiple keys as a single atomic operation. All serial histories across
+replicas therefore continue to map to the same equivalent serial history modulo
+reordering of operations to different keys.
 
-A simple solution to this problem is to use a distributed transaction, using
+One way to do this is to use a distributed transaction, using
 two-phase commit to couple the otherwise independent replication quorums and KV
@@
 store with a single shared memory image holding all the keys. Second, handling
-failures with distributed transactions would have cause an explosion in
+failures with distributed transactions would have caused an explosion in
 complexity in Cyclone, something we wished to avoid in the interests of
 practical deployment to independent end users who need to understand and 
-maintain Cyclone as part of their KV store infrastructure.
+maintain Cyclone as part of their embedded KV store infrastructure.
 
-Our solution instead is to use a technique we termed ganged operations, based on
-the observation that in the failure free case, executing a batched write across
-multiple control plane cores and KV store cores does not look very much
-different from a synchronous shared memory barrier. The challenge is to
+Our solution instead is to use a technique we termed ganged operations. It is
+based on the observation that in the failure free case, executing a batched
+write across multiple control plane cores and KV store cores does not look very
+much different from a synchronous shared memory barrier. The challenge is to
 effectively manage failures in replication without causing participating cores
@@
 the same host. Clients always dispatch batched writes to a fixed control plane
-core (called the injection core) which is then responsible for serially
-forwarding the requests to the participating control plane cores (for
-corresponding physical logs). The injection core uses DPDKs packet cloning
-primitives to avoid dataplane work while forwarding copies of the same batched
-write packet to multiple control plane cores. The injection core also adds a
-``nonce'' - a unique timestamp to the packet. In addition the injection core
-adds a unique view number to the packet corresponding to the term numbers of all
-the participating RAFT instances (read from shared memory of the co-located leaders).
-We describe the remaining flow for ganged operations in terms of the pseudocode
-in Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce
-to later in this section. We also assume that a unique barrier is allocated in
+core (called the injection core) which is then responsible for forwarding the
+request to the participating control plane cores (for corresponding physical
+logs). The injection core uses DPDKs packet cloning primitives to avoid
+dataplane work while forwarding copies of the same batched write packet to
+multiple control plane cores. The injection core also adds a ``nonce'' - a
+unique timestamp to the packet. In addition, the injection core adds a unique
+view number to the packet containing the term numbers of all the participating
+RAFT instances (read from shared memory of the co-located leaders).  We describe
+the remaining flow for ganged operations in terms of the pseudocode in
+Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce to
+later in this section. We also assume that a unique barrier is allocated in
 shared memory for each ganged operation. We discuss later how this is done
@@
 
-In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. One
+In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. Once
 replication is complete on all participating physical logs, the corresponding KV
-store cores execute {\tt event\_replicated\_ganged\_op}. The use a bitmask for
+store cores execute {\tt event\_replicated\_ganged\_op}. They use a bitmask for
 the barrier setting the appropriate bit for themselves. Once all KV store cores
 have arrived at the barrier the operation is executed. A distinguished leader
-core (we use use the minumum numbered participating KV store core) is
+core (we use the minumum numbered participating KV store core) is
 responsible for sending back the response to the client. Two types of failure
-need to be considered. First, one of the non-leader cores could fail. The
+need to be considered. First, replication on the physical log corresponding to
+one of the non-leader cores could fail. The
 remaning cores would never progress past the barrier. To solve this the leader
@@
 detected in a similar manner by the remaining cores by monitoring the published
-view from the leader core in shared memory.
+view from the leader core in shared memory. We add that on a leadership change
+due to timeout we enqueue and execute a nop on the physical log and all
+associated KV store cores to ensure previous entries are committed - a corner
+case in the RAFT protocol.
 
@@
         wait until 
-         core_public_data[leader core].view > packet.view
-         OR barrier.mask == mask of participating KV cores
-        if core_public_data[leader core].view > packet.view
-         OR barrier.failed
+         core_public_data[leader core].view > packet.view OR
+         barrier.mask == mask of participating KV cores
+        if core_public_data[leader core].view > packet.view OR
+           barrier.failed
          Ganged replication failed. Send retry to client. return.
@@
 The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
-mapped to a unique barrier. We achieve this by using a fixed piece memory owned
-by the leader core to hold the barrier and write the nonce to it in order to
-indicate that the barrer is active. Non-leader cores watch for the nonce to know
-when to execute the ganged operation barrier in Figure~\ref{fig:ganged_ops},
-while also monitoring the leader's published view to detect the case where the
-ganged operation fails to replicate on the leader's physical log.
+mapped to a unique barrier. We achieve this by using a fixed piece of memory
+owned by the leader core to hold the barrier and write the nonce to it in order
+to indicate that the barrer is active for the corresponding batched write
+operation. Non-leader cores watch for the nonce to know when to execute the
+ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
+the leader's published view to detect the case where the ganged operation fails
+to replicate on the leader's physical log.
 
@@
 the same machine manages to fail and come back up in less time than the real
-time clock drift (controlled with NTP), a possiblity that we discount.
+time clock drift (controlled with NTP), a possibility that we discount.
 
 Ganged operations possibly consititute the most complex part of Cyclone but the
-code weighs in at well under a couple of hundred lines of code. We believe that
-this complexity is worth paying to avoid distributed transactions.
+code weighs in at well under a couple of hundred lines. We believe that
+this additional complexity is still small compared to distributed transactions.
 

### 2017-04-21 paper/paper.tex
Updates to paper.

@@
 % Paper number and no. of pages as author
-\authorinfo{Paper \textbf{\#224}}{NN pages}
+\authorinfo{Paper \textbf{\#224}}{13 pages}
 
@@
 replication is complete on all participating physical logs, the corresponding KV
-store cores execute {\tt event\_replicated\_ganged\_op}. They use a bitmask for
+store cores execute \\{\tt event\_replicated\_ganged\_op}. They use a bitmask for
 the barrier setting the appropriate bit for themselves. Once all KV store cores
@@
 \label{sec:evaluation}
-We evaluate Cyclone on a 12 node cluster connected via a 10 GigE switch. Three
-of the machines are equipped with 1.6TB Intel DC P3600 SSDs and 4*10 GigE
-ports. The remaining nine machines do not have SSDs and have only one 10 GigE
-port, serving as clients for most of the experiments. As with other
-work~\cite{faast}, we use DRAM on the machines to proxy for NVDIMMs where
-necessary - the persistent memory needed never exceeds 64 MB regardless of the
-size of the key value store or second level log on flash. We divide the
-evaluation into three parts. First, we evaluate Cyclone's performance with a
-single level log as a pure software packet switch. Next, we evaluate performance
-when adding a second level of log on flash. Finally, we evaluate performance
-when integrated with Rocksdb as an alternative to Rocksdb's write ahead
-log. Unless otherwise mentioned, we use a 60 byte header followed by a payload
-for experiments. We log both the header and payload. In all cases the server
-echoes the received entry (header and payload) back to the client. We fix the
-number of application threads at 32 using at most 8 threads for running RAFT
-instances.
+We evaluate Cyclone on a 12 node x86 Xeon cluster connected via a 10 GigE
+switch. Three of the machines are equipped with 1.6TB Intel DC P3600 SSDs and
+4*10 GigE ports. The remaining nine machines do not have SSDs and have only one
+10 GigE port, serving as clients for most of the experiments. We turn on jumbo
+frame support in the 10 GigE switch to enable maximum use of batching in
+Cyclone. As with other work~\cite{faast}, we use DRAM on the machines to proxy
+for NVDIMMs where necessary - the persistent memory needed never exceeds 64 MB
+regardless of the size of the key value store or second level log on flash. We
+divide the evaluation into three parts. First, we evaluate Cyclone's performance
+with a single level log as a pure software packet switch. Next, we evaluate
+performance when adding a second level of log on flash. Finally, we evaluate
+performance when integrated with Rocksdb~\cite{rocksdb} as an alternative to
+Rocksdb's write ahead log. Unless otherwise mentioned, we use a 60 byte header
+followed by an optional payload for experiments. We log both the header and
+payload. In all cases the server echoes the received entry (header and payload)
+back to the client. We fix the number of KV Store threads at 32 using at most 8
+physcial logs with associated control plane cores for running RAFT instances.
 
@@
 the log replication problem. We therefore begin by systematically evaluating
-optimizations applied in Cyclone to replicate the top level log in
+optimizations applied in Cyclone to replicate the NVM log in
 Figure~\ref{fig:network_opts} - with no payload. The y-axis reports latency seen
 at the client (which means two network round trips with replication). Using
-TCP/IP to replicate a RAFT log tops out at around 30K entries/s. Applying
-optimizations improves performance by a large enough amount that we use a log
-scale on the x-axis. Switching to DPDK (the line marked +DPDK) improves the
-throughput by an order of magnitude to around 50K entries/s. Using batching
-(line marked +batching) improves the performance further bringing us close to a
-million entries/s. Scaling horizontally to 8 physical logs (+8 phy logs)
-improves performance to close to 2M entries/s. Finally using all 4 ports on the
-machine to replicate entries improves performance considerably to 6M
-entries/s. In all performance improves by 200X over the TCP/IP single log
-baseline. Cyclone also considerably improves the latency for replication, from
-close to 100us with TCP/IP to around 30us at peak throughput.
+TCP/IP to replicate a RAFT log tops out at around 30K entries/s.  Switching to
+DPDK (the line marked +DPDK) improves the throughput by an order of magnitude to
+around 500K entries/s. Using batching (the line marked +batching) improves the
+performance further bringing us close to a million entries/s. Scaling
+horizontally to 8 physical logs (+8 phy logs) improves performance to close to
+2M entries/s. Finally using all 4 ethernet ports on the machine to replicate
+entries improves performance considerably to 6M entries/s. In all, performance
+improves by 200X over the TCP/IP single log baseline. Cyclone also considerably
+improves the latency for replication, from close to 100us with TCP/IP to around
+30us at peak throughput.
 
@@
 entries/s). Adding replicas decreases the peak throughput down to around 2M
-entries/s. We note that a number of previous pieces of work~\cite{faast, farm}
-use three replicas and therefore we focus on three replicas for the replicated
-cases we consider below. The second factor that dictates Cyclone's performance
-is the size of the log entry being replicated. Figure~\ref{fig:payload} shows
-the effect of increasing the payload size from zero to 512 bytes. Peak
-throughput drops from 6M entries/s to approximately 2M entries/s. At this
-replication rate, the leader replica needs to transmit data at approximately 30
-Gbit/s. Coupled with the cost of network headers all four 10 GigE ports are now
-saturated and therefore Cyclone hits the network line rate bottleneck at this
-point.
+entries/s with 5 replicas. We note that a number of previous pieces of
+work~\cite{faast, farm} use three replicas and therefore we focus on three
+replicas for the replicated cases we consider below. The second factor that
+dictates Cyclone's performance is the size of the log entry being
+replicated. Figure~\ref{fig:payload} shows the effect of increasing the payload
+size from zero to 512 bytes. Peak throughput drops from 6M entries/s to
+approximately 2M entries/s. At this replication rate, the leader replica needs
+to transmit data at approximately 30 Gbit/s. Coupled with the cost of network
+headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
+network line rate bottleneck at this point.
 
@@
 one by adding the second level flashlog. We evaluate the impact of adding
-storage to our hitherto pure packet switching scenario in
+block storage to our hitherto pure packet switching scenario in
 Figure~\ref{fig:flashlog}. Batching entries from the top level NVDIMM log to the
@@
 \includegraphics[scale=0.6]{results2/flashlog_512.pdf}
-\caption{Impact of adding second level log (512 bytes payload)}
+\caption{Second level log with 512 bytes payload}
 \label{fig:flashlog_512}
@@
 replication quorums as well executing our rendezvous protocol on multiple
-cores. We therefore setup an experiment where a single client reflecting the
-unloaded case made ganged requests to the replicas. We varied the number of
+cores. We therefore setup an experiment where a single client - reflecting the
+unloaded case - made ganged requests to the replicas. We varied the number of
 cores participating in the request from one to the full complement of 32
@@
 40 us from the baseline of 20 us. There are two causes of this. First, there is
-a rapid increase from 1 to 8 active cores, the reason being that this
-corresponds to an increase in the number of quorums that must be synchronized
-on. There is some amount of synchronization involved in the userspace DPDK
-driver for access to common resources for accessing the NIC and this affects the
-ability to simultaneously issues replication messages on all quorums. The
-smaller contribution to increasing latency that continues past the 8 core case
-is due to cacheline pingponging when executing the rendevous. Both these sources
-of latency drift could be corrected using replication quorums mapped to
-dedicated NICs and more scalable rendezvous designs (such as with machine aware
-broadcast trees~\cite{broadcast_tree}). However we deemed the complexity of such
+a quicker rate of increase from 1 to 8 active cores, the reason being that this
+corresponds to an increase in the number of physical logs that must be
+synchronized on. There is some amount of synchronization involved in the
+userspace DPDK driver for access to common resources for accessing the NIC and
+this affects the ability to simultaneously issues replication messages on all
+quorums. The smaller contribution to increasing latency that continues past the
+8 core case is due to cacheline pingponging when executing the rendevous of
+Figure~\ref{fig:ganged_ops}. Both these sources of latency drift could be
+corrected using replication quorums mapped to dedicated NICs and more scalable
+rendezvous designs (such as with machine aware broadcast
+trees~\cite{broadcast_tree}). However we deemed the complexity of such
 optimizations unnecessary. The added latency for most cases is well under the
-extra round trip delay, which would the minimum needed for a solution using
-distributed transactions via two phase commit.
+extra round trip delay (including replication for all phases of a 2PC), which
+would the minimum needed for a solution using distributed transactions via two
+phase commit.
 
@@
 We evaluate two different request sizes: 8 byte keys with 8 byte values and 8
-bytes keys with 256 byte values. We run 8 physical logs on 8 cores and devote 32
-cores to making requests to Rocksdb. Since we are interested in performance of
-the log, which is only used for update requests, our workload is 100\% writes.
+byte keys with 256 byte values. We run 8 physical logs on 8 cores and devote 32
+cores to Rocksdb. Since we are interested in performance of the log, which is
+only used for update requests, our workload is 100\% writes.
 
-Before evaluating with Rocksdb we measure the baseline performance of
-replicating the log with Cycloine for the given request sizes - Rocksdb performs
+Before evaluating with Rocksdb, we measure the baseline performance of
+replicating the log with Cyclone for the given request sizes - Rocksdb performs
 a no-op. We note that in addition to the key and value, we are also logging
@@
 The performance of Rocksdb with Cyclone for the small update workload is shown
-in Figure~\ref{fig:rocksdb}. We consider four different settings. The line
-labeled Rocksdb is the key value store running with no logging whatsover - a
-system crash would lead to data loss. The line labeled Rocksdb/WAL is for
-Rocksdb running with its write ahead logging turned on. The large gap between
-these two is the overhead of the existing Rocksdb WAL solution. The line labeled
-Rocksdb/Cyclone 1 way is a two level Cyclone log but without any
-replication. The line almost exactly tracks the performance of Rockdb. As
-suggested by the baseline replication performance, Cyclone is able to provide a
-write ahead log with no overhead to Rocksdb. The line labeled Rocksdb/Cyclone 3
-way is with 3-way replication turned on. Other than a 20us delta due to the
-extra network round trip, the line almost exactly tracks the Rocksdb no logging
-case. Cyclone therefore provides high availability to Rocksdb at a fraction of
-the cost of its existing single machine write ahead log. We also repeat the
-experiment for the larger update size in Figure~\ref{fig:rocksdb_256}. The
-conclusions are identical: Cyclone solves Rocksdb's write ahead logging
-problem.
+in Figure~\ref{fig:rocksdb} - essentially presenting the solution to the problem
+description graph in Figure~\ref{fig:problem}.  We consider four different
+setups. The line labeled 'rocksdb' is the key value store running with no
+logging whatsover - a system crash would lead to data loss. The line labeled
+'rocksdb/wal' is for Rocksdb running with its write ahead logging turned on. The
+large gap between these two is the overhead of the existing Rocksdb WAL
+solution. The line labeled 'rocksdb/Cyclone 1 way' is a two level Cyclone log
+but without any replication. The line almost exactly tracks the performance of
+Rockdb. As suggested by the baseline replication performance, Cyclone is able to
+provide a write ahead log with no overhead to Rocksdb. The line labeled
+'rocksdb/Cyclone 3 way' is with 3-way replication turned on. Other than a 20us
+delta due to the extra network round trip, the line almost exactly tracks
+Rocksdb performance with no logging. Cyclone therefore provides high
+availability to Rocksdb at a fraction of the cost of its existing single machine
+write ahead log. We also repeat the experiment for the larger update size in
+Figure~\ref{fig:rocksdb_256}. The conclusions are identical: Cyclone solves
+Rocksdb's write ahead logging problem.
 
@@
 increasing number of keys in the batch - till 32 keys that covers all
-application cores. The line labeled Rocksdb is with no write-ahead logging. We
+application cores. The line labeled Rocksdb is with no write ahead logging. We
 note an increasing latency for this baseline indicating Rocksdb takes longer
-with larger key batches. For the other options - Rocksdb/wal has considerably
+with larger key batches. The existing option of Rocksdb/wal has considerably
 larger latency. Cyclone does an effective job of cutting down on this latency
@@
 application cores making it somewhat slower than running Rocksdb with no logging
-for batched writes.
+for batched writes. A distributed 2PC transaction still remains more expensive
+just in terms of network delay than using Cyclone.
 
@@
 
-Finally we showcase the benefit of using Cyclone beyond pure performance as
+Finally, we showcase the benefit of using Cyclone beyond pure performance as
 compared to the existing single machine Rocksdb write ahead log. Cyclone brings
@@
 
-Cyclone takes a software only approach to improving replication performance over
-commodity network hardware. At the other end of the spectrum, Consensus in a
-box~\cite{consensus_box} implements the zookeeper atomic broadcast protocol
+Cyclone takes a software only approach to improving replication performance
+using commodity network hardware. At the other end of the spectrum, Consensus in
+a box~\cite{consensus_box} implements the Zookeeper atomic broadcast protocol
 together with a NIC on an FPGA. That work ignored durability focusing purely on
 replication performance. Latency numbers are impressive from the FPGA that cuts
-out the path from CPU caches to the NIC are as low as 3us for a round trip from
-leader to follower replica compared to the 7us we observe with DPDK. Notably
-however they reported 7us when using TCP/IP from the FPGA rather than their own
-connection oriented network protocol. On the other hand peak replication
-throughput appears to be 6M replications/sec on 66 Gbps of aggregate network
-bandwith, comparable to the mark obtained by Cyclone on an aggregate 40 Gbps of
-network bandwidth. We believe the general applicability afforded by our software
-only packes switching solution makes it attractive in comparison to an FPGA
-based solution even given the larger latency.
+out the path from CPU caches to the NIC and are reported to be as low as 3us for
+a round trip from leader to follower replica compared to the 7us we observe with
+DPDK. Notably however they reported 7us when using TCP/IP from the FPGA rather
+than their own connection oriented network protocol. On the other hand peak
+replication throughput appears to be 6M replications/sec on 66 Gbps of aggregate
+network bandwith, comparable to the mark obtained by Cyclone on an aggregate 40
+Gbps of network bandwidth. We believe the general applicability afforded by our
+software only packet switching solution makes it a compelling alternative to an
+FPGA based solution even given the larger latency.
 
-Networked Ordered Paxos~\cite{nopaxos} reprograms top of rack switches to order
+Network Ordered Paxos~\cite{nopaxos} reprograms top of rack switches to order
 client requests and multicast them to replicas - attacking replication latency
@@
 amounts of data - CRANE reports around a 2X slowdown for MySQL likely rendering
-it unsuitable as a replacement for the WAL in key value stores.
+it unsuitable as a replacement for the write ahead log in key value stores.
 
 A number of researchers have proposed using high performance networking hardware
-for improving the performance of various distributed system stacks such as
+for improving the performance of various distributed system primitives such as
 distributed transactions~\cite{farm} and replication~\cite{dare, mojim}. We note
@@
 switching and horizontal scaling can be equally applied to RDMA. However, we
-chose commodity ethernet due to the short term deployment plans that are a
+chose commodity ethernet due to short term deployment plans that are a
 priority design target.
 
-The scalable commutativity rule~\cite{scalable_commutativity} generalized the
+The scalable commutativity rule~\cite{scalable_commutativity} generalizes the
 idea of commutativity in key value store interfaces that we have used for
-horizontal scaling in Cyclone. Conveniently it provides a tool for checking when
+horizontal scaling in Cyclone. Conveniently, it provides a tool for checking when
 API calls can commute in a history. It is a simple exercise to express a key
-value store in COMMUTER and check that operations to different keys
+value store API in COMMUTER and check that operations to different keys
 commute. This provides a theoretical basis for horizontal scaling, something we
 have omitted for brevity from this paper. The scalable commutativity rule also
-suggests that scalable key value store implementations would exploit
-commutativity something we have observed in the case of in-memory layers for key
-value stores such as Flodb and indeed in the way we distributed keys across
-physical logs in Cyclone.
+suggests that scalable key value store implementations would necessarily have
+commutative interfaces, something we confirm and exploit for key
+value stores.
 
-Finally, we note the in progress transformation of durable memory in datacenters
+There is an in progress transformation of durable memory in datacenters
 with the advents of 3DXPoint. Researchers have speculated on the impact of cheap
@@
 3DXPoint would be an idea target for Cyclone because each NVM log entry can be
-discarded as soon as it is applied to the key value store, since the results are
-persistent. This would allow Cyclone to operate at pure packet switching speeds,
-a goal we hope to achieve in the future with a single level log for NVML
-applications.
+discarded as soon as it is applied to the persistent key value store removing
+the need for the flashlog. This would allow Cyclone to operate at pure packet
+switching speeds, a goal we hope to achieve in the future with a single level
+log for NVML data store applications.
+
+Cyclone provides a client library that automatically times out and retries
+requests to deal with failure. Protection against repeated execution of the same
+request if currently left to the key value store and client, since we felt that
+exactly once semantics is optional when updates such as put operations are
+idempotent. We have experimented with exactly once semantics as a basis for
+distributed transactions across shards running Cyclone integrated key value
+stores, along the lines of similar work~\cite{raft_lin}, using a fixed client
+space and maintaining information about last seen sequence numbers from clients
+in NVM. Distributed transactions on Cyclone shards however are outside the scope
+of this paper.
 
 \section{Conclusion}
-
+Persistent key values stores struggle today to provide high availability using a
+write ahead log without a significant impact to performance. Cyclone solves this
+problem by leveraging a small amount of directly attached non-volatile memory
+that transforms the replication problem into a software packet switching
+problem. Cyclone and its Rocksdb bindings are slated for release to the open
+source community and we hope both the code and the ideas in the system itself
+will help key value store builders and researchers deal with their logging
+problem.
 \newcommand\myurl[2]{\url{#1}}

### 2017-04-21 paper/paper.tex
Updates to paper.

@@
 switching problem as the leader sends the exact same packet to all follower
-replicas and we solve the problem as such by using the RAFT protocol implemented
-as a software multicast module on the Data Plane Development Kit (DPDK). Second,
+replicas. We solve this problem by using the RAFT protocol implemented
+as a software multicast module on the Data Plane Development Kit (DPDK). In addition,
 we leverage the observation that key value store interfaces are commutative for
 operations to different keys to scale our two level log horizontally into a
-number of physical two level logs - with a hash of the key being used to choose
+number of physical two level logs. A hash of the key is used to choose
 the physical log. Each log is replicated by an independent instance of RAFT,
@@
 when using multiple CPU cores. Cyclone is able to replicate millions of small
-updates a second using only commodity 10 gigabit ethernet adapters and improves
+updates a second using only commodity 10 gigabit ethernet adapters, improving
 the performance and availability of RocksDB, a popular persistent key value
@@
 availability. Unfortunately the write ahead is a performance achilles heel for
-these systems - eclipsing much of the work on improving performance of the LSM
+these systems - eclipsing much of the work on improving the performance of the LSM
 component. To illustrate the impact of the write ahead log, consider
-Figure~\ref{fig:problem}. The line marked Rocksdb shows the performance of
+Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
 Rocksdb without the write ahead log. The performance when persisting the write
@@
 persisting every update to the log or replicating it using TCP/IP causes
-performance to drop by an order of magnitude (observe the log scale on the
+performance to drop by an order of magnitude (note the log scale on the
 x-axis). It is therefore no suprise that deployments of Rocksdb often turn off
@@
 \includegraphics[scale=0.6]{results2/problem.pdf}
-\caption{Rocksdb Write Ahead Logging Impact}
+\caption{Rocksdb write ahead logging impact}
 \label{fig:problem}
@@
 batches to a log placed on a flash drive. This allows Cyclone to use only a
-limited amount of NVM reasoning that current forms of NVM such as NVDIMMs are
-more expensive than DRAM. Our experiments use an NVM footprint of a mere 64 MB
-to achieve performance comparable to placing the entire log in NVM. This allows
-Cyclone to be used in deployments where the key value store is maintained
-entirely in main memory rather than as a multi-level LSM tree structure, with
-the log serving as the source for recovery in the event of failure.
+limited amount of NVM following the reasoning that current forms of NVM such as
+NVDIMMs are more expensive than DRAM. Our experiments use an NVM footprint of a
+mere 64 MB to achieve performance comparable to placing the entire log in
+NVM. This allows Cyclone to be used in deployments where the key value store is
+maintained entirely in main memory rather than as a multi-level LSM tree
+structure, with the log serving as the source for recovery in the event of
+failure.
 
@@
 being software switched. Our solution is to split the log into two levels. The
-upper level is held in persistent directly attached non-volatile memory (NVM)
+upper level is held in directly attached non-volatile memory (NVM)
 such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
@@
 \includegraphics[scale=0.3]{figures2/nvm_log.pdf}
-\caption{NVM Log Structure}
+\caption{NVM log structure}
 \label{fig:nvm_log}
@@
 \end{verbatim}
-\caption{Cyclone Control Plane Fragment}
+\caption{Cyclone control plane fragment}
 \label{fig:control_plane}
@@
   \includegraphics[scale=0.3]{figures2/network_packet.pdf}
-  \caption{Cyclone Network Packet Layout}
+  \caption{Cyclone network packet layout}
   \label{fig:packet_layout}
@@
 
-We note that by discrding stream oriented protocols such as TCP, the network
-layer is not longer reliable or ordered. This is not a problem for RAFT that is
+We note that by discarding stream oriented protocols such as TCP, the network
+layer is no longer reliable or ordered. This is not a problem for RAFT that is
 designed to tolerate network failures resulting in drops or reordering of
@@
 efficiently flush to persistent memory without the overhead of the serialization
-normally introduced by {\tt clflush}. This allows one to hit full memory
+normally introduced by {\tt clflush}. This allows us to hit full memory
 bandwidth on present generation platforms, a quantity in excess of 200 Gb/s per
@@
 updating pages on the SSD even under power failure i.e. there are no shorn
-writes on a 4KB page~\cite{shorn_writes}. We also assume that the SSD has power
-loss data protection meaning that writes cached in the drive's volatile cache
-are written to the SSD using a backup capacitor in the event of power failure -
-a property of many data center class SSDs today, including the Intel DC P3600
-SSD~\cite{ssd_spec} we use in our evaluation. Together these two assumptions
-mean that we can recover a consistent prefix of the log on a power failure.
+writes (otherwise known as torn writes) on a 4KB page~\cite{shorn_writes}. We
+also assume that the SSD has power loss data protection meaning that writes
+cached in the drive's volatile cache are written to the SSD using a backup
+capacitor in the event of power failure - a property of many data center class
+SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
+evaluation. Together these two assumptions mean that we can recover a consistent
+prefix of the log on a power failure.
 
@@
 responses from a majority quorum is received, the request is passed to the
-targeted KV store core for execution. The KV store core process the request and
-sends back the response to the client. We note that replication is pipelined and
-decoupled from execution and thus multiple updates are usually in
-flight on the network for replication at the same time.
+targeted KV store core for execution. The KV store core processes the request
+and sends back the response to the client. We note that replication is pipelined
+and decoupled from execution and thus multiple updates are usually in flight on
+the network for replication at the same time.
 
@@
 We note that commutativity as assumed by Cyclone is an \emph{API level
-  property}. There are no assumption as to how application cores synchronize to
+  property}. There is no assumption as to how application cores synchronize to
 access the shared key-value state or the underlying data structures for holding
@@
 
-Our solution instead is to use a technique we termed ganged operations. It is
+Our solution instead is to use a technique we term ganged operations. It is
 based on the observation that in the failure free case, executing a batched
@@
 request to the participating control plane cores (for corresponding physical
-logs). The injection core uses DPDKs packet cloning primitives to avoid
+logs). The injection core uses DPDK's packet cloning primitives to avoid
 dataplane work while forwarding copies of the same batched write packet to
@@
 view number to the packet containing the term numbers of all the participating
-RAFT instances (read from shared memory of the co-located leaders).  We describe
-the remaining flow for ganged operations in terms of the pseudocode in
-Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce to
-later in this section. We also assume that a unique barrier is allocated in
-shared memory for each ganged operation. We discuss later how this is done
-without using dynamic allocation.
+RAFT instances (read from shared memory of the co-located leaders).  The flow
+for ganged operations is described in the pseudocode of
+Figure~\ref{fig:ganged_ops}. We defer the discussion of how we generate the
+nonce to later in this section. We also assume that a unique barrier is
+allocated in shared memory for each ganged operation. We discuss later how this
+is done without using dynamic allocation.
 
@@
 replication is complete on all participating physical logs, the corresponding KV
-store cores execute \\{\tt event\_replicated\_ganged\_op}. They use a bitmask for
-the barrier setting the appropriate bit for themselves. Once all KV store cores
-have arrived at the barrier the operation is executed. A distinguished leader
-core (we use the minumum numbered participating KV store core) is
+store cores execute \\{\tt event\_replicated\_ganged\_op}. They use a bitmask
+for the barrier setting the appropriate bit for themselves. Once all KV store
+cores have arrived at the barrier the operation is executed. A distinguished
+leader core (we use the minumum numbered participating KV store core) is
 responsible for sending back the response to the client. Two types of failure
 need to be considered. First, replication on the physical log corresponding to
-one of the non-leader cores could fail. The
-remaning cores would never progress past the barrier. To solve this the leader
-core montors the remaning cores for their current view - which is the RAFT term
-of the last executed log entry. If the view has moved past the view in the
-packet, replication failed on that physical log. The leader then sets the bit
-for the core that will no longer participate in the barrier to aid progress (a
-technique we borrowed from lock-free data structures~\cite{lock-free}). It also
-sets a failed flag to signal failure to the other participating cores. The
-second failure case is that the leader core might not receive the operation if
-it fails during replication from corresponding physical log. This case is
-detected in a similar manner by the remaining cores by monitoring the published
-view from the leader core in shared memory. We add that on a leadership change
-due to timeout we enqueue and execute a nop on the physical log and all
-associated KV store cores to ensure previous entries are committed - a corner
-case in the RAFT protocol.
+one of the non-leader cores could fail. The remaning cores would never progress
+past the barrier. To solve this the leader core monitors the remaning cores for
+their current view - which is the RAFT term of the last executed log entry. If
+the view has moved past the view in the packet, replication failed on that
+physical log. The leader then sets the bit for the core that will no longer
+participate in the barrier to aid progress (a technique we borrowed from
+lock-free data structures~\cite{lock-free}). It also sets a failed flag to
+signal failure to the other participating cores. The second failure case is that
+the leader core might not receive the operation if it fails during replication
+from corresponding physical log. This case is detected in a similar manner by
+the remaining cores by monitoring the published view from the leader core in
+shared memory. We add that on a leadership change due to timeout we enqueue and
+execute a nop on the physical log and all associated KV store cores to ensure
+previous entries are committed - a corner case in the RAFT protocol.
 
@@
 failure detector and repeated retries from the clients. There are - therefore -
-no points on the ``knee'' of the curve as in the pure packet switched one level
+no points on the ``knee'' of the curve as in the pure packet switched one-level
 log case.
@@
 userspace DPDK driver for access to common resources for accessing the NIC and
-this affects the ability to simultaneously issues replication messages on all
+this affects the ability to simultaneously issue replication messages on all
 quorums. The smaller contribution to increasing latency that continues past the
@@
 optimizations unnecessary. The added latency for most cases is well under the
-extra round trip delay (including replication for all phases of a 2PC), which
-would the minimum needed for a solution using distributed transactions via two
-phase commit.
+extra round trip delay (including replication) for all phases of a 2PC.
 
@@
 together with a NIC on an FPGA. That work ignored durability focusing purely on
-replication performance. Latency numbers are impressive from the FPGA that cuts
+replication performance. Latency numbers are impressive due to the FPGA that cuts
 out the path from CPU caches to the NIC and are reported to be as low as 3us for
@@
 requests to deal with failure. Protection against repeated execution of the same
-request if currently left to the key value store and client, since we felt that
+request is currently left to the key value store and client, since we felt that
 exactly once semantics is optional when updates such as put operations are
@@
 
+Finally, we note that scaling a log horizontally for concurrency is also an idea
+adopted by filesystems. NOVA~\cite{nova} is an example that uses a per-inode log
+to improve concurrency when the filesystem is placed on NVM attached to the
+memory bus. Cyclone also uses a distinct NVM log per control plane core in a
+similar vein but does not maintain a namespace since the number of NVM logs is a
+small fixed number equal to the number of control plane cores.
+
 \section{Conclusion}
-Persistent key values stores struggle today to provide high availability using a
+Persistent key value stores struggle today to provide high availability using a
 write ahead log without a significant impact to performance. Cyclone solves this
@@
 source community and we hope both the code and the ideas in the system itself
-will help key value store builders and researchers deal with their logging
+will help key value store builders and researchers overcome their logging
 problem.

### 2017-07-21 paper/paper.tex
Updates to paper.
Switch to ASPLOS format.

@@
-\documentclass[10pt, preprint, nonatbib]{sigplanconf}
-\usepackage{times,epsfig,endnotes,graphicx, verbatim}
-\usepackage{datetime, url, pdfpages, parskip}
+\documentclass[pageno]{jpaper}
 
-\conferenceinfo{SOSP'17}{October 29--31, 2017, Shanghai, China}
-\copyrightyear{2017}
+%replace XXX with the submission number you are given from the ASPLOS submission site.
+\newcommand{\asplossubmissionnumber}{XXX}
 
-% These only appear when the 'preprint' option is specified.
-% Enabling these will cause the first page of the document to fail the 
-% format check on HotCRP :-(
-%\titlebanner{Under submission to SOSP 2017 - do not cite or distribute}
-%\preprintfooter{Draft of {\currenttime}, \today{}}
-
-% No date in title area.
-\date{}
-
-% Paper number and no. of pages as author
-\authorinfo{Paper \textbf{\#224}}{13 pages}
+\usepackage[normalem]{ulem}
 
@@
 
-%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
 \title{\Large \bf Cyclone: Warp Speed Replication for Key Value Stores}
 
+\date{}
 \maketitle
+
+\thispagestyle{empty}
 
@@
 to maintain and replicate a write ahead log to guarantee availability in the
-face of machine and storage failures. Cyclone offers a solution to this problem
-by leveraging a small amount of directly attached non-volatile memory (NVM) such
-as NVDIMMs to maintain a two level log: the upper level in a small amount of NVM
-draining into a larger lower level log maintained on flash. Replicating the
-smaller log kept in NVMs transforms the replication problem into a packet
-switching problem as the leader sends the exact same packet to all follower
-replicas. We solve this problem by using the RAFT protocol implemented
-as a software multicast module on the Data Plane Development Kit (DPDK). In addition,
-we leverage the observation that key value store interfaces are commutative for
-operations to different keys to scale our two level log horizontally into a
-number of physical two level logs. A hash of the key is used to choose
-the physical log. Each log is replicated by an independent instance of RAFT,
-thereby leveraging the scalability afforded by classic software packet switching
-when using multiple CPU cores. Cyclone is able to replicate millions of small
-updates a second using only commodity 10 gigabit ethernet adapters, improving
-the performance and availability of RocksDB, a popular persistent key value
-store by an order of magnitude when compared to its own write ahead log without
-replication.
+face of machine and storage failures. Cyclone is a replicated log plug-in for
+key value stores that systematically addresses various sources of this
+bottleneck. It uses a small amount of non-volatile memory directly addressable
+by the CPU - such as in the form of NVDIMMs - to remove block oriented IO
+devices such as SSDs from the critical path for appending to the log. It then
+addresses network overheads using an implementation of the RAFT consensus
+protocol that is designed around a userspace network stack to relieve the CPU of
+the burden of data copies. Finally, it provides a way to efficiently map the
+commutativity in key-value store APIs to the parallelism available in commodity
+NICs via multiple transmit receive queues to scale replication performance on
+multicore machines. Cyclone is able to replicate millions of small updates a
+second using only commodity 10 gigabit ethernet adapters. As a practical
+application, we use it to improve the performance and availability of RocksDB, a
+popular persistent key value store by an order of magnitude when compared to its
+own write ahead log without replication.
 \end{abstract}
@@
 and kept durable for every appended update provides the necessary high
-availability. Unfortunately the write ahead is a performance achilles heel for
-these systems - eclipsing much of the work on improving the performance of the LSM
-component. To illustrate the impact of the write ahead log, consider
+availability. Unfortunately, the write ahead log is a performance achilles heel
+for these systems - eclipsing much of the work on improving the performance of
+the LSM component. To illustrate the impact of the write ahead log, consider
 Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
@@
 performance to drop by an order of magnitude (note the log scale on the
-x-axis). It is therefore no suprise that deployments of Rocksdb often turn off
+x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
 the write ahead log~\cite{samza}. On the other side of the spectrum, key value
-store research such as FloDB~\cite{flodb} turns off the write ahead log to be
-able to showcase benefits of sophisticated extensions to LSM datastructures.
+store research prototypes such as FloDB~\cite{flodb} turns off the write ahead
+log to be able to showcase benefits of sophisticated extensions to LSM
+datastructures.
 
 \begin{figure}
-\centering
-\includegraphics[scale=0.6]{results2/problem.pdf}
+\centering \includegraphics[scale=0.6]{results2/problem.pdf}
 \caption{Rocksdb write ahead logging impact}
@@
 Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
-\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the
-host - such as in the form of NVDIMMs~\cite{farm} or directly attached
-persistent memory~\cite{pmfs}.
+\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the host
+- such as in the form of NVDIMMs~\cite{farm} or directly attached persistent
+memory~\cite{pmfs}. Cyclone logs update operations to the key value store and
+calls on the key value store to execute the update only when it has been
+successfully replicated to majority quorum of replicas. All reads and write to
+the key value store are always executed by the leader replica. A client side
+stub provided by Cyclone automatically locates and directs read and write
+requests to the leader replica, ensuring linearizable semantics from the key
+value store. Using a replicated write ahead log to build a linearizable and
+highly available key value store is not new [find some examples here]. The
+novelty in Cyclone is that it accomplishes this without compromising on the
+performance baseline of no logging or replication. Cyclone does this by
+systematically addressing the sources of performance degradation in
+Figure~\ref{fig:problem}.
 
-Cyclone makes use of a limited amount of attached NVM using two ideas. First, it
-leverages the observation that with persistence being a first class property of
-a subset of directly addressable DRAM, replicating the same log entry from the
-leader to follower replicas is no different from multicast using software packet
-switching techniques. It directly leverages this observation to implement a
-version of RAFT that replicates the NVM log using classic packet switching
-techniques such as separation of control and data planes. Second, Cyclone uses a
-two level log structure where it drains entries from the NVM log in IO friendly
-batches to a log placed on a flash drive. This allows Cyclone to use only a
-limited amount of NVM following the reasoning that current forms of NVM such as
-NVDIMMs are more expensive than DRAM. Our experiments use an NVM footprint of a
-mere 64 MB to achieve performance comparable to placing the entire log in
-NVM. This allows Cyclone to be used in deployments where the key value store is
-maintained entirely in main memory rather than as a multi-level LSM tree
-structure, with the log serving as the source for recovery in the event of
-failure.
+The primary cause of slowdown when turning on a write ahead log on a single
+machine is the need to do synchronous block IO for every update. For even
+moderately sized updates, this leads to severe write amplification as a 4KB page
+must be synchronously written to the log file on flash. Cyclone avoids this by
+making use of directly attached persistent memory on the server. This can be in
+the form of NVDIMMs (DIMMs with an added ultracapacitor to dump state to a small
+amount of attached flash) or newer forms of persistent memory~\cite{pmfs}. For
+current deployments, NVDIMMs are expensive - retailing at approximately 2X the
+price of a standard DIMM [find ref]. This means that they will likely be
+available in small quantities. We compensate for this by periodically draining
+the NVDIMM log into a log placed on a standard SSD on the system. We describe
+this two level log structure in Section~\ref{sec:storage}.
 
-We discuss the two level structure focusing on the packet switching components
-in Section~\ref{sec:2level}. We then describe how we can horizontally scale this
-basic idea across multiple cores by leveraging the commutativity of key value
-store APIs across different keys in Section~\ref{sec:horizontal}. We then
-demonstrate in the evaluation in Section~\ref{sec:evaluation} that Cyclone can
-replicate millions of updates a second using only commodity 10 Gigabit ethernet
-adapters thereby closing the performance gap between key value store performance
-and write ahead logging performance using only a small amount of directly
-attached NVM. A discussion of related work follows before we conclude.
+The network overheads for replicating a log entry to a quorum of machines stems
+from the overhead of the TCP/IP stack in the linux kernel - itself the topic of
+much research [dataplane OS stuff]. We therefore replase the TCP/IP stack with
+DPDK, a userspace network library that allows direct IO of raw ethernet frames
+to and from the network interface card (NIC) eliminating all overheads. We no
+longer use TCP - since the rare network errors encountered in a typical
+datacenter network are automatically handled by the replication protocol (RAFT)
+as part of its failure model. Finally, we carefully design our implementation of
+RAFT to relieve the CPU of all data copy operations, leaving the core to only
+run the protocol and return an accept/reject decision for each received
+packet. The set of optimizations to the network stack is decribed in
+Section~\ref{sec:network}.
+
+Both the storage (NVM, SSD) as well as the NIC expose significant parallelism
+that is left unused when Cyclone is used to replicate a single log. To exploit
+this parallelism we use Cyclone to replicate multiple logs, each using a
+different instance of the RAFT consensus protocol. We show how to exploit the
+commutativity inherent in key-value store interfaces to make use of this
+parallelism in Cyclone's design in Section~\ref{parallelism}.
+
+A detailed evaluation of Cyclone's replication is provided in the context of
+RocksDB - a popular persistent key value store in
+Section~\ref{sec:evaluation}. We show that Cyclone achieves performance
+transparent replication in that it can provide both a persistent write ahead log
+and replication without compromising on RocksDB's performance.
+
+\section{System Architecture}
+
+\section{Storage}
+
+\section{Network}
+
+\section{Parallelism}
 
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
+  \includegraphics[scale=0.35]{figures2/network_packet.pdf}
   \caption{Cyclone network packet layout}
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \includegraphics[scale=0.4]{figures2/batching.pdf}
   \caption{Batching}
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/flashlog_page.pdf}
+  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
   \caption{Flashlog segment}
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/race.pdf}
+  \includegraphics[scale=0.35]{figures2/race.pdf}
   \caption{Race with multiple physical logs}

### 2017-07-21 paper/paper.tex
Updates to paper.

@@
 Cyclone is a high speed strongly consistent replicated write ahead logging
-service specialized for key value stores such as Rocksdb. Cyclone provides the
-order of magnitude better performance required to close the performance gap in
+service specialized for key value stores such as Rocksdb. We describe Cyclone's
+system architecture in Section~\ref{sec:sysarch}. Cyclone provides the order of
+magnitude better performance required to close the performance gap in
 Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
@@
 as part of its failure model. Finally, we carefully design our implementation of
-RAFT to relieve the CPU of all data copy operations, leaving the core to only
+RAFT to relieve the CPU of all data copy operations, leaving the CPU to only
 run the protocol and return an accept/reject decision for each received
@@
 commutativity inherent in key-value store interfaces to make use of this
-parallelism in Cyclone's design in Section~\ref{parallelism}.
+parallelism in Cyclone's design in Section~\ref{sec:parallelism}.
 
@@
 \section{System Architecture}
+\label{sec:sysarch}
 
 \section{Storage}
-
-\section{Network}
-
-\section{Parallelism}
-
-\section{Two Level Log}
-\label{sec:2level}
+\label{sec:storage}
 A fundamental challenge in building an efficient replicated log is dealing with
@@
 network interface.
+
+\section{Network}
+\label{sec:network}
 
@@
 \end{figure}
+
+\section{Parallelism}
+\label{sec:parallelism}
 

### 2017-07-23 paper/paper.tex
Updates.

@@
 \usepackage[normalem]{ulem}
+\usepackage{parskip}
 
@@
 \thispagestyle{empty}
-
-% Use the following at camera-ready time to suppress page numbers.
-% Comment it out when you first submit the paper for review.
-%\thispagestyle{empty}
 
@@
 ahead log. A machine or storage failure leading to total loss or temporary
-unavailability of data is unacceptable in many internet scale services where
-high availabilty and revenue are often interconnected. Key value stores
-therefore usually incorporate support for a write ahead log that if replicated
-and kept durable for every appended update provides the necessary high
-availability. Unfortunately, the write ahead log is a performance achilles heel
-for these systems - eclipsing much of the work on improving the performance of
-the LSM component. To illustrate the impact of the write ahead log, consider
+unavailability of data is unacceptable in services where high availabilty and
+revenue are interconnected. Key value stores therefore usually incorporate
+support for a write ahead log that if replicated and kept durable for every
+appended update provides the necessary high availability. Unfortunately, the
+write ahead log is a performance achilles heel for these systems - eclipsing
+much of the work on improving the performance of the LSM component. To
+illustrate the impact of the write ahead log, consider
 Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
@@
 requests to the leader replica, ensuring linearizable semantics from the key
-value store. Using a replicated write ahead log to build a linearizable and
-highly available key value store is not new [find some examples here]. The
-novelty in Cyclone is that it accomplishes this without compromising on the
-performance baseline of no logging or replication. Cyclone does this by
-systematically addressing the sources of performance degradation in
+value store. The novelty in Cyclone is that it accomplishes this without
+compromising on the performance baseline of no logging or replication. Cyclone
+does this by systematically addressing the sources of performance degradation in
 Figure~\ref{fig:problem}.
@@
 A detailed evaluation of Cyclone's replication is provided in the context of
-RocksDB - a popular persistent key value store in
+RocksDB - a popular persistent key value store - in
 Section~\ref{sec:evaluation}. We show that Cyclone achieves performance
@@
 \label{sec:sysarch}
+A persistent key value store durably stores key-value mappings accessible through a
+simple interface:
+
+\begin{itemize}
+\item GET(K): returns value corresponding to key K
+\item PUT(K, V): sets the value of key K to value V
+\end{itemize}
+
+A key value store typically serves multiple client requests concurrently and is
+linearizable~\cite{linearizability}. This means any history of requests and
+responses can be rearranged such that responses immediately follow corresponding
+requests and if a response precedes a request in the original history it also
+does so in the rearranged history. Key value stores such as RocksDB also
+support atomic writes to multiple keys and snapshots.
+
+Cyclone integrates with key value stores as both a client and server side
+library. Cyclone replicates the key value store across a set of replica
+servers. The client side library sends requests to a distinguished leader
+replica.  On the server side Cyclone accepts requests from clients and calls
+into the key value store. It then returns a response to the client. Before
+executing any PUT request, Cyclone appends it to a durable log and replicates
+the request to the logs of follower replicas. The request is considered
+replicated once acknowledged by a majority of replicas.
+
+Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On a
+failure of the leader replica a new leader is automatically elected - RAFT
+ensures that the new leader has the most up to date logs. The Cyclone client
+library automatically locates the new leader on a failover - clients therefore
+see only a brief interruption in service. Although read requests (GET
+operations) are not logged, they are serviced in FIFO order with the write requests
+(PUT operations) currently in Cyclone's log. This ensures linearizability in the
+face of failovers.
 
@@
 \label{sec:storage}
-A fundamental challenge in building an efficient replicated log is dealing with
-network overheads. Cyclone's construction is predicated on the observation that
-in the common case, protocols such as RAFT simply multicast the same entry
-received from a client, from the leader replica to all follower replicas. This
-suggests that an efficient way to implement a replicated log is to treat it as a
-software packet switching problem. However peristing logs on flash represents an
-impediment in that block storage devices are not good candidates to hold packets
-being software switched. Our solution is to split the log into two levels. The
-upper level is held in directly attached non-volatile memory (NVM)
-such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
-level log as the flashlog. We periodically drain entries in batches from the NVM
-log to the backing flashlog. An important consequence of this two level approach
-is that we need only a limited amounts of NVM with the bulk of the log residing
-on flash - an arrangement that maps well to the current relative costs of the
-two types of memory. We begin by describing the NVM log and how we implement the
-RAFT consensus protocol in the packet switching paradigm around it. We then
-describe the flashlog and how we drain entries in batches from the NVM level to
-the flash level.
-
-\subsection{NVM Log}
+Storage devices such as flash-based SSDs export a block IO interface. Appending
+a small update to the write-ahead log requires a synchronous update of an entire
+4KB page. This leads to a high baseline latency at even low load, as shown in
+Figure~\ref{fig:problem}.  On the other hand small updates (in the order of 100s
+of bytes for both the key and value taken together) are commonplace in many
+workloads [find refs]. This problem leads many system designers to
+underprovision key-value store shards ensuring that it operates at moderate to
+high load, using group-commit to batch additions to the log. They therefore pay
+a price in latency to ensure the bandwidth to storage is fully utilized to
+prevent it from becoming a bottleneck.
+
+Cyclone adopts a different strategy by making use of novel memory technologies
+to obviate the need to make this tradeoff in the first place. We assume a small
+amount of non-volatile memory directly addressable by the CPU rather than placed
+behind a block oriented IO interface. Log appends are done to the ``NVM log''
+placed in this non-volatile memory. While we designed Cyclone to leverage new
+non-volatile memory technologies that might be available in the
+future~\cite{3dxpoint} current non-volatile memory options such are NVDIMMs
+are more expensive that DRAM. Therefore, for Cyclone to be economical
+we cannot assume that a sufficient amount of NVM is available to hold the entire
+log. The log entries can only be garbage collected when the in-memory component
+of the key value store is checkpointed to secondary storage. The growing size of
+available DRAM means that also provisioning sufficient NVDIMMs to hold the
+corresponding log of updates is not feasible given that NVDIMMs cost at least
+twice as much as volatile DRAM. We therefore resort to using a second larger log
+level placed on a flash SSD. Entries are drained from the NVM log to this flash
+log in conveniently large units that are a multiple of 4KB - the optimum IO unit
+for flash SSDs. This ensures we get the best
+possible throughput from a flash device without paying the price of synchronous
+IO for small key-value pairs. At the same time, the amount of non-volatile
+memory required is small enough to not add undue cost of the server.
+
+The flashlog is written out in segments of configurable size (we use 128KB
+segments). A segment buffer is prepared in memory (volatile DRAM) using the
+layout shown in Figure~\ref{fig:flashlog_page}. We do not allow objects in the
+flashlog to cross a 4KB boundary - linking multiple objects together with a
+special flag encoded into the size if necessary. We flush log segments out to
+the log file using asynchronous direct IO, and therefore we fill one log segment
+buffer while keeping IO to another one outstanding to the flash drive. To avoid
+having to do a synchronous metadata flush, we preallocate (using the
+posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
+the log file whenever we hit its end.
+
+In order to recover from a crash, we make two important assumptions about the
+underlying SSD. First, we assume that 4KB is the \emph{minimum} atomic unit for
+updating pages on the SSD even under power failure i.e. there are no shorn
+writes (otherwise known as torn writes) on a 4KB page~\cite{shorn_writes}. We
+also assume that the SSD has power loss data protection meaning that writes
+cached in the drive's volatile cache are written to the SSD using a backup
+capacitor in the event of power failure - a property of many data center class
+SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
+evaluation. Together these two assumptions mean that we can recover a consistent
+prefix of the log on a power failure.
+
+We move log entries from the head of the NVM log to the flashlog buffers in FIFO
+order. The NVM log entry is only actually removed when the IO for the
+corresponding flashlog page is complete. This means that during recovery we can
+have the same log entry both in the NVM log and the flashlog, a condition that
+can be detected by examining the log sequence number that we embed in each log
+entry.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
+  \caption{Flashlog segment}
+  \label{fig:flashlog_page}
+\end{figure}
+
+The two level log design in Cyclone lays the foundation for improvements to the
+network stack that we discuss next - a significant contributor to the overheads
+in Figure~\ref{fig:problem}. We use RAFT to replicate the NVM log from the
+leader replica to the followers. Each replica \emph{independently} drains
+committed entries (in terms of RAFT consensus) from the NVM log to the
+flashlog. Block IO therefore is no longer a consideration when optimizing data
+movement over the network, unlike systems that need to deal with such
+problems~\cite{reflex}. Replication reduces to efficiently sending a block of
+data already present in directly attached memory over the network.
+
+\section{Network}
+\label{sec:network}
+
+The core network operation in replication is to receive a request from the
+client at the leader replica and send that exact same request (as a log entry)
+to all follower replicas. The leader therefore \emph{multicasts} a received
+request packet to follower replicas, a problem well studied in the networking
+community when building software packet switches. Software packet switches reach
+impressive speeds of millions of packets forwarded per core [find refs], a
+number far in excess of the few hundreds of packets we manage in
+Figure~\ref{fig:problem}. We therefore proceed by approximating a software
+packet switch for the networking component of Cyclone. However software packet
+switching does not need to deal with failure - usually requiring only a simple
+decision for each packet at the node. RAFT introduces significant
+protocol level work for each forwarded packet a challenge that we need to deal
+with to approach the desired software packet switching efficiency in the common
+case.
+
+\subsection{DPDK}
+The Data Plane Development Kit~\cite{dpdk} provides low latency userspace access
+to an ethernet NIC, permitting the application to directly send and receive raw
+ethernet frames via the transmit and receive queues on the NIC. DPDK is often
+used by developers of software packet switches and therefore we also leverage
+DPDK as a library for building Cyclone.
+
+DPDK does not by itself provide a TCP stack. This is not a problem. RAFT (and
+indeed most consensus protocols) tolerate network losses and reordering by
+design due to the need to support asynchronous communication. In addition, most
+datacenter networks rarely drop or reorder packets and provide full bisection
+bandwith between servers that might serve as Cyclone replicas. We therefore
+jettisoned TCP and chose to send raw IP packets encapsulated in ethernet
+frames. We currently follow this communication model both for server to server
+communication between replicas as well as client server communication for
+requests and responses. Although a detailed evaluation is made later, switching
+from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
+testbed from {\tt 24 us} to as low as {\tt 7 us}, providing a significant boost
+to performance.
+
+\subsection{Eliminating Data Movement}
+
+Another key problem we had to solve with DPDK was teaching it about persistent
+memory. DPDK allows one to plug in a memory allocator. We required the allocator
+to support various features such as allowing multiple entities (the cpu core and
+NIC driver) to hold reference to DPDK packets and allocate/free in parallel.
+
 The NVM log is maintained as a circular log of fixed sized pointers (a pointer
@@
 network interface.
-
-\section{Network}
-\label{sec:network}
 
@@
 \subsection{Flash Log}
-Cyclone is designed to use only limited amounts of NVM and drains log entries
-down into a log placed on a flash SSD - that we term the flashlog. The flashlog
-is written out in segments of configurable size (we use 128KB segments). A
-segment buffer is prepared in memory (volatile DRAM) using the layout shown in
-Figure~\ref{fig:flashlog_page}. We do not allow objects in the flashlog to cross
-a 4KB boundary - linking multiple objects together with a special flag encoded
-into the size if necessary. We flush log segments out to the log file using
-asynchronous direct IO, and therefore we fill one log segment buffer while
-keeping IO to another one outstanding to the flash drive. To avoid having to do
-a synchronous metadata flush, we preallocate (using the posix\_fallocate call) a
-gigabyte worth of zero filled disk pages at the end of the log file whenever we
-hit its end.
-
-In order to recover from a crash, we make two important assumptions about the
-underlying SSD. First, we assume that 4KB is the \emph{minimum} atomic unit for
-updating pages on the SSD even under power failure i.e. there are no shorn
-writes (otherwise known as torn writes) on a 4KB page~\cite{shorn_writes}. We
-also assume that the SSD has power loss data protection meaning that writes
-cached in the drive's volatile cache are written to the SSD using a backup
-capacitor in the event of power failure - a property of many data center class
-SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
-evaluation. Together these two assumptions mean that we can recover a consistent
-prefix of the log on a power failure.
-
-We move log entries from the head of the NVM log to the flashlog buffers in FIFO
-order. The NVM log entry is only actually removed when the IO for the
-corresponding flashlog page is complete. This means that during recovery we can
-have the same log entry both in the NVM log and the flashlog, a condition that
-can be detected by examining the RAFT related information that is embedded as
-part of the logged packet.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
-  \caption{Flashlog segment}
-  \label{fig:flashlog_page}
-\end{figure}
 

### 2017-07-24 paper/paper.tex
More updates.

@@
 
-\subsection{Eliminating Data Movement}
-
-Another key problem we had to solve with DPDK was teaching it about persistent
-memory. DPDK allows one to plug in a memory allocator. We required the allocator
-to support various features such as allowing multiple entities (the cpu core and
-NIC driver) to hold reference to DPDK packets and allocate/free in parallel.
-
-The NVM log is maintained as a circular log of fixed sized pointers (a pointer
-log) to memory buffers - each of which holds a client request. This arrangement
-is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of such
-an arragement. Adding a level of indirection in the NVM log allows us to
-separate the FIFO ordered circular log being manipulated by the RAFT from
-variable sized client being managed by the memory allocator of the packet
-switching library we use. Another advantage is that it makes recovery from NVM
-easy - appends to the circular pointer log are atomic and we can use the pointer
-log to recover allocator state i.e. what pieces of NVM are currently in use by
-the log.
+We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
+concurrent memory allocator based on reference counting - used by both the NIC
+driver and CPU cores. This means that packet allocation and deallocation does
+not happen in the same order as their corresponding position in the replicated
+RAFT log. To deal with this, we use a level of indirection as shown in
+Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
+sized pointers. Adding a level of indirection in the NVM log allows us to
+separate the FIFO ordered circular log being manipulated by RAFT from packet
+data being managed by the memory allocator of DPDK. Both the circular log and
+packet data are in NVM. An advantage of this scheme is that it makes recovery
+from NVM easy - appends to the circular pointer log are atomic and we can use
+the pointer log to recover allocator state i.e. what pieces of NVM are currently
+in use by the log.
 
@@
 
-We now describe how we implement the RAFT consensus protocol to replicate the
-NVM log across a set of replicas. The software packet switching idea we apply is
-to separate the control plane from the data plane using the Data Plane
-Development Kit (DPDK~\cite{dpdk}) library for direct userspace access to the
-network interface.
-
-\subsubsection{Dataplane Separation}
-We use RAFT~\cite{raft} as the underlying protocol for log replication. RAFT
-replicates a log across a set of replicas. Its operations are leader-based and
-RAFT incorporates a leader election algorithm that elects a leader with the most
-up to date log. The leader receives log entries from clients, appends them to the
-persistent log (NVM log in this case) and multicasts the new entry to follower
-replicas. On a response from enough follower replicas (constituting a majority
-quorum) the leader replica can then execute the command in the log entry. In
-our case this involves accessing the key value store and sending the response
-back to the client.
-
-We implement the common case flow in RAFT as a software packet switch using
-DPDK. The dataplane now encompasses the Network Interface Card (NIC) and the CPU
-caches with the NIC placing packet data directly in the CPU cache via
-DDIO~\cite{ddio} - a standard feature of most NICs today. The dataplane for the
-RAFT protocol is therefore separated from the control plane. The control plane
-runs in a thread pinned to a dedicated CPU core and takes care of the protocol
-state machine: checking for correct term, log index and that the replica
-receiving the packet from the client is indeed the leader. On the follower side
-it is responsible for checking that the leader is in the correct term before
-accepting the packet by appending it to the NVM log and initiating a
-response. In addition, the control plane is responsible for uncommon case
-activities such as dealing with timeouts and leader elections.
-
-The pseudocode in Figure~\ref{fig:control_plane} describes part of the control
-plane in Cyclone organized as event handlers triggered on receiving a packet at
-the leader. We focus on one event for illustration (and brevity): the event
-where a request is received from a client. A key concern when we built Cyclone
-was adhering as far as possible to the software packet switching principle of
-separating the control plane from the dataplane. This means that the CPU should
-avoid examining every byte of the packet as far as possible.
-
-The sequence of operations when a packet is received from the client needs
-careful design for efficient software packet switching. To illustrate how this
-is achieved, Figure~\ref{fig:packet_layout} shows how Cyclone manipulates packet
+\subsection{Eliminating Data Movement}
+Software packet switches built on DPDK try to touch as little of the packet data
+as possible, minimizing movement of data up and down the cache and memory
+hierarchy. Log replication looks like multicast that does not require deep
+packet inspection. Software packet switches implementing multicast therefore
+simply manipulate packet headers to produce new packets to send on the output
+ports. We designed our implementation of raft around the same principle.
+
+\begin{figure}
+\begin{verbatim}
+    event_receive_client_req()
+    {
+      if(!check_is_leader()) {
+        drop_msg
+        return
+      }
+      Prepend raft log term and index
+      Persist to local log
+      Transmit to follower replicas
+    }
+\end{verbatim}
+\caption{Cyclone event handling}
+\label{fig:control_plane}
+\end{figure}
+
+The pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
+handling code in Cyclone organized as event handlers triggered on receiving a
+packet at the leader. We focus on only one key event for brevity: the event
+where a request is received from a client. The first step is to check that this
+replica is indeed the RAFT leader in the current view (term). If not, the
+message is simply dropped (a timeout causes the client to try a different
+server). It then prepends RAFT related information to the packet - this includes
+the current term and log index for this entry. Next, it appends a pointer to the
+log in Figure~\ref{fig:nvm_log} effectively appending the packet to the
+persistent log. Finally, it transmits the packet to follower replicas.
+
+This entire process is done without making any copies of the received packet.
+To illustrate how this is achieved, Figure~\ref{fig:packet_layout} shows how
+Cyclone manipulates packet
 layouts across the two steps of prepending a RAFT header and transmitting to
@@
 \begin{figure}
-\begin{verbatim}
-    event_receive_client_req()
-    {
-      if(!check_is_leader()) {
-        drop_msg
-        return
-      }
-      Prepend raft log term and index
-      Persist to local log
-      Transmit to follower replicas
-    }
-\end{verbatim}
-\caption{Cyclone control plane fragment}
-\label{fig:control_plane}
-\end{figure}
-
-\begin{figure}
   \centering
@@
 
-We note that by discarding stream oriented protocols such as TCP, the network
-layer is no longer reliable or ordered. This is not a problem for RAFT that is
-designed to tolerate network failures resulting in drops or reordering of
-network packets. However, this rules out consensus protocols such as Zookeeper
-Atomic Broadcast (ZAB~\cite{zab}) that depend on a reliable stream-oriented
-network layer to function correctly.
-
 Finally, we turn our attention to the persistence step in
 Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
-before it is multicast out to follower replicas. Since the NVM is directly
-attached we do this by executing a cacheline flush ({\tt clflush}) instruction
-for every cacheline in the packet and the pointer in the pointer buffer to
-persist these via the memory bus. This is the one exception to our otherwise
-clean separation between the control and data plane - necessitated by the fact
-that at the moment NICs do not support persistent memory. This is not too
-onerous a burden because we can use the newly introduced {\tt
-  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
-efficiently flush to persistent memory without the overhead of the serialization
-normally introduced by {\tt clflush}. This allows us to hit full memory
-bandwidth on present generation platforms, a quantity in excess of 200 Gb/s per
-core, well above the near term speeds of network interface cards.
+before it is multicast out to follower replicas. DPDK userspace NIC drivers
+operate in DDIO mode~\cite{ddio} where the packet is directly written into the
+CPU cache rather than first being DMAed into DRAM and then fetched by the CPU on
+demand. Since the NVM is directly attached we need to persist the packet to it
+by executing a cacheline flush ({\tt clflush}) instruction for every cacheline
+in the packet and the pointer in the pointer buffer to persist these via the
+memory bus. This is not too onerous a burden because we can use the newly
+introduced {\tt clflush-opt}~\cite{clflush_opt} instruction specifically
+intended to efficiently flush to persistent memory without the overhead of the
+serialization normally introduced by {\tt clflush}. This allows us to hit full
+memory bandwidth on present generation platforms, a quantity in excess of 200
+Gb/s per core, well above the near term speeds of network interface cards. We
+execute a single serializing {\tt sfence} before the sequence of {\tt clflushes}
+to make sure any dirty cacheline due to header manipulation related writes from
+the CPU are sent to cache.
 
@@
 
-\subsubsection{Batching}
-Although the control plane code only runs operations related to the RAFT state
-machine it is still slow enough relative to the dataplane to add significant
-overhead for each packet. We tackle this problem by applying another classic
-software packet switching technique: batching - illustrated in
-Figure~\ref{fig:batching}. We use a burst receive call available in the DPDK
-userspace driver to receive a burst of client packets at a time. We then chain
-these packets together and treat them as a single log entry from the perspective
-of RAFT, amortizing the control plane overheads over the packets (at most 32 at
-a time due to current driver limitations).
-
-We note that batching in Cyclone does not involve a latency-throughput
-tradeoff like in many other systems~\cite{ix-dataplane}. The batch receive call
-we use in DPDK returns immediately with whatever number of packets is available,
-including zero. We always flush the transmit buffer after every call to DPDK to
-transmit packets to replicas. Therefore, we never tradeoff latency for
-throughput when batching.
-
-\subsection{Flash Log}
-
-The fact that we can immedately absorb updates into an NVM log and only move
-them to the flashlog in suitable units for block IO sidesteps a common problem
-with logging to secondary storage - the need to do group commit. Logging systems
-often wait to collect a batch of entries to write to secondary storage trading
-latency for throughput. This is unnecessary in Cyclone as the use of directly
-attached NVM with a two level log removes the need to make this tradeoff.
+Although RAFT is an efficient consensus protocol in the common case, the
+protocol state machine still adds significant overhead to each packet, relative
+to the time for the packet to flow in from the NIC and back out to the replicas.
+We address this problem for the \emph{loaded} case using batching - treating a
+whole sequence of client commands as a single RAFT log entry, while avoiding any
+copies to group these packets together. Figure~\ref{fig:batching} illustrates
+how this is done. We use a burst receive call available in the DPDK userspace
+driver to receive a burst of client packets at a time. We then chain these
+packets together and treat them as a single log entry from the perspective of
+RAFT, amortizing the control plane overheads over the packets (at most 32 at a
+time due to current driver limitations). Crucially, batching in Cyclone does not
+involve a latency-throughput tradeoff like in many other
+systems~\cite{ix-dataplane}. The batch receive call we use in DPDK returns
+immediately with whatever number of packets is available, including zero. We
+always flush the transmit buffer after every call to DPDK to transmit packets to
+replicas. Therefore, we never tradeoff latency for throughput when batching.
+
+\section{Parallelism}
+\label{sec:parallelism}
 
@@
 \end{figure}
-
-\section{Parallelism}
-\label{sec:parallelism}
 

### 2017-07-24 paper/paper.tex
Updates.

@@
 
-Finally, we turn our attention to the persistence step in
+We now turn our attention to the persistence step in
 Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.35]{figures2/race.pdf}
-  \caption{Race with multiple physical logs}
-  \label{fig:race}
-\end{figure}
+%\begin{figure}
+%  \centering
+%  \includegraphics[scale=0.35]{figures2/race.pdf}
+%  \caption{Race with multiple physical logs}
+%  \label{fig:race}
+%\end{figure}
 
-\section{Horizontal Scaling}
-\label{sec:horizontal}
-The basic principle in Cyclone is to turn log replication into a software packet
-switching problem. A key ingredient in software packet switching is applying
-multiple CPU cores to parallelize the control plane, scaling message processing
-throughput. In this section, we describe how we can horizontally scale Cyclone's
-single \emph{logical} log across multiple \emph{physical} logs. For each
-physical log, we run an independent version of RAFT with the corresponding
-control plane mapped to a dedicated CPU core. Further, we assume a concurrent
-key value store where a set of CPU cores independently run operations from the
-physical logs on the shared memory key-value store data structure, synchronizing
-among themselves as necessary.
+Althought the work described in the previous section significantly boosts the
+network performance of replication in Cyclone, it still does not come close to
+saturating the capabilities of even the commodity network interface card that we
+use. This is because NICs today encapsulate significant parallelism in terms of
+multiple send and receive queue pairs. Exploiting this capability to improve
+throughput requires us to remove the bottleneck of a single sequential log in
+RAFT. To this end, we extend the implementation described thus far to run
+multiple copies of the RAFT consensus protocol each maintaining and replicating
+its own two level log. We now refer to these logs as physical logs. All these
+instances however exist in the same process address space as the key value store
+application itself and therefore manipulate the same application.
 
-We begin by showing how we divide CPU cores and NIC resources on the plaform across
-replication and key value store work, and show how a KV store request flows
-across them.  We then describe how a key-value store API can be mapped to
-independent physical logs without divergence in replica state. Finally, we
-describe special handling in Cyclone for multi-key operations.
+The first key question we deal with is - how do Cyclone clients decide which
+physical log to send a {\tt PUT} or {\tt GET} request to ? A key constraint here
+is that all replicas should agree on the seqeunce of updates to a key - as
+otherwise they would diverge in terms of state. Allowing updates to the same key
+to go to different physical logs would lead to a divergence as there is no
+guarantee about the order in which the independent RAFT instances would apply
+the update. Therefore all updates to a key must go to the same physcial
+log. A similar constraint applies to reads in order to ensure that any committed
+writes are seen by readers. Allowing reads to a key to go to a different RAFT
+instance than the write means that it could be served on a machine where the
+write is yet to be applied but where the machine is the leader for the RAFT
+instance. All reads and write to a key must therefore go to the same physical
+log. We achieve this by hashing the key to select the physical log.
 
-\subsection{Core Allocation}
-A key-value store integrated with Cyclone runs two types of threads. The first
-type of thread does key-value store work. This includes servicing lookup and
-update requests and background threads that flush in-memory data structures to
-secondary storage and do maintainence work on secondary storage data structures
-such as compacting log-structured merge trees~\cite{lsmtree}. The second type of
-thread is a dedicated thread per-physical log to run the corresponding control
-plane including the associated RAFT state machine. We use static assignment of
-threads to cores as recommended for packet switching libraries such as DPDK for
-the best possible performance. Therefore we use the terms thread and core
-interchangeably in this paper. Client requests are always received by the
-control plane core for the physical log the request is directed to. A response
-for the core is always sent by the KV store core that finally handles the
-request.
+It turns out that this simple \emph{necessary} scheme is also \emph{sufficient}
+to ensure linearizability in the face of failure. Clearly the system is
+linearizable for a single key since it involves only a single physical log. In
+addition, one can switch operations to different keys in a history without
+affecting the correctness of the history because operations to different keys
+are commutative. Any history of the system is therefore linearizable.
 
-An important resource for Cyclone is the NIC that must be shared among the
-control plane cores that receive and replicate requests and the KV store cores
-that send back responses. Most NICs provide independent transmit/receive queue
-pairs and we aim for synchronization free access to these to aid horizontal
-scalability with respect to the network resource. DPDK already provides fairly
-scalable access to NIC queues, with the queue number being a parameter to send
-and receive calls. All that remains for us is to map the queues to the control
-plane and KV store cores, keeping in mind that there may be multiple NICs on the
-system. We do this by iterating over all control plane cores followed by all KV
-store cores. We assign two queue pairs to each control plane core, one to
-receive requests from the client and the other to transmit and receive
-replication protocol related messages. We assign one queue pair to each KV store
-core to send back responses to clients. We are careful to spread the different
-queue pairs for a core across the available NICs as far as possible. This is
-important due to the asymmetry between the different types of traffic. Queue
-pairs carrying replication traffic will likely be more busy than those carrying
-client request responses due to the amplification in message count to multicast
-the same message to different replicas.
+Multiple physical logs in Cyclone operate in a shared nothing manner by
+partitioning the NVM and SSD space evenly between them and by allocating
+dedicated NIC queue pairs to each instance of RAFT. This works well because the
+memory hierarchy including the NVM and the SSD efficiently support concurrent
+operations. The only synchronization necessary is when doing reads or writes to
+the single shared key value stores.  The level of concurrency therefore is
+constrained only by the concurrency available in the software architecture of
+the persistent key value store itself, for which a number of good designs
+already exist~\cite{flodb}.
 
-The flow for an update operation in Cyclone is as follows. The network packet is
-received by the control plane for the targeted physical log. The packet is then
-immediately sent out for replication as described in the previous section. When
-responses from a majority quorum is received, the request is passed to the
-targeted KV store core for execution. The KV store core processes the request
-and sends back the response to the client. We note that replication is pipelined
-and decoupled from execution and thus multiple updates are usually in flight on
-the network for replication at the same time.
-
-The flow for a read operation is simpler. It is still targeted to a physical log
-but instead of replicating it, the control plane core simply passes it on
-straight away to the KV store core for execution and response.
-
-The control plane cores communicate with the KV store cores through a one-way
-lock-free FIFO queue passing pointers to packets. We leverage the convenient
-ring library in DPDK for this.
-
-Finally, an asynchronous source of work in the system is maintenance of the
-two-level log. We depart from a 1:1 mapping between the first level NVM log and
-the second level flashlog here. Each physical log consists of one NVM log
-managed by the corresponding control plane core and multiple flashlogs each of
-which is managed as an additional responsibility by a KV store core. We
-therefore enforce a many to one mapping between key value store cores and a
-front end control plane core - thereby partitioning KV store cores across
-control plane cores. A request picks a control plane core and any one of the
-associated KV store cores for execution. The request is logged in the NVM log of
-the control plane core and garbage collected once it is drained to the flashlog
-of the selected KV store core.
-
-\subsection{Leveraging API Commutativity}
-Update requests to a key-value store include a key and an associated operation -
-such as ``put'' that sets the value for the key or ``merge'' that calls a
-user-defined operator to merge a new value for the key into the existing value.
-A naive approach that maps a key to a random physical log and subsequently to a
-random core on the system for execution runs into the determinism problem -
-replicas could diverge with respect to the final value assigned to a key on
-different replicas. We illustrate this problem with an example in
-Figure~\ref{fig:race} where two updates to the same key race with each other on
-different physical logs and execution cores to end up being applied in different
-orders on different replicas, leading to a divergence of state.
-
-The solution to this problem is observe that the key-value store is a concurrent
-data structure that ultimately admits a serialization of operations to
-it. Further, consecutive operations to different keys in the serial history can
-be interchanged, because operations to different keys are
-\emph{commutative}. The correct way to use different physical logs is therefore
-to simply hash the key to select the physical log \emph{and} the core on which
-the operation executes. This guarantees determinism as the serial history of
-two different replicas can be transformed into the same serial history by
-reordering operations to different keys. Operations to the same key are already
-in the same order in the serial histories by virtue of being replicated on the
-same physical log and being executed on the same KV store core.
-
-Further relaxations to the mapping scheme are possible when operations to the
-same key are also commutative. A good example is when key value stores are used
-to maintain counters. Different increment operations to the same counter are
-commutative if one is prepared to tolerate non-determinism in the sequence of
-updates seen by a reader with the counter reaching the same eventual value. This
-specific use-case, in fact, lead to the addition of the merge operation in
-Rocksdb, a key value store that we use for evaluation. Merge operations
-therefore could be dispatched to a random physical log with no harm. For this
-paper however, we stick to the stricter case that operations to the same
-key are not commutative. However, Cyclone's client side libraries leave it to
-the user to specify the exact physical log and key-value store core to use in
-order to enable more relaxed mappings when necessary.
-
-We note that commutativity as assumed by Cyclone is an \emph{API level
-  property}. There is no assumption as to how application cores synchronize to
-access the shared key-value state or the underlying data structures for holding
-key-value pairs. We emphasize that this means replicas ultimately have the same
-content but not necessarily the same memory state.
-
-Finally, Cyclone provides seraializable reads but does not - at the moment -
-guarantee the ability to read one's own writes, repeatable reads or more
-generally, linearizable reads. The reason is that on a failover the new leader
-replica might still have update operations in its NVM logs that are yet to be
-applied to the key value store. However the original failed leader replica might
-have applied these updates and a client might have already observed them. Users
-who desire linearizable reads can issue a special API call to ensure that reads
-are also routed through the logs. Flushing the NVM logs on failover is a better
-solution that we are currently in the process of engineering into Cyclone.
+In our discussion thus far, we have focused on operations to a single
+key. Supporting atomic writes or reads to multiple keys while maintaining
+linearizability requires special handling with multiple physical logs so as not
+to break linearizability - that we discuss next.
 
 \subsection{Ganged Operations}
-Key value stores (such as Rocksdb) usually provide a batched write operation
-that atomically updates a set of key-value pairs. A batched write is often a key
-primitive for building more sophisticated operations such as
-transactions. Batched writes require special handling in Cyclone due to the
-requirement that we route a write request by hashing the key to select both the
-control plane core (physical log) as well as KV store core. A batched write has
-multiple keys and therefore we need to synchronize the progress of the operation
-across multiple physical logs (and associated replication quorums) as well as KV
-store cores. In effect, the serial history contains a grouped set of operation
-on multiple keys as a single atomic operation. All serial histories across
-replicas therefore continue to map to the same equivalent serial history modulo
-reordering of operations to different keys.
+Supporting reads or writes to multiple keys that map to different physical logs
+requires a ganged operation across those logs such that all the operations
+complete together in the face of failures. Since our starting point is a single
+key value store, we constrained Cyclone's design early on to avoid the need for
+any two phase commit.
 
-One way to do this is to use a distributed transaction, using
-two-phase commit to couple the otherwise independent replication quorums and KV
-store cores. However this method has a number of problems. First, it introduces
-an additional network round trip that is normally not necessary for a key value
-store with a single shared memory image holding all the keys. Second, handling
-failures with distributed transactions would have caused an explosion in
-complexity in Cyclone, something we wished to avoid in the interests of
-practical deployment to independent end users who need to understand and 
-maintain Cyclone as part of their embedded KV store infrastructure.
+The first key problem we need to deal with is that the leaders for the different
+RAFT instances can be located on different machines. To avoid having a client
+co-ordinate different machines for a single operation, we constrain the RAFT
+leader election algorithm for the different RAFT instances to converge to the
+same machine. We do this by triggering re-elections for a RAFT instance other
+than that of the first physical log until it is on the same machine as the
+\emph{first} physical log. This process is therefore resilient to failed
+machines. RAFT requires that the leader have the most up to date log. To ensure
+that the process converges we therefore ensure that the currently elected leader
+brings a majority quroum up to date before triggering a relection. We also do
+not accept any requests from clients until all leaders are co-located. This
+process is guaranteed to converge as long as all no network partitions occur
+\emph{within a machine} i.e. some RAFT instances are able to send and receive
+packets, while others are not, a possiblity we discount due to our single
+process design.
 
-Our solution instead is to use a technique we term ganged operations. It is
-based on the observation that in the failure free case, executing a batched
-write across multiple control plane cores and KV store cores does not look very
-much different from a synchronous shared memory barrier. The challenge is to
-effectively manage failures in replication without causing participating cores
-to stop making progress. The first step is to reconfigure RAFTs leader election
-protocol to ensure that the leaders for all the physical logs are co-located on
-the same host. Clients always dispatch batched writes to a fixed control plane
-core (called the injection core) which is then responsible for forwarding the
-request to the participating control plane cores (for corresponding physical
-logs). The injection core uses DPDK's packet cloning primitives to avoid
-dataplane work while forwarding copies of the same batched write packet to
-multiple control plane cores. The injection core also adds a ``nonce'' - a
-unique timestamp to the packet. In addition, the injection core adds a unique
-view number to the packet containing the term numbers of all the participating
-RAFT instances (read from shared memory of the co-located leaders).  The flow
-for ganged operations is described in the pseudocode of
-Figure~\ref{fig:ganged_ops}. We defer the discussion of how we generate the
-nonce to later in this section. We also assume that a unique barrier is
-allocated in shared memory for each ganged operation. We discuss later how this
-is done without using dynamic allocation.
+Next, we need to simultaneously inject the ganged operation into all
+participating physical logs on the machine hosting the leaders. Clients always
+dispatch batched writes to a fixed co-ordinating physical log/RAFT instance
+(henceforth called the co-ordinator), which is then responsible for forwarding
+the request to the participating logs. We make use of packet cloning primitives
+to avoid making a physical copy of the packet, generating indirect references
+instead. The co-ordinator also adds a ``nonce'' - a unique timestamp to the
+packet. In addition, the co-ordinator adds a unique view number to the packet
+containing the term numbers of all the participating RAFT instances (read from
+shared memory of the co-located leaders).  The event that applies a ganged
+operation is described in the pseudocode of Figure~\ref{fig:ganged_ops}. We
+defer the discussion of how we generate the nonce to later in this section. We
+also assume that a unique barrier is allocated in shared memory for each ganged
+operation. We discuss later how this is done without using dynamic allocation.
 
-In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. Once
-replication is complete on all participating physical logs, the corresponding KV
-store cores execute \\{\tt event\_replicated\_ganged\_op}. They use a bitmask
-for the barrier setting the appropriate bit for themselves. Once all KV store
-cores have arrived at the barrier the operation is executed. A distinguished
-leader core (we use the minumum numbered participating KV store core) is
-responsible for sending back the response to the client. Two types of failure
-need to be considered. First, replication on the physical log corresponding to
-one of the non-leader cores could fail. The remaning cores would never progress
-past the barrier. To solve this the leader core monitors the remaning cores for
-their current view - which is the RAFT term of the last executed log entry. If
-the view has moved past the view in the packet, replication failed on that
-physical log. The leader then sets the bit for the core that will no longer
-participate in the barrier to aid progress (a technique we borrowed from
-lock-free data structures~\cite{lock-free}). It also sets a failed flag to
-signal failure to the other participating cores. The second failure case is that
-the leader core might not receive the operation if it fails during replication
-from corresponding physical log. This case is detected in a similar manner by
-the remaining cores by monitoring the published view from the leader core in
-shared memory. We add that on a leadership change due to timeout we enqueue and
-execute a nop on the physical log and all associated KV store cores to ensure
-previous entries are committed - a corner case in the RAFT protocol.
+The complexity in Figure~\ref{fig:ganged_ops} arises from the need to handle
+both shared memory concurrency and failure during replication. Without failure,
+Figure~\ref{fig:ganged_ops} is straightforward. A barrier is executed on all
+RAFT instances. Once replication is complete on all participating physical logs
+(or this is a read ganged operation), a state indicated by all necessary bits
+being set in the barrier mask, the RAFT instances simultaneously execute the
+ganged operation.
+
+Each RAFT instance always publishes the current term it is operating in. A
+failure causes a new leader to be elected moving forward the term. If the
+participants of a ganged operatation do not detect this case, they could be left
+waiting for ever as ganged operation might never successfully replicate in the
+failed over RAFT instance. We detect failure on any RAFT instance by having each
+instance publish its current term and continuously comparing the view in the
+ganged operation to it. If any RAFT instance has moved past that term, the
+ganged operation is then terminated by having one of the still alive participants set a
+failure flag and write the appropriate bit in the barrier mask on behalf of the
+failed participant, moving execution past the barrier wait - a technique we
+borrow from lock-free data structures~\cite{lock-free}.
 
@@
 \begin{verbatim}
-    // Control plane core
-    event_recv_ganged_op(packet)
+    // Apply ganged operation
+    event_apply_ganged_op (packet, barrier)
     {
-      Begin replication of packet
-    }
-
-    // KV store core
-    event_replicated_ganged_op (packet, barrier)
-    {
-      if(leader KV core)
+      if(co-ordinator)
         atomic set bit me in barrier.mask         
         do
-         for each KV store core in packet
-          if core_public_data[core].view > packet.view
+         for each participant p in operation
+          if public_data[p].view > packet.view
            barrier.failed = true 
-           atomic set bit for core in barrier.mask
-        while barrrer.mask != mask of participating KV cores
+           atomic set bit for p in barrier.mask
+        while barrrer.mask != mask of all participants
         if barrier.failed
@@
         wait until 
-         core_public_data[leader core].view > packet.view OR
+         public_data[co-ordinator].view > packet.view OR
          barrier.mask == mask of participating KV cores
-        if core_public_data[leader core].view > packet.view OR
+        if public_data[co-ordinator].view > packet.view OR
            barrier.failed
@@
 mapped to a unique barrier. We achieve this by using a fixed piece of memory
-owned by the leader core to hold the barrier and write the nonce to it in order
-to indicate that the barrer is active for the corresponding batched write
-operation. Non-leader cores watch for the nonce to know when to execute the
+owned by the co-ordinator to hold the barrier and write the nonce to it in order
+to indicate that the barrer is active for the corresponding ganged
+operation. Participants watch for the nonce to know when to execute the
 ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
 the leader's published view to detect the case where the ganged operation fails
-to replicate on the leader's physical log.
+to replicate on the co-ordinator's physical log.
 
 Finally, we describe how we generate the nonce. The nonce is generated on the
-injector core by concatenating the ethernet MAC ID of the first NIC on the
-system with a 64 bit value that is the number of CPU timestamp counter cycles
-since epoch time (read from the real time clock at startup plus the number of
-cycles from the CPU {\tt rdtsc} instruction). The nonce can only be repeated if
-the same machine manages to fail and come back up in less time than the real
-time clock drift (controlled with NTP), a possibility that we discount.
+co-ordinator RAFT instance by concatenating the ethernet MAC ID of the first NIC
+on the system with a 64 bit value that is the number of CPU timestamp counter
+cycles since epoch time (read from the real time clock at startup plus the
+number of cycles from the CPU {\tt rdtsc} instruction). The nonce can only be
+repeated if the same machine manages to fail and come back up in less time than
+the real time clock drift (controlled with NTP), a possibility that we discount.
 
@@
 code weighs in at well under a couple of hundred lines. We believe that
-this additional complexity is still small compared to distributed transactions.
+this additional complexity is still small compared to distributed transactions
+using two-phase commit that also need to recover from failure used in systems
+such as FARM~\cite{farm}.
 

### 2017-07-26 paper/paper.tex
Updates.

@@
 \begin{abstract}
+\vspace{0.1in}
 Persistent key value stores are rapidly becoming an important component of many
@@
 Cyclone is a high speed strongly consistent replicated write ahead logging
-service specialized for key value stores such as Rocksdb. We describe Cyclone's
-system architecture in Section~\ref{sec:sysarch}. Cyclone provides the order of
-magnitude better performance required to close the performance gap in
-Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
-\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the host
-- such as in the form of NVDIMMs~\cite{farm} or directly attached persistent
-memory~\cite{pmfs}. Cyclone logs update operations to the key value store and
-calls on the key value store to execute the update only when it has been
-successfully replicated to majority quorum of replicas. All reads and write to
-the key value store are always executed by the leader replica. A client side
-stub provided by Cyclone automatically locates and directs read and write
-requests to the leader replica, ensuring linearizable semantics from the key
-value store. The novelty in Cyclone is that it accomplishes this without
-compromising on the performance baseline of no logging or replication. Cyclone
-does this by systematically addressing the sources of performance degradation in
-Figure~\ref{fig:problem}.
+service specialized for key value stores such as Rocksdb. Cyclone entirely
+closes the performance gap due to both the storage and the network shown in
+Figure~\ref{fig:problem}. We show that Cyclone achieves performance transparent
+replication in that it can provide both a persistent write ahead log and
+replication without compromising on RocksDB's performance.
 
-The primary cause of slowdown when turning on a write ahead log on a single
-machine is the need to do synchronous block IO for every update. For even
-moderately sized updates, this leads to severe write amplification as a 4KB page
-must be synchronously written to the log file on flash. Cyclone avoids this by
-making use of directly attached persistent memory on the server. This can be in
-the form of NVDIMMs (DIMMs with an added ultracapacitor to dump state to a small
-amount of attached flash) or newer forms of persistent memory~\cite{pmfs}. For
-current deployments, NVDIMMs are expensive - retailing at approximately 2X the
-price of a standard DIMM [find ref]. This means that they will likely be
-available in small quantities. We compensate for this by periodically draining
-the NVDIMM log into a log placed on a standard SSD on the system. We describe
-this two level log structure in Section~\ref{sec:storage}.
+The first key contribution of Cyclone is in showing how a small amount of
+non-volatile memory can be combined with the flash SSD based persistent logs
+when providing high availability.  The primary cause of slowdown when turning on
+a write ahead log on a single machine in Figure~\ref{fig:problem} is the need to
+do synchronous block IO for every update. For evenmoderately sized updates, this
+leads to severe write amplification as a 4KB page must be synchronously written
+to the log file on flash. Cyclone avoids this by making use of directly attached
+persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
+an added ultracapacitor to dump state to a small amount of attached flash) or
+newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
+are expensive - retailing at approximately 2X the price of a standard DIMM [find
+  ref]. This means that they will likely be available in small quantities. We
+compensate for this by periodically draining the NVDIMM log into a log placed on
+a standard SSD on the system.
 
-The network overheads for replicating a log entry to a quorum of machines stems
-from the overhead of the TCP/IP stack in the linux kernel - itself the topic of
-much research [dataplane OS stuff]. We therefore replase the TCP/IP stack with
-DPDK, a userspace network library that allows direct IO of raw ethernet frames
-to and from the network interface card (NIC) eliminating all overheads. We no
-longer use TCP - since the rare network errors encountered in a typical
-datacenter network are automatically handled by the replication protocol (RAFT)
-as part of its failure model. Finally, we carefully design our implementation of
-RAFT to relieve the CPU of all data copy operations, leaving the CPU to only
-run the protocol and return an accept/reject decision for each received
-packet. The set of optimizations to the network stack is decribed in
-Section~\ref{sec:network}.
+The second key contribution of this work is in showing that the performance
+overheads in the network stack can be mitigated entirely in software. Existing
+work on log replication using state machine replication protocols such as
+Paxos~\cite{paxos} and RAFT~\cite{raft} has not - thus far - addressed high
+performance replication in the local area network of a datacenter using
+commodity networks.  Research focusing on addressing wide area network latencies
+by reducing the number of network hops with new protocols such as Fast
+Paxos~\cite{fast-paxos} or network switch modifications such as No
+Paxos~\cite{nopaxos} do not address throughput bottlenecks due to network stack
+related inefficiencies in the participating nodes of the quorum - a more urgent
+problem with local area replication. On the other hand, work done to address
+network related inefficiencies start with the assumption that the bottleneck to
+good throughput on the network is the CPU and therefore one must either resort
+to offload mechanisms such as RDMA~\cite{dare, farm, faast} or offload the
+protocol entirely to an FPGA~\cite{consensus_box}. We demonstrate in this paper
+that this is not the case. Rather, once persistence is provided by directly
+attached non-volatile memory eliminating block IO from the persistence step in
+replication protocols, the problem becomes equivalent to multicasting the same
+log entry from the leader to all followers. We design Cyclone's network stack
+around well known principles in the software packet switching community. In
+particular, a careful implementation of a consensus protocol that relieves the
+CPU of the responsibility of data movement permits a high performance
+implementation entirely in software that addresses the order of magnitude
+network related performance gap in Figure~\ref{fig:problem} using only commodity
+10 Gigabit ethernet.
 
-Both the storage (NVM, SSD) as well as the NIC expose significant parallelism
-that is left unused when Cyclone is used to replicate a single log. To exploit
-this parallelism we use Cyclone to replicate multiple logs, each using a
-different instance of the RAFT consensus protocol. We show how to exploit the
-commutativity inherent in key-value store interfaces to make use of this
-parallelism in Cyclone's design in Section~\ref{sec:parallelism}.
-
-A detailed evaluation of Cyclone's replication is provided in the context of
-RocksDB - a popular persistent key value store - in
-Section~\ref{sec:evaluation}. We show that Cyclone achieves performance
-transparent replication in that it can provide both a persistent write ahead log
-and replication without compromising on RocksDB's performance.
+The rest of the paper is organized as follows. We describe Cyclone's system
+architecture in Section~\ref{sec:sysarch}. We describe this two level log
+structure in Section~\ref{sec:storage}. The set of optimizations to the network
+stack is decribed in Section~\ref{sec:network}. Both the storage (NVM, SSD) as
+well as the NIC expose significant parallelism that is left unused when Cyclone
+is used to replicate a single log. To exploit this parallelism we use Cyclone to
+replicate multiple logs, each using a different instance of the RAFT consensus
+protocol. We show how to exploit the commutativity inherent in key-value store
+interfaces to make use of this parallelism in Cyclone's design in
+Section~\ref{sec:parallelism}.  A detailed evaluation of Cyclone's replication
+is provided in the context of RocksDB - a popular persistent key value store -
+in Section~\ref{sec:evaluation}. We then discuss related work before concluding.
 

### 2017-07-27 paper/paper.tex
More updates.

@@
 from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
-testbed from {\tt 24 us} to as low as {\tt 7 us}, providing a significant boost
+testbed from {\tt 24 us} to as low as {\tt 5 us}, providing a significant boost
 to performance.
@@
 Cyclone takes a software only approach to improving replication performance
-using commodity network hardware. At the other end of the spectrum, Consensus in
-a box~\cite{consensus_box} implements the Zookeeper atomic broadcast protocol
-together with a NIC on an FPGA. That work ignored durability focusing purely on
-replication performance. Latency numbers are impressive due to the FPGA that cuts
-out the path from CPU caches to the NIC and are reported to be as low as 3us for
-a round trip from leader to follower replica compared to the 7us we observe with
-DPDK. Notably however they reported 7us when using TCP/IP from the FPGA rather
-than their own connection oriented network protocol. On the other hand peak
-replication throughput appears to be 6M replications/sec on 66 Gbps of aggregate
-network bandwith, comparable to the mark obtained by Cyclone on an aggregate 40
-Gbps of network bandwidth. We believe the general applicability afforded by our
-software only packet switching solution makes it a compelling alternative to an
-FPGA based solution even given the larger latency.
+using commodity network hardware. In contrast, Mojim~\cite{mojim} pairs directly
+attached persistent memory with RDMA to replicate persistent data structures. On
+40Gbps infiniband, Mojim reported being able to mirror data at 4GB/s. In
+contrast, Cyclone can replicate data three ways using 4*10 Gbps commodity
+ethernet NICs at a peak of roughly 3.7 GB/s. Notably this is while Cyclone is
+also running the RAFT consensus protocol and draining the log to a high capacity
+flash SSD, while Mojim is doing simple mirroring of main memory. One can
+conclude that network offload via RDMA and Infiniband is not an a-priori
+requirement for high throughput replication. This is made possible by Cyclone's
+approach that relieves the CPU from the burden of data copies. Of course, the
+latency of a hop on commodity ethernet at 5 us with our DPDK stack is larger
+than the close to 2us latency with Infiniband. Depending on the use case for the
+key value store application the extra 6us with ethernet for the round trip to
+the replica might be a concern for some applications. We note that mapping the
+commutativity of key value store interfaces to multiple logs in Cyclone is a
+technique that can also be applied to RDMA. The authors of Mojim confirm that
+concurrency is a significant determinant of performance when replicating over
+RDMA.
 
-Network Ordered Paxos~\cite{nopaxos} reprograms top of rack switches to order
-client requests and multicast them to replicas - attacking replication latency
-by removing the need to direct requests through a leader replica. It is hard to
-compare performance due to the 1Gbps links used to leaf nodes in NoPaxos but we
-note that avoiding changes to switching infrastructure was a specific constraint
-in the design of Cyclone due to the difficulty we encountered in convincing
-public cloud customers to allow such changes to be made.
+Consensus in a box~\cite{consensus_box} implements the Zookeeper atomic
+broadcast protocol together with a NIC on an FPGA. That work ignored durability,
+focusing purely on network performance for replication. Latency numbers are
+excellent due to the FPGA that cuts out the path from CPU caches to the NIC and
+are reported to be as low as 3us for a round trip from leader to follower
+replica compared to the 6us we observe with DPDK. Notably however they reported
+7us when using TCP/IP from the FPGA rather than their own connection oriented
+network protocol. On the other hand their peak replication throughput is 6M
+replications/sec on 66 Gbps of aggregate network bandwith, comparable to the
+same mark obtained by Cyclone on an aggregate 40 Gbps of network bandwidth. We
+believe the general applicability afforded by our software only solution
+together with the fact that we provide persistence by design makes it a
+compelling alternative to an FPGA based solution even given the larger latency.
 
@@
 multithreading comes with high overheads for applications that store large
-amounts of data - CRANE reports around a 2X slowdown for MySQL likely rendering
+amounts of data - CRANE reports around a 2X slowdown for MySQL, likely rendering
 it unsuitable as a replacement for the write ahead log in key value stores.
-
-A number of researchers have proposed using high performance networking hardware
-for improving the performance of various distributed system primitives such as
-distributed transactions~\cite{farm} and replication~\cite{dare, mojim}. We note
-that the ideas in Cyclone are not specific to ethernet. The same ideas of packet
-switching and horizontal scaling can be equally applied to RDMA. However, we
-chose commodity ethernet due to short term deployment plans that are a
-priority design target.
 
@@
 
-Finally, we note that scaling a log horizontally for concurrency is also an idea
-adopted by filesystems. NOVA~\cite{nova} is an example that uses a per-inode log
-to improve concurrency when the filesystem is placed on NVM attached to the
-memory bus. Cyclone also uses a distinct NVM log per control plane core in a
-similar vein but does not maintain a namespace since the number of NVM logs is a
-small fixed number equal to the number of control plane cores.
-
 \section{Conclusion}

### 2017-07-27 paper/paper.tex
Updates.

@@
 
-\title{\Large \bf Cyclone: Warp Speed Replication for Key Value Stores}
+\title{\Large \bf Cyclone: Warp Speed Replication for Persistent Key Value Stores}
 
@@
 
-The first key contribution of Cyclone is in showing how a small amount of
-non-volatile memory can be combined with the flash SSD based persistent logs
-when providing high availability.  The primary cause of slowdown when turning on
-a write ahead log on a single machine in Figure~\ref{fig:problem} is the need to
-do synchronous block IO for every update. For evenmoderately sized updates, this
-leads to severe write amplification as a 4KB page must be synchronously written
-to the log file on flash. Cyclone avoids this by making use of directly attached
-persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
-an added ultracapacitor to dump state to a small amount of attached flash) or
-newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
-are expensive - retailing at approximately 2X the price of a standard DIMM [find
-  ref]. This means that they will likely be available in small quantities. We
-compensate for this by periodically draining the NVDIMM log into a log placed on
-a standard SSD on the system.
+The first key contribution of Cyclone is demonstrating the benefit of a small
+amount of non-volatile memory when combined with the flash SSD based persistent
+logs.  The primary cause of slowdown when turning on a write ahead log on a
+single machine in Figure~\ref{fig:problem} is the need to do synchronous block
+IO for every update. For small to moderately sized updates, this leads to severe
+write amplification as a 4KB page must be synchronously written to the log file
+on flash. Cyclone avoids this by making use of directly attached persistent
+memory on the server. This can be in the form of NVDIMMs (DIMMs with an added
+ultracapacitor to dump state to a small amount of attached flash) or newer forms
+of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs are expensive
+- retailing at approximately 2X the price of a standard DIMM [find ref]. This
+means that they will likely be available in small quantities. We compensate for
+this by periodically draining the NVDIMM log into a log placed on a standard SSD
+on the system.
 
-The second key contribution of this work is in showing that the performance
+The second key contribution of this work is in showing that the throughput
 overheads in the network stack can be mitigated entirely in software. Existing
@@
 using commodity network hardware. In contrast, Mojim~\cite{mojim} pairs directly
-attached persistent memory with RDMA to replicate persistent data structures. On
-40Gbps infiniband, Mojim reported being able to mirror data at 4GB/s. In
-contrast, Cyclone can replicate data three ways using 4*10 Gbps commodity
-ethernet NICs at a peak of roughly 3.7 GB/s. Notably this is while Cyclone is
-also running the RAFT consensus protocol and draining the log to a high capacity
-flash SSD, while Mojim is doing simple mirroring of main memory. One can
-conclude that network offload via RDMA and Infiniband is not an a-priori
-requirement for high throughput replication. This is made possible by Cyclone's
-approach that relieves the CPU from the burden of data copies. Of course, the
-latency of a hop on commodity ethernet at 5 us with our DPDK stack is larger
-than the close to 2us latency with Infiniband. Depending on the use case for the
-key value store application the extra 6us with ethernet for the round trip to
-the replica might be a concern for some applications. We note that mapping the
-commutativity of key value store interfaces to multiple logs in Cyclone is a
-technique that can also be applied to RDMA. The authors of Mojim confirm that
-concurrency is a significant determinant of performance when replicating over
-RDMA.
+attached persistent memory with RDMA to replicate persistent data structures. A
+comparison is hard since Mojim is designed for the hypothetical scenario where
+there is enough persistent memory to hold all persistent data, while Cyclone is
+designed for the practical case where both the write-ahead log and key value
+store data structures need to be maintained largely on flash. Nevertheless we
+can compare their replication performance for the case where we use only the NVM
+log in Cyclone. On 40Gbps infiniband, Mojim reported being able to mirror data
+at 4GB/s. In contrast, Cyclone can replicate data three ways using 4*10 Gbps
+commodity ethernet NICs at a peak of roughly 3.7 GB/s. Notably this is while
+Cyclone is also running the RAFT consensus protocol, while Mojim is doing simple
+mirroring of main memory. One can conclude that network offload via RDMA and
+Infiniband is not an a-priori requirement for high throughput replication. This
+is made possible by Cyclone's approach that relieves the CPU from the burden of
+data copies. Of course, the latency of a hop on commodity ethernet at 5 us with
+our DPDK stack is larger than the close to 2us latency with
+Infiniband. Depending on the use case for the key value store application the
+extra 6us with ethernet for the round trip to the replica might be a concern for
+some applications. We note that mapping the commutativity of key value store
+interfaces to multiple logs in Cyclone is a technique that can also be applied
+to RDMA. The authors of Mojim confirm that concurrency is a significant
+determinant of performance when replicating over RDMA.
 
@@
 replications/sec on 66 Gbps of aggregate network bandwith, comparable to the
-same mark obtained by Cyclone on an aggregate 40 Gbps of network bandwidth. We
-believe the general applicability afforded by our software only solution
-together with the fact that we provide persistence by design makes it a
-compelling alternative to an FPGA based solution even given the larger latency.
+same mark obtained by Cyclone on an aggregate 40 Gbps of network bandwidth when
+not using the log on the SSD. We believe the general applicability afforded by
+our software only solution together with the fact that we provide persistence by
+design makes it a compelling alternative to an FPGA based solution even given
+the larger latency.
 
@@
 
+Using non-volatile memory to cache writes to flash is also employed in the
+internal architecture of flash SSDs - including the SSD we use in the evaluation
+of the paper. This ``write cache'' however does not help with small appends to
+the write ahead log, as experiments with Rocksdb's own write-ahead log has shown
+in this paper.
+
 Cyclone provides a client library that automatically times out and retries
@@
 distributed transactions across shards running Cyclone integrated key value
-stores, along the lines of similar work~\cite{raft_lin}, using a fixed client
-space and maintaining information about last seen sequence numbers from clients
-in NVM. Distributed transactions on Cyclone shards however are outside the scope
-of this paper.
+stores. This is along the lines of similar work~\cite{raft_lin}, using a fixed
+client space and maintaining information about last seen sequence numbers from
+clients in NVM. Distributed transactions on Cyclone shards however are outside
+the scope of this paper.
 

### 2017-07-27 paper/paper.tex
Additions.

@@
 single machine in Figure~\ref{fig:problem} is the need to do synchronous block
-IO for every update. For small to moderately sized updates, this leads to severe
-write amplification as a 4KB page must be synchronously written to the log file
-on flash. Cyclone avoids this by making use of directly attached persistent
+IO for every update. Mordern NVMe SSDs advertise a synchronous write latency of
+the order of around 20us, capping throughput at 50K ops/sec, even with write coalescing 
+in the SSDs own internal write cache. Cyclone avoids this by making use of directly attached persistent
 memory on the server. This can be in the form of NVDIMMs (DIMMs with an added
@@
 Storage devices such as flash-based SSDs export a block IO interface. Appending
-a small update to the write-ahead log requires a synchronous update of an entire
-4KB page. This leads to a high baseline latency at even low load, as shown in
-Figure~\ref{fig:problem}.  On the other hand small updates (in the order of 100s
-of bytes for both the key and value taken together) are commonplace in many
-workloads [find refs]. This problem leads many system designers to
+a small update to the write-ahead log requires a synchronous access to the SSD. Even with 
+write coalescing using a write cache on the SSD, the round  trip time to a current generation NVMe 
+SSD such as the one we use in this paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to under 50K ops/sec.
+
+This leads to a high baseline latency at even low load, as shown in
+Figure~\ref{fig:problem}. This problem leads many system designers to
 underprovision key-value store shards ensuring that it operates at moderate to
@@
 to obviate the need to make this tradeoff in the first place. We assume a small
-amount of non-volatile memory directly addressable by the CPU rather than placed
+amount of non-volatile memory directly addressable by the CPU rather than being placed
 behind a block oriented IO interface. Log appends are done to the ``NVM log''
@@
 
-Using non-volatile memory to cache writes to flash is also employed in the
-internal architecture of flash SSDs - including the SSD we use in the evaluation
-of the paper. This ``write cache'' however does not help with small appends to
-the write ahead log, as experiments with Rocksdb's own write-ahead log has shown
-in this paper.
-
 Cyclone provides a client library that automatically times out and retries

### 2017-07-27 paper/paper.tex
Updates.

@@
 
-\title{\Large \bf Cyclone: Warp Speed Replication for Persistent Key Value Stores}
+\title{\Large \bf Cyclone: High Availability for Persistent Key Value Stores}
 
@@
 single machine in Figure~\ref{fig:problem} is the need to do synchronous block
-IO for every update. Mordern NVMe SSDs advertise a synchronous write latency of
-the order of around 20us, capping throughput at 50K ops/sec, even with write coalescing 
-in the SSDs own internal write cache. Cyclone avoids this by making use of directly attached persistent
-memory on the server. This can be in the form of NVDIMMs (DIMMs with an added
+IO for every update \emph{regardless of the size of the update}. SSDs provide
+the maximum throughput when writing large chunks (usually more than 4KB) of
+data. For small updates, the synchronous write latency dominates the time for
+IO.  Cyclone avoids this by making use of directly attached persistent memory on
+the server. This can be in the form of NVDIMMs (DIMMs with an added
 ultracapacitor to dump state to a small amount of attached flash) or newer forms
@@
 Storage devices such as flash-based SSDs export a block IO interface. Appending
-a small update to the write-ahead log requires a synchronous access to the SSD. Even with 
-write coalescing using a write cache on the SSD, the round  trip time to a current generation NVMe 
-SSD such as the one we use in this paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to under 50K ops/sec.
-
-This leads to a high baseline latency at even low load, as shown in
-Figure~\ref{fig:problem}. This problem leads many system designers to
+a small update to the write-ahead log requires a synchronous access to the
+SSD. Even with write coalescing using a write cache on the SSD, the round trip
+time to a current generation NVMe SSD such as the one we use in this
+paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
+under 50K ops/sec. This leads to a high baseline latency at even low load, as
+shown in Figure~\ref{fig:problem}. This problem leads many system designers to
 underprovision key-value store shards ensuring that it operates at moderate to
@@
 at 4GB/s. In contrast, Cyclone can replicate data three ways using 4*10 Gbps
-commodity ethernet NICs at a peak of roughly 3.7 GB/s. Notably this is while
+commodity ethernet NICs at a peak of roughly 3.7 GB/s. This is with
 Cyclone is also running the RAFT consensus protocol, while Mojim is doing simple
@@
 interfaces to multiple logs in Cyclone is a technique that can also be applied
-to RDMA. The authors of Mojim confirm that concurrency is a significant
+to RDMA. Mojim confirms that concurrency is a significant
 determinant of performance when replicating over RDMA.
@@
 broadcast protocol together with a NIC on an FPGA. That work ignored durability,
-focusing purely on network performance for replication. Latency numbers are
+focusing purely on network performance for replication. Latency is
 excellent due to the FPGA that cuts out the path from CPU caches to the NIC and

### 2017-07-28 core/cyclone_comm_dpdk.hpp
Changes.

@@ -1,7 +1,7 @@
 #ifndef _CYCLONE_COMM_DPDK_
 #define _CYCLONE_COMM_DPDK_
 
-
+#include <cinttypes>
 #include <stdio.h>
 #include <stdlib.h>
 #include <string.h>
@@ -86,6 +86,7 @@ static void init_port_conf()
   //port_conf.rx_adv_conf.rss_conf.rss_hf = ETH_RSS_IP;
 }
 
+/*
 static struct rte_fdir_conf fdir_conf;
 static void init_fdir_conf()
 {
@@ -104,7 +105,7 @@ static void init_fdir_conf()
   fdir_conf.mask.tunnel_id_mask = 0xFFFFFFFF;
   fdir_conf.drop_queue = 127;
 };
-
+*/
 
 
 typedef struct {

### 2017-07-28 core/libcyclone.hpp

@@ -39,7 +39,7 @@ static const int Q_BUFS = 8191;
 static const int R_BUFS = 1023;
 
 // Static client setting for tunnel tests
-static const int num_clients = 10;
+static const int num_clients = 1;
 
 // Maximum clients (1 million)
 static const unsigned int MAX_CLIENTS = 1024U*1024U;

### 2017-07-29 paper/paper.tex
Updates.

@@
 x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
-the write ahead log~\cite{samza}. On the other side of the spectrum, key value
-store research prototypes such as FloDB~\cite{flodb} turns off the write ahead
-log to be able to showcase benefits of sophisticated extensions to LSM
+the write ahead log~\cite{samza}, depending on upper layers to provide
+availability by co-ordinating replicas. On the other side of the spectrum, key
+value store research prototypes such as FloDB~\cite{flodb} turn off the write
+ahead log to be able to showcase benefits of sophisticated extensions to LSM
 datastructures.
@@
 the maximum throughput when writing large chunks (usually more than 4KB) of
-data. For small updates, the synchronous write latency dominates the time for
-IO.  Cyclone avoids this by making use of directly attached persistent memory on
-the server. This can be in the form of NVDIMMs (DIMMs with an added
-ultracapacitor to dump state to a small amount of attached flash) or newer forms
-of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs are expensive
-- retailing at approximately 2X the price of a standard DIMM [find ref]. This
-means that they will likely be available in small quantities. We compensate for
-this by periodically draining the NVDIMM log into a log placed on a standard SSD
-on the system.
+data. This is due to the fixed cost of a round trip to the SSD through the IO
+interface (NVMe in our case), regardless of the volume of data being
+synchronously persisted. Cyclone avoids this by making use of directly attached
+persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
+an added ultracapacitor to dump state to a small amount of attached flash) or
+newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
+are expensive - retailing at approximately 2X the price of a standard DIMM [find
+  ref]. This means that they will likely be available in small quantities. We
+compensate for this by periodically draining the NVDIMM log into a log placed on
+a standard SSD on the system.
 
@@
 Section~\ref{sec:parallelism}.  A detailed evaluation of Cyclone's replication
-is provided in the context of RocksDB - a popular persistent key value store -
-in Section~\ref{sec:evaluation}. We then discuss related work before concluding.
+performance is provided in the context of RocksDB - a popular persistent key
+value store - in Section~\ref{sec:evaluation}. We then discuss related work
+before concluding.
 
@@
 library. Cyclone replicates the key value store across a set of replica
-servers. The client side library sends requests to a distinguished leader
-replica.  On the server side Cyclone accepts requests from clients and calls
+servers. The client side library sends \emph{all} requests to a distinguished
+leader replica.  On the server side Cyclone accepts requests from clients and calls
 into the key value store. It then returns a response to the client. Before
@@
 the request to the logs of follower replicas. The request is considered
-replicated once acknowledged by a majority of replicas.
+replicated once acknowledged by a majority of replicas. Follower replicas apply
+updates from the log in order.
 
@@
 failure of the leader replica a new leader is automatically elected - RAFT
-ensures that the new leader has the most up to date logs. The Cyclone client
+ensures that the new leader has the most up to date log. The Cyclone client
 library automatically locates the new leader on a failover - clients therefore
@@
 operations) are not logged, they are serviced in FIFO order with the write requests
-(PUT operations) currently in Cyclone's log. This ensures linearizability in the
-face of failovers.
+(PUT operations) currently in Cyclone's log at the moment the read request is
+received. This ensures linearizability in the face of failovers.
 
@@
 to obviate the need to make this tradeoff in the first place. We assume a small
-amount of non-volatile memory directly addressable by the CPU rather than being placed
-behind a block oriented IO interface. Log appends are done to the ``NVM log''
-placed in this non-volatile memory. While we designed Cyclone to leverage new
-non-volatile memory technologies that might be available in the
-future~\cite{3dxpoint} current non-volatile memory options such are NVDIMMs
-are more expensive that DRAM. Therefore, for Cyclone to be economical
-we cannot assume that a sufficient amount of NVM is available to hold the entire
-log. The log entries can only be garbage collected when the in-memory component
-of the key value store is checkpointed to secondary storage. The growing size of
+amount of non-volatile memory directly addressable by the CPU rather than being
+placed behind a block oriented IO interface. Log appends are done to the ``NVM
+log'' placed in this non-volatile memory. While we designed Cyclone to leverage
+new non-volatile memory technologies that might be available in the
+future~\cite{3dxpoint}, current non-volatile memory options such are NVDIMMs are
+more expensive that DRAM. Therefore, for Cyclone to be economical we cannot
+assume that a sufficient amount of NVM is available to hold the entire log. The
+log entries can only be garbage collected when the in-memory component of the
+key value store is checkpointed to secondary storage. The growing size of
 available DRAM means that also provisioning sufficient NVDIMMs to hold the
@@
 log in conveniently large units that are a multiple of 4KB - the optimum IO unit
-for flash SSDs. This ensures we get the best
-possible throughput from a flash device without paying the price of synchronous
-IO for small key-value pairs. At the same time, the amount of non-volatile
-memory required is small enough to not add undue cost of the server.
+for flash SSDs. This ensures we get the best possible throughput from a flash
+device without paying the price of synchronous IO for small key-value pairs. At
+the same time, the amount of non-volatile memory required is small enough to not
+add undue cost of the server.
 
@@
 SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
-evaluation. Together these two assumptions mean that we can recover a consistent
+evaluation. Together, these two assumptions mean that we can recover a consistent
 prefix of the log on a power failure.
@@
 leader replica to the followers. Each replica \emph{independently} drains
-committed entries (in terms of RAFT consensus) from the NVM log to the
+committed (in terms of RAFT consensus) entries from the NVM log to the
 flashlog. Block IO therefore is no longer a consideration when optimizing data
@@
 \subsection{Eliminating Data Movement}
+\label{sec:dm}
 Software packet switches built on DPDK try to touch as little of the packet data
@@
 replicas. Therefore, we never tradeoff latency for throughput when batching.
+
+Finally, we discuss our choice of RAFT as a consensus protocol. A key advantage
+of RAFT is that unlike alternatives such as MultiPaxos~\cite{multipaxos} or
+Quorum based replication~\cite{quorum}, the leader is guaranteed to have the
+most up to date logs. There is therefore no case where a leader needs to receive
+log entries to ``fill'' holes in its logs from follower replicas, simplifying
+the protocol state machine we need to implement. A leader can immediately reject
+any responses that are not more current than its view (term). In turn, the
+persistent log operates as a double ended queue, with entries either being
+appended (or possibly deleted at followers) at the end and being deleted from
+the front (after commit). This is critical to efficient operation of Cyclone, as
+it allows a simple top level structure (a circular buffer of pointers) to
+represent the log.
 
@@
 
+DARE~\cite{dare} implements state machine replication over RDMA, using a custom
+replication protocol. It is worthwhile to point out that since DARE allows any
+replica to become the leader it requires a log adjustment protocol for the
+leader to bring its logs up to date, a situation we avoid by using RAFT as
+described in Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication
+DARE reported being able to replicate 500K requests/sec. This is without
+persisting the log of updates. In contrast, Cyclone replicates upwards of 6
+million similarly sized requests each second with 4*10 Gbps ethernet (without the
+second level flashlog). The conclusions we can draw are similar to the
+comparison with Mojim: a carefully implemented log replication system obviates
+the need for network offload.
+
 {CRANE}~\cite{crane} is a system for replicating mutithreaded programs using
@@
 have omitted for brevity from this paper. The scalable commutativity rule also
-suggests that scalable key value store implementations would necessarily have
-commutative interfaces, something we confirm and exploit for key
-value stores.
+suggests that scalable key value store implementations such as
+FloDB~\cite{flodb} would necessarily have commutative interfaces, enabling more
+general applicability of Cyclone.
 

### 2017-07-29 paper/paper.tex
Updates to paper.

@@
 is made possible by Cyclone's approach that relieves the CPU from the burden of
-data copies. Of course, the latency of a hop on commodity ethernet at 5 us with
+data copies. On the other hand, the latency of a hop on commodity ethernet at 5 us with
 our DPDK stack is larger than the close to 2us latency with
@@
 DARE~\cite{dare} implements state machine replication over RDMA, using a custom
-replication protocol. It is worthwhile to point out that since DARE allows any
-replica to become the leader it requires a log adjustment protocol for the
-leader to bring its logs up to date, a situation we avoid by using RAFT as
-described in Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication
-DARE reported being able to replicate 500K requests/sec. This is without
-persisting the log of updates. In contrast, Cyclone replicates upwards of 6
-million similarly sized requests each second with 4*10 Gbps ethernet (without the
-second level flashlog). The conclusions we can draw are similar to the
-comparison with Mojim: a carefully implemented log replication system obviates
-the need for network offload.
+replication protocol. DARE does not consider persistence in its design. It is
+worthwhile to point out that since DARE allows any replica to become the leader
+it requires a log adjustment protocol for the leader to bring its logs up to
+date, a situation we avoid by using RAFT as described in
+Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
+being able to replicate 500K requests/sec. This is without persisting the log of
+updates. In contrast, Cyclone replicates upwards of 6 million similarly sized
+requests each second with 4*10 Gbps ethernet (without the second level
+flashlog). DARE reports an unloaded write latency of 15us, which is lower than
+the unloaded write latency of Cyclone at 20us, thanks to infiniband. The
+conclusions we can draw are similar to the comparison with Mojim: a carefully
+implemented log replication system obviates the need for network offload when
+doing replication.
 

### 2017-07-29 paper/paper.tex
Updates.

@@
 newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
-are expensive - retailing at approximately 2X the price of a standard DIMM [find
-  ref]. This means that they will likely be available in small quantities. We
-compensate for this by periodically draining the NVDIMM log into a log placed on
-a standard SSD on the system.
+are significantly more expensive than standard DIMMs. This means that they will
+likely be available in small quantities. We compensate for this by periodically
+draining the NVDIMM log into a log placed on a standard SSD on the system.
 
@@
 available DRAM means that also provisioning sufficient NVDIMMs to hold the
-corresponding log of updates is not feasible given that NVDIMMs cost at least
-twice as much as volatile DRAM. We therefore resort to using a second larger log
-level placed on a flash SSD. Entries are drained from the NVM log to this flash
-log in conveniently large units that are a multiple of 4KB - the optimum IO unit
-for flash SSDs. This ensures we get the best possible throughput from a flash
-device without paying the price of synchronous IO for small key-value pairs. At
-the same time, the amount of non-volatile memory required is small enough to not
-add undue cost of the server.
+corresponding log of updates is not feasible given that NVDIMMs are
+significantly more expensive than standard volatile DIMMs.  To illustrate this,
+at the time of writing this paper, an 8 GB DDR4 NVDIMM was retailing at \$612,
+while a plain 8 GB DD4 DIMM was retailing at \$178.  We therefore resort to
+using a second larger log level placed on a flash SSD. Entries are drained from
+the NVM log to this flash log in conveniently large units that are a multiple of
+4KB - the optimum IO unit for flash SSDs. This ensures we get the best possible
+throughput from a flash device without paying the price of synchronous IO for
+small key-value pairs. At the same time, the amount of non-volatile memory
+required is small enough to not add undue cost of the server.
 

### 2017-07-29 paper/paper.tex
Updates to paper.

@@
 community when building software packet switches. Software packet switches reach
-impressive speeds of millions of packets forwarded per core [find refs], a
+impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, a
 number far in excess of the few hundreds of packets we manage in

### 2017-07-29 paper/paper.tex
Updates to paper.

@@
 high load, using group-commit to batch additions to the log. They therefore pay
-a price in latency to ensure the bandwidth to storage is fully utilized to
-prevent it from becoming a bottleneck.
+a price in latency to ensure the bandwidth to storage is fully utilized.
 
@@
 from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
-testbed from {\tt 24 us} to as low as {\tt 5 us}, providing a significant boost
+testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
 to performance.
@@
 To illustrate how this is achieved, Figure~\ref{fig:packet_layout} shows how
-Cyclone manipulates packet
-layouts across the two steps of prepending a RAFT header and transmitting to
-follower replicas. DPDK describes packets using an ``mbuf'' data
-structure. Roughly speaking, an ``mbuf'' consists of a flat array of bytes
-actually containing the packet and a fixed size piece of external metadata that
-describes various aspects of the packet, most crucially a pointer to the start
-and end of the packet in the byte array. DPDK's userspace drivers receive
-packets from the NIC such that they are offset in the byte array by a
+Cyclone manipulates packet layouts across the two steps of prepending a RAFT
+header and transmitting to follower replicas. DPDK describes packets using an
+``mbuf'' data structure. Roughly speaking, an ``mbuf'' consists of a flat array
+of bytes actually containing the packet and a fixed size piece of external
+metadata that describes various aspects of the packet, most crucially a pointer
+to the start and end of the packet in the byte array. DPDK's userspace drivers
+receive packets from the NIC such that they are offset in the byte array by a
 configurable amount referred to as ``headroom''. We strip off the existing
@@
 
-Althought the work described in the previous section significantly boosts the
+Although the work described in the previous section significantly boosts the
 network performance of replication in Cyclone, it still does not come close to
@@
 \section{Conclusion}
-Persistent key value stores struggle today to provide high availability using a
-write ahead log without a significant impact to performance. Cyclone solves this
-problem by leveraging a small amount of directly attached non-volatile memory
-that transforms the replication problem into a software packet switching
-problem. Cyclone and its Rocksdb bindings are slated for release to the open
+Cyclone addresses two fundamental difficulties in adding high availability to
+persistent key value stores. It leverage a small amount of non-volatile memory
+to avoid paying the cost of the block IO interface for every update
+operation. It also leverages software packet switching principles to replicate
+the write ahead log thus avoiding any need for offload capable network
+interfaces.  Cyclone and its Rocksdb bindings are slated for release to the open
 source community and we hope both the code and the ideas in the system itself
-will help key value store builders and researchers overcome their logging
-problem.
+will help key value store builders and researchers deal with high availabilty
+without compromising performance.
 \newcommand\myurl[2]{\url{#1}}

### 2017-08-01 test/fb_loader.cpp
fb.hpp

@@ -1,0 +1,124 @@
+/*
+ * Copyright (c) 2015, Intel Corporation
+ *
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted provided that the following conditions
+ * are met:
+ *
+ *     * Redistributions of source code must retain the above copyright
+ *       notice, this list of conditions and the following disclaimer.
+ *
+ *     * Redistributions in binary form must reproduce the above copyright
+ *       notice, this list of conditions and the following disclaimer in
+ *       the documentation and/or other materials provided with the
+ *       distribution.
+ *
+ *     * Neither the name of Intel Corporation nor the names of its
+ *       contributors may be used to endorse or promote products derived
+ *       from this software without specific prior written permission.
+ *
+ * THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
+ * "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
+ * LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
+ * A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
+ * OWNER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
+ * SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
+ * LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
+ * DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
+ * THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
+ * (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+ * OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+ */
+
+
+#include<assert.h>
+#include<errno.h>
+#include<string.h>
+#include<stdlib.h>
+#include "../core/logging.hpp"
+#include "../core/clock.hpp"
+#include<stdio.h>
+#include <time.h>
[…]

### 2017-08-01 test/fb.hpp
Header file.

@@ -1,0 +1,16 @@
+//### New values following distribution in facebook's paper
+//VALUE_SIZE_ARR=( 16 32 64 128 256 512 1024 2048 4096 8192 )
+//NUM_OPS_ARR=( 20132864 10066944 10066944 35232768 25165824 3145728 1888256 2359296 786432 1022976 )
+//LIMIT_OPS_ARR=( 15728640 7864320 5242880 8388608 12582912 1048576 524288 524288 367616 157696 )
+
+// Truncated value sizes
+// Div num keys by factor of 10
+static int values = 6;
+static int VALUE_SIZE_ARR[] = { 16, 32, 64, 128, 256, 512 };
+static int NUM_OPS_ARR [] = { 2013286, 1006694, 1006694, 3523276, 2516582,
+			      314572 };
+// 8 byte key
+// most significant byte is index into value size array
+// remaining bytes unique counter
+			      
+

### 2017-08-01 test/fb_loader.cpp
Updates.

@@ -62,9 +62,9 @@ void load()
       for(int k=j;k<j + BATCH_SIZE && k < keys;k++) {
 	unsigned long key = i << 56;
 	key += k;
-	rocksdb::Slice key((const char *)&key, 8);
-	rocksdb::Slice value((const char *)&value_base[0], VALUE_SZ_ARR[i]);
-	batch.Put(key, value);
+	rocksdb::Slice rdb_key((const char *)&key, 8);
+	rocksdb::Slice rdb_value((const char *)&value_base[0], VALUE_SIZE_ARR[i]);
+	batch.Put(rdb_key, rdb_value);
       }
       rocksdb::WriteOptions write_options;
       write_options.sync = false;
@@ -118,7 +118,7 @@ void closedb()
 int main(int argc, char *argv[])
 {
   opendb();
-  load(rocks_keys);
+  load();
   closedb();
 }
 

### 2017-08-01 test/fb.hpp
More testing code.

@@ -1,3 +1,6 @@
+#ifndef _FB_COMMON_
+#define _FB_COMMON_
+#include "rocksdb.hpp"
 //### New values following distribution in facebook's paper
 //VALUE_SIZE_ARR=( 16 32 64 128 256 512 1024 2048 4096 8192 )
 //NUM_OPS_ARR=( 20132864 10066944 10066944 35232768 25165824 3145728 1888256 2359296 786432 1022976 )
@@ -5,12 +8,94 @@
 
 // Truncated value sizes
 // Div num keys by factor of 10
+#include<random>
+#include<vector>
+#include<algorithm>
+#include<iostream>
+using namespace std;
 static int values = 6;
 static int VALUE_SIZE_ARR[] = { 16, 32, 64, 128, 256, 512 };
-static int NUM_OPS_ARR [] = { 2013286, 1006694, 1006694, 3523276, 2516582,
+static int NUM_OPS_ARR [] = { 2013286, 
+			      1006694, 
+			      1006694, 
+			      3523276, 
+			      2516582,
 			      314572 };
+static double access_probs[] = { 0.309278, 
+				 0.154639, 
+				 0.103093, 
+				 0.164948,
+				 0.247423,
+				 0.020619 };
+
+const double prob_rd = 0.9;
+
 // 8 byte key
 // most significant byte is index into value size array
 // remaining bytes unique counter
-			      
+class load_gen {
+  vector<double> cdf;
+  mt19937* base_rng;
+  uniform_real_distribution<double> * rng;
+  double flip_coin()
+  {
+    return (*rng)(*base_rng);
+  }
+
+  // 0 == read
+  // 1 == wr
+  int select_op()
+  {
+    double ur = flip_coin();
+    if(ur <= prob_rd)
+      return 0;
+    else
+      return 1;
+  }
+
+  int select_index()
+  {
+    double ur = flip_coin();
+    auto index_itr = upper_bound(cdf.begin(), cdf.end(), ur);
+    if(index_itr == cdf.end()) 
+      return values - 1;
+    return index_itr - cdf.begin();
+  }
+
+public:
+  load_gen(int client_seed)
+  {
+    array<unsigned int, mt19937::state_size> seeds;
+    for(int i=0;i<mt19937::state_size;i++) {
+      seeds[i] = client_seed*(i + 1); // seed as a deterministic fn of client_seed
+    }
[…]

### 2017-08-01 test/fb.hpp
More testing code.

@@ -28,7 +28,7 @@ static double access_probs[] = { 0.309278,
 				 0.247423,
 				 0.020619 };
 
-const double prob_rd = 0.9;
+const double prob_rd = 0.95;
 
 // 8 byte key
 // most significant byte is index into value size array
@@ -41,7 +41,7 @@ class load_gen {
   {
     return (*rng)(*base_rng);
   }
-
+public:
   // 0 == read
   // 1 == wr
   int select_op()
@@ -62,7 +62,7 @@ class load_gen {
     return index_itr - cdf.begin();
   }
 
-public:
+
   load_gen(int client_seed)
   {
     array<unsigned int, mt19937::state_size> seeds;

### 2017-08-01 test/fb_client.cpp

@@ -1,0 +1,182 @@
+/*
+ * Copyright (c) 2015, Intel Corporation
+ *
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted provided that the following conditions
+ * are met:
+ *
+ *     * Redistributions of source code must retain the above copyright
+ *       notice, this list of conditions and the following disclaimer.
+ *
+ *     * Redistributions in binary form must reproduce the above copyright
+ *       notice, this list of conditions and the following disclaimer in
+ *       the documentation and/or other materials provided with the
+ *       distribution.
+ *
+ *     * Neither the name of Intel Corporation nor the names of its
+ *       contributors may be used to endorse or promote products derived
+ *       from this software without specific prior written permission.
+ *
+ * THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
+ * "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
+ * LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
+ * A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
+ * OWNER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
+ * SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
+ * LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
+ * DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
+ * THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
+ * (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+ * OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+ */
+
+#include <stdio.h>
+#include <sys/stat.h>
+#include <stdlib.h>
+#include <string.h>
+#include <unistd.h>
+#include <time.h>
+#include <assert.h>
+#include <boost/property_tree/ini_parser.hpp>
[…]

### 2017-08-02 test/fb.hpp
Updates to code.

@@ -37,6 +37,7 @@ class load_gen {
   vector<double> cdf;
   mt19937* base_rng;
   uniform_real_distribution<double> * rng;
+  vector<uniform_int_distribution<unsigned long>*> rng_array;
   double flip_coin()
   {
     return (*rng)(*base_rng);
@@ -62,6 +63,10 @@ public:
     return index_itr - cdf.begin();
   }
 
+  int gen_key(int idx)
+  {
+    return (*(rng_array[idx]))(*base_rng);
+  }
 
   load_gen(int client_seed)
   {
@@ -76,6 +81,7 @@ public:
     for(int i=0;i<values;i++) {
       sum = sum + access_probs[i];
       cdf.push_back(sum);
+      rng_array.push_back(new uniform_int_distribution<unsigned long>(0, NUM_OPS_ARR[i]));
     }
     cdf.back() = 1.0;
   }

### 2017-08-02 test/fb_client.cpp

@@ -92,17 +92,20 @@ int driver(void *arg)
   srand(tx_begin_time);
   int partition;
   while(true) {
+    int idx = test.select_index();
+    int sz;
     if(test.select_op() == 0) {
       rpc_flags = RPC_FLAG_RO;
       kv->op    = OP_GET;
+      sz = sizeof(fb_kv_t);
     }
     else {
       rpc_flags = 0;
       kv->op    = OP_PUT;
+      sz = sizeof(fb_kv_t) + VALUE_SIZE_ARR[idx];
     }
-    int idx = test.select_index();
-    int sz = sizeof(fb_kv_t) + VALUE_SIZE_ARR[idx];
-    kv->key   = 0;
+    kv->key   = ((unsigned long)idx) << 56;
+    kv->key   = kv->key + test.gen_key(idx);
     my_core = kv->key % executor_threads;
     sz = make_rpc(handles[0],
 		  buffer,

### 2017-08-02 test/fb.hpp
Updates.

@@ -36,11 +36,9 @@ const double prob_rd = 0.95;
 class load_gen {
   vector<double> cdf;
   mt19937* base_rng;
-  uniform_real_distribution<double> * rng;
-  vector<uniform_int_distribution<unsigned long>*> rng_array;
   double flip_coin()
   {
-    return (*rng)(*base_rng);
+    return  ((double)(*base_rng)() - mt19937::min())/(mt19937::max() - mt19937::min());
   }
 public:
   // 0 == read
@@ -63,9 +61,9 @@ public:
     return index_itr - cdf.begin();
   }
 
-  int gen_key(int idx)
+  unsigned long gen_key(int idx)
   {
-    return (*(rng_array[idx]))(*base_rng);
+    return NUM_OPS_ARR[idx]*flip_coin();
   }
 
   load_gen(int client_seed)
@@ -76,12 +74,10 @@ public:
     }
     seed_seq seeds_in_seq(begin(seeds), end(seeds));
     base_rng = new mt19937(seeds_in_seq);
-    rng = new uniform_real_distribution<double>(0.0, 1.0);
     double sum = 0.0;
     for(int i=0;i<values;i++) {
       sum = sum + access_probs[i];
       cdf.push_back(sum);
-      rng_array.push_back(new uniform_int_distribution<unsigned long>(0, NUM_OPS_ARR[i]));
     }
     cdf.back() = 1.0;
   }

### 2017-08-02 test/fb.hpp
updates.

@@ -12,6 +12,7 @@
 #include<vector>
 #include<algorithm>
 #include<iostream>
+#include<array>
 using namespace std;
 static int values = 6;
 static int VALUE_SIZE_ARR[] = { 16, 32, 64, 128, 256, 512 };
@@ -38,7 +39,7 @@ class load_gen {
   mt19937* base_rng;
   double flip_coin()
   {
-    return  ((double)(*base_rng)() - mt19937::min())/(mt19937::max() - mt19937::min());
+    return  ((double)(*base_rng)() - base_rng->min())/(base_rng->max() - base_rng->min());
   }
 public:
   // 0 == read
@@ -68,12 +69,7 @@ public:
 
   load_gen(int client_seed)
   {
-    array<unsigned int, mt19937::state_size> seeds;
-    for(int i=0;i<mt19937::state_size;i++) {
-      seeds[i] = client_seed*(i + 1); // seed as a deterministic fn of client_seed
-    }
-    seed_seq seeds_in_seq(begin(seeds), end(seeds));
-    base_rng = new mt19937(seeds_in_seq);
+    base_rng = new mt19937(client_seed);
     double sum = 0.0;
     for(int i=0;i<values;i++) {
       sum = sum + access_probs[i];

### 2017-08-02 paper/paper.tex
Added etcd trace expt.

@@
 
+Our evaluation has, thus far, focused on synthetic write heavy
+workloads. To illustrate that Cyclone can also be useful for read-heavy
+workloads, we configured our clients to generate key value requests that mimic
+the distribution of value sizes in Facebook's ETCD
+trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read
+heavy (95\% reads). Figure~\ref{fig:fb} illustrates that even with a
+read-dominated workload, Cyclone is effective at providing a write ahead log
+with 3 way replication with a lower performance overhead than Rocksdb's own
+write ahead logging.
+
+\begin{figure}
+\includegraphics[scale=0.6]{results2/fb.pdf}
+\caption{Rocksdb - read heavy workload}
+\label{fig:fb}
+\end{figure}
+
 Finally, we showcase the benefit of using Cyclone beyond pure performance as

### 2017-08-03 paper/paper.tex
Updates.

@@
 special flag encoded into the size if necessary. We flush log segments out to
-the log file using asynchronous direct IO, and therefore we fill one log segment
-buffer while keeping IO to another one outstanding to the flash drive. To avoid
-having to do a synchronous metadata flush, we preallocate (using the
+the log file using asynchronous direct IO, and therefore we continue to fill log
+segment buffers while keeping IO to previously filled buffers outstanding to the
+flash drive. We allocate enough buffers to keep a maximum of 32 outstanding
+requests to the SSD. To avoid having to do a synchronous metadata flush, we preallocate (using the
 posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
@@
 this additional complexity is still small compared to distributed transactions
-using two-phase commit that also need to recover from failure used in systems
-such as FARM~\cite{farm}.
+using two-phase commit - especially when taking into account failure recovery -
+as used in systems such as FARM~\cite{farm}.
 
@@
 regardless of the size of the key value store or second level log on flash. We
-divide the evaluation into three parts. First, we evaluate Cyclone's performance
-with a single level log as a pure software packet switch. Next, we evaluate
-performance when adding a second level of log on flash. Finally, we evaluate
-performance when integrated with Rocksdb~\cite{rocksdb} as an alternative to
-Rocksdb's write ahead log. Unless otherwise mentioned, we use a 60 byte header
-followed by an optional payload for experiments. We log both the header and
-payload. In all cases the server echoes the received entry (header and payload)
-back to the client. We fix the number of KV Store threads at 32 using at most 8
-physcial logs with associated control plane cores for running RAFT instances.
+divide the evaluation into three parts.
 
-Cyclone is built on top of DPDK to apply software packet switching techniques to
-the log replication problem. We therefore begin by systematically evaluating
-optimizations applied in Cyclone to replicate the NVM log in
-Figure~\ref{fig:network_opts} - with no payload. The y-axis reports latency seen
-at the client (which means two network round trips with replication). Using
-TCP/IP to replicate a RAFT log tops out at around 30K entries/s.  Switching to
-DPDK (the line marked +DPDK) improves the throughput by an order of magnitude to
-around 500K entries/s. Using batching (the line marked +batching) improves the
-performance further bringing us close to a million entries/s. Scaling
-horizontally to 8 physical logs (+8 phy logs) improves performance to close to
-2M entries/s. Finally using all 4 ethernet ports on the machine to replicate
-entries improves performance considerably to 6M entries/s. In all, performance
-improves by 200X over the TCP/IP single log baseline. Cyclone also considerably
-improves the latency for replication, from close to 100us with TCP/IP to around
-30us at peak throughput.
+First, we evaluate Cyclone's performance with a single level log as a pure
+software packet switch. For this purpose, Cyclone uses a dummy server stub that
+simply echoes the client request back to it. Next, we evaluate
+performance with the dummy stub, but with the addition of the second level of log
+on flash. Finally, we evaluate performance when integrated with
+Rocksdb~\cite{rocksdb} as an alternative to Rocksdb's write ahead log.
+
+Unless otherwise mentioned, we use a 60 byte header followed by an optional
+payload for experiments. We log both the header and payload. Also, unless
+otherwise mentioned we use 8 physical logs (and associated RAFT instances) each
+mapped to a dedicated core. The remaining 32 cores are dedicated to the stub
+server or Rocksdb, as the case may be.
+
+We begin by systematically evaluating network stack related optimizations
+applied in Cyclone to replicate the NVM log in Figure~\ref{fig:network_opts} -
+with no payload. The y-axis reports latency seen at the client (which means two
+network round trips with replication). Using TCP/IP to replicate a RAFT log tops
+out at around 30K entries/s.  Switching to DPDK (the line marked +DPDK) improves
+the throughput by an order of magnitude to around 500K entries/s. Using batching
+(the line marked +batching) improves the performance further bringing us close
+to a million entries/s. Scaling to 8 physical logs (+8 phy logs) improves
+performance to close to 2M entries/s. Finally using all 4 ethernet ports on the
+machine to replicate entries improves performance considerably to 6M
+entries/s. In all, performance improves by 200X over the TCP/IP single log
+baseline. Cyclone also considerably improves the latency for replication, from
+close to 100us with TCP/IP to around 30us at peak throughput.
+
+One can draw two important conclusions from
+Figure~\ref{fig:network_opts}. First, one can indeed treat log replication as a
+software packet switching problem \emph{provided} the persistent log can be held
+in directly attached memory eliminating the block interface from the critical
+path. Second, relieving the CPU from the overhead of data copies has a
+significant positive effect on performance. Finally, using multiple physical
+logs is essential to exploiting the concurrency available at the level of the
+network interface card (even a single one!) that would otherwise go wasted due
+to the serializing abstraction of a single log. Finally, we note that DPDKs raw
+packet forwarding capability on that configuration is close to 12M packets per
+second. We do not reach that mark due to the overhead of running the replication
+protocol rather than a simple switching decision per packet. Although batching
+goes some way in amortizing this overhead it cannot be eliminated entirely
+because consensus protocols are more complex than a simple packet switching
+decision.
 
@@
 
-There are two factors that can have significant impact on Cyclone's
-performance. First, the number of replicas dictates the outgoing message rate
-from the leader replica and therefore increasing the replication factor can
+Both the replica count and payload size can have significant impact on Cyclone's
+network performance. First, the number of replicas dictates the outgoing message
+rate from the leader replica and therefore increasing the replication factor can
 decrease Cyclone's performance. Figure~\ref{fig:replicas} shows the impact of
@@
 headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
-network line rate bottleneck at this point.
+network line rate bottleneck at this point. One can contrast this with the
+previous experiment where the bottleneck to throughput was the cost of running
+the consensus protocol for batches of packets.
 
@@
 We now turn our attention from the network component of Cyclone to the storage
-one by adding the second level flashlog. We evaluate the impact of adding
-block storage to our hitherto pure packet switching scenario in
-Figure~\ref{fig:flashlog}. Batching entries from the top level NVDIMM log to the
-second level flashlog is clearly beneficial as adding flash storage at the
-second level has almost no impact on peak performance in Cyclone for small
-entries. The situation however changes for larger
-entries. Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
+one by adding the second level flashlog. We evaluate the impact of adding the
+flashlog in Figures~\ref{fig:flashlog} and \ref{fig:flashlog_512}. The benefit
+of the second level flashlog is that it lets us limit the amount of needed
+NVDIMM memory. The hypothesis was that batching the movement of data from the
+NVM log to the flashlog would be sufficient to hide the latency of the block IO
+device from network speed replication. The results confirm that in terms of
+latency for the low to moderate load cases, our two level log arrangement is
+effective at hiding the latency of the NVMe SSD. The picture however is
+different for peak throughput. For small updates there is no impact on
+throughput when we add the SSD. On the other hand 
+Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
 significant impact on peak throughput - it drops to approximately 350K
 ops/sec. This corresponds to around 50K 4KB IOPS to the SSD to write out the
-flashlog pages. The peak for the drive is 160K IOPS using a queue depth
-(concurrency) of 128. With 32 application threads we expect a lower peak
-throughput around 40K IOPS explaining our bottleneck at 50K IOPS.  It is
-possible to tune our observed performance further by aligning the flush boundary
-to increase the number of outstanding requests - we do not do so in this paper,
-keeping Cyclone agnostic to the exact size of entry being replicated. A final
-point about Figure~\ref{fig:flashlog_512} is that once we are past the storage
-bottleneck the latency spike is dramatic and large enough to trigger Cyclone's
-failure detector and repeated retries from the clients. There are - therefore -
-no points on the ``knee'' of the curve as in the pure packet switched one-level
-log case.
+flashlog pages. This is in fact the expected IOPS limit with the 32 outstanding
+requests to the drive that we maintain (Section~\ref{sec:storage}).
+A final point about Figure~\ref{fig:flashlog_512} is that once we are past the
+storage bottleneck the latency spike is dramatic and large enough to trigger
+Cyclone's failure detector and repeated retries from the clients. There are -
+therefore - no points on the ``knee'' of the curve as in the pure packet
+switched one-level log case.
+
+One can therefore conclude that using a small amount of directly attached NVM is
+effective at hiding the latency of a background block storage device from the
+criticial path of replication. Further, for small updates the throughput
+bottleneck is the capability of the software to run the consensus protocol but
+shifts to the secondary storage device for larger request sizes.
 
@@
 
-The final dimension we evaluate is of ganged operations. The primary purpose of
+We now evaluate ganged operations. The primary purpose of
 ganged operations is to avoid the need for distributed transactions to
@@
 unloaded case - made ganged requests to the replicas. We varied the number of
-cores participating in the request from one to the full complement of 32
-application cores. Figure~\ref{fig:ganged} shows the results both using a single
-level log as well as a two level log. The primary takeaway is that unloaded
-latency increases slowly as we increase the number of active cores - to around
-40 us from the baseline of 20 us. There are two causes of this. First, there is
-a quicker rate of increase from 1 to 8 active cores, the reason being that this
-corresponds to an increase in the number of physical logs that must be
-synchronized on. There is some amount of synchronization involved in the
-userspace DPDK driver for access to common resources for accessing the NIC and
-this affects the ability to simultaneously issue replication messages on all
-quorums. The smaller contribution to increasing latency that continues past the
-8 core case is due to cacheline pingponging when executing the rendevous of
-Figure~\ref{fig:ganged_ops}. Both these sources of latency drift could be
-corrected using replication quorums mapped to dedicated NICs and more scalable
-rendezvous designs (such as with machine aware broadcast
-trees~\cite{broadcast_tree}). However we deemed the complexity of such
-optimizations unnecessary. The added latency for most cases is well under the
-extra round trip delay (including replication) for all phases of a 2PC.
+physical log participating in the request from one to the full complement of 8.
+Figure~\ref{fig:ganged} shows the results. Unloaded
+latency increases slowly as we increase the number of active physical logs - to
+around 32 us from the baseline of 21 us. The reason for this increase is that
+the DPDK userspace NIC driver pipelines request processing for communication
+between the CPU core and the NIC. Simultaneous replicaton on multiple physical
+logs represents a worst case for this arrangement as all eight cores try to
+send replication messages to their replicas, adding about a microsecond of
+serial latency each, to dispatch requests to the NIC. Regardless, the experiment
+underlines the value of our ganged operation design over a distributed
+transaction that would need 4 network hops for the two phase commit (and
+replication) using an external co-ordinator, a minimum of 40us.
 
@@
 store. Rocksdb is a complex industrial strength persistent key value store and
-this means that it is accompanied by a complex array of performance tuning knobs
-to get the best performance from flash. We were also aware that SSD performance
-is slated for dramatic increases in the coming years with the introduction of
-new memory technology such as 3DXPoint based Optane SSDs. To demonstrate that
-Cyclone is future proof we eliminated flash media performance from the picture
-as far as the key value store is concerned by placing all files for the key
-value store (SSTables) on a RAMdisk, which presumably represents the limit in
-performance for flash in the near future. All log files on secondary storage
-however - both Rocksdb's own write ahead log and the alternative of Cyclone's
-second level flashlog - are placed on the SSDs.
+an important target for practical use of Cyclone.  Rocksdb is accompanied by a
+complex array of performance tuning knobs to get the best performance from
+flash. Designing LSM tree based key value stores to extract good performance
+from flash and make best use of DRAM on the machine is an area of active
+research~\cite{flodb, triad}. We wished to make sure that our conclusions would
+apply even as Rocksdb performance improves with the integration of new
+ideas. Keeping this in mind, we configured RocksDB to place all files for the
+key value store (SSTables) on a RAMdisk, which presumably represents the limit
+in performance for both software enhancements as well as improvements to
+flash-based SSDs. However - both Rocksdb's own write ahead log and the
+alternative of Cyclone's second level flashlog - are placed on the SSDs. In
+effect we therefore make availability a harder problem in this setting,
+guaranteeing our design can handle the immediate future.
 
-We evaluate two different request sizes: 8 byte keys with 8 byte values and 8
-byte keys with 256 byte values. We run 8 physical logs on 8 cores and devote 32
-cores to Rocksdb. Since we are interested in performance of the log, which is
-only used for update requests, our workload is 100\% writes.
+We use 8 bytes keys with either 8 bytes values or 256 byte values.
+Since our focus is replication and only update requests are replicated, we use a
+pure update workload to test the capability of the system.
 
@@
 a no-op. We note that in addition to the key and value, we are also logging
-RocksDB specific request data such as operation type and the request header.
+RocksDB specific request data such as operation type and the request header,
+making this different from the previous experiments.
 Figure~\ref{fig:kv_baseline} shows the baseline performance for the chosen
@@
 Next, we consider the problem of supporting Rocksdb's write-batch operation that
-atomically writes a set of key-value pairs into the KV store. In order to
+atomically writes a set of key-value pairs into the KV store. This is a
+practical application of ganged operations in Cyclone. In order to
 perform the operation with Cyclone managing the log, the client must issue a
-ganged operation across cores owning the keys in the write batch. This is in
-contrast to baseline Rocksdb where the entire write batch can be sent to any
-core. A key concern here was whether Cyclone would add any latency to the
-operation due to the extra synchronization needed across replication quorums and
-participating application cores. We examine the problem for the unloaded case
+ganged operation across physical logs owning the keys in the write batch.
+A key concern here was whether Cyclone would add any latency to the
+operation due to the extra synchronization needed across replication quorums (
+shown in a previous experiment).
+We examine the problem for the unloaded case
 and small updates in Figure~\ref{fig:rocksdb_multi} for a single client with
-increasing number of keys in the batch - till 32 keys that covers all
-application cores. The line labeled Rocksdb is with no write ahead logging. We
-note an increasing latency for this baseline indicating Rocksdb takes longer
+increasing number of keys in the batch - till 32 keys.
+The line labeled Rocksdb is with no write ahead logging. We
+note an increasing latency for this baseline indicating Rocksdb itself takes longer
 with larger key batches. The existing option of Rocksdb/wal has considerably
 larger latency. Cyclone does an effective job of cutting down on this latency
-even as it needs to pay a price for synchronizing multiple quorums and
-application cores making it somewhat slower than running Rocksdb with no logging
-for batched writes. A distributed 2PC transaction still remains more expensive
-just in terms of network delay than using Cyclone.
+even as it needs to pay a price for synchronizing multiple physical
+logs. Cyclone therefore provides effective replicated logging for batched
+operations using its idea of ganged operations, while scaling the performance of
+single operations using its multiple physical logs.
 

### 2017-08-03 paper/paper.tex
Updates.

@@
 flash drive. We allocate enough buffers to keep a maximum of 32 outstanding
-requests to the SSD. To avoid having to do a synchronous metadata flush, we preallocate (using the
-posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
-the log file whenever we hit its end.
+requests to the SSD. To avoid having to do a synchronous metadata flush, we
+preallocate (using the posix\_fallocate call) a gigabyte worth of zero filled
+disk pages at the end of the log file whenever we hit its end.
 
@@
 jettisoned TCP and chose to send raw IP packets encapsulated in ethernet
-frames. We currently follow this communication model both for server to server
+frames. Cyclone takes complete control of an ethernet interface, receiving all
+packets directed to it, while the IP and ethernet addresses provide sufficient
+information to route the packet if necessary through multiple switches. We
+currently follow this communication model both for server to server
 communication between replicas as well as client server communication for
@@
 the packet for transmission to the various follower replicas. To do this we
-prepare a different packet containing an ethernet header for \emph{each}
+prepare a different packet containing a network header for \emph{each}
 targeted replica and ``chain'' the data packet to each of these headers. Each
@@
 requires a ganged operation across those logs such that all the operations
-complete together in the face of failures. Since our starting point is a single
-key value store, we constrained Cyclone's design early on to avoid the need for
-any two phase commit.
+complete together in the face of failures. One option is to have an external
+co-ordinator run a distributed transaction using two phase commit. In addition
+to requiring extra external resources to host the co-ordinator one must also
+worry about failure recovery. In addition this requires two round trips to the
+leader to complete what should have been a single batched write request from a
+client that is stateless and allowed to fail - such as with key value stores
+like Rocksdb~\cite{rocksdb}. We therefore describe a simpler and more performant
+solution called ganged operations integrated into Cyclone. This allows a client
+to issue a batch of {\tt PUT} operations as a gang simultaneously and atomically
+updating a batch of keys in a single call.
 
 The first key problem we need to deal with is that the leaders for the different
-RAFT instances can be located on different machines. To avoid having a client
-co-ordinate different machines for a single operation, we constrain the RAFT
-leader election algorithm for the different RAFT instances to converge to the
-same machine. We do this by triggering re-elections for a RAFT instance other
-than that of the first physical log until it is on the same machine as the
-\emph{first} physical log. This process is therefore resilient to failed
-machines. RAFT requires that the leader have the most up to date log. To ensure
-that the process converges we therefore ensure that the currently elected leader
-brings a majority quroum up to date before triggering a relection. We also do
-not accept any requests from clients until all leaders are co-located. This
-process is guaranteed to converge as long as all no network partitions occur
-\emph{within a machine} i.e. some RAFT instances are able to send and receive
-packets, while others are not, a possiblity we discount due to our single
-process design.
+RAFT instances to which the keys map can be located on different machines. To
+avoid having a client co-ordinate different machines for a single operation, we
+constrain the RAFT leader election algorithm for the different RAFT instances to
+converge to the same machine. We do this by triggering re-elections for a RAFT
+instance other than that of the first physical log until it is on the same
+machine as the \emph{first} physical log. This process is therefore resilient to
+failed machines. RAFT requires that the leader have the most up to date log. To
+ensure that the process converges we therefore ensure that the currently elected
+leader brings a majority quroum up to date before triggering a relection. We
+also do not accept any requests from clients until all leaders are
+co-located. We assume that no network partitions occur \emph{within a machine}
+i.e. some RAFT instances are able to send and receive packets, while others are
+not, a possiblity we discount due to our single process design.
 
@@
 ganged operation to it. If any RAFT instance has moved past that term, the
-ganged operation is then terminated by having one of the still alive participants set a
-failure flag and write the appropriate bit in the barrier mask on behalf of the
-failed participant, moving execution past the barrier wait - a technique we
-borrow from lock-free data structures~\cite{lock-free}.
+ganged operation is then terminated by having one of the still alive
+participants set a failure flag and write the appropriate bit in the barrier
+mask on behalf of the failed participant, moving execution past the barrier wait
+- a technique we borrow from lock-free data structures~\cite{lock-free}.
 
@@
          public_data[co-ordinator].view > packet.view OR
-         barrier.mask == mask of participating KV cores
+         barrier.mask == mask of participating cores
         if public_data[co-ordinator].view > packet.view OR

### 2017-08-04 paper/paper.tex
Updates.

@@
 %replace XXX with the submission number you are given from the ASPLOS submission site.
-\newcommand{\asplossubmissionnumber}{XXX}
+\newcommand{\asplossubmissionnumber}{387}
 
@@
 interface (NVMe in our case), regardless of the volume of data being
-synchronously persisted. Cyclone avoids this by making use of directly attached
+synchronously persisted. Small updates (sum of key and value data) are
+surprisingly common in observed workloads for key value pairs. For example,
+Facebook~\cite{fb_workload_analysis} report that 90\% of the space allocated in
+key value stores is for data items under 500 bytes in size.
+
+Cyclone avoids this by making use of directly attached
 persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
@@
 The second key contribution of this work is in showing that the throughput
-overheads in the network stack can be mitigated entirely in software. Existing
+bottlenecks in the network stack can be mitigated entirely in software. Existing
 work on log replication using state machine replication protocols such as
@@
 The rest of the paper is organized as follows. We describe Cyclone's system
-architecture in Section~\ref{sec:sysarch}. We describe this two level log
+architecture in Section~\ref{sec:sysarch}. We describe the two level log
 structure in Section~\ref{sec:storage}. The set of optimizations to the network
@@
 \item PUT(K, V): sets the value of key K to value V
+\item DELETE(K): deletes the key
 \end{itemize}
@@
 does so in the rearranged history. Key value stores such as RocksDB also
-support atomic writes to multiple keys and snapshots.
+support atomic writes ({\tt PUT} operations) to multiple keys, called batched
+writes and the ability to take snapshots.
 
@@
 
-Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On a
-failure of the leader replica a new leader is automatically elected - RAFT
+Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On
+failure of the leader replica, a new leader is automatically elected - RAFT
 ensures that the new leader has the most up to date log. The Cyclone client
@@
 see only a brief interruption in service. Although read requests (GET
-operations) are not logged, they are serviced in FIFO order with the write requests
-(PUT operations) currently in Cyclone's log at the moment the read request is
-received. This ensures linearizability in the face of failovers.
+operations) are not logged, they are serviced in FIFO order with the write
+requests (PUT/DELETE operations) currently in Cyclone's log at the moment the
+read request is received. This ensures linearizability in the face of failovers.
 
@@
 The two level log design in Cyclone lays the foundation for improvements to the
-network stack that we discuss next - a significant contributor to the overheads
-in Figure~\ref{fig:problem}. We use RAFT to replicate the NVM log from the
-leader replica to the followers. Each replica \emph{independently} drains
+network stack that we discuss next. We use RAFT to replicate the NVM log from
+the leader replica to the followers. Each replica \emph{independently} drains
 committed (in terms of RAFT consensus) entries from the NVM log to the
@@
 number far in excess of the few hundreds of packets we manage in
-Figure~\ref{fig:problem}. We therefore proceed by approximating a software
-packet switch for the networking component of Cyclone. However software packet
-switching does not need to deal with failure - usually requiring only a simple
-decision for each packet at the node. RAFT introduces significant
-protocol level work for each forwarded packet a challenge that we need to deal
-with to approach the desired software packet switching efficiency in the common
-case.
+Figure~\ref{fig:problem}. The key insight in removing network overheads in
+Cyclone is therefore to approximate a software packet switch for the networking
+component of Cyclone.
+
+This is however not a straightforward mapping. Software packet switches make a
+simple (usually stateless or sometimes flow associated) decision for each
+packet. In contrast, with Cyclone the role of replicas is asymmetric. Log
+entries flow from leader replica to followers and acknowledgments flow
+back. Further, the assigned role to a replica can change on reconfiguration
+after a failure. A replica must also reject packets from old configurations. The
+action associated with a received packet therefore is far more complex than
+simply appending it to a log.  The \emph{challenge} is to do this efficiently in
+software.
 
@@
 
+\subsection{Memory Management}
+\label{sec:mm}
 We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
@@
 Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
-sized pointers. Adding a level of indirection in the NVM log allows us to
-separate the FIFO ordered circular log being manipulated by RAFT from packet
-data being managed by the memory allocator of DPDK. Both the circular log and
-packet data are in NVM. An advantage of this scheme is that it makes recovery
-from NVM easy - appends to the circular pointer log are atomic and we can use
-the pointer log to recover allocator state i.e. what pieces of NVM are currently
-in use by the log.
+sized pointers to the actual packets. Adding a level of indirection in the NVM
+log allows us to separate the FIFO ordered circular log being manipulated by
+RAFT from packet data being managed by the memory allocator of DPDK. Both the
+circular log and packet data are in NVM. An advantage of this scheme is that it
+makes recovery from NVM easy - appends to the circular pointer log are atomic
+and we can use the pointer log to recover allocator state i.e. what pieces of
+NVM are currently in use by the log.
 
@@
 
-\subsection{Eliminating Data Movement}
+A key advantage of RAFT is that unlike alternatives such as
+MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
+leader is guaranteed to have the most up to date logs. There is therefore no
+case where a leader needs to receive log entries to ``fill'' holes in its logs
+from follower replicas, simplifying the protocol state machine we need to
+implement. A leader can immediately reject any responses that are not more
+current than its view (term). In turn, the persistent log operates as a double
+ended queue, with entries either being appended (or possibly deleted at
+followers) at the end and being deleted from the front (after commit). This is
+critical to efficient operation of Cyclone, as it allows a simple top level
+structure (a circular buffer of pointers) to represent the log.
+
+\subsection{Data Movement}
 \label{sec:dm}
@@
 simply manipulate packet headers to produce new packets to send on the output
-ports. We designed our implementation of raft around the same principle.
+ports. We designed our implementation of Raft around the same principle.
 
@@
 Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
-before it is multicast out to follower replicas. DPDK userspace NIC drivers
-operate in DDIO mode~\cite{ddio} where the packet is directly written into the
-CPU cache rather than first being DMAed into DRAM and then fetched by the CPU on
-demand. Since the NVM is directly attached we need to persist the packet to it
-by executing a cacheline flush ({\tt clflush}) instruction for every cacheline
-in the packet and the pointer in the pointer buffer to persist these via the
-memory bus. This is not too onerous a burden because we can use the newly
-introduced {\tt clflush-opt}~\cite{clflush_opt} instruction specifically
-intended to efficiently flush to persistent memory without the overhead of the
-serialization normally introduced by {\tt clflush}. This allows us to hit full
-memory bandwidth on present generation platforms, a quantity in excess of 200
-Gb/s per core, well above the near term speeds of network interface cards. We
-execute a single serializing {\tt sfence} before the sequence of {\tt clflushes}
-to make sure any dirty cacheline due to header manipulation related writes from
-the CPU are sent to cache.
+\emph{before} it is multicast out to follower replicas. DPDK userspace NIC
+drivers operate in DDIO mode~\cite{ddio} where the packet is directly written
+into the CPU cache rather than first being DMAed into DRAM and then fetched by
+the CPU on demand. Since the NVM is directly attached we need to persist the
+packet to it by executing a cacheline flush ({\tt clflush}) instruction for
+every cacheline in the packet and the pointer in the pointer buffer to persist
+these via the memory bus. This is not too onerous a burden because we can use
+the newly introduced {\tt clflush-opt}~\cite{clflush_opt} instruction
+specifically intended to efficiently flush to persistent memory without the
+overhead of the serialization normally introduced by {\tt clflush}. This allows
+us to hit full memory bandwidth on present generation platforms, a quantity in
+excess of 200 Gb/s per core, well above the near term speeds of network
+interface cards. We execute a single serializing {\tt sfence} before the
+sequence of {\tt clflushes} to make sure any dirty cacheline due to header
+manipulation related writes from the CPU are sent to cache.
+
+\subsection{Batching}
 
@@
 
-Finally, we discuss our choice of RAFT as a consensus protocol. A key advantage
-of RAFT is that unlike alternatives such as MultiPaxos~\cite{multipaxos} or
-Quorum based replication~\cite{quorum}, the leader is guaranteed to have the
-most up to date logs. There is therefore no case where a leader needs to receive
-log entries to ``fill'' holes in its logs from follower replicas, simplifying
-the protocol state machine we need to implement. A leader can immediately reject
-any responses that are not more current than its view (term). In turn, the
-persistent log operates as a double ended queue, with entries either being
-appended (or possibly deleted at followers) at the end and being deleted from
-the front (after commit). This is critical to efficient operation of Cyclone, as
-it allows a simple top level structure (a circular buffer of pointers) to
-represent the log.
-
 \section{Parallelism}
@@
 
-The first key question we deal with is - how do Cyclone clients decide which
-physical log to send a {\tt PUT} or {\tt GET} request to ? A key constraint here
-is that all replicas should agree on the seqeunce of updates to a key - as
-otherwise they would diverge in terms of state. Allowing updates to the same key
-to go to different physical logs would lead to a divergence as there is no
-guarantee about the order in which the independent RAFT instances would apply
-the update. Therefore all updates to a key must go to the same physcial
-log. A similar constraint applies to reads in order to ensure that any committed
-writes are seen by readers. Allowing reads to a key to go to a different RAFT
-instance than the write means that it could be served on a machine where the
-write is yet to be applied but where the machine is the leader for the RAFT
-instance. All reads and write to a key must therefore go to the same physical
+The first question we deal with is - how do Cyclone clients decide which
+physical log to send a request to ? A key constraint here is that all replicas
+should agree on the seqeunce of updates to a key - as otherwise they would
+diverge in terms of state. Allowing updates to the same key to go to different
+physical logs would lead to a divergence as there is no guarantee about the
+order in which the independent RAFT instances would apply the update. Therefore
+all updates to a key must go to the same physcial log. A similar constraint
+applies to reads in order to ensure that all committed writes are seen by
+readers. Allowing reads to a key to go to a different RAFT instance than the
+write means that it could be served on a machine where the write is yet to be
+applied. All reads and write to a key must therefore go to the same physical
 log. We achieve this by hashing the key to select the physical log.
@@
 operations. The only synchronization necessary is when doing reads or writes to
-the single shared key value stores.  The level of concurrency therefore is
+the single shared key value store.  The level of concurrency therefore is
 constrained only by the concurrency available in the software architecture of
-the persistent key value store itself, for which a number of good designs
-already exist~\cite{flodb}.
+the persistent key value store itself, for which good designs
+exist~\cite{flodb}.
 
@@
 requires a ganged operation across those logs such that all the operations
-complete together in the face of failures. One option is to have an external
-co-ordinator run a distributed transaction using two phase commit. In addition
-to requiring extra external resources to host the co-ordinator one must also
-worry about failure recovery. In addition this requires two round trips to the
+complete together even in the face of failures. One option is to have an
+external co-ordinator run a distributed transaction using two phase commit. In
+addition to requiring extra external resources to host the co-ordinator one must
+also worry about failure recovery. This also requires two round trips to the
 leader to complete what should have been a single batched write request from a
@@
 like Rocksdb~\cite{rocksdb}. We therefore describe a simpler and more performant
-solution called ganged operations integrated into Cyclone. This allows a client
-to issue a batch of {\tt PUT} operations as a gang simultaneously and atomically
-updating a batch of keys in a single call.
+solution called ganged operations integrated into Cyclone. This allows a
+stateless client to issue a batch of update operations as a gang, simultaneously
+and atomically updating a batch of keys in a single call.
 
@@
 the request to the participating logs. We make use of packet cloning primitives
-to avoid making a physical copy of the packet, generating indirect references
+to avoid making physical copies of the packet, generating indirect references
 instead. The co-ordinator also adds a ``nonce'' - a unique timestamp to the
@@
 Figure~\ref{fig:ganged_ops} is straightforward. A barrier is executed on all
-RAFT instances. Once replication is complete on all participating physical logs
-(or this is a read ganged operation), a state indicated by all necessary bits
-being set in the barrier mask, the RAFT instances simultaneously execute the
-ganged operation.
-
-Each RAFT instance always publishes the current term it is operating in. A
-failure causes a new leader to be elected moving forward the term. If the
-participants of a ganged operatation do not detect this case, they could be left
-waiting for ever as ganged operation might never successfully replicate in the
-failed over RAFT instance. We detect failure on any RAFT instance by having each
-instance publish its current term and continuously comparing the view in the
-ganged operation to it. If any RAFT instance has moved past that term, the
-ganged operation is then terminated by having one of the still alive
-participants set a failure flag and write the appropriate bit in the barrier
-mask on behalf of the failed participant, moving execution past the barrier wait
-- a technique we borrow from lock-free data structures~\cite{lock-free}.
-
-\begin{figure*}
+RAFT instances. Once replication is complete on all participating physical logs,
+a state indicated by all necessary bits being set in the barrier mask, the RAFT
+instances simultaneously execute the ganged operation.
+
+A failure causes a new leader to be elected for the affected physical log,
+moving forward the term. If the participants of a ganged operatation do not
+detect this case, they could be left waiting forever. We detect failure on any
+RAFT instance by having each instance publish its current term and continuously
+comparing the view in the ganged operation to it. If any RAFT instance has moved
+past that term, the ganged operation is then terminated.
+
+\begin{figure}
   \centering
+  \scriptsize
 \begin{verbatim}
@@
            atomic set bit for p in barrier.mask
-        while barrrer.mask != mask of all participants
+        while barrier.mask != mask of all participants
         if barrier.failed
-         Ganged replication failed. Send retry to client. return.
+         send retry to client
         else       
@@
            barrier.failed
-         Ganged replication failed. Send retry to client. return.
+         send retry to client
         else
@@
 \label{fig:ganged_ops}
-\end{figure*}
+\end{figure}
 
@@
 otherwise mentioned we use 8 physical logs (and associated RAFT instances) each
-mapped to a dedicated core. The remaining 32 cores are dedicated to the stub
+mapped to a dedicated core. The remaining cores are dedicated to the stub
 server or Rocksdb, as the case may be.
@@
 underlines the value of our ganged operation design over a distributed
-transaction that would need 4 network hops for the two phase commit (and
-replication) using an external co-ordinator, a minimum of 40us.
+transaction that would need 8 network hops for the two phase commit (and
+replication) using an external co-ordinator, a minimum of 80us.
 
@@
 
-We now evaluate Cyclone integrated with the Rocksdb persistent key value
+Next, we evaluate Cyclone integrated with the Rocksdb persistent key value
 store. Rocksdb is a complex industrial strength persistent key value store and
@@
 
-We use 8 bytes keys with either 8 bytes values or 256 byte values.
-Since our focus is replication and only update requests are replicated, we use a
-pure update workload to test the capability of the system.
+We use 8 byte keys with either 8 byte values or 256 byte values.  Since our
+focus is replication and only update requests are replicated, we use a pure
+update workload to test the capability of the system. This involves loading 100
+million key value pairs and during the test updating the value associated with a
+key picked at random.
 
@@
 but without any replication. The line almost exactly tracks the performance of
-Rockdb. As suggested by the baseline replication performance, Cyclone is able to
-provide a write ahead log with no overhead to Rocksdb. The line labeled
-'rocksdb/Cyclone 3 way' is with 3-way replication turned on. Other than a 20us
-delta due to the extra network round trip, the line almost exactly tracks
-Rocksdb performance with no logging. Cyclone therefore provides high
-availability to Rocksdb at a fraction of the cost of its existing single machine
-write ahead log. We also repeat the experiment for the larger update size in
-Figure~\ref{fig:rocksdb_256}. The conclusions are identical: Cyclone solves
-Rocksdb's write ahead logging problem.
+Rockdb. Cyclone is able to provide a write ahead log with no overhead to
+Rocksdb. The line labeled 'rocksdb/Cyclone 3 way' is with 3-way replication
+turned on. Other than a 20us delta due to the extra network round trip, the line
+almost exactly tracks Rocksdb performance with no logging. Cyclone therefore
+provides high availability to Rocksdb at a fraction of the cost of its existing
+single machine write ahead log. We also repeat the experiment for the larger
+update size in Figure~\ref{fig:rocksdb_256}. The conclusions are identical:
+Cyclone solves Rocksdb's write ahead logging problem.
 
@@
 extra 6us with ethernet for the round trip to the replica might be a concern for
-some applications. We note that mapping the commutativity of key value store
+some. We note that mapping the commutativity of key value store
 interfaces to multiple logs in Cyclone is a technique that can also be applied
@@
 date, a situation we avoid by using RAFT as described in
-Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
+Section~\ref{sec:mm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
 being able to replicate 500K requests/sec. This is without persisting the log of
@@
 operation. It also leverages software packet switching principles to replicate
-the write ahead log thus avoiding any need for offload capable network
-interfaces.  Cyclone and its Rocksdb bindings are slated for release to the open
-source community and we hope both the code and the ideas in the system itself
-will help key value store builders and researchers deal with high availabilty
-without compromising performance.
+the write ahead log, while making the best possible use of commodity networking
+hardware. These techniques enable Cyclone to provide high availability to
+persistent key value stores, without compromising on their performance. 
 \newcommand\myurl[2]{\url{#1}}

### 2017-08-05 paper/paper.tex
Updates.

@@
 network interface card (even a single one!) that would otherwise go wasted due
-to the serializing abstraction of a single log. Finally, we note that DPDKs raw
-packet forwarding capability on that configuration is close to 12M packets per
-second. We do not reach that mark due to the overhead of running the replication
-protocol rather than a simple switching decision per packet. Although batching
-goes some way in amortizing this overhead it cannot be eliminated entirely
-because consensus protocols are more complex than a simple packet switching
-decision.
+to the serializing abstraction of a single log.
 
@@
 headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
-network line rate bottleneck at this point. One can contrast this with the
-previous experiment where the bottleneck to throughput was the cost of running
-the consensus protocol for batches of packets.
+network line rate bottleneck at this point.
 

### 2017-08-05 paper/paper.tex
Updates.

@@
 
-One can draw two important conclusions from
+One can draw three important conclusions from
 Figure~\ref{fig:network_opts}. First, one can indeed treat log replication as a
@@
 path. Second, relieving the CPU from the overhead of data copies has a
-significant positive effect on performance. Finally, using multiple physical
+significant positive effect on performance. Second, using multiple physical
 logs is essential to exploiting the concurrency available at the level of the
 network interface card (even a single one!) that would otherwise go wasted due
-to the serializing abstraction of a single log.
+to the serializing abstraction of a single log. Finally we note that Cyclone
+achieves about 50\% of the linerate across the four ethernet
+ports. This is essentially the cost of running a consensus protocol in
+software.
 
@@
 headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
-network line rate bottleneck at this point.
+network line rate bottleneck at this point. It is worthwhile to compare this to
+the case with small updates, where running the consensus protocol was a
+bottleneck to reaching line rate.
 

### 2017-08-05 paper/paper.tex
Updates.

@@
 broadcast protocol together with a NIC on an FPGA. That work ignored durability,
-focusing purely on network performance for replication. Latency is
-excellent due to the FPGA that cuts out the path from CPU caches to the NIC and
-are reported to be as low as 3us for a round trip from leader to follower
-replica compared to the 6us we observe with DPDK. Notably however they reported
-7us when using TCP/IP from the FPGA rather than their own connection oriented
-network protocol. On the other hand their peak replication throughput is 6M
-replications/sec on 66 Gbps of aggregate network bandwith, comparable to the
-same mark obtained by Cyclone on an aggregate 40 Gbps of network bandwidth when
-not using the log on the SSD. We believe the general applicability afforded by
-our software only solution together with the fact that we provide persistence by
-design makes it a compelling alternative to an FPGA based solution even given
-the larger latency.
+focusing purely on network performance for replication. Latency is excellent due
+to the FPGA that cuts out the path from CPU caches to the NIC and are reported
+to be as low as 3us for a round trip from leader to follower replica compared to
+the 6us we observe with DPDK. Notably however they reported 7us when using
+TCP/IP from the FPGA rather than their own connection oriented network
+protocol. On the other hand their peak replication throughput is 4M
+replications/sec on 66 Gbps of aggregate network bandwith for small messages,
+comparable to the 6M replications/sec mark obtained by Cyclone on an aggregate
+40 Gbps of network bandwidth when \emph{also persisting the log on the SSD}. We
+believe the general applicability afforded by our software only solution
+together with the fact that we provide persistence by design makes it a
+compelling alternative to an FPGA based solution even given the larger latency.
 

### 2017-08-05 paper/paper.tex
Updates.

@@
 with 3 way replication with a lower performance overhead than Rocksdb's own
-write ahead logging.
+write ahead logging. We consider the case of a workload that more write
+intensive at 80\% writes with a smaller range of value sizes, derived from
+Facebook's VAR trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
+Figure~\ref{fig:fb2}. One can similarly conclude that Cyclone provides more
+functionality (replicated write ahead log) at a far lower performance penalty
+that Rocksdb's existing single machine write ahead log solution.
 
@@
 \label{fig:fb}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/fb2.pdf}
+\caption{Rocksdb - write heavy workload}
+\label{fig:fb2}
 \end{figure}

### 2017-08-06 paper/paper.tex
Updates.

@@
   \scriptsize
+  \bf
 \begin{verbatim}
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/flashlog_512.pdf}
-\caption{Second level log with 512 bytes payload}
-\label{fig:flashlog_512}
-\end{figure}
-
 We now evaluate ganged operations. The primary purpose of
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/multi.pdf}
+\begin{figure*}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
+  \caption{Second level log with 512 bytes payload}
+  \label{fig:flashlog_512}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/multi.pdf}
   \caption{Ganged Operations}
   \label{fig:ganged}
-\end{figure}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/kv_baseline.pdf}
+  \caption{Baseline KV replication performance}
+  \label{fig:kv_baseline}
+\end{minipage}
+\end{figure*}
 
@@
 how well Cyclone performs with Rocksdb.
-
-\begin{figure}
-\includegraphics[scale=0.6]{results2/kv_baseline.pdf}
-\caption{Baseline KV replication performance}
-\label{fig:kv_baseline}
-\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/rocksdb.pdf}
+\begin{figure*}
+  \begin{minipage}{0.33\textwidth}
+    \centering
+    \includegraphics[scale=0.45]{results2/rocksdb.pdf}
     \caption{Rocksdb - small updates}
     \label{fig:rocksdb}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.6]{results2/rocksdb_256.pdf}
+  \end{minipage} 
+  \begin{minipage}{0.33\textwidth}
+    \includegraphics[scale=0.45]{results2/rocksdb_256.pdf}
     \caption{Rocksdb - large updates}
     \label{fig:rocksdb_256}
-\end{figure}
+  \end{minipage} 
+  \begin{minipage}{0.33\textwidth}
+    \includegraphics[scale=0.45]{results2/rocksdb_multi.pdf}
+    \caption{Rocksdb - batched writes}
+    \label{fig:rocksdb_multi}
+  \end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/rocksdb_multi.pdf}
-\caption{Rocksdb - batched writes}
-\label{fig:rocksdb_multi}
-\end{figure}
-
 Our evaluation has, thus far, focused on synthetic write heavy
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/fb.pdf}
+\begin{figure*}
+\begin{minipage}{0.33\textwidth}
+\includegraphics[scale=0.45]{results2/fb.pdf}
 \caption{Rocksdb - read heavy workload}
 \label{fig:fb}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/fb2.pdf}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/fb2.pdf}
   \caption{Rocksdb - write heavy workload}
   \label{fig:fb2}
-\end{figure}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/failover.pdf}
+  \caption{Rocksdb - failover}
+  \label{fig:timeline}
+\end{minipage}
+\end{figure*}
 
@@
 60ms.
-
-\begin{figure}
-\includegraphics[scale=0.6]{results2/failover.pdf}
-\caption{Rocksdb - failover}
-\label{fig:timeline}
-\end{figure}
 
@@
 
-The scalable commutativity rule~\cite{scalable_commutativity} generalizes the
-idea of commutativity in key value store interfaces that we have used for
-horizontal scaling in Cyclone. Conveniently, it provides a tool for checking when
-API calls can commute in a history. It is a simple exercise to express a key
-value store API in COMMUTER and check that operations to different keys
-commute. This provides a theoretical basis for horizontal scaling, something we
-have omitted for brevity from this paper. The scalable commutativity rule also
-suggests that scalable key value store implementations such as
-FloDB~\cite{flodb} would necessarily have commutative interfaces, enabling more
-general applicability of Cyclone.
+The scalable commutativity rule~\cite{scalable_commutativity} states that for a
+shared memory program to scale in performance with concurrency it must also have
+an interface that is commutative for common case invocations. We have also used
+interface level commutativity to extract parallelism for the purpose of scaling
+replication performance. Others have used the same commutativity to create
+per-cpu data structures to scale KV store performance~\cite{flodb}.
 

### 2017-08-07 paper/paper.tex
Minor updates

@@
 \vspace{0.1in}
-Persistent key value stores are rapidly becoming an important component of many
+Persistent key value stores are an important component of many
 distributed data serving solutions with innovations targeted at taking advantage
@@
 \section{Introduction}
-Persistent key value stores are an increasingly important component of
+Persistent key value stores are an important component of
 datacenter scale storage services. Key value stores such as
 Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
-large amounts of effort on both the engineering and research fronts. These key
+significant efforts on both engineering and research fronts. These key
 value stores include sophisticated in-memory data structures built around Log
@@
 appended update provides the necessary high availability. Unfortunately, the
-write ahead log is a performance achilles heel for these systems - eclipsing
+write ahead log is a performance achilles heel for these systems, eclipsing
 much of the work on improving the performance of the LSM component. To
@@
 
-Cyclone avoids this by making use of directly attached
+Cyclone avoids this overhead by making use of directly attached
 persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
@@
 newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
-are significantly more expensive than standard DIMMs. This means that they will
-likely be available in small quantities. We compensate for this by periodically
-draining the NVDIMM log into a log placed on a standard SSD on the system.
+are significantly more expensive than standard DIMMs, which means that they will
+likely be available only in smaller quantities compared to flash. We compensate 
+for NVDIMM's size by periodically draining the NVDIMM log into a log placed on a
+standard SSD on the system.
 
@@
 Paxos~\cite{nopaxos} do not address throughput bottlenecks due to network stack
-related inefficiencies in the participating nodes of the quorum - a more urgent
+related inefficiencies in the participating nodes of the quorum -- a more urgent
 problem with local area replication. On the other hand, work done to address
@@
 \label{sec:sysarch}
-A persistent key value store durably stores key-value mappings accessible through a
-simple interface:
+A persistent key value store durably stores key-value mappings accessible
+through a simple interface:
 
@@
 support atomic writes ({\tt PUT} operations) to multiple keys, called batched
-writes and the ability to take snapshots.
+writes, and the ability to take snapshots.
 
@@
 servers. The client side library sends \emph{all} requests to a distinguished
-leader replica.  On the server side Cyclone accepts requests from clients and calls
-into the key value store. It then returns a response to the client. Before
+leader replica.  On the server side Cyclone accepts requests from clients, calls
+into the key value store, and returns a response to the client. Before
 executing any PUT request, Cyclone appends it to a durable log and replicates
@@
 
-Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On
-failure of the leader replica, a new leader is automatically elected - RAFT
+Cyclone uses RAFT protocol~\cite{raft} to keep the logs in sync across the
+replicas. On
+failure of the leader replica, a new leader is automatically elected -- RAFT
 ensures that the new leader has the most up to date log. The Cyclone client
-library automatically locates the new leader on a failover - clients therefore
-see only a brief interruption in service. Although read requests (GET
+library automatically locates the new leader on a failover, resulting in only a
+brief interruption in service for clients. Although read requests (GET
 operations) are not logged, they are serviced in FIFO order with the write
-requests (PUT/DELETE operations) currently in Cyclone's log at the moment the
+requests (PUT/DELETE operations) in Cyclone's log at the moment the
 read request is received. This ensures linearizability in the face of failovers.
@@
 paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
-under 50K ops/sec. This leads to a high baseline latency at even low load, as
-shown in Figure~\ref{fig:problem}. This problem leads many system designers to
+under 50K ops/sec and resulting in a high baseline latency at even low load (as
+shown in Figure~\ref{fig:problem}). This problem leads many system designers to
 underprovision key-value store shards ensuring that it operates at moderate to
@@
 placed behind a block oriented IO interface. Log appends are done to the ``NVM
-log'' placed in this non-volatile memory. While we designed Cyclone to leverage
+log'' placed in this non-volatile memory.
+%% TODO: Messaging has to be improved here. It appears as if the problems
+%% with NVDIMM (as far as Cyclone is concerned) will be solved once 3DXPoint
+%% appears -- lower cost and higher capacity than DRAM.
+While we designed Cyclone to leverage
 new non-volatile memory technologies that might be available in the
@@
 small key-value pairs. At the same time, the amount of non-volatile memory
-required is small enough to not add undue cost of the server.
+required is small enough to not add undue cost to the server.
 

### 2017-08-08 paper/paper.tex
Updates

@@
 bottleneck. It uses a small amount of non-volatile memory directly addressable
-by the CPU - such as in the form of NVDIMMs - to remove block oriented IO
+by the CPU - such as in the form of NVDIMMs or Intel 3DXPoint - to remove
+block oriented IO
 devices such as SSDs from the critical path for appending to the log. It then
@@
 NICs via multiple transmit receive queues to scale replication performance on
-multicore machines. Cyclone is able to replicate millions of small updates a
+multicore machines. Cyclone is able to replicate millions of small updates per
 second using only commodity 10 gigabit ethernet adapters. As a practical
@@
 value stores include sophisticated in-memory data structures built around Log
-Structured Merge trees~\cite{lsmtree} and are heavily tuned to extract maximum
-performance from flash-based solid state drives (SSDs).
+Structured Merge (LSM) trees~\cite{lsmtree} and are heavily tuned to extract
+maximum performance from flash-based solid state drives (SSDs).
 
@@
 ahead log to be able to showcase benefits of sophisticated extensions to LSM
-datastructures.
+data structures.
 
@@
 interface (NVMe in our case), regardless of the volume of data being
-synchronously persisted. Small updates (sum of key and value data) are
+synchronously persisted. Small updates (sum of key and value sizes) are
 surprisingly common in observed workloads for key value pairs. For example,
-Facebook~\cite{fb_workload_analysis} report that 90\% of the space allocated in
+Facebook~\cite{fb_workload_analysis} reports that 90\% of the space allocated in
 key value stores is for data items under 500 bytes in size.
 
-Cyclone avoids this overhead by making use of directly attached
-persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
-an added ultracapacitor to dump state to a small amount of attached flash) or
-newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
-are significantly more expensive than standard DIMMs, which means that they will
-likely be available only in smaller quantities compared to flash. We compensate 
-for NVDIMM's size by periodically draining the NVDIMM log into a log placed on a
-standard SSD on the system.
+Cyclone avoids this overhead by making use of a small amount of directly
+attached non-volatile memory (NVM), also called persistent memory, on the
+server.
+This can be in the form of NVDIMMs
+(DIMMs with an added ultracapacitor to dump state to a small amount of attached 
+flash) -or- newer and potentially cheaper (but slower) forms of
+persistent memory~\cite{pmfs}.
+%For current deployments, NVDIMMs are significantly more expensive than standard
+%DIMMs, which means that they will %likely be available only in smaller
+%quantities compared to flash.
+In cases where the log does not entirely fit in NVM, we periodically drain
+the NVM log into a log placed on a standard SSD on the system.
 
@@
 
+Both the storage (NVM, SSD) as well as the NIC expose significant parallelism
+that is left unused when Cyclone is used to replicate a single log. The third
+key contribution of this work is to show how one can exploit this parallelism in
+Cyclone by making use of the commutativity inherent in key-value store
+interfaces and mapping it to the available parallelism in the system via
+multiple instances of the same consensus protocol.
+
 The rest of the paper is organized as follows. We describe Cyclone's system
@@
 structure in Section~\ref{sec:storage}. The set of optimizations to the network
-stack is decribed in Section~\ref{sec:network}. Both the storage (NVM, SSD) as
-well as the NIC expose significant parallelism that is left unused when Cyclone
-is used to replicate a single log. To exploit this parallelism we use Cyclone to
-replicate multiple logs, each using a different instance of the RAFT consensus
-protocol. We show how to exploit the commutativity inherent in key-value store
-interfaces to make use of this parallelism in Cyclone's design in
-Section~\ref{sec:parallelism}.  A detailed evaluation of Cyclone's replication
+stack is decribed in Section~\ref{sec:network}. We show how Cyclone can be scaled to make use of available parallelism on the system in
+Section~\ref{sec:parallelism}..
+A detailed evaluation of Cyclone's replication
 performance is provided in the context of RocksDB - a popular persistent key
@@
 \item PUT(K, V): sets the value of key K to value V
-\item DELETE(K): deletes the key
+\item DELETE(K): deletes the key K
 \end{itemize}
@@
 
-Cyclone uses RAFT protocol~\cite{raft} to keep the logs in sync across the
+Cyclone uses the RAFT protocol~\cite{raft} to keep the logs in sync across the
 replicas. On
@@
 to obviate the need to make this tradeoff in the first place. We assume a small
-amount of non-volatile memory directly addressable by the CPU rather than being
+amount of non-volatile memory directly addressable by the CPU, rather than being
 placed behind a block oriented IO interface. Log appends are done to the ``NVM
-log'' placed in this non-volatile memory.
-%% TODO: Messaging has to be improved here. It appears as if the problems
-%% with NVDIMM (as far as Cyclone is concerned) will be solved once 3DXPoint
-%% appears -- lower cost and higher capacity than DRAM.
-While we designed Cyclone to leverage
-new non-volatile memory technologies that might be available in the
-future~\cite{3dxpoint}, current non-volatile memory options such are NVDIMMs are
-more expensive that DRAM. Therefore, for Cyclone to be economical we cannot
-assume that a sufficient amount of NVM is available to hold the entire log. The
-log entries can only be garbage collected when the in-memory component of the
-key value store is checkpointed to secondary storage. The growing size of
-available DRAM means that also provisioning sufficient NVDIMMs to hold the
-corresponding log of updates is not feasible given that NVDIMMs are
-significantly more expensive than standard volatile DIMMs.  To illustrate this,
-at the time of writing this paper, an 8 GB DDR4 NVDIMM was retailing at \$612,
-while a plain 8 GB DD4 DIMM was retailing at \$178.  We therefore resort to
-using a second larger log level placed on a flash SSD. Entries are drained from
-the NVM log to this flash log in conveniently large units that are a multiple of
-4KB - the optimum IO unit for flash SSDs. This ensures we get the best possible
+log'' placed in this non-volatile memory. However, directly attached persistent 
+memory in the form of NVDIMMs today or new memory technologies in the
+future~\cite{pmfs} will be of lesser capacity than traditional (NAND based)
+flash available through an
+SSD. Providing the same capacity as existing flash based logs for key value
+stores therefore requires us to provide a second level of the log
+placed on a flash SSD (called flashlog).  Entries are drained from
+the NVM log to this flashlog in conveniently large units that are a multiple of
+4KB -- the optimum IO unit for flash SSDs. This ensures we get the best possible
 throughput from a flash device without paying the price of synchronous IO for
@@
 cached in the drive's volatile cache are written to the SSD using a backup
-capacitor in the event of power failure - a property of many data center class
+capacitor in the event of power failure -- a property of many data center class
 SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
-evaluation. Together, these two assumptions mean that we can recover a consistent
-prefix of the log on a power failure.
+evaluation. Together, these two assumptions mean that we can recover a
+consistent prefix of the log on a power failure.
 
@@
 The two level log design in Cyclone lays the foundation for improvements to the
-network stack that we discuss next. We use RAFT to replicate the NVM log from
+network stack that we discuss next. We use the RAFT protocol to replicate the
+NVM log from
 the leader replica to the followers. Each replica \emph{independently} drains
@@
 movement over the network, unlike systems that need to deal with such
-problems~\cite{reflex}. Replication reduces to efficiently sending a block of
-data already present in directly attached memory over the network.
+problems~\cite{reflex}. The task of replication reduces to that of efficiently
+sending a block of data already present in directly attached memory over the
+network.
 
@@
 community when building software packet switches. Software packet switches reach
-impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, a
-number far in excess of the few hundreds of packets we manage in
+impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, 
+a number far in excess of the few hundreds of packets we manage in
 Figure~\ref{fig:problem}. The key insight in removing network overheads in
@@
 simple (usually stateless or sometimes flow associated) decision for each
-packet. In contrast, with Cyclone the role of replicas is asymmetric. Log
-entries flow from leader replica to followers and acknowledgments flow
+packet. In contrast, the role of replicas in Cyclone is asymmetric. Log
+entries flow from the leader replica to followers and acknowledgments flow
 back. Further, the assigned role to a replica can change on reconfiguration
@@
 action associated with a received packet therefore is far more complex than
-simply appending it to a log.  The \emph{challenge} is to do this efficiently in
-software.
+simply forwarding it to a log.  The \emph{challenge} is to do this efficiently
+in software.
 
@@
 ethernet frames via the transmit and receive queues on the NIC. DPDK is often
-used by developers of software packet switches and therefore we also leverage
+used by developers of software packet switches and therefore we leverage
 DPDK as a library for building Cyclone.
 
-DPDK does not by itself provide a TCP stack. This is not a problem. RAFT (and
+DPDK does not by itself provide a TCP stack. But this is not a problem, since
+RAFT (and
 indeed most consensus protocols) tolerate network losses and reordering by
@@
 currently follow this communication model both for server to server
-communication between replicas as well as client server communication for
+communication between replicas as well as client to server communication for
 requests and responses. Although a detailed evaluation is made later, switching
@@
 circular log and packet data are in NVM. An advantage of this scheme is that it
-makes recovery from NVM easy - appends to the circular pointer log are atomic
+makes recovery from NVM easy -- appends to the circular pointer log are atomic
 and we can use the pointer log to recover allocator state i.e. what pieces of
@@
 simply manipulate packet headers to produce new packets to send on the output
-ports. We designed our implementation of Raft around the same principle.
+ports. We designed our implementation of the RAFT protocol based on the same
+principle.
 
@@
 This entire process is done without making any copies of the received packet.
-To illustrate how this is achieved, Figure~\ref{fig:packet_layout} shows how
+Figure~\ref{fig:packet_layout} illustrates how
 Cyclone manipulates packet layouts across the two steps of prepending a RAFT
@@
 overhead of the serialization normally introduced by {\tt clflush}. This allows
-us to hit full memory bandwidth on present generation platforms, a quantity in
-excess of 200 Gb/s per core, well above the near term speeds of network
-interface cards. We execute a single serializing {\tt sfence} before the
-sequence of {\tt clflushes} to make sure any dirty cacheline due to header
-manipulation related writes from the CPU are sent to cache.
+us to hit full memory bandwidth on current generation platforms, a quantity in
+excess of 200 Gb/s per core which is well above the near term speeds of 
+commodity NICs. We execute a single serializing {\tt sfence} 
+before the sequence of {\tt clflushes} to make sure any dirty cachelines due to
+header manipulation related writes from the CPU are sent to cache.
 
@@
 network performance of replication in Cyclone, it still does not come close to
-saturating the capabilities of even the commodity network interface card that we
-use. This is because NICs today encapsulate significant parallelism in terms of
+saturating the capabilities of even the commodity NICs that we use. This is
+because NICs today encapsulate significant parallelism in terms of
 multiple send and receive queue pairs. Exploiting this capability to improve
@@
 multiple copies of the RAFT consensus protocol each maintaining and replicating
-its own two level log. We now refer to these logs as physical logs. All these
+its own two level log. We refer to these logs as physical logs. All these
 instances however exist in the same process address space as the key value store
@@
 
-The first question we deal with is - how do Cyclone clients decide which
-physical log to send a request to ? A key constraint here is that all replicas
-should agree on the seqeunce of updates to a key - as otherwise they would
+The first question we deal with is -- how do Cyclone clients decide which
+physical log to send a request to? A key constraint here is that all replicas
+should agree on the seqeunce of updates to a key, as otherwise they would
 diverge in terms of state. Allowing updates to the same key to go to different
@@
 write means that it could be served on a machine where the write is yet to be
-applied. All reads and write to a key must therefore go to the same physical
+applied. Therefore, all reads and writes to a key must go to the same physical
 log. We achieve this by hashing the key to select the physical log.
@@
 to ensure linearizability in the face of failure. Clearly the system is
-linearizable for a single key since it involves only a single physical log. In
-addition, one can switch operations to different keys in a history without
-affecting the correctness of the history because operations to different keys
-are commutative. Any history of the system is therefore linearizable.
+linearizable for a single key since it involves only a single physical log an
+d we always service existing requests in the physical log before responding to read requests. In addition, one can switch operations to different keys in a
+history without affecting its correctness because operations to different keys
+are commutative. This allows us to arrive at a history that explains the key
+value store responses, is correct and makes sure that a request responded to
+before subsequent requests in the original history follows that same order in
+the linearized history. We discuss how to handle requests that atomically
+manipulate multiple keys in the next section.
 
@@
 like Rocksdb~\cite{rocksdb}. We therefore describe a simpler and more performant
-solution called ganged operations integrated into Cyclone. This allows a
-stateless client to issue a batch of update operations as a gang, simultaneously
-and atomically updating a batch of keys in a single call.
+solution called ``ganged operations" in Cyclone. Ganged operations allow a
+stateless client to issue a batch of update operations as a ``gang",
+simultaneously and atomically updating a batch of keys in a single call.
 
@@
 Cyclone. As with other work~\cite{faast}, we use DRAM on the machines to proxy
-for NVDIMMs where necessary - the persistent memory needed never exceeds 64 MB
+for NVDIMMs where necessary - the NVM needed never exceeds 64 MB
 regardless of the size of the key value store or second level log on flash. We

### 2017-08-08 paper/paper.tex
Updates.

@@
 \vspace{0.1in}
-Persistent key value stores are an important component of many
-distributed data serving solutions with innovations targeted at taking advantage
-of growing flash speeds. Unfortunately their performance is hampered by the need
-to maintain and replicate a write ahead log to guarantee availability in the
-face of machine and storage failures. Cyclone is a replicated log plug-in for
-key value stores that systematically addresses various sources of this
-bottleneck. It uses a small amount of non-volatile memory directly addressable
-by the CPU - such as in the form of NVDIMMs or Intel 3DXPoint - to remove
-block oriented IO
-devices such as SSDs from the critical path for appending to the log. It then
-addresses network overheads using an implementation of the RAFT consensus
-protocol that is designed around a userspace network stack to relieve the CPU of
-the burden of data copies. Finally, it provides a way to efficiently map the
-commutativity in key-value store APIs to the parallelism available in commodity
-NICs via multiple transmit receive queues to scale replication performance on
-multicore machines. Cyclone is able to replicate millions of small updates per
-second using only commodity 10 gigabit ethernet adapters. As a practical
-application, we use it to improve the performance and availability of RocksDB, a
-popular persistent key value store by an order of magnitude when compared to its
-own write ahead log without replication.
+Persistent key value stores are an important component of many distributed data
+serving solutions with innovations targeted at taking advantage of growing flash
+speeds. Unfortunately their performance is hampered by the need to maintain and
+replicate a write ahead log to guarantee availability in the face of machine and
+storage failures. Cyclone is a replicated log plug-in for key value stores that
+systematically addresses various sources of this bottleneck. It uses a small
+amount of non-volatile memory directly addressable by the CPU - such as in the
+form of NVDIMMs or Intel 3DXPoint - to remove block oriented IO devices such as
+SSDs from the critical path for appending to the log. It then addresses network
+overheads using an implementation of the RAFT consensus protocol that is
+designed around a userspace network stack to relieve the CPU of the burden of
+data copies. Finally, it provides a way to efficiently map the commutativity in
+key-value store APIs to the parallelism available in commodity NICs via multiple
+transmit receive queues to scale replication performance on multicore
+machines. Cyclone is able to replicate millions of small updates per second
+using only commodity 10 gigabit ethernet adapters. As a practical application,
+we use it to improve the performance and availability of RocksDB, a popular
+persistent key value store by an order of magnitude when compared to its own
+write ahead log without replication.
 \end{abstract}
@@
 \section{Introduction}
-Persistent key value stores are an important component of
-datacenter scale storage services. Key value stores such as
-Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
-significant efforts on both engineering and research fronts. These key
-value stores include sophisticated in-memory data structures built around Log
-Structured Merge (LSM) trees~\cite{lsmtree} and are heavily tuned to extract
-maximum performance from flash-based solid state drives (SSDs).
+Persistent key value stores are an important component of datacenter scale
+storage services. Key value stores such as Rocksdb~\cite{rocksdb},
+LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent significant efforts on
+both engineering and research fronts. These key value stores include
+sophisticated in-memory data structures built around Log Structured Merge (LSM)
+trees~\cite{lsmtree} and are heavily tuned to extract maximum performance from
+flash-based solid state drives (SSDs).
 
@@
 appended update provides the necessary high availability. Unfortunately, the
-write ahead log is a performance achilles heel for these systems, eclipsing
-much of the work on improving the performance of the LSM component. To
-illustrate the impact of the write ahead log, consider
-Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
-Rocksdb without the write ahead log. The performance when persisting the write
-ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
-line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
-without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
-persisting every update to the log or replicating it using TCP/IP causes
-performance to drop by an order of magnitude (note the log scale on the
-x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
-the write ahead log~\cite{samza}, depending on upper layers to provide
-availability by co-ordinating replicas. On the other side of the spectrum, key
-value store research prototypes such as FloDB~\cite{flodb} turn off the write
-ahead log to be able to showcase benefits of sophisticated extensions to LSM
-data structures.
+write ahead log is a performance achilles heel for these systems, eclipsing much
+of the work on improving the performance of the LSM component. To illustrate the
+impact of the write ahead log, consider Figure~\ref{fig:problem}. The line
+marked 'Rocksdb' shows the performance of Rocksdb without the write ahead
+log. The performance when persisting the write ahead log without replicating it
+is shown as the line marked 'rocksdb/WAL'. The line marked 'rocksdb/3 way rep.'
+is for simply replicating the log three ways without persisting it, using
+RAFT~\cite{raft} running over TCP/IP. Either persisting every update to the log
+or replicating it using TCP/IP causes performance to drop by an order of
+magnitude (note the log scale on the x-axis). It is therefore no surprise that
+deployments of Rocksdb often turn off the write ahead log~\cite{samza},
+depending on upper layers to provide availability by co-ordinating replicas. On
+the other side of the spectrum, key value store research prototypes such as
+FloDB~\cite{flodb} turn off the write ahead log to be able to showcase benefits
+of sophisticated extensions to LSM data structures.
 
@@
 \end{figure}
+
+These two performance bottlenecks illustrated in Figure~\ref{fig:problem} are
+not \emph{independent}. In particular, durability semantics mean that the
+key-value store  cannot respond to the client until the operation has been
+persisted to the log, since it must be completed even on a crash and subsequent
+recovery. Therefore, addressing the network bottleneck is pointless without also
+addressing the storage bottleneck. Cyclone shows how a small amount of
+non-volatile memory can be used to address both bottlenecks.
 
@@
 attached non-volatile memory (NVM), also called persistent memory, on the
-server.
-This can be in the form of NVDIMMs
-(DIMMs with an added ultracapacitor to dump state to a small amount of attached 
-flash) -or- newer and potentially cheaper (but slower) forms of
-persistent memory~\cite{pmfs}.
+server.  This can be in the form of NVDIMMs (DIMMs with an added ultracapacitor
+to dump state to a small amount of attached flash) -or- newer and potentially
+cheaper (but slower) forms of persistent memory~\cite{pmfs}.
 %For current deployments, NVDIMMs are significantly more expensive than standard
@@
 %quantities compared to flash.
-In cases where the log does not entirely fit in NVM, we periodically drain
-the NVM log into a log placed on a standard SSD on the system.
+In cases where the log does not entirely fit in NVM, we periodically drain the
+NVM log into a log placed on a standard SSD on the system.
 
@@
 structure in Section~\ref{sec:storage}. The set of optimizations to the network
-stack is decribed in Section~\ref{sec:network}. We show how Cyclone can be scaled to make use of available parallelism on the system in
-Section~\ref{sec:parallelism}..
-A detailed evaluation of Cyclone's replication
+stack is decribed in Section~\ref{sec:network}. We show how Cyclone can be
+scaled to make use of available parallelism on the system in
+Section~\ref{sec:parallelism}..  A detailed evaluation of Cyclone's replication
 performance is provided in the context of RocksDB - a popular persistent key
@@
 
-Cyclone uses the RAFT protocol~\cite{raft} to keep the logs in sync across the
-replicas. On
-failure of the leader replica, a new leader is automatically elected -- RAFT
-ensures that the new leader has the most up to date log. The Cyclone client
-library automatically locates the new leader on a failover, resulting in only a
-brief interruption in service for clients. Although read requests (GET
-operations) are not logged, they are serviced in FIFO order with the write
-requests (PUT/DELETE operations) in Cyclone's log at the moment the
-read request is received. This ensures linearizability in the face of failovers.
+Cyclone uses the RAFT consensus protocol~\cite{raft} to keep the logs in sync
+across the replicas. On failure of the leader replica, a new leader is
+automatically elected -- RAFT ensures that the new leader has the most up to
+date log. The Cyclone client library automatically locates the new leader on a
+failover, resulting in only a brief interruption in service for
+clients. Although read requests (GET operations) are not logged, they are
+serviced in FIFO order with the write requests (PUT/DELETE operations) in
+Cyclone's log at the moment the read request is received. This ensures
+linearizability in the face of failovers.
 
@@
 
-The two level log design in Cyclone lays the foundation for improvements to the
-network stack that we discuss next. We use the RAFT protocol to replicate the
-NVM log from
-the leader replica to the followers. Each replica \emph{independently} drains
-committed (in terms of RAFT consensus) entries from the NVM log to the
-flashlog. Block IO therefore is no longer a consideration when optimizing data
-movement over the network, unlike systems that need to deal with such
-problems~\cite{reflex}. The task of replication reduces to that of efficiently
-sending a block of data already present in directly attached memory over the
-network.
+In this section we demonstrated how Cyclone uses a small amount of non-volatile
+memory to improve storage bottlenecks when maintaining a durable log on a single
+machine. In the next section, we deal with replicating this log efficiently
+across multiple machines.
 
@@
 \label{sec:network}
+Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the log of
+operations across machines. It promises the client that any operation responded
+to is durable and its results \emph{available} as long as a majority of replicas
+continue to function. Cyclone's failure model therefore is \emph{fail-recovery},
+where nodes that can recover from failure simply continue the protocol. This
+requires RAFT to persist every log entry \emph{before} it sends it out for
+replication. Storage overheads are therefore not independent of network
+overheads and we must also deal with the storage overhead to improve the
+performance of replication.
+
+We leverage the two level log from the previous section to remove storage
+bottlenecks from the critical path of replication in much the same way we did
+for logging on a single machine. RAFT is only aware of and replicates the top
+level NVM log. Each replica \emph{independently} drains committed (in terms of
+RAFT consensus) entries from the NVM log to the flashlog. Block IO therefore is
+no longer a consideration when optimizing data movement over the network, unlike
+systems that need to deal with such problems~\cite{reflex}. The task of
+replication reduces to that of efficiently sending a block of data already
+present in directly attached memory over the network.
 
@@
 simple (usually stateless or sometimes flow associated) decision for each
-packet. In contrast, the role of replicas in Cyclone is asymmetric. Log
-entries flow from the leader replica to followers and acknowledgments flow
+packet. In contrast, the role of replicas in Cyclone is asymmetric. Log entries
+flow from the leader replica to followers and acknowledgments flow
 back. Further, the assigned role to a replica can change on reconfiguration
@@
 DPDK does not by itself provide a TCP stack. But this is not a problem, since
-RAFT (and
-indeed most consensus protocols) tolerate network losses and reordering by
-design due to the need to support asynchronous communication. In addition, most
-datacenter networks rarely drop or reorder packets and provide full bisection
-bandwith between servers that might serve as Cyclone replicas. We therefore
-jettisoned TCP and chose to send raw IP packets encapsulated in ethernet
-frames. Cyclone takes complete control of an ethernet interface, receiving all
-packets directed to it, while the IP and ethernet addresses provide sufficient
-information to route the packet if necessary through multiple switches. We
-currently follow this communication model both for server to server
+RAFT (and indeed most consensus protocols) tolerate network losses and
+reordering by design due to the need to support asynchronous communication. In
+addition, most datacenter networks rarely drop or reorder packets and provide
+full bisection bandwith between servers that might serve as Cyclone replicas. We
+therefore jettisoned TCP and chose to send raw IP packets encapsulated in
+ethernet frames. Cyclone takes complete control of an ethernet interface,
+receiving all packets directed to it, while the IP and ethernet addresses
+provide sufficient information to route the packet if necessary through multiple
+switches. We currently follow this communication model both for server to server
 communication between replicas as well as client to server communication for
@@
 to ensure linearizability in the face of failure. Clearly the system is
-linearizable for a single key since it involves only a single physical log an
-d we always service existing requests in the physical log before responding to read requests. In addition, one can switch operations to different keys in a
+linearizable for a single key since it involves only a single physical log an d
+we always service existing requests in the physical log before responding to
+read requests. In addition, one can switch operations to different keys in a
 history without affecting its correctness because operations to different keys
@@
 
-Our evaluation has, thus far, focused on synthetic write heavy
-workloads. To illustrate that Cyclone can also be useful for read-heavy
-workloads, we configured our clients to generate key value requests that mimic
-the distribution of value sizes in Facebook's ETCD
-trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read
-heavy (95\% reads). Figure~\ref{fig:fb} illustrates that even with a
-read-dominated workload, Cyclone is effective at providing a write ahead log
-with 3 way replication with a lower performance overhead than Rocksdb's own
-write ahead logging. We consider the case of a workload that more write
-intensive at 80\% writes with a smaller range of value sizes, derived from
-Facebook's VAR trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
+Our evaluation has, thus far, focused on synthetic write heavy workloads. To
+illustrate that Cyclone can also be useful for read-heavy workloads, we
+configured our clients to generate key value requests that mimic the
+distribution of value sizes in Facebook's ETCD
+trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read heavy (95\%
+reads). Figure~\ref{fig:fb} illustrates that even with a read-dominated
+workload, Cyclone is effective at providing a write ahead log with 3 way
+replication with a lower performance overhead than Rocksdb's own write ahead
+logging. We consider the case of a workload that more write intensive at 80\%
+writes with a smaller range of value sizes, derived from Facebook's VAR
+trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
 Figure~\ref{fig:fb2}. One can similarly conclude that Cyclone provides more
@@
 at 4GB/s. In contrast, Cyclone can replicate data three ways using 4*10 Gbps
-commodity ethernet NICs at a peak of roughly 3.7 GB/s. This is with
-Cyclone is also running the RAFT consensus protocol, while Mojim is doing simple
-mirroring of main memory. One can conclude that network offload via RDMA and
-Infiniband is not an a-priori requirement for high throughput replication. This
-is made possible by Cyclone's approach that relieves the CPU from the burden of
-data copies. On the other hand, the latency of a hop on commodity ethernet at 5 us with
-our DPDK stack is larger than the close to 2us latency with
+commodity ethernet NICs at a peak of roughly 3.7 GB/s. This is with Cyclone is
+also running the RAFT consensus protocol, while Mojim is doing simple mirroring
+of main memory. One can conclude that network offload via RDMA and Infiniband is
+not an a-priori requirement for high throughput replication. This is made
+possible by Cyclone's approach that relieves the CPU from the burden of data
+copies. On the other hand, the latency of a hop on commodity ethernet at 5 us
+with our DPDK stack is larger than the close to 2us latency with
 Infiniband. Depending on the use case for the key value store application the
 extra 6us with ethernet for the round trip to the replica might be a concern for
-some. We note that mapping the commutativity of key value store
-interfaces to multiple logs in Cyclone is a technique that can also be applied
-to RDMA. Mojim confirms that concurrency is a significant
-determinant of performance when replicating over RDMA.
+some. We note that mapping the commutativity of key value store interfaces to
+multiple logs in Cyclone is a technique that can also be applied to RDMA. Mojim
+confirms that concurrency is a significant determinant of performance when
+replicating over RDMA.
 
@@
 
-There is an in progress transformation of durable memory in datacenters
-with the advents of 3DXPoint. Researchers have speculated on the impact of cheap
-and high volume directly attached 3DXPoint~\cite{tiering}. Memory vendors have
-also begun shipping software libraries such as NVML~\cite{nvml} for building durable
-in-memory data structures using undo logging. A durable key value store using
-3DXPoint would be an idea target for Cyclone because each NVM log entry can be
-discarded as soon as it is applied to the persistent key value store removing
-the need for the flashlog. This would allow Cyclone to operate at pure packet
-switching speeds, a goal we hope to achieve in the future with a single level
-log for NVML data store applications.
+Database practitioners have considered the utility of byte adressable
+non-volatile memory in improving logging for
+databases~\cite{nvram_log}. However, that work proposes placing the entire log
+in NVM, presuming sufficient availability of such memory. It also does not
+consider replication.
 
@@
 \section{Conclusion}
-Cyclone addresses two fundamental difficulties in adding high availability to
-persistent key value stores. It leverage a small amount of non-volatile memory
-to avoid paying the cost of the block IO interface for every update
-operation. It also leverages software packet switching principles to replicate
-the write ahead log, while making the best possible use of commodity networking
-hardware. These techniques enable Cyclone to provide high availability to
+Cyclone shows how one can leverage a small amount of non-volatile memory to
+address two fundamental difficulties in adding high availability to persistent
+key value stores. It leverage a small amount of non-volatile memory to avoid
+paying the cost of the block IO interface for every update operation. This
+removes block IO latency from the task of appending updates to the log as well
+as sending it out over the network for replication. In turn, this enables us to
+apply software packet switching techniques and concurrency to improve the
+performance of and make full use of available resources when replicating log
+entries over the network. This enables Cyclone to provide high availability to
 persistent key value stores, without compromising on their performance. 

### 2017-08-08 paper/paper.tex
more updates.

@@
 These two performance bottlenecks illustrated in Figure~\ref{fig:problem} are
-not \emph{independent}. In particular, durability semantics mean that the
-key-value store  cannot respond to the client until the operation has been
-persisted to the log, since it must be completed even on a crash and subsequent
-recovery. Therefore, addressing the network bottleneck is pointless without also
-addressing the storage bottleneck. Cyclone shows how a small amount of
-non-volatile memory can be used to address both bottlenecks.
+\emph{not independent}, they cannot be addressed in isolation. In particular,
+durability semantics mean that the key-value store cannot respond to the client
+until the operation has been persisted to the log, since it must be completed
+even on a crash and subsequent recovery. Therefore, addressing the network
+bottleneck is pointless without also addressing the storage bottleneck. Cyclone
+shows how a small amount of non-volatile memory can be used to address both
+bottlenecks.
 
@@
 
-\begin{figure}
+\begin{figure*}
+  \begin{minipage}{.2\textwidth}
+    \scriptsize
 \begin{verbatim}
@@
 \end{verbatim}
-\caption{Cyclone event handling}
+\caption{Event handling}
 \label{fig:control_plane}
-\end{figure}
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
+  \caption{Cyclone network packet layout}
+  \label{fig:packet_layout}
+\end{minipage}
+\begin{minipage}{.3\textwidth}
+  \centering
+  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \caption{Batching}
+  \label{fig:batching}
+\end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.35]{figures2/network_packet.pdf}
-  \caption{Cyclone network packet layout}
-  \label{fig:packet_layout}
-\end{figure}
-
 We now turn our attention to the persistence step in
@@
 \subsection{Batching}
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.4]{figures2/batching.pdf}
-  \caption{Batching}
-  \label{fig:batching}
-\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/replicas.pdf}
+\begin{figure*}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/replicas.pdf}
   \caption{Impact of replica count}
   \label{fig:replicas}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.6]{results2/512.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/512.pdf}
   \caption{Impact of payload size}
   \label{fig:payload}
-\end{figure}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/flashlog.pdf}
+  \caption{Adding second level log}
+  \label{fig:flashlog}
+\end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/flashlog.pdf}
-\caption{Impact of adding second level log}
-\label{fig:flashlog}
-\end{figure}
-
 We now evaluate ganged operations. The primary purpose of
@@
   \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
-  \caption{Second level log with 512 bytes payload}
+  \caption{Second level log - 512 bytes payload}
   \label{fig:flashlog_512}
@@
   \includegraphics[scale=0.45]{results2/kv_baseline.pdf}
-  \caption{Baseline KV replication performance}
+  \caption{Baseline KV replication}
   \label{fig:kv_baseline}
@@
 using commodity network hardware. In contrast, Mojim~\cite{mojim} pairs directly
-attached persistent memory with RDMA to replicate persistent data structures. A
-comparison is hard since Mojim is designed for the hypothetical scenario where
-there is enough persistent memory to hold all persistent data, while Cyclone is
-designed for the practical case where both the write-ahead log and key value
-store data structures need to be maintained largely on flash. Nevertheless we
-can compare their replication performance for the case where we use only the NVM
-log in Cyclone. On 40Gbps infiniband, Mojim reported being able to mirror data
-at 4GB/s. In contrast, Cyclone can replicate data three ways using 4*10 Gbps
-commodity ethernet NICs at a peak of roughly 3.7 GB/s. This is with Cyclone is
-also running the RAFT consensus protocol, while Mojim is doing simple mirroring
-of main memory. One can conclude that network offload via RDMA and Infiniband is
-not an a-priori requirement for high throughput replication. This is made
-possible by Cyclone's approach that relieves the CPU from the burden of data
-copies. On the other hand, the latency of a hop on commodity ethernet at 5 us
-with our DPDK stack is larger than the close to 2us latency with
-Infiniband. Depending on the use case for the key value store application the
-extra 6us with ethernet for the round trip to the replica might be a concern for
-some. We note that mapping the commutativity of key value store interfaces to
-multiple logs in Cyclone is a technique that can also be applied to RDMA. Mojim
-confirms that concurrency is a significant determinant of performance when
-replicating over RDMA.
+attached persistent memory with RDMA to replicate persistent data
+structures. Mojim is a weak replication scheme in that none of the numerous
+\emph{evaluated} modes of Mojim provide any means to keep more than two replicas in
+strong synchronization. Its multi-tiered replication could lose committed
+updates and therefore its guarantees are too weak to be used as a replacement
+for RocksDB's write ahead log. Further, Mojim makes the somewhat strange claim
+that ``Paxos-like protocols'' require two networking round trips and therefore
+are somehow inefficient compared to mirroring. This is not true, as we have
+demonstrated with Cyclone where the leader responds as soon as it hears back
+from a majority quorum, the critical path to replication therefore being one
+network round trip.  A performance comparison with Mojim is hard, since it is
+designed for the hypothetical scenario where there is enough persistent memory
+to hold all persistent data, while Cyclone is designed for the practical case
+where both the write-ahead log and key value store data structures need to be
+maintained largely on flash. Nevertheless we can compare their replication
+performance for the case where we use only the NVM log in Cyclone. On 40Gbps
+infiniband, Mojim reported being able to mirror data at 4GB/s. In contrast,
+Cyclone can replicate data three ways using 4*10 Gbps commodity ethernet NICs at
+a peak of roughly 3.7 GB/s. This is with Cyclone is also running the RAFT
+consensus protocol, while Mojim is doing simple mirroring of main memory. One
+can conclude that network offload via RDMA and Infiniband is not an a-priori
+requirement for high throughput replication. This is made possible by Cyclone's
+approach that relieves the CPU from the burden of data copies. On the other
+hand, the latency of a hop on commodity ethernet at 5 us with our DPDK stack is
+larger than the close to 2us latency with Infiniband. Depending on the use case
+for the key value store application the extra 6us with ethernet for the round
+trip to the replica might be a concern for some. We note that mapping the
+commutativity of key value store interfaces to multiple logs in Cyclone is a
+technique that can also be applied to RDMA. Mojim confirms that concurrency is a
+significant determinant of performance when replicating over RDMA.
 
@@
 Paxos. It uses deterministic multi-threading to ensure replicas converge to the
-same state as opposed to semantic equivalence in Cyclone using the commutativity
-of the key value store interface. Although CRANE is more generally applicable in
-that it makes no assumptions about the application API, deterministic
-multithreading comes with high overheads for applications that store large
-amounts of data - CRANE reports around a 2X slowdown for MySQL, likely rendering
+same state as opposed to semantic equivalence in Cyclone using API
+commutativity.  CRANE reports around a 2X slowdown for MySQL, likely rendering
 it unsuitable as a replacement for the write ahead log in key value stores.

### 2017-08-09 paper/paper.tex
Updates.

@@
 using only commodity 10 gigabit ethernet adapters. As a practical application,
-we use it to improve the performance and availability of RocksDB, a popular
+we use it to improve the performance (and availability) of RocksDB, a popular
 persistent key value store by an order of magnitude when compared to its own
@@
 that this is not the case. Rather, once persistence is provided by directly
-attached non-volatile memory eliminating block IO from the persistence step in
+attached non-volatile memory, thereby eliminating block IO from the persistence 
+step in
 replication protocols, the problem becomes equivalent to multicasting the same
@@
 scaled to make use of available parallelism on the system in
-Section~\ref{sec:parallelism}..  A detailed evaluation of Cyclone's replication
+Section~\ref{sec:parallelism}. A detailed evaluation of Cyclone's replication
 performance is provided in the context of RocksDB - a popular persistent key
@@
 Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the log of
-operations across machines. It promises the client that any operation responded
+operations across machines. Cyclone provides guarantee to the client that any
+operation responded
 to is durable and its results \emph{available} as long as a majority of replicas
 continue to function. Cyclone's failure model therefore is \emph{fail-recovery},
-where nodes that can recover from failure simply continue the protocol. This
+where nodes that can recover from failure simply continue to participate in
+the protocol. This
 requires RAFT to persist every log entry \emph{before} it sends it out for
 replication. Storage overheads are therefore not independent of network
-overheads and we must also deal with the storage overhead to improve the
-performance of replication.
+overheads, and we must deal with both storage and networking overheads
+simultaneously to improve the performance of replication.
 
@@
 logs is essential to exploiting the concurrency available at the level of the
-network interface card (even a single one!) that would otherwise go wasted due
-to the serializing abstraction of a single log. Finally we note that Cyclone
-achieves about 50\% of the linerate across the four ethernet
+NIC (even a single one!) that would otherwise go wasted due
+to the serializing abstraction of a single log. Finally, we note that Cyclone
+achieves about 50\% of the line rate across the four ethernet
 ports. This is essentially the cost of running a consensus protocol in
@@
 flashlog in Figures~\ref{fig:flashlog} and \ref{fig:flashlog_512}. The benefit
-of the second level flashlog is that it lets us limit the amount of needed
-NVDIMM memory. The hypothesis was that batching the movement of data from the
+of the second level flashlog is that it lets us keep the same amount of log
+space as with the flash based implementation, and not limit it to the amount of
+NVM available in the system.
+The hypothesis was that batching the movement of data from the
 NVM log to the flashlog would be sufficient to hide the latency of the block IO
@@
 requests to the drive that we maintain (Section~\ref{sec:storage}).
-A final point about Figure~\ref{fig:flashlog_512} is that once we are past the
-storage bottleneck the latency spike is dramatic and large enough to trigger
-Cyclone's failure detector and repeated retries from the clients. There are -
-therefore - no points on the ``knee'' of the curve as in the pure packet
+A final observation on Figure~\ref{fig:flashlog_512} is that once we are past
+the storage bottleneck the latency spike is dramatic and large enough to trigger
+Cyclone's failure detector and repeated retries from the clients. There are
+therefore no points on the ``knee'' of the curve as in the pure packet
 switched one-level log case.
@@
 effective at hiding the latency of a background block storage device from the
-criticial path of replication. Further, for small updates the throughput
-bottleneck is the capability of the software to run the consensus protocol but
-shifts to the secondary storage device for larger request sizes.
+criticial path of replication. Further, for small updates the
+bottleneck is the capability of the software to run the consensus protocol, but
+the bottleneck shifts to the secondary storage device for larger request sizes.
 
@@
 replication quorums as well executing our rendezvous protocol on multiple
-cores. We therefore setup an experiment where a single client - reflecting the
-unloaded case - made ganged requests to the replicas. We varied the number of
-physical log participating in the request from one to the full complement of 8.
+cores. We therefore setup an experiment where a single client -- reflecting the
+unloaded case -- made ganged requests to the replicas. We varied the number of
+physical logs participating in the request from one to the full complement of 8.
 Figure~\ref{fig:ganged} shows the results. Unloaded
@@
 Next, we evaluate Cyclone integrated with the Rocksdb persistent key value
-store. Rocksdb is a complex industrial strength persistent key value store and
+store. Rocksdb is a complex commercial grade persistent key value store and
 an important target for practical use of Cyclone.  Rocksdb is accompanied by a
@@
 from flash and make best use of DRAM on the machine is an area of active
-research~\cite{flodb, triad}. We wished to make sure that our conclusions would
-apply even as Rocksdb performance improves with the integration of new
-ideas. Keeping this in mind, we configured RocksDB to place all files for the
+research~\cite{flodb, triad}. One of our main goals is to demonstrate that
+our conclusions will apply even as Rocksdb performance continues to improve
+with the integration of new ideas and better SSDs. Keeping this in mind, we
+configured RocksDB to place all files for the
 key value store (SSTables) on a RAMdisk, which presumably represents the limit
 in performance for both software enhancements as well as improvements to
-flash-based SSDs. However - both Rocksdb's own write ahead log and the
-alternative of Cyclone's second level flashlog - are placed on the SSDs. In
-effect we therefore make availability a harder problem in this setting,
-guaranteeing our design can handle the immediate future.
-
-We use 8 byte keys with either 8 byte values or 256 byte values.  Since our
-focus is replication and only update requests are replicated, we use a pure
+flash-based SSDs. However, both Rocksdb's own write ahead log and the
+alternative of Cyclone's second level flashlog are placed on the SSDs. In
+effect we make availability a harder problem in this setting to ensure that our
+design is future-proof.
+
+For the first experiment with Rocksdb we use 8 byte keys with either
+8 byte values or 256 byte values.  Since our
+focus is replication and only update requests are replicated, we use a 100\%
 update workload to test the capability of the system. This involves loading 100
@@
 in Figure~\ref{fig:rocksdb} - essentially presenting the solution to the problem
-description graph in Figure~\ref{fig:problem}.  We consider four different
+demonstrated in Figure~\ref{fig:problem}.  We consider four different
 setups. The line labeled 'rocksdb' is the key value store running with no
@@
 We examine the problem for the unloaded case
-and small updates in Figure~\ref{fig:rocksdb_multi} for a single client with
-increasing number of keys in the batch - till 32 keys.
-The line labeled Rocksdb is with no write ahead logging. We
-note an increasing latency for this baseline indicating Rocksdb itself takes longer
+and small updates in Figure~\ref{fig:rocksdb_multi}, for a single client with
+increasing number of keys in the batch (up to 32 keys).
+The line labeled Rocksdb is with no write ahead logging. We note an increasing
+latency for this baseline indicating Rocksdb itself takes longer
 with larger key batches. The existing option of Rocksdb/wal has considerably
@@
 logs. Cyclone therefore provides effective replicated logging for batched
-operations using its idea of ganged operations, while scaling the performance of
+operations using the idea of ganged operations, while scaling the performance of
 single operations using its multiple physical logs.
@@
 Our evaluation has, thus far, focused on synthetic write heavy workloads. To
-illustrate that Cyclone can also be useful for read-heavy workloads, we
-configured our clients to generate key value requests that mimic the
+illustrate that Cyclone can also be useful for real-world read-heavy workloads, 
+we configured our clients to generate key value requests that mimic the
 distribution of value sizes in Facebook's ETCD
 trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read heavy (95\%
-reads). Figure~\ref{fig:fb} illustrates that even with a read-dominated
+reads). Figure~\ref{fig:fb} illustrates that even with a read dominated
 workload, Cyclone is effective at providing a write ahead log with 3 way
 replication with a lower performance overhead than Rocksdb's own write ahead
-logging. We consider the case of a workload that more write intensive at 80\%
-writes with a smaller range of value sizes, derived from Facebook's VAR
-trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
+logging. We consider the case of another real-world workload that is more write 
+intensive (at 80\%) and with a smaller range of value sizes -- derived from
+Facebook's VAR trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
 Figure~\ref{fig:fb2}. One can similarly conclude that Cyclone provides more
@@
 demonstrate this in Figure ~\ref{fig:timeline} that shows the timeline of a run
-where we kill the server process on the leader replica. Cyclone is configured to
+where we kill the server process on the leader replica. Cyclone is configured 
 with a 30ms failure detection timeout after which the client library tries
-replicas in turn to locate the leader - in this case it fails over in about
+replicas in turn to locate the leader. In this case it fails over in about
 60ms.
@@
 structures. Mojim is a weak replication scheme in that none of the numerous
-\emph{evaluated} modes of Mojim provide any means to keep more than two replicas in
-strong synchronization. Its multi-tiered replication could lose committed
+\emph{evaluated} modes of Mojim provide any means to keep more than two replicas
+in strong synchronization. Mojim's multi-tiered replication could lose committed
 updates and therefore its guarantees are too weak to be used as a replacement
-for RocksDB's write ahead log. Further, Mojim makes the somewhat strange claim
-that ``Paxos-like protocols'' require two networking round trips and therefore
-are somehow inefficient compared to mirroring. This is not true, as we have
-demonstrated with Cyclone where the leader responds as soon as it hears back
-from a majority quorum, the critical path to replication therefore being one
+for RocksDB's write ahead log. Further, Mojim's claim that
+``Paxos-like protocols'' require two networking round trips and are
+therefore somehow inefficient compared to mirroring is not true. For instance,
+we have demonstrated with Cyclone, where the leader responds as soon as it hears
+back from a majority quorum, that the critical path to replication is just one
 network round trip.  A performance comparison with Mojim is hard, since it is
-designed for the hypothetical scenario where there is enough persistent memory
-to hold all persistent data, while Cyclone is designed for the practical case
-where both the write-ahead log and key value store data structures need to be
-maintained largely on flash. Nevertheless we can compare their replication
+designed for the (hypothetical) scenario where there is enough persistent memory
+to hold all persistent data, while Cyclone is designed for the more practical
+case where both the write-ahead log and key value store data structures need to 
+be maintained largely on flash. Nevertheless we can compare their replication
 performance for the case where we use only the NVM log in Cyclone. On 40Gbps
@@
 Cyclone can replicate data three ways using 4*10 Gbps commodity ethernet NICs at
-a peak of roughly 3.7 GB/s. This is with Cyclone is also running the RAFT
-consensus protocol, while Mojim is doing simple mirroring of main memory. One
-can conclude that network offload via RDMA and Infiniband is not an a-priori
+roughly 3.7 GB/s. Note that this result is with Cyclone running the provable
+RAFT consensus protocol, while Mojim is doing simple mirroring of main memory.
+One can conclude that network offload via RDMA and Infiniband is not an a-priori
 requirement for high throughput replication. This is made possible by Cyclone's
@@
 focusing purely on network performance for replication. Latency is excellent due
-to the FPGA that cuts out the path from CPU caches to the NIC and are reported
-to be as low as 3us for a round trip from leader to follower replica compared to
-the 6us we observe with DPDK. Notably however they reported 7us when using
-TCP/IP from the FPGA rather than their own connection oriented network
+to the fact that FPGA cuts out the path from CPU caches to the NIC. Authors
+report latency as low as 3us for a round trip from leader to follower replica,
+compared to the 6us we observe with DPDK. Notably however they reported 7us when
+using TCP/IP from the FPGA rather than their own connection oriented network
 protocol. On the other hand their peak replication throughput is 4M
@@
 40 Gbps of network bandwidth when \emph{also persisting the log on the SSD}. We
-believe the general applicability afforded by our software only solution
+believe the general applicability afforded by our software only approach
 together with the fact that we provide persistence by design makes it a
-compelling alternative to an FPGA based solution even given the larger latency.
+compelling alternative to an FPGA based solution, even given the larger latency.
 
@@
 updates. In contrast, Cyclone replicates upwards of 6 million similarly sized
-requests each second with 4*10 Gbps ethernet (without the second level
+requests per second with 4*10 Gbps ethernet (without the second level
 flashlog). DARE reports an unloaded write latency of 15us, which is lower than
@@
 client space and maintaining information about last seen sequence numbers from
-clients in NVM. Distributed transactions on Cyclone shards however are outside
+clients in NVM. Distributed transactions on Cyclone shards is however outside
 the scope of this paper.
@@
 address two fundamental difficulties in adding high availability to persistent
-key value stores. It leverage a small amount of non-volatile memory to avoid
+key value stores. First, Cyclone avoids
 paying the cost of the block IO interface for every update operation. This
@@
 apply software packet switching techniques and concurrency to improve the
-performance of and make full use of available resources when replicating log
-entries over the network. This enables Cyclone to provide high availability to
-persistent key value stores, without compromising on their performance. 
+performance of replication, and making full use of available resources when
+replicating log entries over the network. By optimizing the storage and network
+components simultaneously, Cyclone enables high availability to persistent key
+value stores without compromising on their performance.
 \newcommand\myurl[2]{\url{#1}}

### 2017-08-09 paper/paper.tex
Updates.

@@
 Cyclone is therefore to approximate a software packet switch for the networking
-component of Cyclone.
-
-This is however not a straightforward mapping. Software packet switches make a
-simple (usually stateless or sometimes flow associated) decision for each
-packet. In contrast, the role of replicas in Cyclone is asymmetric. Log entries
-flow from the leader replica to followers and acknowledgments flow
-back. Further, the assigned role to a replica can change on reconfiguration
-after a failure. A replica must also reject packets from old configurations. The
-action associated with a received packet therefore is far more complex than
-simply forwarding it to a log.  The \emph{challenge} is to do this efficiently
-in software.
+component of Cyclone. We do this by implementing Cyclone's network stack on top
+of the Data Plane Development Kit (DPDK)~\cite{dpdk}, that provides low latency
+userspace access to an ethernet NIC. We discuss Cyclone's network protocol in
+Section~\ref{sec:netprot}. There are two key challenges in implementing a
+consensus protocol as a software packet switch.
+
+Software packet switches make a simple (usually stateless or
+sometimes flow associated) decision for each packet. In contrast, running a
+consensus protocol requires a relatively complex decision process for each
+packet. The direction of flow of log entries depends on the role of the
+replicas. Further, the assigned role to a replica can change on reconfiguration
+after a failure. A replica must also reject packets from old configurations. In
+Section~\ref{sec:dm} we cover the techniques we use in Cyclone to ameliorate
+these overheads.
 
 \subsection{DPDK}
+\label{sec:netprot}
 The Data Plane Development Kit~\cite{dpdk} provides low latency userspace access
@@
 
-\subsection{Memory Management}
-\label{sec:mm}
-We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
-concurrent memory allocator based on reference counting - used by both the NIC
-driver and CPU cores. This means that packet allocation and deallocation does
-not happen in the same order as their corresponding position in the replicated
-RAFT log. To deal with this, we use a level of indirection as shown in
-Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
-sized pointers to the actual packets. Adding a level of indirection in the NVM
-log allows us to separate the FIFO ordered circular log being manipulated by
-RAFT from packet data being managed by the memory allocator of DPDK. Both the
-circular log and packet data are in NVM. An advantage of this scheme is that it
-makes recovery from NVM easy -- appends to the circular pointer log are atomic
-and we can use the pointer log to recover allocator state i.e. what pieces of
-NVM are currently in use by the log.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
-\caption{NVM log structure}
-\label{fig:nvm_log}
-\end{figure}
-
-A key advantage of RAFT is that unlike alternatives such as
-MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
-leader is guaranteed to have the most up to date logs. There is therefore no
-case where a leader needs to receive log entries to ``fill'' holes in its logs
-from follower replicas, simplifying the protocol state machine we need to
-implement. A leader can immediately reject any responses that are not more
-current than its view (term). In turn, the persistent log operates as a double
-ended queue, with entries either being appended (or possibly deleted at
-followers) at the end and being deleted from the front (after commit). This is
-critical to efficient operation of Cyclone, as it allows a simple top level
-structure (a circular buffer of pointers) to represent the log.
-
-\subsection{Data Movement}
+\subsection{Addressing Overheads}
 \label{sec:dm}
@@
 
-We now turn our attention to the persistence step in
-Figure~\ref{fig:control_plane}. RAFT requires that the log entry be persisted
-\emph{before} it is multicast out to follower replicas. DPDK userspace NIC
-drivers operate in DDIO mode~\cite{ddio} where the packet is directly written
-into the CPU cache rather than first being DMAed into DRAM and then fetched by
-the CPU on demand. Since the NVM is directly attached we need to persist the
-packet to it by executing a cacheline flush ({\tt clflush}) instruction for
-every cacheline in the packet and the pointer in the pointer buffer to persist
-these via the memory bus. This is not too onerous a burden because we can use
-the newly introduced {\tt clflush-opt}~\cite{clflush_opt} instruction
-specifically intended to efficiently flush to persistent memory without the
-overhead of the serialization normally introduced by {\tt clflush}. This allows
-us to hit full memory bandwidth on current generation platforms, a quantity in
-excess of 200 Gb/s per core which is well above the near term speeds of 
-commodity NICs. We execute a single serializing {\tt sfence} 
-before the sequence of {\tt clflushes} to make sure any dirty cachelines due to
-header manipulation related writes from the CPU are sent to cache.
-
-\subsection{Batching}
+RAFT requires that the log entry be persisted \emph{before} it is multicast out
+to follower replicas. DPDK userspace NIC drivers operate in DDIO
+mode~\cite{ddio} where the packet is directly written into the CPU cache rather
+than first being DMAed into DRAM and then fetched by the CPU on demand. Since
+the NVM is directly attached we need to persist the packet to it by executing a
+cacheline flush ({\tt clflush}) instruction for every cacheline in the packet
+and the pointer in the pointer buffer to persist these via the memory bus. This
+is not too onerous a burden because we can use the newly introduced {\tt
+  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
+efficiently flush to persistent memory without the overhead of the serialization
+normally introduced by {\tt clflush}. This allows us to hit full memory
+bandwidth on current generation platforms, a quantity in excess of 200 Gb/s per
+core which is well above the near term speeds of commodity NICs. We execute a
+single serializing {\tt sfence} before the sequence of {\tt clflushes} to make
+sure any dirty cachelines due to header manipulation related writes from the CPU
+are sent to cache.
 
@@
 replicas. Therefore, we never tradeoff latency for throughput when batching.
+
+We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
+concurrent memory allocator based on reference counting - used by both the NIC
+driver and CPU cores. This means that packet allocation and deallocation does
+not happen in the same order as their corresponding position in the replicated
+RAFT log. To deal with this, we use a level of indirection as shown in
+Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
+sized pointers to the actual packets. Adding a level of indirection in the NVM
+log allows us to separate the FIFO ordered circular log being manipulated by
+RAFT from packet data being managed by the memory allocator of DPDK. Both the
+circular log and packet data are in NVM. An advantage of this scheme is that it
+makes recovery from NVM easy -- appends to the circular pointer log are atomic
+and we can use the pointer log to recover allocator state i.e. what pieces of
+NVM are currently in use by the log.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+\caption{NVM log structure}
+\label{fig:nvm_log}
+\end{figure}
+
+Our choice of RAFT as a consensus protocol is driven by the need to efficiently
+maintain this log datastructure. Unlike alternatives such as
+MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
+leader is guaranteed to have the most up to date logs. There is therefore no
+case where a leader needs to receive log entries to ``fill'' holes in its logs
+from follower replicas, simplifying the protocol state machine we need to
+implement. A leader can immediately reject any responses that are not at least as
+current as its view (term). In turn, the persistent log operates as a double
+ended queue, with entries either being appended (or possibly deleted at
+followers) at the end and being deleted from the front (after commit). This is
+critical to efficient operation of Cyclone, as it allows a simple top level
+structure (a circular buffer of pointers) to represent the log.
 
@@
 together with the fact that we provide persistence by design makes it a
-compelling alternative to an FPGA based solution, even given the larger latency.
+compelling alternative to an FPGA based solution, even given the larger
+latency. Also, these results demonstrate that a general purpose CPU core can be
+as efficient as an FPGA in running a consensus protocol using the techniques
+described in this paper.
 
@@
 date, a situation we avoid by using RAFT as described in
-Section~\ref{sec:mm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
+Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
 being able to replicate 500K requests/sec. This is without persisting the log of
@@
 Cyclone provides a client library that automatically times out and retries
-requests to deal with failure. Protection against repeated execution of the same
-request is currently left to the key value store and client, since we felt that
-exactly once semantics is optional when updates such as put operations are
-idempotent. We have experimented with exactly once semantics as a basis for
-distributed transactions across shards running Cyclone integrated key value
-stores. This is along the lines of similar work~\cite{raft_lin}, using a fixed
-client space and maintaining information about last seen sequence numbers from
-clients in NVM. Distributed transactions on Cyclone shards is however outside
-the scope of this paper.
+requests to deal with failure. We do not provide exactly once semantics as we
+believe that exactly once semantics are better provided by the key value store
+itself by persisting necessary request markers - as done in systems such as
+Kafka~\cite{kafka}. A possible direction of future work however is to add
+exactly once semantics to Cyclone's client-server protocol itself to serve as a
+building block for distributed transactions across key-value shards, each
+running Cyclone, in a manner similar to other systems~\cite{raft_lin}.
 

### 2017-08-09 paper/paper.tex
Updates.

@@
 
-\thispagestyle{empty}
+%\thispagestyle{empty}
 
 \begin{abstract}
-\vspace{0.1in}
+\vspace{0.05in}
 Persistent key value stores are an important component of many distributed data
@@
 data copies. Finally, it provides a way to efficiently map the commutativity in
-key-value store APIs to the parallelism available in commodity NICs via multiple
-transmit receive queues to scale replication performance on multicore
-machines. Cyclone is able to replicate millions of small updates per second
+key-value store APIs to the parallelism available in commodity NICs.
+Cyclone is able to replicate millions of small updates per second
 using only commodity 10 gigabit ethernet adapters. As a practical application,
@@
 impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, 
-a number far in excess of the few hundreds of packets we manage in
+a number far in excess of the few thousands of packets we manage in
 Figure~\ref{fig:problem}. The key insight in removing network overheads in

### 2017-08-09 paper/paper.tex
updates.

@@
 form of NVDIMMs or Intel 3DXPoint - to remove block oriented IO devices such as
-SSDs from the critical path for appending to the log. It then addresses network
+SSDs from the critical path for appending to the log. This enables it to address network
 overheads using an implementation of the RAFT consensus protocol that is
@@
 The second key contribution of this work is in showing that the throughput
-bottlenecks in the network stack can be mitigated entirely in software. Existing
-work on log replication using state machine replication protocols such as
-Paxos~\cite{paxos} and RAFT~\cite{raft} has not - thus far - addressed high
-performance replication in the local area network of a datacenter using
-commodity networks.  Research focusing on addressing wide area network latencies
-by reducing the number of network hops with new protocols such as Fast
-Paxos~\cite{fast-paxos} or network switch modifications such as No
-Paxos~\cite{nopaxos} do not address throughput bottlenecks due to network stack
-related inefficiencies in the participating nodes of the quorum -- a more urgent
-problem with local area replication. On the other hand, work done to address
-network related inefficiencies start with the assumption that the bottleneck to
-good throughput on the network is the CPU and therefore one must either resort
-to offload mechanisms such as RDMA~\cite{dare, farm, faast} or offload the
-protocol entirely to an FPGA~\cite{consensus_box}. We demonstrate in this paper
-that this is not the case. Rather, once persistence is provided by directly
-attached non-volatile memory, thereby eliminating block IO from the persistence 
-step in
-replication protocols, the problem becomes equivalent to multicasting the same
-log entry from the leader to all followers. We design Cyclone's network stack
-around well known principles in the software packet switching community. In
-particular, a careful implementation of a consensus protocol that relieves the
-CPU of the responsibility of data movement permits a high performance
-implementation entirely in software that addresses the order of magnitude
-network related performance gap in Figure~\ref{fig:problem} using only commodity
-10 Gigabit ethernet.
+bottlenecks in the network stack can be mitigated entirely in software, once
+block IO is removed from the critical path. Existing work on log replication
+using state machine replication protocols such as Paxos~\cite{paxos} and
+RAFT~\cite{raft} has not - thus far - addressed high performance replication in
+the local area network of a datacenter using commodity networks.  Research
+focusing on addressing wide area network latencies by reducing the number of
+network hops with new protocols such as Fast Paxos~\cite{fast-paxos} or network
+switch modifications such as No Paxos~\cite{nopaxos} do not address throughput
+bottlenecks due to network stack related inefficiencies in the participating
+nodes of the quorum -- a more urgent problem with local area replication. On the
+other hand, work done to address network related inefficiencies start with the
+assumption that the bottleneck to good throughput on the network is the CPU and
+therefore one must either resort to network offload mechanisms such as RDMA~\cite{dare,
+  farm, faast} or offload the protocol entirely to an
+FPGA~\cite{consensus_box}. We demonstrate in this paper that this is not the
+case. Rather, once persistence is provided by directly attached non-volatile
+memory, thereby eliminating block IO from the persistence step in replication
+protocols, the problem becomes akin to multicasting the same log entry
+from the leader to all followers. We design Cyclone's network stack around well
+known principles in the software packet switching community. In particular, a
+careful implementation of a consensus protocol that relieves the CPU of the
+responsibility of data movement permits a high performance implementation
+entirely in software that addresses the order of magnitude network related
+performance gap in Figure~\ref{fig:problem} using only commodity 10 Gigabit
+ethernet.
 
@@
 into the key value store, and returns a response to the client. Before
-executing any PUT request, Cyclone appends it to a durable log and replicates
+executing any PUT or DELETE request, Cyclone appends it to a durable log and replicates
 the request to the logs of follower replicas. The request is considered
@@
 to is durable and its results \emph{available} as long as a majority of replicas
-continue to function. Cyclone's failure model therefore is \emph{fail-recovery},
+continue to function. Cyclone's failure model is \emph{fail-recovery},
 where nodes that can recover from failure simply continue to participate in

### 2017-08-10 paper/paper.tex
Updates.

@@
 library. Cyclone replicates the key value store across a set of replica
-servers. The client side library sends \emph{all} requests to a distinguished
-leader replica.  On the server side Cyclone accepts requests from clients, calls
-into the key value store, and returns a response to the client. Before
-executing any PUT or DELETE request, Cyclone appends it to a durable log and replicates
-the request to the logs of follower replicas. The request is considered
-replicated once acknowledged by a majority of replicas. Follower replicas apply
-updates from the log in order.
+servers. The client side library sends \emph{all} requests - reads and updates -
+to a distinguished leader replica.  On the server side Cyclone accepts requests
+from clients, calls into the key value store, and returns a response to the
+client. Before executing any PUT or DELETE request, Cyclone appends it to a
+durable log and replicates the request to the logs of follower replicas. The
+request is considered replicated once acknowledged by a majority of
+replicas. Follower replicas apply updates from the log in order.
 
@@
 manipulate multiple keys in the next section.
+
+We note that spreading keys across the physical logs \emph{does not} break
+referential integrity. If key1 is updated followed by key2 and key2 refers to
+key1, even in the face of failure linearizability in Cyclone ensures that key1
+is available once key2 is seen by a client. This is true even if key1 is not
+committed yet at the leader replica, since we execute reads in FIFO order with
+pending entries in the log (Section~\ref{sec:sysarch}).
 
@@
 attached persistent memory with RDMA to replicate persistent data
-structures. Mojim is a weak replication scheme in that none of the numerous
-\emph{evaluated} modes of Mojim provide any means to keep more than two replicas
-in strong synchronization. Mojim's multi-tiered replication could lose committed
-updates and therefore its guarantees are too weak to be used as a replacement
-for RocksDB's write ahead log. Further, Mojim's claim that
-``Paxos-like protocols'' require two networking round trips and are
-therefore somehow inefficient compared to mirroring is not true. For instance,
-we have demonstrated with Cyclone, where the leader responds as soon as it hears
-back from a majority quorum, that the critical path to replication is just one
-network round trip.  A performance comparison with Mojim is hard, since it is
-designed for the (hypothetical) scenario where there is enough persistent memory
-to hold all persistent data, while Cyclone is designed for the more practical
-case where both the write-ahead log and key value store data structures need to 
-be maintained largely on flash. Nevertheless we can compare their replication
-performance for the case where we use only the NVM log in Cyclone. On 40Gbps
-infiniband, Mojim reported being able to mirror data at 4GB/s. In contrast,
-Cyclone can replicate data three ways using 4*10 Gbps commodity ethernet NICs at
-roughly 3.7 GB/s. Note that this result is with Cyclone running the provable
-RAFT consensus protocol, while Mojim is doing simple mirroring of main memory.
-One can conclude that network offload via RDMA and Infiniband is not an a-priori
+structures. None of the \emph{evaluated} modes of Mojim provide any means to
+keep more than two replicas in strong synchronization and therefore its
+guarantees are weak compared to Cyclone.  Further, Mojim's claim that
+``Paxos-like protocols'' require two networking round trips and are therefore
+somehow inefficient compared to mirroring is not true. For instance, we have
+demonstrated with Cyclone, where the leader responds as soon as it hears back
+from a majority quorum, that the critical path to replication is just one
+network round trip. We can compare Mojim's replication performance to Cyclone
+for the case where we use only the NVM log in Cyclone. On 40Gbps Infiniband,
+Mojim reported being able to mirror data at 4GB/s. In contrast, Cyclone can
+replicate data three ways using 4*10 Gbps commodity ethernet NICs at roughly 3.7
+GB/s. Note that this result is with Cyclone running the provable RAFT consensus
+protocol, while Mojim is doing simple mirroring of main memory.  One can
+conclude that network offload via RDMA and Infiniband is not an a-priori
 requirement for high throughput replication. This is made possible by Cyclone's
@@
 focusing purely on network performance for replication. Latency is excellent due
-to the fact that FPGA cuts out the path from CPU caches to the NIC. Authors
+to the fact that FPGA cuts out the path from CPU caches to the NIC. The authors
 report latency as low as 3us for a round trip from leader to follower replica,
@@
 flashlog). DARE reports an unloaded write latency of 15us, which is lower than
-the unloaded write latency of Cyclone at 20us, thanks to infiniband. The
+the unloaded write latency of Cyclone at 20us, thanks to Infiniband. The
 conclusions we can draw are similar to the comparison with Mojim: a carefully
@@
 
-Cyclone provides a client library that automatically times out and retries
-requests to deal with failure. We do not provide exactly once semantics as we
+Cyclone does not provide exactly once semantics as we
 believe that exactly once semantics are better provided by the key value store
@@
 Kafka~\cite{kafka}. A possible direction of future work however is to add
-exactly once semantics to Cyclone's client-server protocol itself to serve as a
+exactly once semantics to Cyclone's client-server protocol to serve as a
 building block for distributed transactions across key-value shards, each

### 2017-08-10 paper/paper.tex
Spellcheck.

@@
 ahead log. A machine or storage failure leading to total loss or temporary
-unavailability of data is unacceptable in services where high availabilty and
+unavailability of data is unacceptable in services where high availability and
 revenue are interconnected. Key value stores therefore usually incorporate
@@
 structure in Section~\ref{sec:storage}. The set of optimizations to the network
-stack is decribed in Section~\ref{sec:network}. We show how Cyclone can be
+stack is described in Section~\ref{sec:network}. We show how Cyclone can be
 scaled to make use of available parallelism on the system in
@@
 shown in Figure~\ref{fig:problem}). This problem leads many system designers to
-underprovision key-value store shards ensuring that it operates at moderate to
+under-provision key-value store shards ensuring that it operates at moderate to
 high load, using group-commit to batch additions to the log. They therefore pay
@@
 addition, most datacenter networks rarely drop or reorder packets and provide
-full bisection bandwith between servers that might serve as Cyclone replicas. We
+full bisection bandwidth between servers that might serve as Cyclone replicas. We
 therefore jettisoned TCP and chose to send raw IP packets encapsulated in
@@
 physical log to send a request to? A key constraint here is that all replicas
-should agree on the seqeunce of updates to a key, as otherwise they would
+should agree on the sequence of updates to a key, as otherwise they would
 diverge in terms of state. Allowing updates to the same key to go to different
@@
 order in which the independent RAFT instances would apply the update. Therefore
-all updates to a key must go to the same physcial log. A similar constraint
+all updates to a key must go to the same physical log. A similar constraint
 applies to reads in order to ensure that all committed writes are seen by
@@
 ensure that the process converges we therefore ensure that the currently elected
-leader brings a majority quroum up to date before triggering a relection. We
+leader brings a majority quorum up to date before triggering a reelection. We
 also do not accept any requests from clients until all leaders are
@@
 i.e. some RAFT instances are able to send and receive packets, while others are
-not, a possiblity we discount due to our single process design.
+not, a possibility we discount due to our single process design.
 
@@
 A failure causes a new leader to be elected for the affected physical log,
-moving forward the term. If the participants of a ganged operatation do not
+moving forward the term. If the participants of a ganged operation do not
 detect this case, they could be left waiting forever. We detect failure on any
@@
 owned by the co-ordinator to hold the barrier and write the nonce to it in order
-to indicate that the barrer is active for the corresponding ganged
+to indicate that the barrier is active for the corresponding ganged
 operation. Participants watch for the nonce to know when to execute the
@@
 
-Ganged operations possibly consititute the most complex part of Cyclone but the
+Ganged operations possibly constitute the most complex part of Cyclone but the
 code weighs in at well under a couple of hundred lines. We believe that
@@
 effective at hiding the latency of a background block storage device from the
-criticial path of replication. Further, for small updates the
+critical path of replication. Further, for small updates the
 bottleneck is the capability of the software to run the consensus protocol, but
@@
 the DPDK userspace NIC driver pipelines request processing for communication
-between the CPU core and the NIC. Simultaneous replicaton on multiple physical
+between the CPU core and the NIC. Simultaneous replication on multiple physical
 logs represents a worst case for this arrangement as all eight cores try to
@@
 setups. The line labeled 'rocksdb' is the key value store running with no
-logging whatsover - a system crash would lead to data loss. The line labeled
+logging whatsoever - a system crash would lead to data loss. The line labeled
 'rocksdb/wal' is for Rocksdb running with its write ahead logging turned on. The
@@
 but without any replication. The line almost exactly tracks the performance of
-Rockdb. Cyclone is able to provide a write ahead log with no overhead to
+Rocksdb. Cyclone is able to provide a write ahead log with no overhead to
 Rocksdb. The line labeled 'rocksdb/Cyclone 3 way' is with 3-way replication
@@
 protocol. On the other hand their peak replication throughput is 4M
-replications/sec on 66 Gbps of aggregate network bandwith for small messages,
+replications/sec on 66 Gbps of aggregate network bandwidth for small messages,
 comparable to the 6M replications/sec mark obtained by Cyclone on an aggregate
@@
 
-{CRANE}~\cite{crane} is a system for replicating mutithreaded programs using
+{CRANE}~\cite{crane} is a system for replicating multithreaded programs using
 Paxos. It uses deterministic multi-threading to ensure replicas converge to the
@@
 
-Database practitioners have considered the utility of byte adressable
+Database practitioners have considered the utility of byte addressable
 non-volatile memory in improving logging for

### 2017-08-10 paper/paper.tex
Updates.

@@
 userspace access to an ethernet NIC. We discuss Cyclone's network protocol in
-Section~\ref{sec:netprot}. There are two key challenges in implementing a
-consensus protocol as a software packet switch.
+Section~\ref{sec:netprot}.
 
@@
 \centering
-  \includegraphics[scale=0.3]{figures2/batching.pdf}
-  \caption{Batching}
-  \label{fig:batching}
+\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+\caption{NVM log structure}
+\label{fig:nvm_log}
 \end{minipage}
@@
 
+We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
+concurrent memory allocator based on reference counting - used by both the NIC
+driver and CPU cores. This means that packet allocation and deallocation does
+not happen in the same order as their corresponding position in the replicated
+RAFT log. To deal with this, we use a level of indirection as shown in
+Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
+sized pointers to the actual packets. Adding a level of indirection in the NVM
+log allows us to separate the FIFO ordered circular log being manipulated by
+RAFT from packet data being managed by the memory allocator of DPDK. Both the
+circular log and packet data are in NVM. An advantage of this scheme is that it
+makes recovery from NVM easy -- appends to the circular pointer log are atomic
+and we can use the pointer log to recover allocator state i.e. what pieces of
+NVM are currently in use by the log.
+
 RAFT requires that the log entry be persisted \emph{before} it is multicast out
@@
 
-We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
-concurrent memory allocator based on reference counting - used by both the NIC
-driver and CPU cores. This means that packet allocation and deallocation does
-not happen in the same order as their corresponding position in the replicated
-RAFT log. To deal with this, we use a level of indirection as shown in
-Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
-sized pointers to the actual packets. Adding a level of indirection in the NVM
-log allows us to separate the FIFO ordered circular log being manipulated by
-RAFT from packet data being managed by the memory allocator of DPDK. Both the
-circular log and packet data are in NVM. An advantage of this scheme is that it
-makes recovery from NVM easy -- appends to the circular pointer log are atomic
-and we can use the pointer log to recover allocator state i.e. what pieces of
-NVM are currently in use by the log.
-
 \begin{figure}
   \centering
-\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
-\caption{NVM log structure}
-\label{fig:nvm_log}
+  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \caption{Batching}
+  \label{fig:batching}
 \end{figure}
@@
 Our choice of RAFT as a consensus protocol is driven by the need to efficiently
-maintain this log datastructure. Unlike alternatives such as
+maintain the log datastructure. Unlike alternatives such as
 MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
@@
 from follower replicas, simplifying the protocol state machine we need to
-implement. A leader can immediately reject any responses that are not at least as
-current as its view (term). In turn, the persistent log operates as a double
+implement. A leader can immediately reject any responses that are not at least
+as current as its view (term). In turn, the persistent log operates as a double
 ended queue, with entries either being appended (or possibly deleted at
@@
 \includegraphics[scale=0.45]{results2/fb.pdf}
-\caption{Rocksdb - read heavy workload}
+\caption{Rocksdb - read heavy}
 \label{fig:fb}
@@
   \includegraphics[scale=0.45]{results2/fb2.pdf}
-  \caption{Rocksdb - write heavy workload}
+  \caption{Rocksdb - write heavy}
   \label{fig:fb2}

### 2017-08-10 paper/paper.tex
Tweak.
Updates.

@@
 form of NVDIMMs or Intel 3DXPoint - to remove block oriented IO devices such as
-SSDs from the critical path for appending to the log. This enables it to address network
-overheads using an implementation of the RAFT consensus protocol that is
+SSDs from the critical path for appending to the log. This enables it to address
+network overheads using an implementation of the RAFT consensus protocol that is
 designed around a userspace network stack to relieve the CPU of the burden of
 data copies. Finally, it provides a way to efficiently map the commutativity in
-key-value store APIs to the parallelism available in commodity NICs.
-Cyclone is able to replicate millions of small updates per second
-using only commodity 10 gigabit ethernet adapters. As a practical application,
-we use it to improve the performance (and availability) of RocksDB, a popular
-persistent key value store by an order of magnitude when compared to its own
-write ahead log without replication.
+key-value store APIs to the parallelism available in commodity NICs.  Cyclone is
+able to replicate millions of small updates per second using only commodity 10
+gigabit ethernet adapters. As a practical application, we use it to improve the
+performance (and availability) of RocksDB, a popular persistent key value store
+by an order of magnitude when compared to its own write ahead log without
+replication.
 \end{abstract}
@@
 durability semantics mean that the key-value store cannot respond to the client
-until the operation has been persisted to the log, since it must be completed
-even on a crash and subsequent recovery. Therefore, addressing the network
-bottleneck is pointless without also addressing the storage bottleneck. Cyclone
-shows how a small amount of non-volatile memory can be used to address both
-bottlenecks.
+until the operation has been persisted to the log at a majority of replicas,
+since it must be completed even on a crash and subsequent recovery. Therefore,
+addressing the network bottleneck is pointless without also addressing the
+storage bottleneck. Cyclone shows how a small amount of non-volatile memory can
+be used to address both bottlenecks.
 
@@
 operations across machines. Cyclone provides guarantee to the client that any
-operation responded
-to is durable and its results \emph{available} as long as a majority of replicas
-continue to function. Cyclone's failure model is \emph{fail-recovery},
-where nodes that can recover from failure simply continue to participate in
-the protocol. This
-requires RAFT to persist every log entry \emph{before} it sends it out for
-replication. Storage overheads are therefore not independent of network
-overheads, and we must deal with both storage and networking overheads
+operation responded to is durable and its results \emph{available} as long as a
+majority of replicas continue to function. Cyclone's failure model is
+\emph{fail-recovery}, where nodes that can recover from failure simply continue
+to participate in the protocol. This requires RAFT to persist every log entry
+\emph{before} it sends it out for replication and before a follower replica
+responds to the leader. Storage overheads are therefore not independent of
+network overheads, and we must deal with both storage and networking overheads
 simultaneously to improve the performance of replication.
@@
 
-Software packet switches make a simple (usually stateless or
-sometimes flow associated) decision for each packet. In contrast, running a
-consensus protocol requires a relatively complex decision process for each
-packet. The direction of flow of log entries depends on the role of the
-replicas. Further, the assigned role to a replica can change on reconfiguration
-after a failure. A replica must also reject packets from old configurations. In
-Section~\ref{sec:dm} we cover the techniques we use in Cyclone to ameliorate
-these overheads.
+However, the flow of packets through a software packet switch is usually very
+simple: the packet enters through an ethernet port and after a simple (usually
+stateless) decision is sent out through a set of chosen ports.  In contrast, log
+replication requires a complex protocol state machine to decide what to do with
+the packet. In addition, unlike software switches we cannot simply forget a
+packet after transmitting it. We must append it to the log and act on it after
+consensus is reached by a majority quorum. In Section~\ref{sec:dm} we cover the
+techniques we use in Cyclone to ameliorate these overheads.
 
@@
 mode~\cite{ddio} where the packet is directly written into the CPU cache rather
-than first being DMAed into DRAM and then fetched by the CPU on demand. Since
-the NVM is directly attached we need to persist the packet to it by executing a
-cacheline flush ({\tt clflush}) instruction for every cacheline in the packet
-and the pointer in the pointer buffer to persist these via the memory bus. This
-is not too onerous a burden because we can use the newly introduced {\tt
+than first being DMAed into DRAM and then fetched by the CPU on demand. We need
+to persist the packet to directly attached NVM by executing a cacheline flush
+({\tt clflush}) instruction for every cacheline in the packet and the pointer in
+the pointer buffer to persist these via the memory bus. This is not too onerous
+a burden because we can use the newly introduced {\tt
   clflush-opt}~\cite{clflush_opt} instruction specifically intended to
@@
 always flush the transmit buffer after every call to DPDK to transmit packets to
-replicas. Therefore, we never tradeoff latency for throughput when batching.
+replicas. Therefore, we never tradeoff latency for throughput when batching. We
+also receive log entries in batches at follower replicas and return a single
+acknowledgment for the entire batch, speeding up the progress of the protocol.
 
@@
 
-The scalable commutativity rule~\cite{scalable_commutativity} states that for a
-shared memory program to scale in performance with concurrency it must also have
-an interface that is commutative for common case invocations. We have also used
-interface level commutativity to extract parallelism for the purpose of scaling
-replication performance. Others have used the same commutativity to create
-per-cpu data structures to scale KV store performance~\cite{flodb}.
+%The scalable commutativity rule~\cite{scalable_commutativity} states that for a
+%shared memory program to scale in performance with concurrency it must also have
+%an interface that is commutative for common case invocations. We have also used
+%interface level commutativity to extract parallelism for the purpose of scaling
+%replication performance. Others have used the same commutativity to create
+%per-cpu data structures to scale KV store performance~\cite{flodb}.
 

### 2017-08-29 paper/paper.tex
Clean up semantics in the paper.

@@
 
-A key value store typically serves multiple client requests concurrently and is
-linearizable~\cite{linearizability}. This means any history of requests and
-responses can be rearranged such that responses immediately follow corresponding
-requests and if a response precedes a request in the original history it also
-does so in the rearranged history. Key value stores such as RocksDB also
-support atomic writes ({\tt PUT} operations) to multiple keys, called batched
-writes, and the ability to take snapshots.
+Key value stores such as RocksDB also support atomic writes ({\tt PUT}
+operations) to multiple keys, called batched writes, and the ability to take
+snapshots.
 
@@
 serviced in FIFO order with the write requests (PUT/DELETE operations) in
-Cyclone's log at the moment the read request is received. This ensures
-linearizability in the face of failovers.
+Cyclone's log at the moment the read request is received.
+
+Cyclone provides the following guarantees to clients:
+
+\begin{itemize}
+\item Every acknowledged update is applied
+\item Every key sees the same sequence of updates across all replicas
+\item A client sees a monotonically extending prefix of the sequence of updates
+  to any key at that always includes its own updates
+\end{itemize}
+
+RAFT ensures that the set of committed updates to any key is in the same order
+as in the log, satisfying the first two conditions. For the last condition,
+clients maintain the term of the RAFT leader locally and ensure they always talk
+to a (possibley different) RAFT leader that has at least the same term. This
+helps us avoid split brain problems where a network partitions allows clients to
+talk to old RAFT leaders after having committed changes to a new RAFT leader.
+
+We do not provide linearizable reads, although that can be easily supported by
+logging the read and running conensus on it like a regular mutation operation.
+Finally, Cyclone allows operations to multiple keys in an atomic batch, where
+either all updates in the batch are applied or no operation in the batch is
+applied, regardless of failures. We also use a lightweight batched operation to
+support snapshots.
 
@@
 The first question we deal with is -- how do Cyclone clients decide which
-physical log to send a request to? A key constraint here is that all replicas
-should agree on the sequence of updates to a key, as otherwise they would
-diverge in terms of state. Allowing updates to the same key to go to different
-physical logs would lead to a divergence as there is no guarantee about the
-order in which the independent RAFT instances would apply the update. Therefore
-all updates to a key must go to the same physical log. A similar constraint
-applies to reads in order to ensure that all committed writes are seen by
-readers. Allowing reads to a key to go to a different RAFT instance than the
-write means that it could be served on a machine where the write is yet to be
-applied. Therefore, all reads and writes to a key must go to the same physical
-log. We achieve this by hashing the key to select the physical log.
-
-It turns out that this simple \emph{necessary} scheme is also \emph{sufficient}
-to ensure linearizability in the face of failure. Clearly the system is
-linearizable for a single key since it involves only a single physical log an d
-we always service existing requests in the physical log before responding to
-read requests. In addition, one can switch operations to different keys in a
-history without affecting its correctness because operations to different keys
-are commutative. This allows us to arrive at a history that explains the key
-value store responses, is correct and makes sure that a request responded to
-before subsequent requests in the original history follows that same order in
-the linearized history. We discuss how to handle requests that atomically
-manipulate multiple keys in the next section.
-
-We note that spreading keys across the physical logs \emph{does not} break
-referential integrity. If key1 is updated followed by key2 and key2 refers to
-key1, even in the face of failure linearizability in Cyclone ensures that key1
-is available once key2 is seen by a client. This is true even if key1 is not
-committed yet at the leader replica, since we execute reads in FIFO order with
-pending entries in the log (Section~\ref{sec:sysarch}).
+physical log to send a request to? The guarantees from Cyclone cover ordering of
+updates to a single key (Section~\ref{sec:sysarch}). They can be satisfied by
+ensuring that all reads and writes to a key go to the same physical log
+(i.e. RAFT instance). We achieve this by hashing the key to select the physical
+log.
 
@@
 
-In our discussion thus far, we have focused on operations to a single
-key. Supporting atomic writes or reads to multiple keys while maintaining
-linearizability requires special handling with multiple physical logs so as not
-to break linearizability - that we discuss next.
-
 \subsection{Ganged Operations}
-Supporting reads or writes to multiple keys that map to different physical logs
+Supporting batched updates to multiple keys that map to different physical logs
 requires a ganged operation across those logs such that all the operations

### 2017-11-17 paper/paper.tex
Updates.

@@
-\documentclass[pageno]{jpaper}
-
-%replace XXX with the submission number you are given from the ASPLOS submission site.
-\newcommand{\asplossubmissionnumber}{387}
+\documentclass[twocolumn]{article}
 
@@
 \usepackage{parskip}
+\usepackage{pdfpages}
+\usepackage{url}
 
@@
 
-\title{\Large \bf Cyclone: High Availability for Persistent Key Value Stores}
+\title{Cyclone: High Availability for Persistent Key Value Stores}
+\author{Amitabha Roy \and Subramanya R. Dulloor}
 
@@
 failover, resulting in only a brief interruption in service for
-clients. Although read requests (GET operations) are not logged, they are
-serviced in FIFO order with the write requests (PUT/DELETE operations) in
-Cyclone's log at the moment the read request is received.
-
-Cyclone provides the following guarantees to clients:
-
-\begin{itemize}
-\item Every acknowledged update is applied
-\item Every key sees the same sequence of updates across all replicas
-\item A client sees a monotonically extending prefix of the sequence of updates
-  to any key at that always includes its own updates
-\end{itemize}
-
-RAFT ensures that the set of committed updates to any key is in the same order
-as in the log, satisfying the first two conditions. For the last condition,
-clients maintain the term of the RAFT leader locally and ensure they always talk
-to a (possibley different) RAFT leader that has at least the same term. This
-helps us avoid split brain problems where a network partitions allows clients to
+clients.  Cyclone guarantees that the order of operations to any key are linearizable. 
+A key sees the exact same sequence of update operations (including deletes) 
+on all replicas. These per-key guarantees follow immediately from the 
+properties of the RAFT log and by requiring clients to perform a quorum read (logged
+as an entry in the RAFT log).
+
+In addition, Cyclone also provides clients with a weak read that does not require a 
+quorum read thereby requiring only one round trip to a server.  A weak read of a key sees a prefix of the linearizable history of a key that 
+includes all successful update operations made by the client before it issues the 
+weak read. To do a weak read, 
+clients maintain the last known term of the RAFT leader locally and ensure they always talk
+to a RAFT leader that has at least the same term. This ensures that clients see their own writes
+avoiding split brain problems where a network partitions allows clients to
 talk to old RAFT leaders after having committed changes to a new RAFT leader.
-
-We do not provide linearizable reads, although that can be easily supported by
-logging the read and running conensus on it like a regular mutation operation.
+In addition, a weak read waits until all pending RAFT log entries have been committed and applied before executing.
+This ensures that in the event of a failover, any previous committed updates from the same client have been applied and are visible to the weak read.
+A weak read is not linearizable as the RAFT leader responding to the weak read might actually no longer be the leader - a new one having been elected
+without it realizing, perhaps due to a network partition.
+This means that while a weak read from a client sees its own writes in this (perhaps pathological) case it might not see later writes by other clients.
+
 Finally, Cyclone allows operations to multiple keys in an atomic batch, where
 either all updates in the batch are applied or no operation in the batch is
-applied, regardless of failures. We also use a lightweight batched operation to
-support snapshots.
+applied - a property that holds \emph{at all replicas}, regardless of failures.
+We also use a lightweight batched operation to support snapshots.
+We discuss how batched operations are implemented later in the paper.
 
@@
 The first question we deal with is -- how do Cyclone clients decide which
-physical log to send a request to? The guarantees from Cyclone cover ordering of
+physical log to send a request to? The guarantees from Cyclone (except batched
+writes)  cover ordering of
 updates to a single key (Section~\ref{sec:sysarch}). They can be satisfied by
@@
 complete together even in the face of failures. One option is to have an
-external co-ordinator run a distributed transaction using two phase commit. In
-addition to requiring extra external resources to host the co-ordinator one must
-also worry about failure recovery. This also requires two round trips to the
-leader to complete what should have been a single batched write request from a
+external co-ordinator run a distributed transaction using two phase commit. This
+requires extra external resources to host the co-ordinator what should have been a single batched write request from a
 client that is stateless and allowed to fail - such as with key value stores
-like Rocksdb~\cite{rocksdb}. We therefore describe a simpler and more performant
+like Rocksdb~\cite{rocksdb}. This also 
+requires two round trips to the
+leader to complete  We therefore describe a simpler and more performant
 solution called ``ganged operations" in Cyclone. Ganged operations allow a
 stateless client to issue a batch of update operations as a ``gang",
-simultaneously and atomically updating a batch of keys in a single call.
+simultaneously and atomically updating a batch of keys in a single call. This preserves
+the system architecture of existing deployments of single image key value store such as Rocksdb.
 
@@
 value stores without compromising on their performance.
+
+Cyclone is available as open source software at:\\
+\url{https://github.com/sdulloor/cyclone/}
 \newcommand\myurl[2]{\url{#1}}

### 2017-11-19 paper/paper.tex
Updates.

@@
 \title{Cyclone: High Availability for Persistent Key Value Stores}
-\author{Amitabha Roy \and Subramanya R. Dulloor}
+\author{Amitabha Roy \\ amitabha.roy@gmail.com 
+	     \and Subramanya R. Dulloor \\ dulloor@gmail.com}
 
@@
 designed around a userspace network stack to relieve the CPU of the burden of
-data copies. Finally, it provides a way to efficiently map the commutativity in
-key-value store APIs to the parallelism available in commodity NICs.  Cyclone is
-able to replicate millions of small updates per second using only commodity 10
+data copies.  Finally, it provides a way to exploit the parallelism available in commodity NICs.  
+Cyclone is able to replicate millions of small updates per second using only commodity 10
 gigabit ethernet adapters. As a practical application, we use it to improve the
@@
 the local area network of a datacenter using commodity networks.  Research
-focusing on addressing wide area network latencies by reducing the number of
+focusing on addressing network latencies by reducing the number of
 network hops with new protocols such as Fast Paxos~\cite{fast-paxos} or network
@@
 key contribution of this work is to show how one can exploit this parallelism in
-Cyclone by making use of the commutativity inherent in key-value store
-interfaces and mapping it to the available parallelism in the system via
-multiple instances of the same consensus protocol.
+Cyclone automatically creates and manages multiple logs, mapping them to the 
+available parallelism in the system via multiple instances of the same consensus 
+protocol.
 
@@
 failover, resulting in only a brief interruption in service for
-clients.  Cyclone guarantees that the order of operations to any key are linearizable. 
-A key sees the exact same sequence of update operations (including deletes) 
-on all replicas. These per-key guarantees follow immediately from the 
-properties of the RAFT log and by requiring clients to perform a quorum read (logged
-as an entry in the RAFT log).
+clients.  Cyclone guarantees that the order of operations to any particular key 
+are linearizable. A key sees the exact same sequence of update operations 
+(including deletes) on all replicas. These per-key guarantees follow immediately 
+from the properties of the RAFT log and by requiring clients to perform a quorum 
+read (logged as an entry in the RAFT log).
 
 In addition, Cyclone also provides clients with a weak read that does not require a 
-quorum read thereby requiring only one round trip to a server.  A weak read of a key sees a prefix of the linearizable history of a key that 
-includes all successful update operations made by the client before it issues the 
-weak read. To do a weak read, 
+quorum read thereby requiring only one round trip to a server.  A weak read of a key sees a 
+prefix of the linearized sequence of updates to a key that includes all successful update operations made 
+by the client before it issues the weak read. To do a weak read, 
 clients maintain the last known term of the RAFT leader locally and ensure they always talk
 to a RAFT leader that has at least the same term. This ensures that clients see their own writes
-avoiding split brain problems where a network partitions allows clients to
-talk to old RAFT leaders after having committed changes to a new RAFT leader.
-In addition, a weak read waits until all pending RAFT log entries have been committed and applied before executing.
-This ensures that in the event of a failover, any previous committed updates from the same client have been applied and are visible to the weak read.
-A weak read is not linearizable as the RAFT leader responding to the weak read might actually no longer be the leader - a new one having been elected
-without it realizing, perhaps due to a network partition.
-This means that while a weak read from a client sees its own writes in this (perhaps pathological) case it might not see later writes by other clients.
+avoiding split brain problems where a network partition allows clients to
+talk to old RAFT leaders after having committed changes to a new RAFT leader. In addition, a weak 
+read waits until all pending RAFT log entries have been committed and applied before executing. 
+This ensures that in the event of a failover, any previous committed updates from the same client 
+have been applied and are visible to the weak read. A weak read is not linearizable as the RAFT 
+leader responding to the weak read might actually no longer be the leader - a new one having been 
+elected without it realizing, perhaps due to a network partition. This means that while a weak read 
+from a client sees its own writes, in some rare cases it might not see later writes by other clients.
 
@@
 either all updates in the batch are applied or no operation in the batch is
-applied - a property that holds \emph{at all replicas}, regardless of failures.
-We also use a lightweight batched operation to support snapshots.
-We discuss how batched operations are implemented later in the paper.
+applied - a property that holds \emph{at all replicas}, regardless of failures. Batched operations 
+are supported in Rocksdb to allow referential integrity between keys - where the value associated
+with a  key is a reference to another key. We also use a lightweight batched operation to support 
+snapshots - that can be viewed as an atomic no-op to all keys. We discuss how batched operations are 
+supported by Cyclone later in the paper.
 
@@
 small key-value pairs. At the same time, the amount of non-volatile memory
-required is small enough to not add undue cost to the server.
+required is small enough to not add undue cost to the server. This is in contrast to
+systems like FARM~\cite{farm} that require both the key value store and write ahead log
+to be held entirely in non-volatile memory.
 
@@
   \centering
-  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
+  \includegraphics[scale=0.6]{figures2/flashlog_page.pdf}
   \caption{Flashlog segment}
@@
 
-\begin{figure*}
-  \begin{minipage}{.2\textwidth}
-    \scriptsize
+\begin{figure}
+\small
 \begin{verbatim}
@@
 \label{fig:control_plane}
-\end{minipage}
-\begin{minipage}{.5\textwidth}
+\end{figure}
+
+\begin{figure*}
   \centering
-  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
+  \includegraphics[scale=0.6]{figures2/network_packet.pdf}
   \caption{Cyclone network packet layout}
   \label{fig:packet_layout}
-\end{minipage}
-\begin{minipage}{.3\textwidth}
+\end{figure*}
+
+\begin{figure}
 \centering
-\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+\includegraphics[scale=0.5]{figures2/nvm_log.pdf}
 \caption{NVM log structure}
 \label{fig:nvm_log}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \includegraphics[scale=0.5]{figures2/batching.pdf}
   \caption{Batching}
@@
 instances however exist in the same process address space as the key value store
-application itself and therefore manipulate the same application.
+application itself and therefore manipulate the same application. The number of 
+instances is a fixed property of the Cyclone service and cannot be changed after 
+startup.
 
@@
 physical log to send a request to? The guarantees from Cyclone (except batched
-writes)  cover ordering of
-updates to a single key (Section~\ref{sec:sysarch}). They can be satisfied by
-ensuring that all reads and writes to a key go to the same physical log
+writes)  cover ordering of updates to a single key (Section~\ref{sec:sysarch}). They can 
+be satisfied by ensuring that all reads and writes to a key go to the same physical log
 (i.e. RAFT instance). We achieve this by hashing the key to select the physical
@@
 \subsection{Ganged Operations}
-Supporting batched updates to multiple keys that map to different physical logs
-requires a ganged operation across those logs such that all the operations
-complete together even in the face of failures. One option is to have an
-external co-ordinator run a distributed transaction using two phase commit. This
-requires extra external resources to host the co-ordinator what should have been a single batched write request from a
-client that is stateless and allowed to fail - such as with key value stores
-like Rocksdb~\cite{rocksdb}. This also 
-requires two round trips to the
-leader to complete  We therefore describe a simpler and more performant
-solution called ``ganged operations" in Cyclone. Ganged operations allow a
-stateless client to issue a batch of update operations as a ``gang",
-simultaneously and atomically updating a batch of keys in a single call. This preserves
-the system architecture of existing deployments of single image key value store such as Rocksdb.
-
+Using multiple instances of RAFT leads to a serious problem with batched operations, 
+since we need to split up the batch into a mini-batch for each of the physical logs. There
+is however no guarantee that the request for a mini-batch will succeed breaking the atomicity
+requirement for batched updates. The obvious and simple solution is to do two phase commit across
+the logs, ensuring all or nothing semantics. However this was unacceptable to some customers who
+pointed out that ensuring that an external co-ordinator implementing two phase commit does not itself
+fail requires running a separate replicated service - with its attendant resource and reliability headaches.
+We were therefore challenged to come up with a solution that did not require two phase commit.
+
+Our final solution is based on two observations. First, we \emph {do not} require that either all mini-batches
+icommit to their raft logs or none do, What we actually requires is that either all are \emph{applied} to Rocksdb or 
+none of them are. Put another way we should not apply a mini-batch from a ganged operation if any of the other mini-batches
+have failed to commit to their RAFT log. This means that the client can be stateless. The second observation is that
+checking whether the other minibatches have committed is easy because the different RAFT instances can communicate 
+through shared memory on the same machines unlike the more general case of distributed participants in two phase 
+commit.
+
+We now describe our solution called ``ganged operations" in Cyclone. 
 The first key problem we need to deal with is that the leaders for the different
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \scriptsize
+  \small
   \bf
@@
 \label{fig:ganged_ops}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/replicas.pdf}
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/replicas.pdf}
   \caption{Impact of replica count}
   \label{fig:replicas}
-\end{minipage}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/512.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/512.pdf}
   \caption{Impact of payload size}
   \label{fig:payload}
-\end{minipage}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/flashlog.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/flashlog.pdf}
   \caption{Adding second level log}
   \label{fig:flashlog}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/flashlog_512.pdf}
   \caption{Second level log - 512 bytes payload}
   \label{fig:flashlog_512}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/multi.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/multi.pdf}
   \caption{Ganged Operations}
   \label{fig:ganged}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/kv_baseline.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/kv_baseline.pdf}
   \caption{Baseline KV replication}
   \label{fig:kv_baseline}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-  \begin{minipage}{0.33\textwidth}
+\begin{figure}
     \centering
-    \includegraphics[scale=0.45]{results2/rocksdb.pdf}
+    \includegraphics[scale=0.6]{results2/rocksdb.pdf}
     \caption{Rocksdb - small updates}
     \label{fig:rocksdb}
-  \end{minipage} 
-  \begin{minipage}{0.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_256.pdf}
+  \end{figure}
+
+  \begin{figure}
+    \includegraphics[scale=0.6]{results2/rocksdb_256.pdf}
     \caption{Rocksdb - large updates}
     \label{fig:rocksdb_256}
-  \end{minipage} 
-  \begin{minipage}{0.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_multi.pdf}
+  \end{figure}
+
+  \begin{figure}
+    \includegraphics[scale=0.6]{results2/rocksdb_multi.pdf}
     \caption{Rocksdb - batched writes}
     \label{fig:rocksdb_multi}
-  \end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read heavy (95\%
-reads). Figure~\ref{fig:fb} illustrates that even with a read dominated
+reads). We use weak reads in Cyclone to avoid the extra round trip for quorum reads.
+Figure~\ref{fig:fb} illustrates that even with a read dominated
 workload, Cyclone is effective at providing a write ahead log with 3 way
@@
 
-\begin{figure*}
-\begin{minipage}{0.33\textwidth}
-\includegraphics[scale=0.45]{results2/fb.pdf}
+\begin{figure}
+\includegraphics[scale=0.6]{results2/fb.pdf}
 \caption{Rocksdb - read heavy}
 \label{fig:fb}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/fb2.pdf}
+\end{figure}
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/fb2.pdf}
   \caption{Rocksdb - write heavy}
   \label{fig:fb2}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/failover.pdf}
+\end{figure}
+\begin{figure}
+  \includegraphics[scale=0.6]{results2/failover.pdf}
   \caption{Rocksdb - failover}
   \label{fig:timeline}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 %shared memory program to scale in performance with concurrency it must also have
-%an interface that is commutative for common case invocations. We have also used
+%an interface that is  for common case invocations. We have also used
 %interface level commutativity to extract parallelism for the purpose of scaling

### 2017-11-19 paper/paper_arxiv.tex
Arxiv submission sources.
Updates.

--- 2017-11-17 paper/paper.tex
@@
 \begin{figure}
-\centering \includegraphics[scale=0.6]{results2/problem.pdf}
+\centering \includegraphics[scale=0.6]{problem.pdf}
 \caption{Rocksdb write ahead logging impact}
@@
   \centering
-  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
+  \includegraphics[scale=0.6]{flashlog_page.pdf}
   \caption{Flashlog segment}
@@
 \label{fig:control_plane}
-\end{minipage}
-\begin{minipage}{.5\textwidth}
+\end{figure}
+
+\begin{figure*}
   \centering
-  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
+  \includegraphics[scale=0.6]{network_packet.pdf}
   \caption{Cyclone network packet layout}
   \label{fig:packet_layout}
-\end{minipage}
-\begin{minipage}{.3\textwidth}
+\end{figure*}
+
+\begin{figure}
 \centering
-\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+\includegraphics[scale=0.5]{nvm_log.pdf}
 \caption{NVM log structure}
 \label{fig:nvm_log}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
   \centering
-  \includegraphics[scale=0.3]{figures2/batching.pdf}
+  \includegraphics[scale=0.5]{batching.pdf}
   \caption{Batching}
@@
 %  \centering
-%  \includegraphics[scale=0.35]{figures2/race.pdf}
+%  \includegraphics[scale=0.35]{race.pdf}
 %  \caption{Race with multiple physical logs}
@@
 \begin{figure}
-\includegraphics[scale=0.6]{results2/network_opts.pdf}
+\includegraphics[scale=0.6]{network_opts.pdf}
 \caption{Network optimizations for top level log}
@@
 
-\begin{figure*}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/replicas.pdf}
+\begin{figure}
+  \includegraphics[scale=0.6]{replicas.pdf}
   \caption{Impact of replica count}
   \label{fig:replicas}
-\end{minipage}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/512.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{512.pdf}
   \caption{Impact of payload size}
   \label{fig:payload}
-\end{minipage}
-\begin{minipage}{.33\textwidth}
-  \includegraphics[scale=0.45]{results2/flashlog.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{flashlog.pdf}
   \caption{Adding second level log}
   \label{fig:flashlog}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
+\begin{figure}
+  \includegraphics[scale=0.6]{flashlog_512.pdf}
   \caption{Second level log - 512 bytes payload}
   \label{fig:flashlog_512}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/multi.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{multi.pdf}
   \caption{Ganged Operations}
   \label{fig:ganged}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/kv_baseline.pdf}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[scale=0.6]{kv_baseline.pdf}
   \caption{Baseline KV replication}
   \label{fig:kv_baseline}
-\end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-  \begin{minipage}{0.33\textwidth}
+\begin{figure}
     \centering
-    \includegraphics[scale=0.45]{results2/rocksdb.pdf}
+    \includegraphics[scale=0.6]{rocksdb.pdf}
     \caption{Rocksdb - small updates}
     \label{fig:rocksdb}
-  \end{minipage} 
-  \begin{minipage}{0.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_256.pdf}
+  \end{figure}
+
+  \begin{figure}
+    \includegraphics[scale=0.6]{rocksdb_256.pdf}
     \caption{Rocksdb - large updates}
     \label{fig:rocksdb_256}
-  \end{minipage} 
-  \begin{minipage}{0.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_multi.pdf}
+  \end{figure}
+
+  \begin{figure}
+    \includegraphics[scale=0.6]{rocksdb_multi.pdf}
     \caption{Rocksdb - batched writes}
     \label{fig:rocksdb_multi}
-  \end{minipage}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-\begin{minipage}{0.33\textwidth}
-\includegraphics[scale=0.45]{results2/fb.pdf}
+\begin{figure}
+\includegraphics[scale=0.6]{fb.pdf}
 \caption{Rocksdb - read heavy}
 \label{fig:fb}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/fb2.pdf}
+\end{figure}
+\begin{figure}
+  \includegraphics[scale=0.6]{fb2.pdf}
   \caption{Rocksdb - write heavy}
   \label{fig:fb2}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-  \includegraphics[scale=0.45]{results2/failover.pdf}
+\end{figure}
+\begin{figure}
+  \includegraphics[scale=0.6]{failover.pdf}
   \caption{Rocksdb - failover}
   \label{fig:timeline}
-\end{minipage}
-\end{figure*}
+\end{figure}
 

### 2017-12-21 paper/paper.tex
Switch to usenix format.
Update.

@@
-\documentclass[twocolumn]{article}
+\documentclass[letterpaper,twocolumn,10pt]{article}
+\usepackage{usenix}
 
@@
 \title{Cyclone: High Availability for Persistent Key Value Stores}
-\author{Amitabha Roy \\ amitabha.roy@gmail.com 
-	     \and Subramanya R. Dulloor \\ dulloor@gmail.com}
+
+\author{{\rm Amitabha Roy} \\ amitabha.roy@gmail.com 
+	     \and {\rm Subramanya R. Dulloor} \\ dulloor@gmail.com}
 
@@
 \vspace{0.05in}
-Persistent key value stores are an important component of many distributed data
-serving solutions with innovations targeted at taking advantage of growing flash
-speeds. Unfortunately their performance is hampered by the need to maintain and
-replicate a write ahead log to guarantee availability in the face of machine and
-storage failures. Cyclone is a replicated log plug-in for key value stores that
-systematically addresses various sources of this bottleneck. It uses a small
-amount of non-volatile memory directly addressable by the CPU - such as in the
-form of NVDIMMs or Intel 3DXPoint - to remove block oriented IO devices such as
-SSDs from the critical path for appending to the log. This enables it to address
-network overheads using an implementation of the RAFT consensus protocol that is
-designed around a userspace network stack to relieve the CPU of the burden of
-data copies.  Finally, it provides a way to exploit the parallelism available in commodity NICs.  
-Cyclone is able to replicate millions of small updates per second using only commodity 10
-gigabit ethernet adapters. As a practical application, we use it to improve the
-performance (and availability) of RocksDB, a popular persistent key value store
-by an order of magnitude when compared to its own write ahead log without
-replication.
+Persistent key value stores are an important component of many
+distributed data serving solutions with innovations targeted at taking
+advantage of growing flash speeds. Unfortunately their performance is
+hampered by the need to maintain and replicate a write ahead log to
+guarantee availability in the face of machine and storage
+failures. Cyclone is a replicated log plug-in for key value stores
+that systematically addresses various sources of this bottleneck. It
+uses a small amount of non-volatile memory directly addressable by the
+CPU - such as in the form of NVDIMMs or Intel 3DXPoint - to remove
+block oriented IO devices such as SSDs from the critical path for
+appending to the log. This then enables it to address network
+overheads using an implementation of the RAFT consensus protocol that
+is designed around a userspace network stack to relieve the CPU of the
+burden of data copies.  Finally, it provides a way to exploit the
+parallelism available in commodity NICs.  Cyclone is able to replicate
+millions of small updates per second using only commodity 10 gigabit
+ethernet adapters. As a practical application, we use it to improve
+the performance (and availability) of RocksDB, a popular persistent
+key value store by an order of magnitude when compared to its own
+write ahead log without replication.
 \end{abstract}
@@
 FloDB~\cite{flodb} turn off the write ahead log to be able to showcase benefits
-of sophisticated extensions to LSM data structures.
+of extensions to LSM data structures.
 
 \begin{figure}
-\centering \includegraphics[scale=0.6]{results2/problem.pdf}
+\centering \includegraphics[scale=0.5]{results2/problem.pdf}
 \caption{Rocksdb write ahead logging impact}
@@
 
-These two performance bottlenecks illustrated in Figure~\ref{fig:problem} are
-\emph{not independent}, they cannot be addressed in isolation. In particular,
-durability semantics mean that the key-value store cannot respond to the client
-until the operation has been persisted to the log at a majority of replicas,
-since it must be completed even on a crash and subsequent recovery. Therefore,
-addressing the network bottleneck is pointless without also addressing the
-storage bottleneck. Cyclone shows how a small amount of non-volatile memory can
-be used to address both bottlenecks.
-
 Cyclone is a high speed strongly consistent replicated write ahead logging
@@
 closes the performance gap due to both the storage and the network shown in
-Figure~\ref{fig:problem}. We show that Cyclone achieves performance transparent
-replication in that it can provide both a persistent write ahead log and
-replication without compromising on RocksDB's performance.
+Figure~\ref{fig:problem}.
 
@@
 amount of non-volatile memory when combined with the flash SSD based persistent
-logs.  The primary cause of slowdown when turning on a write ahead log on a
-single machine in Figure~\ref{fig:problem} is the need to do synchronous block
-IO for every update \emph{regardless of the size of the update}. SSDs provide
-the maximum throughput when writing large chunks (usually more than 4KB) of
-data. This is due to the fixed cost of a round trip to the SSD through the IO
-interface (NVMe in our case), regardless of the volume of data being
-synchronously persisted. Small updates (sum of key and value sizes) are
-surprisingly common in observed workloads for key value pairs. For example,
-Facebook~\cite{fb_workload_analysis} reports that 90\% of the space allocated in
-key value stores is for data items under 500 bytes in size.
-
-Cyclone avoids this overhead by making use of a small amount of directly
-attached non-volatile memory (NVM), also called persistent memory, on the
-server.  This can be in the form of NVDIMMs (DIMMs with an added ultracapacitor
-to dump state to a small amount of attached flash) -or- newer and potentially
-cheaper (but slower) forms of persistent memory~\cite{pmfs}.
-%For current deployments, NVDIMMs are significantly more expensive than standard
-%DIMMs, which means that they will %likely be available only in smaller
-%quantities compared to flash.
-In cases where the log does not entirely fit in NVM, we periodically drain the
-NVM log into a log placed on a standard SSD on the system.
-
-The second key contribution of this work is in showing that the throughput
-bottlenecks in the network stack can be mitigated entirely in software, once
-block IO is removed from the critical path. Existing work on log replication
-using state machine replication protocols such as Paxos~\cite{paxos} and
-RAFT~\cite{raft} has not - thus far - addressed high performance replication in
-the local area network of a datacenter using commodity networks.  Research
-focusing on addressing network latencies by reducing the number of
-network hops with new protocols such as Fast Paxos~\cite{fast-paxos} or network
-switch modifications such as No Paxos~\cite{nopaxos} do not address throughput
-bottlenecks due to network stack related inefficiencies in the participating
-nodes of the quorum -- a more urgent problem with local area replication. On the
-other hand, work done to address network related inefficiencies start with the
-assumption that the bottleneck to good throughput on the network is the CPU and
-therefore one must either resort to network offload mechanisms such as RDMA~\cite{dare,
+logs. We show that this entirely eliminates the storage bottleneck due to block IO.
+
+The second key contribution of this work is in showing that the
+throughput bottlenecks in the network stack can be mitigated entirely
+in software, once block IO is removed from the critical path. Existing
+work on log replication makes the assumption that the bottleneck to
+good throughput on the network is the CPU and therefore one must
+either resort to network offload mechanisms such as RDMA~\cite{dare,
   farm, faast} or offload the protocol entirely to an
-FPGA~\cite{consensus_box}. We demonstrate in this paper that this is not the
-case. Rather, once persistence is provided by directly attached non-volatile
-memory, thereby eliminating block IO from the persistence step in replication
-protocols, the problem becomes akin to multicasting the same log entry
-from the leader to all followers. We design Cyclone's network stack around well
-known principles in the software packet switching community. In particular, a
-careful implementation of a consensus protocol that relieves the CPU of the
-responsibility of data movement permits a high performance implementation
-entirely in software that addresses the order of magnitude network related
-performance gap in Figure~\ref{fig:problem} using only commodity 10 Gigabit
-ethernet.
-
-Both the storage (NVM, SSD) as well as the NIC expose significant parallelism
-that is left unused when Cyclone is used to replicate a single log. The third
-key contribution of this work is to show how one can exploit this parallelism in
-Cyclone automatically creates and manages multiple logs, mapping them to the 
-available parallelism in the system via multiple instances of the same consensus 
-protocol.
+FPGA~\cite{consensus_box}. We demonstrate in this paper that this is
+not the case. Rather, once persistence is provided by directly
+attached non-volatile memory, thereby eliminating block IO from the
+persistence step in replication protocols, the problem becomes akin to
+multicasting the same log entry from the leader to all followers. We
+design Cyclone's network stack around well known principles in the
+software packet switching community. In particular, a careful
+implementation of a consensus protocol that relieves the CPU of the
+responsibility of data movement permits a high performance
+implementation entirely in software that addresses the order of
+magnitude network related performance gap in Figure~\ref{fig:problem}
+using only commodity 10 Gigabit ethernet.
+
+Both the storage (NVM, SSD) as well as the NIC expose significant
+parallelism that is left unused when Cyclone is used to replicate a
+single log. The third key contribution of this work is to show how one
+can exploit this parallelism for replicating the write ahead log in
+key value stores.  Cyclone automatically creates and manages multiple
+logs, mapping them to the available parallelism in the system via
+multiple instances of the same consensus protocol.
 
@@
 
-Cyclone uses the RAFT consensus protocol~\cite{raft} to keep the logs in sync
-across the replicas. On failure of the leader replica, a new leader is
-automatically elected -- RAFT ensures that the new leader has the most up to
-date log. The Cyclone client library automatically locates the new leader on a
-failover, resulting in only a brief interruption in service for
-clients.  Cyclone guarantees that the order of operations to any particular key 
-are linearizable. A key sees the exact same sequence of update operations 
-(including deletes) on all replicas. These per-key guarantees follow immediately 
-from the properties of the RAFT log and by requiring clients to perform a quorum 
-read (logged as an entry in the RAFT log).
-
-In addition, Cyclone also provides clients with a weak read that does not require a 
-quorum read thereby requiring only one round trip to a server.  A weak read of a key sees a 
-prefix of the linearized sequence of updates to a key that includes all successful update operations made 
-by the client before it issues the weak read. To do a weak read, 
-clients maintain the last known term of the RAFT leader locally and ensure they always talk
-to a RAFT leader that has at least the same term. This ensures that clients see their own writes
-avoiding split brain problems where a network partition allows clients to
-talk to old RAFT leaders after having committed changes to a new RAFT leader. In addition, a weak 
-read waits until all pending RAFT log entries have been committed and applied before executing. 
-This ensures that in the event of a failover, any previous committed updates from the same client 
-have been applied and are visible to the weak read. A weak read is not linearizable as the RAFT 
-leader responding to the weak read might actually no longer be the leader - a new one having been 
-elected without it realizing, perhaps due to a network partition. This means that while a weak read 
-from a client sees its own writes, in some rare cases it might not see later writes by other clients.
-
-Finally, Cyclone allows operations to multiple keys in an atomic batch, where
-either all updates in the batch are applied or no operation in the batch is
-applied - a property that holds \emph{at all replicas}, regardless of failures. Batched operations 
-are supported in Rocksdb to allow referential integrity between keys - where the value associated
-with a  key is a reference to another key. We also use a lightweight batched operation to support 
-snapshots - that can be viewed as an atomic no-op to all keys. We discuss how batched operations are 
-supported by Cyclone later in the paper.
+Cyclone uses the RAFT consensus protocol~\cite{raft} to keep the logs
+in sync across the replicas. On failure of the leader replica, a new
+leader is automatically elected -- RAFT ensures that the new leader
+has the most up to date log. The Cyclone client library automatically
+locates the new leader on a failover, resulting in only a brief
+interruption in service for clients.  Cyclone guarantees that the
+order of operations to any particular key are linearizable. A key sees
+the exact same sequence of update operations (including deletes) on
+all replicas. These per-key guarantees follow immediately from the
+properties of the RAFT log and by requiring clients reading a key to
+perform a quorum read (logged as an entry in the RAFT log).
+
+In addition, Cyclone also provides allows clients to do a weak read
+that returns the current value of the key at the last known leader
+replica. Weak reads can see a possibly stale value
+(i.e. is not linearizable) but is howevere guaranteed to have seen all
+of its own previous writes. On the positive side, weak reads avoid an
+extra round trip for quorum consensus.
+
+Finally, Cyclone allows operations to multiple keys in an atomic
+batch, where either all updates in the batch are applied or no
+operation in the batch is applied - a property that holds \emph{at all
+  replicas}, regardless of failures. Batched operations are supported
+in Rocksdb to allow referential integrity between keys - where the
+value associated with a key is a reference to another key. We also use
+a lightweight batched operation to support snapshots - that can be
+viewed as an atomic no-op to all keys. We discuss how batched
+operations are supported by Cyclone later in the paper.
 
@@
 \label{sec:storage}
-Storage devices such as flash-based SSDs export a block IO interface. Appending
-a small update to the write-ahead log requires a synchronous access to the
-SSD. Even with write coalescing using a write cache on the SSD, the round trip
-time to a current generation NVMe SSD such as the one we use in this
-paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
-under 50K ops/sec and resulting in a high baseline latency at even low load (as
-shown in Figure~\ref{fig:problem}). This problem leads many system designers to
-under-provision key-value store shards ensuring that it operates at moderate to
-high load, using group-commit to batch additions to the log. They therefore pay
-a price in latency to ensure the bandwidth to storage is fully utilized.
-
-Cyclone adopts a different strategy by making use of novel memory technologies
-to obviate the need to make this tradeoff in the first place. We assume a small
-amount of non-volatile memory directly addressable by the CPU, rather than being
-placed behind a block oriented IO interface. Log appends are done to the ``NVM
-log'' placed in this non-volatile memory. However, directly attached persistent 
-memory in the form of NVDIMMs today or new memory technologies in the
-future~\cite{pmfs} will be of lesser capacity than traditional (NAND based)
-flash available through an
-SSD. Providing the same capacity as existing flash based logs for key value
-stores therefore requires us to provide a second level of the log
-placed on a flash SSD (called flashlog).  Entries are drained from
-the NVM log to this flashlog in conveniently large units that are a multiple of
-4KB -- the optimum IO unit for flash SSDs. This ensures we get the best possible
-throughput from a flash device without paying the price of synchronous IO for
-small key-value pairs. At the same time, the amount of non-volatile memory
-required is small enough to not add undue cost to the server. This is in contrast to
-systems like FARM~\cite{farm} that require both the key value store and write ahead log
-to be held entirely in non-volatile memory.
+The storage bottleneck for the write ahead log in key value stores
+such as Rocksdb is - surprisingly - not the storage media (i.e. SSD)
+itself but rather the interface (i.e block IO). This is a particular
+pain point for small writes.  For example,
+Facebook~\cite{fb_workload_analysis} reports that 90\% of the space
+allocated in key value stores is for data items under 500 bytes in
+size.  The SSD used in Figure~\ref{fig:problem}, we achieve 50
+Kops/sec to the SSD being limited to a 20us latency on the PCIe
+interface. This translates to 25 MB/s for 500 byte key-value
+pairs. The SSD is capable of achieving over 1 GB/s of sequential write
+bandwidth to an append only log. Clearly the SSDs performance is not
+the problem, but rather the block IO interface that imposes the round
+trip latency \emph{regardless of the size of the update}.
+
+Cyclone solves this problem by making use of novel persistent main
+memory technologies. We assume a small amount of non-volatile memory
+directly addressable by the CPU, rather than being placed behind a
+block oriented IO interface. Log appends are done to the ``NVM log''
+placed in this non-volatile memory. However, directly attached
+persistent memory in the form of NVDIMMs today or new memory
+technologies in the future~\cite{pmfs} will be of lesser capacity than
+traditional (NAND based) flash available through an SSD. Providing the
+same capacity as existing flash based logs for key value stores
+therefore requires us to provide a second level of the log placed on a
+flash SSD (called flashlog).  Entries are drained from the NVM log to
+this flashlog in conveniently large units that are a multiple of 4KB
+-- the optimum IO unit for flash SSDs. This ensures we get the best
+possible throughput from a flash device without paying the price of
+synchronous IO for small key-value pairs. At the same time, the amount
+of non-volatile memory required is small enough to not add undue cost
+to the server. This is in contrast to systems like FARM~\cite{farm}
+that require both the key value store and write ahead log to be held
+entirely in non-volatile memory.
 
@@
   \centering
-  \includegraphics[scale=0.6]{figures2/flashlog_page.pdf}
+  \includegraphics[scale=0.4]{figures2/flashlog_page.pdf}
   \caption{Flashlog segment}
@@
 
-In this section we demonstrated how Cyclone uses a small amount of non-volatile
-memory to improve storage bottlenecks when maintaining a durable log on a single
-machine. In the next section, we deal with replicating this log efficiently
-across multiple machines.
-
 \section{Network}
 \label{sec:network}
-Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the log of
-operations across machines. Cyclone provides guarantee to the client that any
+Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the
+log of operations across machines. The key bottleneck in log
+replication in Figure~\ref{fig:problem} is the CPU - the achieved
+network throughput at load is well under the capacity of the 10 Gbps
+link.  The key to improving CPU utilization for log replication is the
+insight that log replcation protocols such as RAFT are oblivious to
+the actual contents of the log entry being replicated.  This
+observation allowed us to greatly improve replication performance by
+building a (near) zero copy implementation of RAFT. Cyclone's
+implementation of RAFT does only a small \emph{constant} amount of
+work per replicated log entry on the CPU leaving the (commodity) NIC
+to do the rest of the heavy lifting.
+
+Cyclone provides the guarantee to the client that any
 operation responded to is durable and its results \emph{available} as long as a
@@
 
-\begin{figure}
-\small
+\begin{figure*}
+\begin{minipage}{.5\textwidth}
+\scriptsize
 \begin{verbatim}
@@
 \label{fig:control_plane}
-\end{figure}
-
-\begin{figure*}
+\end{minipage}
+\begin{minipage}{.5\textwidth}
   \centering
-  \includegraphics[scale=0.6]{figures2/network_packet.pdf}
+  \includegraphics[scale=0.3]{figures2/network_packet.pdf}
   \caption{Cyclone network packet layout}
   \label{fig:packet_layout}
+\end{minipage}
 \end{figure*}
@@
 \centering
-\includegraphics[scale=0.5]{figures2/nvm_log.pdf}
+\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
 \caption{NVM log structure}
@@
   \centering
-  \includegraphics[scale=0.5]{figures2/batching.pdf}
+  \includegraphics[scale=0.3]{figures2/batching.pdf}
   \caption{Batching}
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \small
+  \scriptsize
   \bf
@@
 bottleneck to reaching line rate.
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/replicas.pdf}
+\begin{figure*}
+\begin{minipage}{0.33\textwidth}
+  \includegraphics[scale=0.45]{results2/replicas.pdf}
   \caption{Impact of replica count}
   \label{fig:replicas}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/512.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/512.pdf}
   \caption{Impact of payload size}
   \label{fig:payload}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/flashlog.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/flashlog.pdf}
   \caption{Adding second level log}
   \label{fig:flashlog}
-\end{figure}
+\end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/flashlog_512.pdf}
+\begin{figure*}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
   \caption{Second level log - 512 bytes payload}
   \label{fig:flashlog_512}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/multi.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/multi.pdf}
   \caption{Ganged Operations}
   \label{fig:ganged}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/kv_baseline.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/kv_baseline.pdf}
   \caption{Baseline KV replication}
   \label{fig:kv_baseline}
-\end{figure}
+\end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
+\begin{minipage}{.33\textwidth}
     \centering
-    \includegraphics[scale=0.6]{results2/rocksdb.pdf}
+    \includegraphics[scale=0.45]{results2/rocksdb.pdf}
     \caption{Rocksdb - small updates}
     \label{fig:rocksdb}
-  \end{figure}
-
-  \begin{figure}
-    \includegraphics[scale=0.6]{results2/rocksdb_256.pdf}
+  \end{minipage}
+  \begin{minipage}{.33\textwidth}
+    \includegraphics[scale=0.45]{results2/rocksdb_256.pdf}
     \caption{Rocksdb - large updates}
     \label{fig:rocksdb_256}
-  \end{figure}
-
-  \begin{figure}
-    \includegraphics[scale=0.6]{results2/rocksdb_multi.pdf}
+  \end{minipage}
+  \begin{minipage}{.33\textwidth}
+    \includegraphics[scale=0.45]{results2/rocksdb_multi.pdf}
     \caption{Rocksdb - batched writes}
     \label{fig:rocksdb_multi}
-\end{figure}
+\end{minipage}
+\end{figure*}
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{results2/fb.pdf}
+\begin{figure*}
+\begin{minipage}{.33\textwidth}
+\includegraphics[scale=0.45]{results2/fb.pdf}
 \caption{Rocksdb - read heavy}
 \label{fig:fb}
-\end{figure}
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/fb2.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/fb2.pdf}
   \caption{Rocksdb - write heavy}
   \label{fig:fb2}
-\end{figure}
-\begin{figure}
-  \includegraphics[scale=0.6]{results2/failover.pdf}
+\end{minipage}
+\begin{minipage}{.33\textwidth}
+  \includegraphics[scale=0.45]{results2/failover.pdf}
   \caption{Rocksdb - failover}
   \label{fig:timeline}
-\end{figure}
+\end{minipage}
+\end{figure*}
 

### 2017-12-21 paper/paper.tex
Update.

@@
 Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the
-log of operations across machines. The key bottleneck in log
-replication in Figure~\ref{fig:problem} is the CPU - the achieved
-network throughput at load is well under the capacity of the 10 Gbps
-link.  The key to improving CPU utilization for log replication is the
-insight that log replcation protocols such as RAFT are oblivious to
-the actual contents of the log entry being replicated.  This
-observation allowed us to greatly improve replication performance by
-building a (near) zero copy implementation of RAFT. Cyclone's
-implementation of RAFT does only a small \emph{constant} amount of
-work per replicated log entry on the CPU leaving the (commodity) NIC
-to do the rest of the heavy lifting.
+log of operations across machines. The bottleneck for log replication
+in Figure~\ref{fig:problem} is the CPU - the achieved network
+throughput at load is well under the capacity of the 10 Gbps link. The
+key to improving CPU utilization for log replication is the insight
+that log replcation protocols such as RAFT are oblivious to the actual
+contents of the log entry being replicated.  This observation allowed
+us to greatly improve replication performance by building a (near)
+zero copy implementation of RAFT. Cyclone's implementation of RAFT
+does only a small \emph{constant} amount of work per replicated log
+entry on the CPU to run the RAFT protocol state machine.  Copying work
+that proportional to the amount of data being replicated is done by
+the (commodity) NIC.  We therefore shift the bottleneck to the NIC,
+making full use of the available network link.
 

### 2017-12-23 paper/paper.tex
Updates.

@@
 ethernet adapters. As a practical application, we use it to improve
-the performance (and availability) of RocksDB, a popular persistent
-key value store by an order of magnitude when compared to its own
-write ahead log without replication.
+the performance of RocksDB, a popular persistent key value store by an
+order of magnitude when compared to its own write ahead log without
+replication.
 \end{abstract}
@@
 \section{Introduction}
-Persistent key value stores are an important component of datacenter scale
-storage services. Key value stores such as Rocksdb~\cite{rocksdb},
-LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent significant efforts on
-both engineering and research fronts. These key value stores include
-sophisticated in-memory data structures built around Log Structured Merge (LSM)
-trees~\cite{lsmtree} and are heavily tuned to extract maximum performance from
-flash-based solid state drives (SSDs).
+Persistent key value stores are an important component of datacenter
+scale storage services. Key value stores such as
+Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb}
+represent significant efforts on both engineering and research
+fronts. They include sophisticated in-memory data structures built
+around Log Structured Merge (LSM) trees~\cite{lsmtree} and are heavily
+tuned to extract maximum performance from flash-based solid state
+drives (SSDs).
 
@@
 
-The first key contribution of Cyclone is demonstrating the benefit of a small
-amount of non-volatile memory when combined with the flash SSD based persistent
-logs. We show that this entirely eliminates the storage bottleneck due to block IO.
+The first key contribution of Cyclone is demonstrating the benefit of
+a small amount of non-volatile memory when combined with the flash SSD
+based persistent logs. We show that this entirely eliminates the
+storage bottleneck due to block IO.
 
@@
 throughput bottlenecks in the network stack can be mitigated entirely
-in software, once block IO is removed from the critical path. Existing
-work on log replication makes the assumption that the bottleneck to
-good throughput on the network is the CPU and therefore one must
-either resort to network offload mechanisms such as RDMA~\cite{dare,
-  farm, faast} or offload the protocol entirely to an
+in software. Existing work on log replication makes the assumption
+that the bottleneck to good throughput on the network is the CPU and
+therefore one must either resort to network offload mechanisms such as
+RDMA~\cite{dare, farm, faast} or offload the protocol entirely to an
 FPGA~\cite{consensus_box}. We demonstrate in this paper that this is
-not the case. Rather, once persistence is provided by directly
-attached non-volatile memory, thereby eliminating block IO from the
-persistence step in replication protocols, the problem becomes akin to
-multicasting the same log entry from the leader to all followers. We
-design Cyclone's network stack around well known principles in the
-software packet switching community. In particular, a careful
-implementation of a consensus protocol that relieves the CPU of the
-responsibility of data movement permits a high performance
-implementation entirely in software that addresses the order of
-magnitude network related performance gap in Figure~\ref{fig:problem}
-using only commodity 10 Gigabit ethernet.
-
-Both the storage (NVM, SSD) as well as the NIC expose significant
-parallelism that is left unused when Cyclone is used to replicate a
-single log. The third key contribution of this work is to show how one
-can exploit this parallelism for replicating the write ahead log in
-key value stores.  Cyclone automatically creates and manages multiple
-logs, mapping them to the available parallelism in the system via
-multiple instances of the same consensus protocol.
+not the case. In particular, a careful implementation of a consensus
+protocol that relieves the CPU of the responsibility of data movement
+permits a high performance implementation entirely in software that
+addresses the order of magnitude network related performance gap in
+Figure~\ref{fig:problem} using only commodity 10 Gigabit ethernet.
+
+Commodity network interfaces expose significant parallelism via
+multiple transmit receive queues that is left unused when Cyclone is
+used to replicate a single log. The third key contribution of this
+work is to show how one can exploit this parallelism for replicating
+the write ahead log in key value stores. Cyclone creates and manages
+multiple logs, mapping them to the available parallelism via multiple
+instances of the same consensus protocol.
 
@@
 Key value stores such as RocksDB also support atomic writes ({\tt PUT}
-operations) to multiple keys, called batched writes, and the ability to take
-snapshots.
-
-Cyclone integrates with key value stores as both a client and server side
-library. Cyclone replicates the key value store across a set of replica
-servers. The client side library sends \emph{all} requests - reads and updates -
-to a distinguished leader replica.  On the server side Cyclone accepts requests
-from clients, calls into the key value store, and returns a response to the
-client. Before executing any PUT or DELETE request, Cyclone appends it to a
-durable log and replicates the request to the logs of follower replicas. The
-request is considered replicated once acknowledged by a majority of
+operations) to multiple keys, called batched writes, and the ability
+to take snapshots. Batched operations are supported in Rocksdb to
+allow referential integrity between keys - where the value associated
+with a key is a reference to another key.
+
+Cyclone integrates with key value stores as both a client and server
+side library. Cyclone replicates the key value store across a set of
+replica servers. The client side library sends \emph{all} requests -
+reads and updates - to a distinguished leader replica.  On the server
+side Cyclone accepts requests from clients, calls into the key value
+store, and returns a response to the client. Before executing any PUT
+or DELETE request, Cyclone appends it to a durable log and replicates
+the request to the logs of follower replicas. The request is
+considered replicated once acknowledged by a majority of
 replicas. Follower replicas apply updates from the log in order.
@@
 locates the new leader on a failover, resulting in only a brief
-interruption in service for clients.  Cyclone guarantees that the
-order of operations to any particular key are linearizable. A key sees
-the exact same sequence of update operations (including deletes) on
+interruption in service for clients.  Cyclone guarantees that each key 
+sees the same linearizable order of operations. (including deletes) on
 all replicas. These per-key guarantees follow immediately from the
@@
 
-In addition, Cyclone also provides allows clients to do a weak read
-that returns the current value of the key at the last known leader
-replica. Weak reads can see a possibly stale value
-(i.e. is not linearizable) but is howevere guaranteed to have seen all
-of its own previous writes. On the positive side, weak reads avoid an
-extra round trip for quorum consensus.
-
-Finally, Cyclone allows operations to multiple keys in an atomic
-batch, where either all updates in the batch are applied or no
+In addition, Cyclone allows clients to do a weak read that returns the
+current value of the key at the last known leader replica. Weak reads
+from a client can see a possibly stale value (are not linearizable)
+but are guaranteed to have seen all of its own previous writes. On the
+positive side, weak reads avoid an extra round trip for quorum
+consensus.
+
+For completeness, Cyclone allows operations to multiple keys in an
+atomic batch, where either all updates in the batch are applied or no
 operation in the batch is applied - a property that holds \emph{at all
-  replicas}, regardless of failures. Batched operations are supported
-in Rocksdb to allow referential integrity between keys - where the
-value associated with a key is a reference to another key. We also use
-a lightweight batched operation to support snapshots - that can be
-viewed as an atomic no-op to all keys. We discuss how batched
-operations are supported by Cyclone later in the paper.
+  replicas}, regardless of failures.  We also use a lightweight
+batched operation to support snapshots - that can be viewed as an
+atomic no-op to all keys.
 
@@
 contents of the log entry being replicated.  This observation allowed
-us to greatly improve replication performance by building a (near)
+us to greatly improve replication performance by building a
 zero copy implementation of RAFT. Cyclone's implementation of RAFT
@@
 entry on the CPU to run the RAFT protocol state machine.  Copying work
-that proportional to the amount of data being replicated is done by
+is proportional to the amount of data being replicated and is done by
 the (commodity) NIC.  We therefore shift the bottleneck to the NIC,
@@
 
-Cyclone provides the guarantee to the client that any
-operation responded to is durable and its results \emph{available} as long as a
-majority of replicas continue to function. Cyclone's failure model is
-\emph{fail-recovery}, where nodes that can recover from failure simply continue
-to participate in the protocol. This requires RAFT to persist every log entry
-\emph{before} it sends it out for replication and before a follower replica
-responds to the leader. Storage overheads are therefore not independent of
-network overheads, and we must deal with both storage and networking overheads
-simultaneously to improve the performance of replication.
-
-We leverage the two level log from the previous section to remove storage
-bottlenecks from the critical path of replication in much the same way we did
-for logging on a single machine. RAFT is only aware of and replicates the top
-level NVM log. Each replica \emph{independently} drains committed (in terms of
-RAFT consensus) entries from the NVM log to the flashlog. Block IO therefore is
-no longer a consideration when optimizing data movement over the network, unlike
-systems that need to deal with such problems~\cite{reflex}. The task of
-replication reduces to that of efficiently sending a block of data already
-present in directly attached memory over the network.
-
-The core network operation in replication is to receive a request from the
-client at the leader replica and send that exact same request (as a log entry)
-to all follower replicas. The leader therefore \emph{multicasts} a received
-request packet to follower replicas, a problem well studied in the networking
-community when building software packet switches. Software packet switches reach
-impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, 
-a number far in excess of the few thousands of packets we manage in
-Figure~\ref{fig:problem}. The key insight in removing network overheads in
-Cyclone is therefore to approximate a software packet switch for the networking
-component of Cyclone. We do this by implementing Cyclone's network stack on top
-of the Data Plane Development Kit (DPDK)~\cite{dpdk}, that provides low latency
-userspace access to an ethernet NIC. We discuss Cyclone's network protocol in
-Section~\ref{sec:netprot}.
-
-However, the flow of packets through a software packet switch is usually very
-simple: the packet enters through an ethernet port and after a simple (usually
-stateless) decision is sent out through a set of chosen ports.  In contrast, log
-replication requires a complex protocol state machine to decide what to do with
-the packet. In addition, unlike software switches we cannot simply forget a
-packet after transmitting it. We must append it to the log and act on it after
-consensus is reached by a majority quorum. In Section~\ref{sec:dm} we cover the
-techniques we use in Cyclone to ameliorate these overheads.
+RAFT is only aware of and replicates the top level NVM log. Each
+replica \emph{independently} drains committed (in terms of RAFT
+consensus) entries from the NVM log to the flashlog. Block IO
+therefore is no longer a consideration when optimizing data movement
+over the network, unlike systems that need to deal with such
+problems~\cite{reflex}.
 
@@
 ethernet frames via the transmit and receive queues on the NIC. DPDK is often
-used by developers of software packet switches and therefore we leverage
-DPDK as a library for building Cyclone.
-
-DPDK does not by itself provide a TCP stack. But this is not a problem, since
+used by developers of software packet switches and is designed to allow the CPU 
+to make a switching decision by only examining the first few bytes of the packet 
+(containing the header).The NIC does all copying work minimizing the memory access 
+required from the CPU. This makes it an ideal candidate to build Cyclone.
+
+DPDK does not by itself provide a TCP stack. This is not a problem, since
 RAFT (and indeed most consensus protocols) tolerate network losses and
@@
 as possible, minimizing movement of data up and down the cache and memory
-hierarchy. Log replication looks like multicast that does not require deep
-packet inspection. Software packet switches implementing multicast therefore
+hierarchy. Software packet switches implementing multicast therefore
 simply manipulate packet headers to produce new packets to send on the output
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
-\caption{NVM log structure}
-\label{fig:nvm_log}
-\end{figure}
-
-The pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
+%\caption{NVM log structure}
+%\label{fig:nvm_log}
+%\end{figure}
+
+he pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
 handling code in Cyclone organized as event handlers triggered on receiving a
@@
 server). It then prepends RAFT related information to the packet - this includes
-the current term and log index for this entry. Next, it appends a pointer to the
-log in Figure~\ref{fig:nvm_log} effectively appending the packet to the
-persistent log. Finally, it transmits the packet to follower replicas.
+the current term and log index for this entry. Next, it appends a pointer to the packet to 
+the NVM log. Finally, it transmits the packet to follower replicas.
 
@@
 
-We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
-concurrent memory allocator based on reference counting - used by both the NIC
-driver and CPU cores. This means that packet allocation and deallocation does
-not happen in the same order as their corresponding position in the replicated
-RAFT log. To deal with this, we use a level of indirection as shown in
-Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
-sized pointers to the actual packets. Adding a level of indirection in the NVM
-log allows us to separate the FIFO ordered circular log being manipulated by
-RAFT from packet data being managed by the memory allocator of DPDK. Both the
-circular log and packet data are in NVM. An advantage of this scheme is that it
-makes recovery from NVM easy -- appends to the circular pointer log are atomic
-and we can use the pointer log to recover allocator state i.e. what pieces of
-NVM are currently in use by the log.
-
-RAFT requires that the log entry be persisted \emph{before} it is multicast out
-to follower replicas. DPDK userspace NIC drivers operate in DDIO
-mode~\cite{ddio} where the packet is directly written into the CPU cache rather
-than first being DMAed into DRAM and then fetched by the CPU on demand. We need
-to persist the packet to directly attached NVM by executing a cacheline flush
-({\tt clflush}) instruction for every cacheline in the packet and the pointer in
-the pointer buffer to persist these via the memory bus. This is not too onerous
-a burden because we can use the newly introduced {\tt
-  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
-efficiently flush to persistent memory without the overhead of the serialization
-normally introduced by {\tt clflush}. This allows us to hit full memory
-bandwidth on current generation platforms, a quantity in excess of 200 Gb/s per
-core which is well above the near term speeds of commodity NICs. We execute a
-single serializing {\tt sfence} before the sequence of {\tt clflushes} to make
-sure any dirty cachelines due to header manipulation related writes from the CPU
-are sent to cache.
-
-Although RAFT is an efficient consensus protocol in the common case, the
-protocol state machine still adds significant overhead to each packet, relative
-to the time for the packet to flow in from the NIC and back out to the replicas.
-We address this problem for the \emph{loaded} case using batching - treating a
-whole sequence of client commands as a single RAFT log entry, while avoiding any
-copies to group these packets together. Figure~\ref{fig:batching} illustrates
-how this is done. We use a burst receive call available in the DPDK userspace
-driver to receive a burst of client packets at a time. We then chain these
-packets together and treat them as a single log entry from the perspective of
-RAFT, amortizing the control plane overheads over the packets (at most 32 at a
-time due to current driver limitations). Crucially, batching in Cyclone does not
-involve a latency-throughput tradeoff like in many other
-systems~\cite{ix-dataplane}. The batch receive call we use in DPDK returns
-immediately with whatever number of packets is available, including zero. We
-always flush the transmit buffer after every call to DPDK to transmit packets to
-replicas. Therefore, we never tradeoff latency for throughput when batching. We
-also receive log entries in batches at follower replicas and return a single
-acknowledgment for the entire batch, speeding up the progress of the protocol.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.3]{figures2/batching.pdf}
-  \caption{Batching}
-  \label{fig:batching}
-\end{figure}
-
-Our choice of RAFT as a consensus protocol is driven by the need to efficiently
-maintain the log datastructure. Unlike alternatives such as
-MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
-leader is guaranteed to have the most up to date logs. There is therefore no
-case where a leader needs to receive log entries to ``fill'' holes in its logs
-from follower replicas, simplifying the protocol state machine we need to
-implement. A leader can immediately reject any responses that are not at least
-as current as its view (term). In turn, the persistent log operates as a double
-ended queue, with entries either being appended (or possibly deleted at
-followers) at the end and being deleted from the front (after commit). This is
-critical to efficient operation of Cyclone, as it allows a simple top level
-structure (a circular buffer of pointers) to represent the log.
+The NVM log is maintained as a circular log of fixed sized pointers to
+the actual packets. Adding a level of
+indirection in the NVM log allows us to separate the FIFO ordered
+circular log being manipulated by RAFT from packet data being managed
+by the memory allocator of DPDK. We use the pointer log to recover
+allocator state i.e. what pieces of NVM are currently in use by the
+log.
+
+%DPDK userspace NIC drivers operate in DDIO mode~\cite{ddio} where the
+%packet is directly written into the CPU cache rather than first being
+%DMAed into DRAM and then fetched by the CPU on demand. We need to
+%persist the packet to directly attached NVM by executing a cacheline
+%flush ({\tt clflush}) instruction for every cacheline in the packet
+%and the pointer in the pointer buffer to persist these via the memory
+%bus. This is not too onerous a burden because we can use the newly
+%introduced {\tt clflush-opt}~\cite{clflush_opt} instruction
+%specifically intended to efficiently flush to persistent memory
+%without the overhead of the serialization normally introduced by {\tt
+%  clflush}. This allows us to hit full memory bandwidth on current
+%generation platforms, a quantity in excess of 200 Gb/s per core which
+%is well above the near term speeds of commodity NICs. This departure
+%from a pure zero copy implementation of RAFT is necessitated by the
+%lack of current platform support for DMA to persistent memory.
+%Although RAFT is an efficient consensus protocol in the common case,
+%the protocol state machine still adds significant overhead to each
+%packet, relative to the time for the packet to flow in from the NIC
+%and back out to the replicas.
+
+We further reduce work per-packet for the \emph{loaded} case using
+batching - treating a whole sequence of client commands as a single
+RAFT log entry, while avoiding any copies to group these packets
+together. We do this by making use of a DPDK feature that allows
+chaining received packet mbufs together and treating the chain as a
+single packet.  Finally, we pipline the RAFT protocol state machine by
+contnuing to transmit log entries to follower replicas without waiting
+for acknowledgments. The RAFT protocol allows tracking multiple log
+entries in flight via their entry index in the log and this allows us
+to keep the link busy without being limited by the round trip time.
+At follower replicas we also receive packets containing log entries in
+batches and send back a single acknowledgment for the whole batch.
+
+Our choice of RAFT as a consensus protocol is driven by the need to
+efficiently maintain the log datastructure. Unlike alternatives such
+as MultiPaxos~\cite{multipaxos} or Quorum based
+replication~\cite{quorum}, the leader is guaranteed to have the most
+up to date logs. There is therefore no case where a leader needs to
+receive log entries to ``fill'' holes in its logs from follower
+replicas, simplifying the protocol state machine we need to implement
+and permitting further optimization.
 

### 2017-12-23 paper/paper.tex
Updates.

@@
 instances however exist in the same process address space as the key value store
-application itself and therefore manipulate the same application. The number of 
-instances is a fixed property of the Cyclone service and cannot be changed after 
-startup.
+application itself and therefore manipulate the same application.
 
-The first question we deal with is -- how do Cyclone clients decide which
-physical log to send a request to? The guarantees from Cyclone (except batched
-writes)  cover ordering of updates to a single key (Section~\ref{sec:sysarch}). They can 
-be satisfied by ensuring that all reads and writes to a key go to the same physical log
-(i.e. RAFT instance). We achieve this by hashing the key to select the physical
+The first question we deal with is -- how do Cyclone clients decide
+which physical log to send a request to? The guarantees from Cyclone
+(except batched writes) cover ordering of updates to a single key
+(Section~\ref{sec:sysarch}). They can be satisfied by ensuring that
+all reads and writes to a key go to the same physical log (i.e. RAFT
+instance). We achieve this by hashing the key to select the physical
 log.
 
-Multiple physical logs in Cyclone operate in a shared nothing manner by
-partitioning the NVM and SSD space evenly between them and by allocating
-dedicated NIC queue pairs to each instance of RAFT. This works well because the
-memory hierarchy including the NVM and the SSD efficiently support concurrent
-operations. The only synchronization necessary is when doing reads or writes to
-the single shared key value store.  The level of concurrency therefore is
-constrained only by the concurrency available in the software architecture of
-the persistent key value store itself, for which good designs
-exist~\cite{flodb}.
+Multiple physical logs in Cyclone operate in a shared nothing manner
+by partitioning the NVM and SSD space evenly between them and by
+allocating dedicated NIC queue pairs to each instance of RAFT.  The
+only synchronization necessary is when doing reads or writes to the
+single shared key value store - something already provided by most
+key-value stores including Rocksdb.  The level of concurrency
+therefore is constrained only by the concurrency available in the
+software architecture of the persistent key value store itself, for
+which good designs exist~\cite{flodb}.
 
 \subsection{Ganged Operations}
-Using multiple instances of RAFT leads to a serious problem with batched operations, 
-since we need to split up the batch into a mini-batch for each of the physical logs. There
-is however no guarantee that the request for a mini-batch will succeed breaking the atomicity
-requirement for batched updates. The obvious and simple solution is to do two phase commit across
-the logs, ensuring all or nothing semantics. However this was unacceptable to some customers who
-pointed out that ensuring that an external co-ordinator implementing two phase commit does not itself
-fail requires running a separate replicated service - with its attendant resource and reliability headaches.
-We were therefore challenged to come up with a solution that did not require two phase commit.
+Using multiple instances of RAFT leads to a serious problem with
+batched operations, since we need to split up the batch into a
+mini-batch for each of the physical logs. There is no guarantee that
+the request for a mini-batch will succeed breaking the atomicity
+requirement for batched updates. The obvious and simple solution is to
+do two phase commit across the logs, ensuring all or nothing
+semantics. However this was unacceptable to some use-cases of Cyclone
+as it requires ensuring that an external co-ordinator implementing two
+phase commit does not itself fail. This requires running a separate
+replicated service - with its attendant resource and reliability
+headaches - too high a price to pay for the performance benefits of
+Cyclone.  We therefore had to come up with a solution that did not
+require two phase commit.
 
-Our final solution is based on two observations. First, we \emph {do not} require that either all mini-batches
-icommit to their raft logs or none do, What we actually requires is that either all are \emph{applied} to Rocksdb or 
-none of them are. Put another way we should not apply a mini-batch from a ganged operation if any of the other mini-batches
-have failed to commit to their RAFT log. This means that the client can be stateless. The second observation is that
-checking whether the other minibatches have committed is easy because the different RAFT instances can communicate 
-through shared memory on the same machines unlike the more general case of distributed participants in two phase 
-commit.
+Our solution is based on two observations. First, we \emph {do not}
+require that either all mini-batches icommit to their raft logs or
+none do. What we actually requires is that either all are
+\emph{applied} to Rocksdb or none of them are. Put another way we
+should not apply a mini-batch from a ganged operation if any of the
+other mini-batches have failed to commit to their RAFT log. This means
+that the client can be stateless. The second observation is that
+checking whether the other minibatches have committed is easy because
+the different RAFT instances can communicate through shared memory on
+the same machines unlike the more general case of distributed
+participants in two phase commit.
 
@@
 converge to the same machine. We do this by triggering re-elections for a RAFT
-instance other than that of the first physical log until it is on the same
+instance \emph{other} than that of the first physical log until it is on the same
 machine as the \emph{first} physical log. This process is therefore resilient to
@@
 to avoid making physical copies of the packet, generating indirect references
-instead. The co-ordinator also adds a ``nonce'' - a unique timestamp to the
-packet. In addition, the co-ordinator adds a unique view number to the packet
+instead. The co-ordinator adds a ``nonce'' - a unique timestamp - to the
+packet. In addition, the co-ordinator adds a view number to the packet
 containing the term numbers of all the participating RAFT instances (read from
@@
 
-The complexity in Figure~\ref{fig:ganged_ops} arises from the need to handle
-both shared memory concurrency and failure during replication. Without failure,
+The complexity in Figure~\ref{fig:ganged_ops} arises from the need to implement
+a shared memory barrier and handle failure during replication. Without failure,
 Figure~\ref{fig:ganged_ops} is straightforward. A barrier is executed on all
@@
 
-The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
-mapped to a unique barrier. We achieve this by using a fixed piece of memory
-owned by the co-ordinator to hold the barrier and write the nonce to it in order
-to indicate that the barrier is active for the corresponding ganged
-operation. Participants watch for the nonce to know when to execute the
-ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
-the leader's published view to detect the case where the ganged operation fails
+The assumption in Figure~\ref{fig:ganged_ops} is that each ganged
+operation is mapped to a unique barrier. We achieve this by using a
+fixed piece of memory owned by the co-ordinator to hold the barrier
+and write the nonce to it in order to indicate that the barrier is
+active for the corresponding ganged operation. Participants watch for
+the nonce to know when to execute the ganged operation barrier in
+Figure~\ref{fig:ganged_ops}, while also monitoring the co-ordinators
+'s published view to detect the case where the ganged operation fails
 to replicate on the co-ordinator's physical log.
 
-Finally, we describe how we generate the nonce. The nonce is generated on the
-co-ordinator RAFT instance by concatenating the ethernet MAC ID of the first NIC
-on the system with a 64 bit value that is the number of CPU timestamp counter
-cycles since epoch time (read from the real time clock at startup plus the
-number of cycles from the CPU {\tt rdtsc} instruction). The nonce can only be
-repeated if the same machine manages to fail and come back up in less time than
-the real time clock drift (controlled with NTP), a possibility that we discount.
+Finally, we describe how we generate the nonce. The nonce is generated
+on the co-ordinator RAFT instance by concatenating the ethernet MAC ID
+of the first NIC on the system with a 64 bit value that is the number
+of CPU timestamp counter cycles since epoch time (read from the real
+time clock at startup plus the number of cycles from the CPU {\tt
+  rdtsc} instruction). The nonce can only be repeated if the same
+machine manages to fail and come back up in less time than the real
+time clock drift (controlled with NTP), a possibility that we
+discount.
 
-Ganged operations possibly constitute the most complex part of Cyclone but the
-code weighs in at well under a couple of hundred lines. We believe that
-this additional complexity is still small compared to distributed transactions
-using two-phase commit - especially when taking into account failure recovery -
-as used in systems such as FARM~\cite{farm}.
+Ganged operations possibly constitute the most complex part of Cyclone
+but the code weighs in at well under a couple of hundred lines. We
+believe that this additional complexity is still small compared to
+distributed transactions using two-phase commit - especially when
+taking into account failure recovery - as used in systems such as
+FARM~\cite{farm}.
 

### 2017-12-23 paper/paper.tex
Updates.

@@
 \author{{\rm Amitabha Roy} \\ amitabha.roy@gmail.com
-	     \and {\rm Subramanya R. Dulloor} \\ dulloor@gmail.com}
+	     \and {\rm Subramanya R. Dulloor} \\ subramanya.r.dulloor@intel.com}
 
@@
 
-DPDK does not by itself provide a TCP stack. This is not a problem, since
-RAFT (and indeed most consensus protocols) tolerate network losses and
-reordering by design due to the need to support asynchronous communication. In
-addition, most datacenter networks rarely drop or reorder packets and provide
-full bisection bandwidth between servers that might serve as Cyclone replicas. We
-therefore jettisoned TCP and chose to send raw IP packets encapsulated in
-ethernet frames. Cyclone takes complete control of an ethernet interface,
-receiving all packets directed to it, while the IP and ethernet addresses
-provide sufficient information to route the packet if necessary through multiple
-switches. We currently follow this communication model both for server to server
-communication between replicas as well as client to server communication for
-requests and responses. Although a detailed evaluation is made later, switching
-from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
-testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
-to performance.
+DPDK does not by itself provide a TCP stack. This is not a problem,
+since RAFT (and indeed most consensus protocols) tolerate network
+losses and reordering by design due to the need to support
+asynchronous communication. In addition, most datacenter networks
+rarely drop or reorder packets and provide full bisection bandwidth
+between servers that might serve as Cyclone replicas, obviating the
+need for congestion control. We therefore jettisoned TCP and chose to
+send raw IP packets encapsulated in ethernet frames. Cyclone takes
+complete control of an ethernet interface, receiving all packets
+directed to it, while the IP and ethernet addresses provide sufficient
+information to route the packet if necessary through multiple
+switches. We currently follow this communication model both for server
+to server communication between replicas as well as client to server
+communication for requests and responses. Although a detailed
+evaluation is made later, switching from the kernel TCP/IP stack to
+DPDK reduces the latency between machines in our testbed from {\tt 18
+  us} to as low as {\tt 5 us}, providing a significant boost to
+performance.
 
@@
 
-he pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
-handling code in Cyclone organized as event handlers triggered on receiving a
-packet at the leader. We focus on only one key event for brevity: the event
-where a request is received from a client. The first step is to check that this
-replica is indeed the RAFT leader in the current view (term). If not, the
-message is simply dropped (a timeout causes the client to try a different
-server). It then prepends RAFT related information to the packet - this includes
-the current term and log index for this entry. Next, it appends a pointer to the packet to 
-the NVM log. Finally, it transmits the packet to follower replicas.
-
-This entire process is done without making any copies of the received packet.
-Figure~\ref{fig:packet_layout} illustrates how
-Cyclone manipulates packet layouts across the two steps of prepending a RAFT
-header and transmitting to follower replicas. DPDK describes packets using an
-``mbuf'' data structure. Roughly speaking, an ``mbuf'' consists of a flat array
-of bytes actually containing the packet and a fixed size piece of external
-metadata that describes various aspects of the packet, most crucially a pointer
-to the start and end of the packet in the byte array. DPDK's userspace drivers
-receive packets from the NIC such that they are offset in the byte array by a
-configurable amount referred to as ``headroom''. We strip off the existing
-network headers in the packet and prepend RAFT related information specific to
-each log entry in the headroom by shifting the start pointer appropriately.
-These operations are standard enough for software packet switches that DPDK
-provides convenient library calls for it. For the final step, we need to prepare
-the packet for transmission to the various follower replicas. To do this we
-prepare a different packet containing a network header for \emph{each}
-targeted replica and ``chain'' the data packet to each of these headers. Each
-header is then separately handed off to the driver for transmission via the NIC,
-carrying the data packet with it by association.
+The pseudocode in Figure~\ref{fig:control_plane} describes part of the
+packet handling code in Cyclone organized as event handlers triggered
+on receiving a packet at the leader. We focus on only one key event
+for brevity: the event where a request is received from a client. The
+first step is to check that this replica is indeed the RAFT leader in
+the current view (term). If not, the message is simply dropped (a
+timeout causes the client to try a different server). It then prepends
+RAFT related information to the packet - this includes the current
+term and log index for this entry. Next, it appends a pointer to the
+packet to the NVM log. Finally, it transmits the packet to follower
+replicas.
+
+This entire process is done without making any copies of the received
+packet.  Figure~\ref{fig:packet_layout} illustrates how Cyclone
+manipulates packet layouts across the two steps of prepending a RAFT
+header and transmitting to follower replicas. DPDK describes packets
+using an ``mbuf'' data structure. Roughly speaking, an ``mbuf''
+consists of a flat array of bytes actually containing the packet and a
+fixed size piece of external metadata that describes various aspects
+of the packet, most crucially a pointer to the start and end of the
+packet in the byte array. DPDK's userspace drivers receive packets
+from the NIC such that they are offset in the byte array by a
+configurable amount referred to as ``headroom''. We strip off the
+existing network headers in the packet and prepend RAFT related
+information specific to each log entry in the headroom by shifting the
+start pointer appropriately.  These operations are standard enough for
+software packet switches that DPDK provides convenient library calls
+for it. For the final step, we need to prepare the packet for
+transmission to the various follower replicas. To do this we prepare a
+different packet containing a network header for \emph{each} targeted
+replica and ``chain'' the data packet to each of these headers. Each
+header is then separately handed off to the driver for transmission
+via the NIC, carrying the data packet with it by association.
 
 The NVM log is maintained as a circular log of fixed sized pointers to
-the actual packets. Adding a level of
-indirection in the NVM log allows us to separate the FIFO ordered
-circular log being manipulated by RAFT from packet data being managed
-by the memory allocator of DPDK. We use the pointer log to recover
-allocator state i.e. what pieces of NVM are currently in use by the
-log.
+the actual packets. Adding a level of indirection in the NVM log
+allows us to separate the FIFO ordered circular log being manipulated
+by RAFT from packet data being managed by the memory allocator of
+DPDK. We use the pointer log to recover allocator state i.e. what
+pieces of NVM are currently in use by the log.
 
@@
 
-Although the work described in the previous section significantly boosts the
-network performance of replication in Cyclone, it still does not come close to
-saturating the capabilities of even the commodity NICs that we use. This is
-because NICs today encapsulate significant parallelism in terms of
-multiple send and receive queue pairs. Exploiting this capability to improve
-throughput requires us to remove the bottleneck of a single sequential log in
-RAFT. To this end, we extend the implementation described thus far to run
-multiple copies of the RAFT consensus protocol each maintaining and replicating
-its own two level log. We refer to these logs as physical logs. All these
-instances however exist in the same process address space as the key value store
-application itself and therefore manipulate the same application.
+Although the work described in the previous section significantly
+boosts the network performance of replication in Cyclone, it still
+does not come close to saturating the capabilities of even the
+commodity NICs that we use. This is because NICs today encapsulate
+significant parallelism in terms of multiple send and receive queue
+pairs. Exploiting this capability to improve throughput requires us to
+remove the bottleneck of a single sequential log in RAFT. To this end,
+we extend the implementation described thus far to run multiple copies
+of the RAFT consensus protocol each maintaining and replicating its
+own two level log. We refer to these logs as physical logs. All these
+instances however exist in the same process address space as the key
+value store application itself and therefore manipulate the same
+application.
 
@@
 Our solution is based on two observations. First, we \emph {do not}
-require that either all mini-batches icommit to their raft logs or
-none do. What we actually requires is that either all are
-\emph{applied} to Rocksdb or none of them are. Put another way we
-should not apply a mini-batch from a ganged operation if any of the
-other mini-batches have failed to commit to their RAFT log. This means
-that the client can be stateless. The second observation is that
-checking whether the other minibatches have committed is easy because
-the different RAFT instances can communicate through shared memory on
-the same machines unlike the more general case of distributed
-participants in two phase commit.
+require that either all mini-batches commit to their raft logs or none
+do. What we actually require is that either all are \emph{applied} to
+Rocksdb or none of them are. Put another way we should not apply a
+mini-batch from a ganged operation if any of the other mini-batches
+have failed to commit to their RAFT log. This means that the client
+can be stateless, with all the decision making done at the server
+side. The second observation is that checking whether the other
+minibatches have committed is easy because the different RAFT
+instances can communicate through shared memory on the same machines
+unlike the more general case of distributed participants in two phase
+commit.
 
 We now describe our solution called ``ganged operations" in Cyclone.
-The first key problem we need to deal with is that the leaders for the different
-RAFT instances to which the keys map can be located on different machines. To
-avoid having a client co-ordinate different machines for a single operation, we
-constrain the RAFT leader election algorithm for the different RAFT instances to
-converge to the same machine. We do this by triggering re-elections for a RAFT
-instance \emph{other} than that of the first physical log until it is on the same
-machine as the \emph{first} physical log. This process is therefore resilient to
-failed machines. RAFT requires that the leader have the most up to date log. To
-ensure that the process converges we therefore ensure that the currently elected
-leader brings a majority quorum up to date before triggering a reelection. We
-also do not accept any requests from clients until all leaders are
-co-located. We assume that no network partitions occur \emph{within a machine}
-i.e. some RAFT instances are able to send and receive packets, while others are
-not, a possibility we discount due to our single process design.
+The first key problem we need to deal with is that the leaders for the
+different RAFT instances to which the keys map can be located on
+different machines. To avoid having a client co-ordinate different
+machines for a single operation, we constrain the RAFT leader election
+algorithm for the different RAFT instances to converge to the same
+machine. We do this by triggering re-elections for a RAFT instance
+\emph{other} than that of the first physical log until it is on the
+same machine as the \emph{first} physical log. This process is
+therefore resilient to failed machines. RAFT requires that the leader
+have the most up to date log. To ensure that the process converges we
+therefore ensure that the currently elected leader brings a majority
+quorum up to date before triggering a reelection. We also do not
+accept any requests from clients until all leaders are co-located. We
+assume that no network partitions occur \emph{within a machine}
+i.e. some RAFT instances are able to send and receive packets, while
+others are not, a possibility we discount due to our single process
+design.
 
 Next, we need to simultaneously inject the ganged operation into all
-participating physical logs on the machine hosting the leaders. Clients always
-dispatch batched writes to a fixed co-ordinating physical log/RAFT instance
-(henceforth called the co-ordinator), which is then responsible for forwarding
-the request to the participating logs. We make use of packet cloning primitives
-to avoid making physical copies of the packet, generating indirect references
-instead. The co-ordinator adds a ``nonce'' - a unique timestamp - to the
-packet. In addition, the co-ordinator adds a view number to the packet
-containing the term numbers of all the participating RAFT instances (read from
-shared memory of the co-located leaders).  The event that applies a ganged
-operation is described in the pseudocode of Figure~\ref{fig:ganged_ops}. We
-defer the discussion of how we generate the nonce to later in this section. We
-also assume that a unique barrier is allocated in shared memory for each ganged
-operation. We discuss later how this is done without using dynamic allocation.
-
-The complexity in Figure~\ref{fig:ganged_ops} arises from the need to implement
-a shared memory barrier and handle failure during replication. Without failure,
-Figure~\ref{fig:ganged_ops} is straightforward. A barrier is executed on all
-RAFT instances. Once replication is complete on all participating physical logs,
-a state indicated by all necessary bits being set in the barrier mask, the RAFT
-instances simultaneously execute the ganged operation.
-
-A failure causes a new leader to be elected for the affected physical log,
-moving forward the term. If the participants of a ganged operation do not
-detect this case, they could be left waiting forever. We detect failure on any
-RAFT instance by having each instance publish its current term and continuously
-comparing the view in the ganged operation to it. If any RAFT instance has moved
-past that term, the ganged operation is then terminated.
+participating physical logs on the machine hosting the
+leaders. Clients always dispatch batched writes to a fixed
+co-ordinating physical log/RAFT instance (henceforth called the
+co-ordinator), which is then responsible for forwarding the request to
+the participating logs. We make use of packet cloning primitives to
+avoid making physical copies of the packet, generating indirect
+references instead. The co-ordinator adds a ``nonce'' - a unique
+timestamp - to the packet. In addition, the co-ordinator adds a view
+number to the packet containing the term numbers of all the
+participating RAFT instances (read from shared memory of the
+co-located leaders).  The event that applies a ganged operation is
+described in the pseudocode of Figure~\ref{fig:ganged_ops}. We defer
+the discussion of how we generate the nonce to later in this
+section. We also assume that a unique barrier is allocated in shared
+memory for each ganged operation. We discuss later how this is done
+without using dynamic allocation.
+
+The complexity in Figure~\ref{fig:ganged_ops} arises from the need to
+implement a shared memory barrier and handle failure during
+replication. Without failure, Figure~\ref{fig:ganged_ops} is
+straightforward. A barrier is executed on all RAFT instances. Once
+replication is complete on all participating physical logs, a state
+indicated by all necessary bits being set in the barrier mask, the
+RAFT instances simultaneously execute the ganged operation.
+
+A failure causes a new leader to be elected for the affected physical
+log, moving forward the term. If the participants of a ganged
+operation do not detect this case, they could be left waiting
+forever. We detect failure on any RAFT instance by having each
+instance publish its current term and continuously comparing the view
+in the ganged operation to it. If any RAFT instance has moved past
+that term, the ganged operation is then terminated.
 
@@
 \label{sec:evaluation}
-We evaluate Cyclone on a 12 node x86 Xeon cluster connected via a 10 GigE
-switch. Three of the machines are equipped with 1.6TB Intel DC P3600 SSDs and
-4*10 GigE ports. The remaining nine machines do not have SSDs and have only one
-10 GigE port, serving as clients for most of the experiments. We turn on jumbo
-frame support in the 10 GigE switch to enable maximum use of batching in
-Cyclone. As with other work~\cite{faast}, we use DRAM on the machines to proxy
-for NVDIMMs where necessary - the NVM needed never exceeds 64 MB
-regardless of the size of the key value store or second level log on flash. We
-divide the evaluation into three parts.
-
-First, we evaluate Cyclone's performance with a single level log as a pure
-software packet switch. For this purpose, Cyclone uses a dummy server stub that
-simply echoes the client request back to it. Next, we evaluate
-performance with the dummy stub, but with the addition of the second level of log
-on flash. Finally, we evaluate performance when integrated with
-Rocksdb~\cite{rocksdb} as an alternative to Rocksdb's write ahead log.
-
-Unless otherwise mentioned, we use a 60 byte header followed by an optional
-payload for experiments. We log both the header and payload. Also, unless
-otherwise mentioned we use 8 physical logs (and associated RAFT instances) each
-mapped to a dedicated core. The remaining cores are dedicated to the stub
-server or Rocksdb, as the case may be.
-
-We begin by systematically evaluating network stack related optimizations
-applied in Cyclone to replicate the NVM log in Figure~\ref{fig:network_opts} -
-with no payload. The y-axis reports latency seen at the client (which means two
-network round trips with replication). Using TCP/IP to replicate a RAFT log tops
-out at around 30K entries/s.  Switching to DPDK (the line marked +DPDK) improves
-the throughput by an order of magnitude to around 500K entries/s. Using batching
-(the line marked +batching) improves the performance further bringing us close
-to a million entries/s. Scaling to 8 physical logs (+8 phy logs) improves
-performance to close to 2M entries/s. Finally using all 4 ethernet ports on the
-machine to replicate entries improves performance considerably to 6M
-entries/s. In all, performance improves by 200X over the TCP/IP single log
-baseline. Cyclone also considerably improves the latency for replication, from
-close to 100us with TCP/IP to around 30us at peak throughput.
+We evaluate Cyclone on a 12 node x86 Xeon cluster connected via a 10
+GigE switch. Three of the machines are equipped with 1.6TB Intel DC
+P3600 SSDs and 4*10 GigE ports. The remaining nine machines do not
+have SSDs and have only one 10 GigE port, serving as clients for most
+of the experiments. We turn on jumbo frame support in the 10 GigE
+switch to enable maximum use of batching in Cyclone. As with other
+work~\cite{faast}, we use DRAM on the machines to proxy for NVDIMMs
+where necessary - the NVM needed never exceeds 64 MB regardless of the
+size of the key value store or second level log on flash. We divide
+the evaluation into three parts.
+
+First, we evaluate Cyclone's performance with a single level log as a
+pure software packet switch. For this purpose, Cyclone uses a dummy
+server stub that simply echoes the client request back to it. Next, we
+evaluate performance with the dummy stub, but with the addition of the
+second level of log on flash. Finally, we evaluate performance when
+integrated with Rocksdb~\cite{rocksdb} as an alternative to Rocksdb's
+write ahead log.
+
+Unless otherwise mentioned, we use a 60 byte header followed by an
+optional payload for experiments. We log both the header and
+payload. Also, unless otherwise mentioned we use 8 physical logs (and
+associated RAFT instances) each mapped to a dedicated core. The
+remaining cores are dedicated to the stub server or Rocksdb, as the
+case may be.
+
+We begin by systematically evaluating network stack related
+optimizations applied in Cyclone to replicate the NVM log in
+Figure~\ref{fig:network_opts} - with no payload. The y-axis reports
+latency seen at the client (which means two network round trips with
+replication). Using TCP/IP to replicate a RAFT log tops out at around
+30K entries/s.  Switching to DPDK (the line marked +DPDK) improves the
+throughput by an order of magnitude to around 500K entries/s. Using
+batching (the line marked +batching) improves the performance further
+bringing us close to a million entries/s. Scaling to 8 physical logs
+(+8 phy logs) improves performance to close to 2M entries/s. Finally
+using all 4 ethernet ports on the machine to replicate entries
+improves performance considerably to 6M entries/s. In all, performance
+improves by 200X over the TCP/IP single log baseline. Cyclone also
+considerably improves the latency for replication, from close to 100us
+with TCP/IP to around 30us at peak throughput.
 
 One can draw three important conclusions from
-Figure~\ref{fig:network_opts}. First, one can indeed treat log replication as a
-software packet switching problem \emph{provided} the persistent log can be held
-in directly attached memory eliminating the block interface from the critical
-path. Second, relieving the CPU from the overhead of data copies has a
-significant positive effect on performance. Second, using multiple physical
-logs is essential to exploiting the concurrency available at the level of the
-NIC (even a single one!) that would otherwise go wasted due
-to the serializing abstraction of a single log. Finally, we note that Cyclone
-achieves about 50\% of the line rate across the four ethernet
-ports. This is essentially the cost of running a consensus protocol in
-software.
+Figure~\ref{fig:network_opts}. First, one can indeed treat log
+replication as a software packet switching problem \emph{provided} the
+persistent log can be held in directly attached memory eliminating the
+block interface from the critical path. Second, relieving the CPU from
+the overhead of data copies has a significant positive effect on
+performance. Second, using multiple physical logs is essential to
+exploiting the concurrency available at the level of the NIC (even a
+single one!) that would otherwise go wasted due to the serializing
+abstraction of a single log. Finally, we note that Cyclone achieves
+about 50\% of the line rate across the four ethernet ports. This is
+essentially the cost of running a consensus protocol in software.
 
@@
 
-Both the replica count and payload size can have significant impact on Cyclone's
-network performance. First, the number of replicas dictates the outgoing message
-rate from the leader replica and therefore increasing the replication factor can
-decrease Cyclone's performance. Figure~\ref{fig:replicas} shows the impact of
-varying replica count. Using only a single replica cuts out a network round trip
-and shows the best unloaded latency (10 us) and peak throughput (near 10M
-entries/s). Adding replicas decreases the peak throughput down to around 2M
-entries/s with 5 replicas. We note that a number of previous pieces of
-work~\cite{faast, farm} use three replicas and therefore we focus on three
-replicas for the replicated cases we consider below. The second factor that
-dictates Cyclone's performance is the size of the log entry being
-replicated. Figure~\ref{fig:payload} shows the effect of increasing the payload
-size from zero to 512 bytes. Peak throughput drops from 6M entries/s to
-approximately 2M entries/s. At this replication rate, the leader replica needs
-to transmit data at approximately 30 Gbit/s. Coupled with the cost of network
-headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
-network line rate bottleneck at this point. It is worthwhile to compare this to
-the case with small updates, where running the consensus protocol was a
+Both the replica count and payload size can have significant impact on
+Cyclone's network performance. First, the number of replicas dictates
+the outgoing message rate from the leader replica and therefore
+increasing the replication factor can decrease Cyclone's
+performance. Figure~\ref{fig:replicas} shows the impact of varying
+replica count. Using only a single replica cuts out a network round
+trip and shows the best unloaded latency (10 us) and peak throughput
+(near 10M entries/s). Adding replicas decreases the peak throughput
+down to around 2M entries/s with 5 replicas. We note that a number of
+previous pieces of work~\cite{faast, farm} use three replicas and
+therefore we focus on three replicas for the replicated cases we
+consider below. The second factor that dictates Cyclone's performance
+is the size of the log entry being
+replicated. Figure~\ref{fig:payload} shows the effect of increasing
+the payload size from zero to 512 bytes. Peak throughput drops from 6M
+entries/s to approximately 2M entries/s. At this replication rate, the
+leader replica needs to transmit data at approximately 30
+Gbit/s. Coupled with the cost of network headers all four 10 GigE
+ports are now saturated and therefore Cyclone hits the network line
+rate bottleneck at this point. It is worthwhile to compare this to the
+case with small updates, where running the consensus protocol was a
 bottleneck to reaching line rate.
@@
 
-We now turn our attention from the network component of Cyclone to the storage
-one by adding the second level flashlog. We evaluate the impact of adding the
-flashlog in Figures~\ref{fig:flashlog} and \ref{fig:flashlog_512}. The benefit
-of the second level flashlog is that it lets us keep the same amount of log
-space as with the flash based implementation, and not limit it to the amount of
-NVM available in the system.
-The hypothesis was that batching the movement of data from the
-NVM log to the flashlog would be sufficient to hide the latency of the block IO
-device from network speed replication. The results confirm that in terms of
-latency for the low to moderate load cases, our two level log arrangement is
-effective at hiding the latency of the NVMe SSD. The picture however is
-different for peak throughput. For small updates there is no impact on
-throughput when we add the SSD. On the other hand 
-Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
-significant impact on peak throughput - it drops to approximately 350K
-ops/sec. This corresponds to around 50K 4KB IOPS to the SSD to write out the
-flashlog pages. This is in fact the expected IOPS limit with the 32 outstanding
-requests to the drive that we maintain (Section~\ref{sec:storage}).
-A final observation on Figure~\ref{fig:flashlog_512} is that once we are past
-the storage bottleneck the latency spike is dramatic and large enough to trigger
-Cyclone's failure detector and repeated retries from the clients. There are
-therefore no points on the ``knee'' of the curve as in the pure packet
-switched one-level log case.
-
-One can therefore conclude that using a small amount of directly attached NVM is
-effective at hiding the latency of a background block storage device from the
-critical path of replication. Further, for small updates the
-bottleneck is the capability of the software to run the consensus protocol, but
-the bottleneck shifts to the secondary storage device for larger request sizes.
-
-We now evaluate ganged operations. The primary purpose of
-ganged operations is to avoid the need for distributed transactions to
-manipulate what is a single shared memory image and therefore we were most
-concerned about unloaded latency given the complexity of synchronizing different
-replication quorums as well executing our rendezvous protocol on multiple
-cores. We therefore setup an experiment where a single client -- reflecting the
-unloaded case -- made ganged requests to the replicas. We varied the number of
-physical logs participating in the request from one to the full complement of 8.
-Figure~\ref{fig:ganged} shows the results. Unloaded
-latency increases slowly as we increase the number of active physical logs - to
-around 32 us from the baseline of 21 us. The reason for this increase is that
-the DPDK userspace NIC driver pipelines request processing for communication
-between the CPU core and the NIC. Simultaneous replication on multiple physical
-logs represents a worst case for this arrangement as all eight cores try to
-send replication messages to their replicas, adding about a microsecond of
-serial latency each, to dispatch requests to the NIC. Regardless, the experiment
-underlines the value of our ganged operation design over a distributed
-transaction that would need 8 network hops for the two phase commit (and
-replication) using an external co-ordinator, a minimum of 80us.
+We now turn our attention from the network component of Cyclone to the
+storage one by adding the second level flashlog. We evaluate the
+impact of adding the flashlog in Figures~\ref{fig:flashlog} and
+\ref{fig:flashlog_512}. The benefit of the second level flashlog is
+that it lets us keep the same amount of log space as with the flash
+based implementation, and not limit it to the amount of NVM available
+in the system.  The hypothesis was that batching the movement of data
+from the NVM log to the flashlog would be sufficient to hide the
+latency of the block IO device from network speed replication. The
+results confirm that in terms of latency for the low to moderate load
+cases, our two level log arrangement is effective at hiding the
+latency of the NVMe SSD. The picture however is different for peak
+throughput. For small updates there is no impact on throughput when we
+add the SSD. On the other hand Figure~\ref{fig:flashlog_512} shows
+that using a 512 byte payload has a significant impact on peak
+throughput - it drops to approximately 350K ops/sec. This corresponds
+to around 50K 4KB IOPS to the SSD to write out the flashlog
+pages. This is in fact the expected IOPS limit with the 32 outstanding
+requests to the drive that we maintain (Section~\ref{sec:storage}).  A
+final observation on Figure~\ref{fig:flashlog_512} is that once we are
+past the storage bottleneck the latency spike is dramatic and large
+enough to trigger Cyclone's failure detector and repeated retries from
+the clients. There are therefore no points on the ``knee'' of the
+curve as in the pure packet switched one-level log case.
+
+One can therefore conclude that using a small amount of directly
+attached NVM is effective at hiding the latency of a background block
+storage device from the critical path of replication. Further, for
+small updates the bottleneck is the capability of the software to run
+the consensus protocol, but the bottleneck shifts to the secondary
+storage device for larger request sizes.
+
+We now evaluate ganged operations. The primary purpose of ganged
+operations is to avoid the need for distributed transactions to
+manipulate what is a single shared memory image and therefore we were
+most concerned about unloaded latency given the complexity of
+synchronizing different replication quorums as well executing our
+rendezvous protocol on multiple cores. We therefore setup an
+experiment where a single client -- reflecting the unloaded case --
+made ganged requests to the replicas. We varied the number of physical
+logs participating in the request from one to the full complement of
+8.  Figure~\ref{fig:ganged} shows the results. Unloaded latency
+increases slowly as we increase the number of active physical logs -
+to around 32 us from the baseline of 21 us. The reason for this
+increase is that the DPDK userspace NIC driver pipelines request
+processing for communication between the CPU core and the
+NIC. Simultaneous replication on multiple physical logs represents a
+worst case for this arrangement as all eight cores try to send
+replication messages to their replicas, adding about a microsecond of
+serial latency each, to dispatch requests to the NIC. Regardless, the
+experiment underlines the value of our ganged operation design over a
+distributed transaction that would need 8 network hops for the two
+phase commit (and replication) using an external co-ordinator, a
+minimum of 80us.
 
@@
   \includegraphics[scale=0.45]{results2/flashlog_512.pdf}
-  \caption{Second level log - 512 bytes payload}
+  \caption{Two level: 512B payload}
   \label{fig:flashlog_512}
@@
 
-Next, we evaluate Cyclone integrated with the Rocksdb persistent key value
-store. Rocksdb is a complex commercial grade persistent key value store and
-an important target for practical use of Cyclone.  Rocksdb is accompanied by a
-complex array of performance tuning knobs to get the best performance from
-flash. Designing LSM tree based key value stores to extract good performance
-from flash and make best use of DRAM on the machine is an area of active
-research~\cite{flodb, triad}. One of our main goals is to demonstrate that
-our conclusions will apply even as Rocksdb performance continues to improve
-with the integration of new ideas and better SSDs. Keeping this in mind, we
-configured RocksDB to place all files for the
-key value store (SSTables) on a RAMdisk, which presumably represents the limit
-in performance for both software enhancements as well as improvements to
+Next, we evaluate Cyclone integrated with the Rocksdb persistent key
+value store. Rocksdb is a complex commercial grade persistent key
+value store and an important target for practical use of Cyclone.
+Rocksdb is accompanied by a complex array of performance tuning knobs
+to get the best performance from flash. Designing LSM tree based key
+value stores to extract good performance from flash and make best use
+of DRAM on the machine is an area of active research~\cite{flodb,
+  triad}. One of our main goals is to demonstrate that our conclusions
+will apply even as Rocksdb performance continues to improve with the
+integration of new ideas and better SSDs. Keeping this in mind, we
+configured RocksDB to place all files for the key value store
+(SSTables) on a RAMdisk, which presumably represents the limit in
+performance for both software enhancements as well as improvements to
 flash-based SSDs. However, both Rocksdb's own write ahead log and the
-alternative of Cyclone's second level flashlog are placed on the SSDs. In
-effect we make availability a harder problem in this setting to ensure that our
-design is future-proof.
-
-For the first experiment with Rocksdb we use 8 byte keys with either
-8 byte values or 256 byte values.  Since our
-focus is replication and only update requests are replicated, we use a 100\%
-update workload to test the capability of the system. This involves loading 100
-million key value pairs and during the test updating the value associated with a
-key picked at random.
+alternative of Cyclone's second level flashlog are placed on the
+SSDs. In effect we make availability a harder problem in this setting
+to ensure that our design is future-proof.
+
+For the first experiment with Rocksdb we use 8 byte keys with either 8
+byte values or 256 byte values.  Since our focus is replication and
+only update requests are replicated, we use a 100\% update workload to
+test the capability of the system. This involves loading 100 million
+key value pairs and during the test updating the value associated with
+a key picked at random.
 
 Before evaluating with Rocksdb, we measure the baseline performance of
-replicating the log with Cyclone for the given request sizes - Rocksdb performs
-a no-op. We note that in addition to the key and value, we are also logging
-RocksDB specific request data such as operation type and the request header,
-making this different from the previous experiments.
-Figure~\ref{fig:kv_baseline} shows the baseline performance for the chosen
-request sizes. With the smaller request size, Cyclone can conservatively sustain
-close to a million requests a second at a latency of just under 25us. With the
-larger request size, Cyclone can sustain around 350K requests a second, again at
-a latency of just under 25 us. Armed with these baseline numbers we now examine
+replicating the log with Cyclone for the given request sizes - Rocksdb
+performs a no-op. We note that in addition to the key and value, we
+are also logging RocksDB specific request data such as operation type
+and the request header, making this different from the previous
+experiments.  Figure~\ref{fig:kv_baseline} shows the baseline
+performance for the chosen request sizes. With the smaller request
+size, Cyclone can conservatively sustain close to a million requests a
+second at a latency of just under 25us. With the larger request size,
+Cyclone can sustain around 350K requests a second, again at a latency
+of just under 25 us. Armed with these baseline numbers we now examine
 how well Cyclone performs with Rocksdb.
 
-The performance of Rocksdb with Cyclone for the small update workload is shown
-in Figure~\ref{fig:rocksdb} - essentially presenting the solution to the problem
-demonstrated in Figure~\ref{fig:problem}.  We consider four different
-setups. The line labeled 'rocksdb' is the key value store running with no
-logging whatsoever - a system crash would lead to data loss. The line labeled
-'rocksdb/wal' is for Rocksdb running with its write ahead logging turned on. The
-large gap between these two is the overhead of the existing Rocksdb WAL
-solution. The line labeled 'rocksdb/Cyclone 1 way' is a two level Cyclone log
-but without any replication. The line almost exactly tracks the performance of
-Rocksdb. Cyclone is able to provide a write ahead log with no overhead to
-Rocksdb. The line labeled 'rocksdb/Cyclone 3 way' is with 3-way replication
-turned on. Other than a 20us delta due to the extra network round trip, the line
-almost exactly tracks Rocksdb performance with no logging. Cyclone therefore
-provides high availability to Rocksdb at a fraction of the cost of its existing
-single machine write ahead log. We also repeat the experiment for the larger
-update size in Figure~\ref{fig:rocksdb_256}. The conclusions are identical:
-Cyclone solves Rocksdb's write ahead logging problem.
-
-\begin{figure*}
-\begin{minipage}{.33\textwidth}
+The performance of Rocksdb with Cyclone for the small update workload
+is shown in Figure~\ref{fig:rocksdb} - essentially presenting the
+solution to the problem demonstrated in Figure~\ref{fig:problem}.  We
+consider four different setups. The line labeled 'rocksdb' is the key
+value store running with no logging whatsoever - a system crash would
+lead to data loss. The line labeled 'rocksdb/wal' is for Rocksdb
+running with its write ahead logging turned on. The large gap between
+these two is the overhead of the existing Rocksdb WAL solution. The
+line labeled 'rocksdb/Cyclone 1 way' is a two level Cyclone log but
+without any replication. The line almost exactly tracks the
+performance of Rocksdb. Cyclone is able to provide a write ahead log
+with no overhead to Rocksdb. The line labeled 'rocksdb/Cyclone 3 way'
+is with 3-way replication turned on. Other than a 20us delta due to
+the extra network round trip, the line almost exactly tracks Rocksdb
+performance with no logging. Cyclone therefore provides high
+availability to Rocksdb at a fraction of the cost of its existing
+single machine write ahead log. We also repeat the experiment for the
+larger update size in Figure~\ref{fig:rocksdb_256}. The conclusions
+are identical: Cyclone solves Rocksdb's write ahead logging problem.
+
+\begin{figure}
   \centering
-    \includegraphics[scale=0.45]{results2/rocksdb.pdf}
+  \includegraphics[scale=0.6]{results2/rocksdb.pdf}
   \caption{Rocksdb - small updates}
   \label{fig:rocksdb}
-  \end{minipage}
-  \begin{minipage}{.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_256.pdf}
+\end{figure}
+\begin{figure}
+    \includegraphics[scale=0.6]{results2/rocksdb_256.pdf}
     \caption{Rocksdb - large updates}
     \label{fig:rocksdb_256}
-  \end{minipage}
-  \begin{minipage}{.33\textwidth}
-    \includegraphics[scale=0.45]{results2/rocksdb_multi.pdf}
+\end{figure}
+\begin{figure}
+    \includegraphics[scale=0.6]{results2/rocksdb_multi.pdf}
     \caption{Rocksdb - batched writes}
     \label{fig:rocksdb_multi}
-\end{minipage}
-\end{figure*}
-
-Next, we consider the problem of supporting Rocksdb's write-batch operation that
-atomically writes a set of key-value pairs into the KV store. This is a
-practical application of ganged operations in Cyclone. In order to
-perform the operation with Cyclone managing the log, the client must issue a
-ganged operation across physical logs owning the keys in the write batch.
-A key concern here was whether Cyclone would add any latency to the
-operation due to the extra synchronization needed across replication quorums (
-shown in a previous experiment).
-We examine the problem for the unloaded case
-and small updates in Figure~\ref{fig:rocksdb_multi}, for a single client with
-increasing number of keys in the batch (up to 32 keys).
-The line labeled Rocksdb is with no write ahead logging. We note an increasing
-latency for this baseline indicating Rocksdb itself takes longer
-with larger key batches. The existing option of Rocksdb/wal has considerably
-larger latency. Cyclone does an effective job of cutting down on this latency
-even as it needs to pay a price for synchronizing multiple physical
-logs. Cyclone therefore provides effective replicated logging for batched
-operations using the idea of ganged operations, while scaling the performance of
-single operations using its multiple physical logs.
-
-Our evaluation has, thus far, focused on synthetic write heavy workloads. To
-illustrate that Cyclone can also be useful for real-world read-heavy workloads, 
-we configured our clients to generate key value requests that mimic the
-distribution of value sizes in Facebook's ETCD
-trace~\cite{fb_workload_analysis,fb_scaling_memcache} and are read heavy (95\%
-reads). We use weak reads in Cyclone to avoid the extra round trip for quorum reads.
-Figure~\ref{fig:fb} illustrates that even with a read dominated
-workload, Cyclone is effective at providing a write ahead log with 3 way
-replication with a lower performance overhead than Rocksdb's own write ahead
-logging. We consider the case of another real-world workload that is more write 
-intensive (at 80\%) and with a smaller range of value sizes -- derived from
-Facebook's VAR trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
-Figure~\ref{fig:fb2}. One can similarly conclude that Cyclone provides more
-functionality (replicated write ahead log) at a far lower performance penalty
-that Rocksdb's existing single machine write ahead log solution.
+\end{figure}
+
+Next, we consider the problem of supporting Rocksdb's write-batch
+operation that atomically writes a set of key-value pairs into the KV
+store. This is a practical application of ganged operations in
+Cyclone. In order to perform the operation with Cyclone managing the
+log, the client must issue a ganged operation across physical logs
+owning the keys in the write batch.  A key concern here was whether
+Cyclone would add any latency to the operation due to the extra
+synchronization needed across replication quorums ( shown in a
+previous experiment).  We examine the problem for the unloaded case
+and small updates in Figure~\ref{fig:rocksdb_multi}, for a single
+client with increasing number of keys in the batch (up to 32 keys).
+The line labeled Rocksdb is with no write ahead logging. We note an
+increasing latency for this baseline indicating Rocksdb itself takes
+longer with larger key batches. The existing option of Rocksdb/wal has
+considerably larger latency. Cyclone does an effective job of cutting
+down on this latency even as it needs to pay a price for synchronizing
+multiple physical logs. Cyclone therefore provides effective
+replicated logging for batched operations using the idea of ganged
+operations, while scaling the performance of single operations using
+its multiple physical logs.
+
+Our evaluation has, thus far, focused on synthetic write heavy
+workloads. To illustrate that Cyclone can also be useful for
+real-world read-heavy workloads, we configured our clients to generate
+key value requests that mimic the distribution of value sizes in
+Facebook's ETCD trace~\cite{fb_workload_analysis,fb_scaling_memcache}
+and are read heavy (95\% reads). We use weak reads in Cyclone to avoid
+the extra round trip for quorum reads.  Figure~\ref{fig:fb}
+illustrates that even with a read dominated workload, Cyclone is
+effective at providing a write ahead log with 3 way replication with a
+lower performance overhead than Rocksdb's own write ahead logging. We
+consider the case of another real-world workload that is more write
+intensive (at 80\%) and with a smaller range of value sizes -- derived
+from Facebook's VAR
+trace~~\cite{fb_workload_analysis,fb_scaling_memcache}, in
+Figure~\ref{fig:fb2}. One can similarly conclude that Cyclone provides
+more functionality (replicated write ahead log) at a far lower
+performance penalty that Rocksdb's existing single machine write ahead
+log solution.
 
@@
 
-Finally, we showcase the benefit of using Cyclone beyond pure performance as
-compared to the existing single machine Rocksdb write ahead log. Cyclone brings
-multi-machine availability with the ability to automatically failover. We
-demonstrate this in Figure ~\ref{fig:timeline} that shows the timeline of a run
-where we kill the server process on the leader replica. Cyclone is configured 
-with a 30ms failure detection timeout after which the client library tries
-replicas in turn to locate the leader. In this case it fails over in about
-60ms.
+Finally, we showcase the benefit of using Cyclone beyond pure
+performance as compared to the existing single machine Rocksdb write
+ahead log. Cyclone brings multi-machine availability with the ability
+to automatically failover. We demonstrate this in Figure
+~\ref{fig:timeline} that shows the timeline of a run where we kill the
+server process on the leader replica. Cyclone is configured with a
+30ms failure detection timeout after which the client library tries
+replicas in turn to locate the leader. In this case it fails over in
+about 60ms.
 
@@
 
-Cyclone takes a software only approach to improving replication performance
-using commodity network hardware. In contrast, Mojim~\cite{mojim} pairs directly
-attached persistent memory with RDMA to replicate persistent data
-structures. None of the \emph{evaluated} modes of Mojim provide any means to
-keep more than two replicas in strong synchronization and therefore its
-guarantees are weak compared to Cyclone.  Further, Mojim's claim that
-``Paxos-like protocols'' require two networking round trips and are therefore
-somehow inefficient compared to mirroring is not true. For instance, we have
-demonstrated with Cyclone, where the leader responds as soon as it hears back
-from a majority quorum, that the critical path to replication is just one
-network round trip. We can compare Mojim's replication performance to Cyclone
-for the case where we use only the NVM log in Cyclone. On 40Gbps Infiniband,
-Mojim reported being able to mirror data at 4GB/s. In contrast, Cyclone can
-replicate data three ways using 4*10 Gbps commodity ethernet NICs at roughly 3.7
-GB/s. Note that this result is with Cyclone running the provable RAFT consensus
-protocol, while Mojim is doing simple mirroring of main memory.  One can
-conclude that network offload via RDMA and Infiniband is not an a-priori
-requirement for high throughput replication. This is made possible by Cyclone's
-approach that relieves the CPU from the burden of data copies. On the other
-hand, the latency of a hop on commodity ethernet at 5 us with our DPDK stack is
-larger than the close to 2us latency with Infiniband. Depending on the use case
-for the key value store application the extra 6us with ethernet for the round
-trip to the replica might be a concern for some. We note that mapping the
-commutativity of key value store interfaces to multiple logs in Cyclone is a
-technique that can also be applied to RDMA. Mojim confirms that concurrency is a
-significant determinant of performance when replicating over RDMA.
-
-Consensus in a box~\cite{consensus_box} implements the Zookeeper atomic
-broadcast protocol together with a NIC on an FPGA. That work ignored durability,
-focusing purely on network performance for replication. Latency is excellent due
-to the fact that FPGA cuts out the path from CPU caches to the NIC. The authors
-report latency as low as 3us for a round trip from leader to follower replica,
-compared to the 6us we observe with DPDK. Notably however they reported 7us when
-using TCP/IP from the FPGA rather than their own connection oriented network
-protocol. On the other hand their peak replication throughput is 4M
-replications/sec on 66 Gbps of aggregate network bandwidth for small messages,
-comparable to the 6M replications/sec mark obtained by Cyclone on an aggregate
-40 Gbps of network bandwidth when \emph{also persisting the log on the SSD}. We
-believe the general applicability afforded by our software only approach
-together with the fact that we provide persistence by design makes it a
-compelling alternative to an FPGA based solution, even given the larger
-latency. Also, these results demonstrate that a general purpose CPU core can be
-as efficient as an FPGA in running a consensus protocol using the techniques
-described in this paper.
-
-DARE~\cite{dare} implements state machine replication over RDMA, using a custom
-replication protocol. DARE does not consider persistence in its design. It is
-worthwhile to point out that since DARE allows any replica to become the leader
-it requires a log adjustment protocol for the leader to bring its logs up to
-date, a situation we avoid by using RAFT as described in
-Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication DARE reported
-being able to replicate 500K requests/sec. This is without persisting the log of
-updates. In contrast, Cyclone replicates upwards of 6 million similarly sized
+Cyclone takes a software only approach to improving replication
+performance using commodity network hardware. In contrast,
+Mojim~\cite{mojim} pairs directly attached persistent memory with RDMA
+to replicate persistent data structures. None of the \emph{evaluated}
+modes of Mojim provide any means to keep more than two replicas in
+strong synchronization and therefore its guarantees are weak compared
+to Cyclone.  Further, Mojim's claim that ``Paxos-like protocols''
+require two networking round trips and are therefore somehow
+inefficient compared to mirroring is not true. For instance, we have
+demonstrated with Cyclone, where the leader responds as soon as it
+hears back from a majority quorum, that the critical path to
+replication is just one network round trip. We can compare Mojim's
+replication performance to Cyclone for the case where we use only the
+NVM log in Cyclone. On 40Gbps Infiniband, Mojim reported being able to
+mirror data at 4GB/s. In contrast, Cyclone can replicate data three
+ways using 4*10 Gbps commodity ethernet NICs at roughly 3.7 GB/s. Note
+that this result is with Cyclone running the provable RAFT consensus
+protocol, while Mojim is doing simple mirroring of main memory.  One
+can conclude that network offload via RDMA and Infiniband is not an
+a-priori requirement for high throughput replication. This is made
+possible by Cyclone's approach that relieves the CPU from the burden
+of data copies. On the other hand, the latency of a hop on commodity
+ethernet at 5 us with our DPDK stack is larger than the close to 2us
+latency with Infiniband. Depending on the use case for the key value
+store application the extra 6us with ethernet for the round trip to
+the replica might be a concern for some. We note that mapping the
+commutativity of key value store interfaces to multiple logs in
+Cyclone is a technique that can also be applied to RDMA. Mojim
+confirms that concurrency is a significant determinant of performance
+when replicating over RDMA.
+
+Consensus in a box~\cite{consensus_box} implements the Zookeeper
+atomic broadcast protocol together with a NIC on an FPGA. That work
+ignored durability, focusing purely on network performance for
+replication. Latency is excellent due to the fact that FPGA cuts out
+the path from CPU caches to the NIC. The authors report latency as low
+as 3us for a round trip from leader to follower replica, compared to
+the 6us we observe with DPDK. Notably however they reported 7us when
+using TCP/IP from the FPGA rather than their own connection oriented
+network protocol. On the other hand their peak replication throughput
+is 4M replications/sec on 66 Gbps of aggregate network bandwidth for
+small messages, comparable to the 6M replications/sec mark obtained by
+Cyclone on an aggregate 40 Gbps of network bandwidth when \emph{also
+  persisting the log on the SSD}. We believe the general applicability
+afforded by our software only approach together with the fact that we
+provide persistence by design makes it a compelling alternative to an
+FPGA based solution, even given the larger latency. Also, these
+results demonstrate that a general purpose CPU core can be as
+efficient as an FPGA in running a consensus protocol using the
+techniques described in this paper.
+
+DARE~\cite{dare} implements state machine replication over RDMA, using
+a custom replication protocol. DARE does not consider persistence in
+its design. It is worthwhile to point out that since DARE allows any
+replica to become the leader it requires a log adjustment protocol for
+the leader to bring its logs up to date, a situation we avoid by using
+RAFT as described in Section~\ref{sec:dm}. Using 40 Gb/s speeds, and
+3-way replication DARE reported being able to replicate 500K
+requests/sec. This is without persisting the log of updates. In
+contrast, Cyclone replicates upwards of 6 million similarly sized
 requests per second with 4*10 Gbps ethernet (without the second level
-flashlog). DARE reports an unloaded write latency of 15us, which is lower than
-the unloaded write latency of Cyclone at 20us, thanks to Infiniband. The
-conclusions we can draw are similar to the comparison with Mojim: a carefully
-implemented log replication system obviates the need for network offload when
-doing replication.
-
-{CRANE}~\cite{crane} is a system for replicating multithreaded programs using
-Paxos. It uses deterministic multi-threading to ensure replicas converge to the
-same state as opposed to semantic equivalence in Cyclone using API
-commutativity.  CRANE reports around a 2X slowdown for MySQL, likely rendering
-it unsuitable as a replacement for the write ahead log in key value stores.
+flashlog). DARE reports an unloaded write latency of 15us, which is
+lower than the unloaded write latency of Cyclone at 20us, thanks to
+Infiniband. The conclusions we can draw are similar to the comparison
+with Mojim: a carefully implemented log replication system obviates
+the need for network offload when doing replication.
+
+{CRANE}~\cite{crane} is a system for replicating multithreaded
+programs using Paxos. It uses deterministic multi-threading to ensure
+replicas converge to the same state as opposed to semantic equivalence
+in Cyclone using API commutativity.  CRANE reports around a 2X
+slowdown for MySQL, likely rendering it unsuitable as a replacement
+for the write ahead log in key value stores.
 
@@
 non-volatile memory in improving logging for
-databases~\cite{nvram_log}. However, that work proposes placing the entire log
-in NVM, presuming sufficient availability of such memory. It also does not
-consider replication.
-
-Cyclone does not provide exactly once semantics as we
-believe that exactly once semantics are better provided by the key value store
-itself by persisting necessary request markers - as done in systems such as
-Kafka~\cite{kafka}. A possible direction of future work however is to add
-exactly once semantics to Cyclone's client-server protocol to serve as a
-building block for distributed transactions across key-value shards, each
-running Cyclone, in a manner similar to other systems~\cite{raft_lin}.
+databases~\cite{nvram_log}. However, that work proposes placing the
+entire log in NVM, presuming sufficient availability of such
+memory. It also does not consider replication.
+
+Cyclone does not provide exactly once semantics as we believe that
+exactly once semantics are better provided by the key value store
+itself by persisting necessary request markers - as done in systems
+such as Kafka~\cite{kafka}. A possible direction of future work
+however is to add exactly once semantics to Cyclone's client-server
+protocol to serve as a building block for distributed transactions
+across key-value shards, each running Cyclone, in a manner similar to
+other systems~\cite{raft_lin}.
 
 \section{Conclusion}
-Cyclone shows how one can leverage a small amount of non-volatile memory to
-address two fundamental difficulties in adding high availability to persistent
-key value stores. First, Cyclone avoids
-paying the cost of the block IO interface for every update operation. This
-removes block IO latency from the task of appending updates to the log as well
-as sending it out over the network for replication. In turn, this enables us to
-apply software packet switching techniques and concurrency to improve the
-performance of replication, and making full use of available resources when
-replicating log entries over the network. By optimizing the storage and network
-components simultaneously, Cyclone enables high availability to persistent key
-value stores without compromising on their performance.
-
+Cyclone shows how one can leverage a small amount of non-volatile
+memory to address two fundamental difficulties in adding high
+availability to persistent key value stores. First, Cyclone avoids
+paying the cost of the block IO interface for every update
+operation. In turn, this enables us to apply software packet
+switching techniques and concurrency to improve the performance of
+replication, and making full use of available resources when
+replicating log entries over the network. It enables deployments of
+Rocksdb with native performance without any compromise on availability.
 Cyclone is available as open source software at:\\

### 2018-02-07 paper/paper.tex
More tweaks.
Minor tweaks.

@@
 
-\author{{\rm Amitabha Roy} \\ amitabha.roy@gmail.com
+\author{{\rm Amitabha Roy} \\ amitabha.roy@gmail.com \thanks{Work done while the author was at Intel Labs}
 	     \and {\rm Subramanya R. Dulloor} \\ subramanya.r.dulloor@intel.com}
@@
 allocated in key value stores is for data items under 500 bytes in
-size.  The SSD used in Figure~\ref{fig:problem}, we achieve 50
-Kops/sec to the SSD being limited to a 20us latency on the PCIe
+size. For the SSD used in Figure~\ref{fig:problem}, we achieve 50
+Kops/sec to the SSD due to being limited by the 20us latency on the PCIe
 interface. This translates to 25 MB/s for 500 byte key-value
@@
 chaining received packet mbufs together and treating the chain as a
-single packet.  Finally, we pipline the RAFT protocol state machine by
-contnuing to transmit log entries to follower replicas without waiting
+single packet.  Finally, we pipeline the RAFT protocol state machine by
+continuing to transmit log entries to follower replicas without waiting
 for acknowledgments. The RAFT protocol allows tracking multiple log
@@
 do two phase commit across the logs, ensuring all or nothing
-semantics. However this was unacceptable to some use-cases of Cyclone
+semantics. However this was unacceptable for some use-cases of Cyclone
 as it requires ensuring that an external co-ordinator implementing two
@@
 Cyclone shows how one can leverage a small amount of non-volatile
-memory to address two fundamental difficulties in adding high
-availability to persistent key value stores. First, Cyclone avoids
+memory to address fundamental difficulties in adding high
+availability to persistent key value stores. Cyclone avoids
 paying the cost of the block IO interface for every update
@@
 switching techniques and concurrency to improve the performance of
-replication, and making full use of available resources when
+replication, and make full use of available resources when
 replicating log entries over the network. It enables deployments of
